### 2017-06-07 TPAIS.tex
initialisation

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks,soul}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}

\setlength{\parindent}{0in}
\usepackage{amsmath,amsfonts,amsbsy,amssymb}
\newcommand{\RARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longrightarrow}^{#3}\; #2}
\newcommand{\RARRlong}[3]{#1
  \;\displaystyle\mathop{-\!\!\!-\!\!\!-\!\!\!-\!\!\!-\!\!\!\!\displaystyle
  \longrightarrow}^{#3}\; #2}
\newcommand{\LARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longleftarrow}^{#3}\; #2}
\newcommand{\LRARR}[4]{{\mbox{ \raise 0.4 mm \hbox{$#1$}}} \;
  \mathop{\stackrel{\displaystyle\longrightarrow}\longleftarrow}^{#3}_{#4}
  \; {\mbox{\raise 0.4 mm\hbox{$#2$}}}}
\newcommand{\bX}{{\bf X}}
\newcommand{\vecx}{{\mathbf x}}
\newcommand{\vecy}{{\mathbf y}}
\newcommand{\vecz}{{\mathbf z}}
\newcommand{\vecq}{{\mathbf q}}
\newcommand{\bs}{{\mathbf s}}
\newcommand{\vecr}{{\mathbf r}}
\newcommand{\vecX}{{\mathbf X}}
\newcommand{\vecv}{{\mathbf v}}
\newcommand{\tick}{\ding{52}}
\newcommand{\cross}{\ding{54}}
\newcommand{\vecn}{{\mathbf n}}
\newcommand{\vecp}{{\mathbf p}}
\newcommand{\cT}{{\mathcal T}}
\newcommand{\dt}{{\mbox{d}t}}
\newcommand{\dx}{{\mbox{d} \vecx}}
\newcommand{\boldnu}{{\boldsymbol \nu}}
\newcommand{\er}{{\mathbb R}}
\renewcommand{\div}{{\rm div}}
\newcommand{\bnu}{{\bf \nu}}
\newcommand{\divergence}{\mathop{\mbox{div}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\FP}{P_{\rm{FP}}}
\newcommand{\ME}{P_{\rm{ME}}}
\newcommand{\MEs}{P_{\rm{ME}_{S}}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\X}{{\mathbf X}}
\newcommand{\Y}{{\mathbf Y}}
\newcommand{\W}{{\mathbf W}}
\newcommand{\data}{D}
\newcommand{\neff}{n_{\text{eff}}}
\newcommand{\E}{{\mathbb E}}
\renewcommand{\b}[1]{{\bf #1}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\picturesAB}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\newcommand{\picturesCD}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}

\makeatletter  
\newcommand{\xleftrightarrows}[2][]{\mathrel{%  
 \raise.40ex\hbox{$  
       \ext@arrow 3095\leftarrowfill@{\phantom{#1}}{#2}$}%  
 \setbox0=\hbox{$\ext@arrow 0359\rightarrowfill@{#1}{\phantom{#2}}$}%  
 \kern-\wd0 \lower.4ex\box0}}

\newcommand{\xrightleftarrows}[2][]{\mathrel{%  
 \raise.40ex\hbox{$\ext@arrow 3095\rightarrowfill@{\phantom{#1}}{#2}$}%  
 \setbox0=\hbox{$\ext@arrow 0359\leftarrowfill@{#1}{\phantom{#2}}$}%  
 \kern-\wd0 \lower.4ex\box0}}

\def\leftrightarrowfill@{%
 \arrowfill@\leftarrow\relbar\rightarrow%
 }
\newcommand*{\centerfloat}{%
  \parindent \z@
  \leftskip \z@ \@plus 1fil \@minus \textwidth
  \rightskip\leftskip
  \parfillskip \z@skip}
\makeatother
\makeatother

\title{Transport map accelerated-PAIS, and application to inverse problems arising from
  multiscale stochastic reaction networks}
\author{Simon Cotter, Yannis Kevrekidis, Paul Russell}
\begin{document}
\maketitle
\begin{abstract}
In many applications, inverse problems arise where where there are
complex correlations between the different parameters which we wish to
infer from data. The correlations often manifest themselves as lower
dimensional manifolds on which the likelihood function is
invariant, or varies very little. This can be due to trying to infer
unobservable parameters, or due to sloppiness in the model which is
being used to describe the data. In such a situation, standard
sampling methods for characterising the posterior distribution which
do not incorporate information about this structure will be highly
inefficient. Moreover, most methods are inherently serial in nature,
and as such are not expoiting the parallelised  nature of modern
computer infrastructure. In this paper, we seek to develop a method to
tackle this problem, using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds.

We demonstrate the approach by considering inverse problems arising
from partially observed stochastic reaction networks. In particular,
we consider systems which exhibit multiscale behaviour, but for which
only the slow variables in the system are observable. We demonstrate
that certain multiscale approximations lead to more consistent
approximations of the posterior than others.
\end{abstract}

\section{Introduction}
%Topics to cover: parallel MCMC, pMC, PAIS
%Transport maps, transport map MCMC
%Stochastic reaction networks
%Multiscale approximations, QSSA/QEA, CMA

In Section \ref{sec:map} we show how an appropriate transport map can
be constructed from importance samples which maps the posterior close
to a reference Gaussian measure. In Section \ref{sec:TPAIS} we show
how such a map can be incorporated into a sophisticated parallel MCMC
infrastructure in order to accelerate mixing. In Section
\ref{sec:multi} we consider how likelihoods can be approximated using
multiscale methodologies in order to carry out inference for
multiscale and/or partially observed stochastic reaction networks. In
Section \ref{sec:num} we present some numerical examples, which serve
to demonstrate the increased efficiency of the described sampling
methodologies, as well as investigating the posterior approximations
discussed in the previous section. We conclude with a discussion in
Section \ref{sec:conc}.

\section{Construction of transport maps in importance sampling} \label{sec:map}
%Follow the other papers, show adaptation(s)

\section{T-PAIS}\label{sec:TPAIS}
%Plug into PAIS

\section{Multiscale approximations of likelihoods in stochastic
  reaction networks}\label{sec:multi}

\section{Numerical Examples}\label{sec:num}

\section{Conclusions}\label{sec:conc}

\bibliographystyle{siam}
\bibliography{refs}
\end{document}

### 2017-06-15 TPAIS.tex
copy of thesis chunks into paper

@@
 \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
+\usepackage{tikz,hhline}
+\usepackage{tabularx}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 
+\newcommand\irregularcircle[2]{% radius, irregularity
+  \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
+  +(0:\len pt)
+  \foreach \a in {10,20,...,350}{
+    \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
+    -- +(\a:\len pt)
+  } -- cycle
+}
+
+\newcommand{\into}{\operatornamewithlimits{\longrightarrow}}
+\newtheorem{dfn}{Definition}[section]
+
+\newcommand{\ltri}{%
+\,\resizebox{!}{0.25\baselineskip}{%
+\begin{tikzpicture}%
+\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
+\end{tikzpicture}%  
+}\xspace%
+}%
+
+\newcommand{\smallltri}{%
+\,\resizebox{!}{0.15\baselineskip}{%
+\begin{tikzpicture}%
+\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
+\end{tikzpicture}%  
+}\xspace%
+}%
+
 \title{Transport map accelerated-PAIS, and application to inverse problems arising from
@@
 
+In this section we give details on how \cite{parno2014transport} propose to couple the target,
+$\mu_{\theta}$, with the reference distribution, $\mu_r$.
+When sampling, ideally we would like to draw a sample directly from the target distribution.
+When we have the exact map, $T\in\mathcal{T}$, $T\colon \mathcal{X}\rightarrow\mathbb{R}^d$, we are able to map samples from the posterior distribution onto a reference density, $\pi_r$ which is a standard Gaussian.
+We can then draw a proposal sample from $\pi_r$, before mapping these samples back onto target space using $T^{-1}$.
+These proposed samples are distributed according to the target distribution.
+On target space, the proposal distribution is induced by the pullback of the standard Gaussian through the map,
+\begin{equation}\label{eq:pullback}
+	\tilde{\pi}(\theta) = \pi_r(T(\theta))|J_T(\theta)|,
+\end{equation}
+which is equal to the target density when $T$ is exact.
+Clearly we require $T$ to be monotonic and have continuous first derivatives, which means that our space of feasible transference plans, $\mathcal{T}^\uparrow$, may not include the exact map.
+Since it is unlikely that $T\in\mathcal{T}^\uparrow$, we are motivated to formulate an optimisation problem as discussed in Section~\ref{sec:intro_optim}.
+
+We attempt to find a deterministic coupling of two continuous probability distributions, $(\mu_\theta, \mu_r)$, such that $\mu_r = T\mu_\theta$ is satisfied and also that $T \in \mathcal{T}^\uparrow$.
+There may be infinitely many such couplings, so we look to find the coupling which minimises the difference between the proposal density and the target distribution, i.e.
+the distance between $\pi(\theta)$ and $\tilde{\pi}(\theta)$.
+In this case, we measure this distance using the Kullback-Leibler divergence.
+
+The map does not necessarily satisfy the monotonicity condition when we optimise the cost function,
+\[
+	C(T) = D_\text{KL}(\pi\|\tilde{\pi}),
+\]
+so it must be enforced in the design of each map component.
+To guarantee these properties we carefully choose the basis functions for each component, and we restrict the map to be lower triangular, i.e.
+$T \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$.
+This lower triangular map has the form,
+\[
+	T(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
+		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
+\]
+where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We also assume that the target and reference probability densities are absolutely continuous on
+$\mathbb{R}^d$ which guarantees the maps existence.
+
+\subsection{The optimisation problem}
+
+With these constraints in mind we formulate the optimisation problem explicitly. The cost function
+is chosen to be the Kullback-Leibler divergence between the posterior density and the pullback density,
+\[
+	D_\text{KL}(\pi\|\tilde{\pi}) =
+		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
+\]
+Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map, the divergence becomes
+\[
+	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
+		\log\left|J_{\tilde{T}}(\theta)\right|\right].
+\]
+We note the posterior density is constant in $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
+This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown, however this is exactly the scenario for which we would turn to MCMC methods for a solution.
+
+To find the best coupling, $\tilde{T} \in \mathcal{T}^{\ltri}$, we solve the optimisation problem,
+\[
+	\tilde{T} = \arg\min_{T \in \mathcal{T}^{\smallltri}} \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
+		\log\left|J_T(\theta)\right|\right]
+\]
+which has a unique solution. However we would like to ensure that this solution is regular and monotonic. For this
+reason we regularise the problem further. The optimisation problem now takes the form
+\begin{equation}\label{eq:gen_map_optim}
+	\tilde{T} = \arg\min_{T\in\mathcal{T}^{\smallltri}} \left[
+		 \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
+		\log\left|J_T(\theta)\right|\right] + \beta\mathbb{E}(T(\theta)- \theta)^2 \right].
+\end{equation}
+This parameter $\beta$ does not need to be tuned, experimentation has shown that the choice
+$\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
+close to the identity, and hence are likely to be monotonic.
+
+\subsection{The structure of the map}
+
+Before we continue with the derivation of the optimisation problem, we consider the structure
+of the map in more detail.
+The lower lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density as well as the inverse of the map, $T^{-1}$.
+The Jacobian of $T$ is a lower triangular matrix,
+\[
+	DT(\theta) = \begin{bmatrix}
+		\partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta)\\
+		\vdots & \ddots & \vdots \\
+		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
+	\end{bmatrix} = \begin{bmatrix}
+		\partial_{\theta_1} T_1(\theta) & \dots & 0\\
+		\vdots & \ddots & \vdots \\
+		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
+	\end{bmatrix}
+\]
+since $\partial_{\theta_n} T_k(\theta) = 0$ for all $n > k$.
+This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
+\begin{equation}\label{eqn:separable_jacobian}
+	\log\left|J_T(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} T_i(\theta).
+\end{equation}
+Here we note that this term is separable in terms of the dimension $i$.
+
+Inverting $T$ is simplified by the lower triangular structure in the sense that each dimension, $i$, we need only invert a univariate polynomial.
+This simplification comes about because we invert each dimension of the map independently, beginning with $T_1$ which is a function of only $\theta_1$.
+Finding the root of $T_1(\theta_1) = r_1$, gives us the value of $\theta_1$ which we can then use to find the root of $T_2(\theta_2;\theta_1) = r_2$, and so on.
+These roots, $r_1,\dots,r_d$, are restricted to be real since $\mathcal{X} \subset \mathbb{R}^d$, which allows us to find the unique point, $\theta$, which corresponds to the point $r$.
+
+We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions.
+Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, although not necessarily orthogonal with respect to the target distribution.
+Each component of the map is defined as a multivariate polynomial expansion,
+\begin{equation}\label{eq:map_defn}
+	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
+\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
+\end{equation}
+The parameter $\gamma_i$ is a vector of coefficients in $\mathbb{R}^{M_i}$. Each component of $\gamma_i$ corresponds to a basis function
+$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$, here $d$ is the dimension of
+$\theta$. These multi-indices are elements of the multi-index set $\mathcal{J}_i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
+\[
+	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
+\]
+and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$.
+Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.
+
+The cardinalities of the multi-index sets $M_i = \text{card}(\mathcal{J}_i)$ give the number of unknowns in our
+optimisation problem, and so we would like to keep this number as small as possible. One option is
+to use polynomials of total order $p$,
+\[
+	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
+\]
+which is optimal in terms of the amount of information captured by the map about the target.
+The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = {i+p \choose p}$ which increases rapidly in $d$ and $p$.
+Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$.
+These index sets are discussed further
+in~\cite{parno2014transport}, here we stick with polynomials of total order $p$ since we work with low dimensional problems with the PAIS algorithm.
+
+\subsection{Implementation of the optimisation problem}\label{sec:transport_implementation}
+
+We now return to Equation~\eqref{eq:gen_map_optim}, and show how this can be efficiently formulated
+for level-2 BLAS routines~\cite{choi1992scalapack}. The Monte Carlo estimator of a function is the
+sample average of the function evaluated at each sample, i.e.
+\begin{align*}
+	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
+			\beta\mathbb{E}(T(\theta)-\theta)^2 \\
+		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
+			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right].
+\end{align*}
+This cost function will create a map from $\pi$ to any reference density $\pi_r$.
+We now restrict ourselves to a Gaussian reference by explicitly using a standardised Gaussian density.
+The above expression simplifies to
+\begin{equation}\label{eq:gauss_map_optim}
+	C(T) = \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[\frac{1}{2}
+		T_i^2(\theta^{(k)}) - \log\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}) +
+		\beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right].
+\end{equation}
+Note that since we assume that the map is monotonic, the derivatives of each component are
+non-negative and so we don't need to take the absolute values in the log.
+In practise it is infeasible to enforce this condition across the whole parameter space.
+We instead enforce this condition by ensuring that the derivatives are non-negative at each sample point.
+This means that when we sample away from these support points while in reference space, it is possible to enter a region of space where the map is not monotonic.
+
+We now return to the structure of the map components given in Equation~\eqref{eq:map_defn}. Since the basis functions are
+fixed, the optimisation problem in \eqref{eq:gen_map_optim} is really over the map components $\bar{\gamma} = (\gamma_1, \dots,
+\gamma_d)$ where $\gamma_i \in \mathbb{R}^{M_i}$.
+Note that $C(T)$ is the sum of $d$ expectations, and these expectations each only concern one dimension.
+Therefore we can rewrite \eqref{eq:gen_map_optim} as $d$ separable optimisation problems.
+\begin{align}\label{eq:gamma_map_optim}
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{K}\sum\limits_{k=1}^K
+		\left[\frac{1}{2}T_i^2(\theta^{(k)}; \gamma_i) - \log\frac{\partial T_i}{\partial\theta_i}(\theta^{(k)}; \gamma_i) + \beta(T_i(\theta^{(k)};
+		\gamma_i)-\theta^{(k)})^2\right], \\
+	&\text{subject to} \quad \frac{\partial T_i}{\partial\theta_i}(\theta^{(k)};
+		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,n.
+		\notag
+\end{align}
+The sum in Equation~\eqref{eq:map_defn} is really the inner
+product between the vector of map coefficients, and the evaluations of the basis function at a
+particular $\theta^{(k)}$. This means that if we organise our basis evaluations into two matrices,
+$(F_i)_{k,\mathbf{j}} = \psi_\mathbf{j}(\theta^{(k)})$, and $(G_i)_{k,\mathbf{j}} =
+\frac{\partial\psi_\mathbf{j}}{\partial\theta_i}(\theta^{(k)})$, for all $\mathbf{j}
+\in \mathcal{J}_i^\text{TO}$, and $k = 1,\dots,K$, then we have that $T_i(\theta^{(k)}) = (F_i)_{k\cdot}\gamma_i$ and $\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}; \gamma_i) = (G_i)_{k\cdot}\gamma_i$, so \eqref{eq:gen_map_optim} becomes
+\begin{align}\label{eq:blas_map_optim}
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2}(F_i\gamma_i)^\top(F_i\gamma_i) -
+		c^\top\log(G_i\gamma_i) + \beta\sum\limits_{k=1}^K \!
+		(F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}), \\
+	&\text{subject to} \quad G_i\gamma_i > 0. \notag
+\end{align}
+In this expression, the vector $c$ is a $K\times 1$ vector of ones, and $\log(G_i\gamma_i)$ is to be
+evaluated element-wise. As the Monte Carlo simulations advance, new rows can be appended to the
+$F_i$ and $G_i$ matrices, and $F_i^\top F_i$ can be efficiently updated via the addition of rank-1 matrices.
+
+The regularisation term in Equation~\ref{eq:blas_map_optim} can be approximated using Parseval's identity,
+\[
+	\sum\limits_{k=1}^K \! (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}) \approx
+		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\mu_\theta =
+		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
+\]
+where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
+the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\mu_\theta$; however this
+approximation accomplishes the goal of efficiently regularising the optimisation problem.
+
+These simplifications result in the efficiently implementable, regularised optimisation problem for
+computing the map coefficients,
+\begin{align}\label{eq:final_map_optim}
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2K}\gamma_i^\top F_i^\top F_i\gamma_i -
+		\frac{c^\top}{K}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
+	&\text{subject to} \quad G_i\gamma_i > 0. \notag
+\end{align}
+This optimisation problem can be efficiently solved using Newton iterations. It is suggested
+in~\cite{parno2014transport} that this method usually converges in around 10-15 iterations, and we
+have seen no evidence that this is not a reasonable estimate. When calculating the map several times
+during a Monte Carlo run, using previous guesses of the optimal map to seed the Newton algorithm
+results in much faster convergence, usually taking only a couple of iterations to satisfy the stopping
+criteria.
+
+\subsection{Implementation of the optimisation problem in PAIS}
+
+In the PAIS algorithm, we use weighted samples to approximate the posterior rather than equally weighted samples. Fortunately, the majority of the derivation of a weighted optimisation problem
+follows unchanged. We look at the importance sampling Monte Carlo estimate of $C(T)$, compared with
+Equation~\eqref{eq:gauss_map_optim},
+\[
+	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^n \! \sum\limits_{k=1}^K
+		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
+		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
+\]
+here $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
+all these weights. This necessitates a minor alteration to the optimisation problem in
+Equation~\eqref{eq:final_map_optim},
+\begin{align}\label{eq:weighted_map_optim}
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
+		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
+	&\text{subject to} \quad G_i\gamma_i > 0. \notag
+\end{align}
+We introduce a diagonal matrix $W = \text{diag}(w)$, where $w = (w_1,\dots,w_K)^\top$ into the log of the reference density, and replace the ones vector, $c$, with $w$, otherwise the cost function is unchanged.
+We should obtain the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation.
+The weights are strictly
+positive resulting in a positive definite matrix $W$. This means that the Hessian of the objective
+function is still positive definite, and so the optimisation problem remains convex. Again this
+problem can be efficiently solved using the Newton optimisation algorithm. The Hessian takes the
+form
+\[
+	HC_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i + G_i^\top
+		W\text{diag}([G_i\gamma_i]^{-2})G_i\right] + \beta I,
+\]
+where $[G_i\gamma_i]^{-2}$ is to be taken element-wise, and $I$ is the $M_i\times M_i$
+identity matrix. The first derivative of $C_i(T)$ is
+\[
+	\nabla C_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i\gamma_i - G_i^\top
+		W[G_i\gamma_i]^{-1}\right] + \beta(\gamma_i - \iota),
+\]
+again $[G_i\gamma_i]^{-1}$ is taken element-wise.
+
 \section{T-PAIS}\label{sec:TPAIS}
 %Plug into PAIS
+
+In a similar way, we can use the Transport map derived in Equation~\eqref{eq:weighted_map_optim} to
+design a proposal scheme for the PAIS algorithm. In this case we have a choice in how to proceed; we
+can either use the reference space only to propose new values, or we can run the algorithm on reference space, including resampling, and only map samples onto the target space to calculate expectations.
+The first option allows us to reuse much of the framework
+from the standard PAIS algorithm and in the numerics later we see that this performs better than
+both the Transport MH algorithm, and the standard PAIS algorithm. The second option requires some
+restructuring but results in improved resampling, especially in higher dimensions.
+
+\begin{table}
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
+Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
+\For{$k \leftarrow 1, \dots, L-1$}{
+	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
+	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
+	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
+	Calculate:
+	\[
+		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
+	\]
+
+	Resample $\theta^{(k+1)} \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{\hat{\theta}^{(k)}_j}(\cdot)$.\;
+
+	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
+		\For{$i \leftarrow 1, \dots, n$}{
+			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
+				and update $\gamma_i^{(k+1)}$.\;
+		}
+	}{
+		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
+	}
+}
+\caption{PAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportPAIS1}}
+\end{algorithm}
+\end{table}
+
+The first option is given in Algorithm~\ref{alg:TransportPAIS1}.
+We denote the ensembles of states in target space $\theta^{(k)} = \{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble size.
+Similarly, the proposal states are denoted $r' = \{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) = \{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M, \hat{\theta}^{(k)}_M)\}$, where these pairs are the states which we consider to be our sample from the target distribution.
+As in the standard version of the PAIS algorithm we use the deterministic mixture weights.
+
+\begin{table}
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
+Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
+\For{$k \leftarrow 1, \dots, N-1$}{
+	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
+	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
+	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
+	Calculate:
+	\[
+		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
+	\]
+
+	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\;
+	Invert $\theta^{(k+1)}_i = \tilde{T}^{-1}(r^*_i)$.\;
+	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
+		\For{$i \leftarrow 1, \dots, n$}{
+			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
+				and update $\gamma_i^{(k+1)}$.\;
+		}
+	}{
+		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
+	}
+}
+\caption{PAIS algorithm with adaptive transport map. Option 2.\label{alg:TransportPAIS2}}
+\end{algorithm}
+\end{table}
+
+The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to the first except that rather than resampling in target space we resample in reference space.
+In reference space the dimensions are roughly uncorrelated, and the Gaussian marginals are easy to approximate with fewer ensemble members.
+This means that the resampling step will be more efficient in higher dimensions without needing to increase the ensemble size as much as we have needed to in the standard PAIS algorithm.
 
@@
 
+We now move onto our main application area; the modelling of discrete chemical populations and the discovery of reaction rates within these systems.
+Often when modelling populations of chemical or animal species, the population is large enough that we can consider continuous models, such as coupled differential equations, to describe the evolution of the populations over time.
+However, this approximation is not always appropriate.
+These continuous models, when applied to smaller populations, result in time points where there are fractional numbers of a species.
+In the real world there can only be an integer number of individuals, and the effect on other species in the system might be very different to the continuous model's prediction.
+It is instead more appropriate to model these systems with stochastic simulation algorithms (SSAs) which model population counts as non-negative integers and take account of the intrinsic noise.
+The most common SSA is the Gillespie algorithm proposed in 1977~\cite{gillespie1977exact}.
+
+\subsection{Stochastic Simulation Algorithms}\label{sec:SSAs}
+
+SSAs are a powerful tool for describing the dynamics of discrete systems, most commonly used in the description of small scale chemical systems such as in cell biology.
+These algorithms simulate reactions between different species.
+A reaction has occured whenever the populations of the different species in the system change.
+For example, a simple reaction where two chemicals from the same species join together to form a chemical from a different species, this is denoted
+\begin{equation}\label{eqn:chem_reaction}
+	X_1 + X_1 \into^{k} X_2
+\end{equation}
+where the variable $k$ is the rate at which the reaction occurs.
+These reactions are considered to occur instantaneously.
+By randomly sampling the time between reactions as well as which reactions occur we can simulate a potential trajectory for the population of a species during a given time interval.
+Methods for the simulation of these systems assume that the system is in thermal equilibrium and that the density of each species is constant across the entire system domain.
+
+\begin{dfn}[Stochastically exact SSA]
+An SSA for a quantity $S_t$ is stochastically exact if at every time point, $t$, the probability that a particular trajectory, $T_t$, takes a value $s$ is equal to the probability of the quantity $S_t$ taking that value, i.e.
+\[
+	\mathbb{P}(S_t = s) = \mathbb{P}(T_t = s)\quad \forall t \ \text{and} \ s.
+\]
+At the time $t$, the distribution of $S_t$ is denoted as $\mu_t$.
+\end{dfn}
+If an SSA is stochastically exact, these trajectories can be used to estimate probability distributions for the population of a species.
+Empirical distributions for the value of $S_t$ throughout a time interval can be produced by a sufficiently large number of simulated trajectories.
+
+Population models can contain several sources of variability.
+Large scale heterogeneity such as genetic diversity and environmental factors are often ignored in deterministic modelling.
+On a smaller scale the population has intrinsic noise which can come from thermal fluctuations at the molecular level~\cite{szekely2014stochastic,mcadams1999sa}.
+This small scale variability is inherently incorporated into the model dynamics of an SSA, whereas genetic and environmental factors are not.
+During the time span of a model simulation, we can assume that the genetic factors are constant and, if necessary, environmental factors can be explicitly included by evolving the probabilities of which reactions occur~\cite{shahrezaei2008colored}.
+
+We now describe some methods of approximating $\mu_t$~\cite{gillespie1991markov}.
+The deterministic master equation is a system of DEs which defines a chemical system, when it is possible to solve this system of differential equations (DEs) we can obtain an exact form for $\mu_t$~\cite{anderson2016product,jahnke2007solving,anderson2010product}.
+However analytic solutions are rare and it is common to approximate $\mu_t$ using stochastic methods, such as the Gillespie SSA~\cite{gillespie1977exact}, $\tau$-leap methods~\cite{cao2006efficient,cao2005avoiding,chatterjee2005binomial} or SDEs such as the chemical Langevin equation~\cite{gillespie2000chemical}.
+
+%%% Gillespie SSA
+
+The Gillespie algorithm is an example of an exact SSA.
+The problem is formulated as a set of all possible reactions of the type in Equation~\eqref{eqn:chem_reaction}, the choice of which reaction and when it occurs is decided by a pair of random numbers whose distribution is determined by the relative densities of the chemical species.
+
+\begin{dfn}[Propensity function]
+Given $n$ chemical species with populations
+\[
+	\mathbf{X}(t) = [X_1(t),\dots,X_n(t)]^\top,
+\]
+the likelihood of a reaction, $R_i$, occuring is determined by the \emph{propensity function} $\alpha_i(\mathbf{X}(t); \mathbf{k})$ and depends on the current state of the system and the reaction rates $\mathbf{k}$.
+\end{dfn}
+
+The propensity function also takes into account the volume of the domain, the temperature, and many other factors.
+These functions $\alpha_i(\mathbf{X}(t);\mathbf{k})$ are normalised to give the probabilities, $p_i(\mathbf{X}(t);\mathbf{k})$, that $R_i$ will be the next reaction to occur,
+\[
+	p_i(\mathbf{X}(t); \mathbf{k}) = \frac{\alpha_i(\mathbf{X}(t);\mathbf{k})}{\alpha_0(\mathbf{X}(t);\mathbf{k})}, \quad \text{where} \quad \alpha_0(\mathbf{X}(t);\mathbf{k}) = \sum_i \!
+\alpha_i(\mathbf{X}(t);\mathbf{k}).
+\]
+There is a propensity function associated with each of the $d$ possible reactions.
+The form of these functions is dependent on the the type of reaction which occurs.
+This notation can become cumbersome and so the dependence on the current state and reaction rates may be omitted.
+The normalisation constant $\alpha_0$ is known as the \emph{total system propensity} or \emph{total propensity}.
+
+The Gillespie algorithm simulates a Markov jump process about the chemical population state space.
+The algorithm assumes that reactions occur at times which follow an exponential distribution.
+The rate associated with this exponential distribution is linked to the total system propensity $\alpha_0$.
+Which reaction occurs at each of these times depends on the distribution $\mathbf{p} = [p_1, \dots, p_d]^\top$.
+This assumption is the same as that in the formulation of the chemical master equation.
+
+Simulating a trajectory by sampling every reaction in this way allows us to generate stochastically exact draws from the true distribution $(\mu_t, \forall t)$, however it can be extremely costly, especially when reactions occur on different timescales.
+Variations on the algorithm have been proposed which reduce the computational cost, these are described in the review \cite{gillespie2007stochastic}.
+
+\begin{table}[!htpb]
+\centering
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+	Choose $\mathbf{X}(0) = [x_1, \ldots, x_n]^\top$.\;
+	Define $\mathbf{k} = [k_1, \ldots, k_d]^\top$.\;
+	Set $j= 0,\ t_0 = 0$.\;
+	\While{$t_j < \tau$}{
+		Update $\alpha_i(\mathbf{X}(t_j);\mathbf{k})$, for $i=1,\ldots, d$.\;
+		Set $\alpha_0 = \sum_i \alpha_i(\mathbf{X}(t_j); \mathbf{k})$ and calculate $\mathbf{p}$.\;
+		Sample $\Delta t \sim \text{Exp}(\alpha_0)$ and $r(t_j+\Delta t) \sim \text{Multinomial}(1, \mathbf{p})$.\;
+		Set $t_{j+1} \leftarrow t_j + \Delta t$.\;
+		Update populations $\mathbf{X}(t_{j+1}) = f_{r(t_{j+1})}(\mathbf{X}(t_j))$.\;
+		$j \leftarrow j+1$.\;
+	}
+\caption{The Gillespie Algorithm~\cite{gillespie2007stochastic}.\label{alg:gillespie}}
+\end{algorithm}
+\end{table}
+
+The Gillespie algorithm is outlined in Algorithm~\ref{alg:gillespie}.
+At each iteration we calculate the propensities, $\alpha_i(\mathbf{X}(t_j);\mathbf{k})$, based on the current state of each of the populations.
+We use these propensities to calculate $\alpha_0$ as well as the probabilities, $\mathbf{p}(\mathbf{X}(t_j); \mathbf{k})$.
+We now generate two random numbers, the first, $r(t_j)$, is from the multinomial distribution defined by the probability vector $\mathbf{p}$, which tells us which reaction occurs next.
+The second, $\Delta t$, is from the exponential distribution with rate $\alpha_0$, and tells us how long until this reaction occurs.
+Finally we update the system's state, $\mathbf{X}(t_{j+1})$, taking into account which reaction has occured, using the transition function $f_{r(t_j)}$.
+These steps result in a stochastically exact trajectory from which we can determine the probability of being in any particular state.
+
+%%% Multiscale
+
+In real world chemical systems, reactions will occur on more than one time scale.
+It is often possible to isolate which reactions are occuring more frequently (the \textit{fast reactions}) and those which are occuring less frequently (the \textit{slow reactions}).
+In many cases we are interested in the dynamics of the slow reactions, and the full simulation of the fast reactions can be considered a waste of computational resources.
+In this thesis we consider two separate simplifying approaches for these multiscale systems.
+The first approach uses the quasi-equilibrium assumption.
+
+\begin{dfn}[Quasi-equilibrium assumption]
+	The quasi-equilibrium assumption (QEA) is the assumption that the fast reactions converge in distribution on a timescale which is negligible with respect to the rate of occurence of the slow reactions.
+\end{dfn}
+
+This assumption allows us to approximate the dynamics of the slowly changing quantities in the system by assuming that the fast quantities are in equilibrium with respect to the fast reactions in isolation.
+We also consider the constrained multiscale approach (CMA) described in \cite{cotter2011constrained,cotter2016constrained}.
+The CMA accounts for the differences in the invariant distribution of the fast species which are due to the occurence of slow reactions.
+When the fast and slow subsystems operate on very different timescales, the CMA and QEA produce near identical results, however when the systems are not so clearly separated the CMA provides superior accuracy.
+Other simulation methods such as the $\tau$-leap method introduce significant bias into the approximations.
+
+In this section we are interested in chemical systems where we have a small number of chemical species.
+We assume that we have thermal equilibrium and that the chemicals are well mixed so that we have a homogeneous system.
+We assume that we have accurate observations of the populations of the chemicals and the times of the reactions, and we are concerned with recovering the probability distribution which describes the rates of the reactions.
+The posteriors arising from situations such as these often have non-Gaussian tails which, as discussed in previous chapters, can cause problems when using kernel density estimations of the posterior distribution as we do in the standard PAIS algorithm.
+The dynamics of these systems are also highly non-linear resulting in complex correlation structures between parameters.
+
+%%%
+
+\subsection{Sufficient Statistics for Rate Recovery}
+
+Following the assumptions for the modelling of the stochastic process $\mathbf{X}(t),\ t \in (t_0, \tau]$ outlined in the previous section, we can determine sufficient statistics for the probability distributions of the model parameters $\mathbf{k}$.
+Here we treat the problem in the Bayesian context and so we define a posterior distribution, $\pi(\mathbf{k}|D)$, which incorporates the model likelihoods with prior information for these reaction rates.
+The data, $D$, is a complete trajectory of a molecular system which includes the times of each reaction as well as which reaction occurs.
+In practise it might not be obvious which reaction has occured just by looking at the changes in chemical populations, however we ignore this problem for now.
+
+For a problem with $n$ chemical species, $X_i$,  and $d$ reactions, $R_j$, the data, $D$, has the form given in Table~\ref{tab:chem_data}.
+The data is measured over a time interval which contains $M$ reactions.
+We assume that the times between reactions have the exponential distribution, so we define the differences $\Delta t_j = t_{j}-t_{j-1} \sim \text{Exp}\left(\alpha_0(\mathbf{X}(t_{j-1});\mathbf{k})\right)$.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{l|r|r|r|}
+	\hhline{~|--|}
+	Data $D$ \quad & Time & Reaction \\ \hhline{~|=|=|}
+	& $t_1$	& $r(t_1)$ \\
+	& $t_2$	& $r(t_2)$ \\
+	& \vdots & \vdots \\
+	& $t_M$	& $r(t_M)$ \\ \hhline{~|--|}
+ \end{tabular}
+\caption{Form of the data $D$, in the notation given in Algorithm~\ref{alg:gillespie}. The initial time $t_0$ is set to 0.}
+\label{tab:chem_data}
+\end{table}
+
+At a time $t_j$, the reaction $R_i$ occurs with probability $p_i(\mathbf{X}(t_{j-1});\mathbf{k})$, where the propensities are calculated using the system state at the previous time step,
+\[
+	r(t_j) \sim \text{Multinomial}\left(1, \mathbf{p}(\mathbf{X}(t_{j-1});\mathbf{k})\right)
+\]
+All the possible states of the process, $\mathbf{X}(t),\ t \in (t_0,\tau]$, belong in the state space $\mathcal{S}$, the space itself does not depend on time for the problems which we are considering.
+
+From this formulation, we see that the random variables $(\Delta t_j, r(t_j))$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are independent of each other.
+This independence means that we can group events together by what state the system was in when the event happened.
+We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
+\[
+	T(\mathbf{Y}) = \sum\limits_{j=1}^M \Delta t_j\; \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y}).
+\]
+This random variable, $T(\mathbf{Y})$, is a sum of Exponential distributions, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
+\begin{equation}\label{eqn:chem_time_dist}
+	T(\mathbf{Y}) \sim \text{Gamma}\left(\alpha=K(\mathbf{Y}),~\beta = \alpha_0(\mathbf{Y}; \mathbf{k})\right),
+\end{equation}
+where $K(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y})$.
+
+Similarly, we can define the reactions which occured when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_d(\mathbf{Y})]^\top$ where
+\[
+	r_i(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(r(t_j) = i \textbf{ and }\mathbf{X}(t_{j-1}) = \mathbf{Y}).
+\]
+Here all the random variables $r(t_j)$ follow the same multinomial distribution, and so
+\begin{equation}\label{eqn:chem_react_dist}
+	\mathbf{r}(\mathbf{Y}) \sim \text{Multinomial}(K(\mathbf{Y}),~\mathbf{p}(\mathbf{Y})). 
+\end{equation}
+The random variables defined in Equations~\eqref{eqn:chem_time_dist} and \eqref{eqn:chem_react_dist} are sufficient statistics for the posterior distribution $\pi(\mathbf{k}|D)$.
+With these definitions we define two new structures
+\[
+	\mathbf{T} = [T(\mathbf{Y}_1), \dots, T(\mathbf{Y}_K)]^\top, \quad \text{and} \quad \mathbf{R} = [\mathbf{r}(\mathbf{Y}_1), \dots, \mathbf{r}(\mathbf{Y}_K)]^\top,
+\]
+where $K = |\mathcal{S}|$, the number of states in $\mathcal{S}$, and each state $\mathbf{Y}_i \in \mathcal{S}$ has been enumerated. We use these structures to define shorter notation,
+\[
+	\mathbf{T}_i = T(\mathbf{Y}_i), \quad \mathbf{R}_{ij} = r_j(\mathbf{Y}_i), \quad \text{and} \quad \mathbf{K}_i = K(\mathbf{Y}_i).
+\]
+
+\subsection{Posterior Distribution}\label{sec:chem_stoch_posterior}
+
+To construct the posterior distribution for the reaction rates, $\mathbf{k}$, in the chemical system, we formulate the likelihood using the sufficient statistics derived in the previous section.
+Due to the positivity of these reaction rates, we assign a Gamma prior distribution to each rate.
+Given the distributions in Equations~\eqref{eqn:chem_time_dist} and~\eqref{eqn:chem_react_dist} for our data, the likelihood of observing the data $\mathbf{R}$ and $\mathbf{T}$ is
+\[
+	\ell(\mathbf{R},~ \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
+\]
+where again $\mathbf{Y}_i \in \mathcal{S}$ and the propensities $\alpha_i$ and probabilities $p_i$ depend on the reaction rates $\mathbf{k}$.
+
+Our choice of Gamma prior distributions with hyper-parameters $(a_i, b_i)$ results in the posterior distribution of the form
+\begin{align}
+	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
+	\prod_{i=1}^d \! \text{Gamma}(k_i; a_i, b_i) \notag \\
+		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
+				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
+%		& \qquad\qquad\qquad
+				+\sum_{j=1}^d \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
+			\Bigg]  \notag\\
+		&	\qquad\qquad\qquad{} + \sum_{i=1}^d ((a_i-1)\log k_i - b_ik_i)\Bigg\}. \label{eqn:chem_posterior}
+\end{align}
+
 \section{Numerical Examples}\label{sec:num}
+
+In this section, we consider a simple multiscale example involving two chemical species $S_1$ and $S_2$:
+\begin{equation}\label{eqn:full_system}
+	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
+\end{equation}
+Each arrow represents a reaction from a reactant to a product, with
+some rate constant $k_i$, and where the rates of the reactions are
+assumed to follow mass action kinetics. The parameters $k_i$ are strictly positive, and $\mathbf{k} = [k_1,\dots,k_4]^\top \in \mathbb{R}_+^4 = \mathcal{X}$. We denote the concentration of
+species $S_i$ by $X_i$. We assume that we are in a
+parameter regime such that the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
+much much more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
+chemical species are involved in fast reactions. However, the quantity
+$S = X_1 + X_2$ is conserved by both of the fast reactions,
+and as such, this is the slowly changing quantity in this system.
+The effective dynamics of $S$ can be represented as follows
+\begin{equation}\label{eqn:QSSA_system}
+	\emptyset \xrightarrow{k_1} S \xrightarrow{\hat{k}_4} \emptyset.
+\end{equation}
+The new reaction rate $\hat{k}_4$ is approximated through
+application of the two approaches mentioned in Section~\ref{sec:SSAs}.
+Under the QEA, the value of $\mathbb{E}\left[k_4X_2|S=s\right]$ is approximated by finding the steady state of the ODE representing the fast subsystem of
+reactions:
+\[
+	S_1 \xleftrightarrows[k_3]{k_2} S_2, \qquad X_1 + X_2 = s
+\]
+we arrive at
+\[
+	\hat{k}_4^{\text{QEA}}(s) = \mathbb{E}_{\text{QEA}}\left[k_4X_2|S=s\right] = \frac{k_2k_4s}{k_2+k_3}.
+\]
+Similarly, the analysis of the constrained system as discussed in~\cite{cotter2016constrained} yields the effective propensity
+\begin{equation}\label{eqn:chem_CMA_rate}
+	\hat{k}_4^{\text{CMA}}(s) = \mathbb{E}_{\text{CMA}}\left[k_4X_2|S=s\right] = \frac{k_2k_4s}{k_2+k_3+k_4}.
+\end{equation}
+
+Our observations are uninformative about the reaction rates $k_2, k_3$, and $k_4$, as there
+is a submanifold $\mathcal{M} \subset \mathcal{X}$ along which the effective rate
+$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
+problem. This type of problem is notoriously difficult to
+sample from using standard MH algorithms, as the algorithms quickly
+find a point on $\mathcal{M}$ but then
+exploration around $\mathcal{M}$ is slow.
+
+\subsubsection{Target Distribution}\label{sec:multiscale_posterior}
+
+We can use the posterior distribution derived in Section~\ref{sec:chem_stoch_posterior} to specify the posterior distribution for each of the three scenarios we are interested in.
+We are interested in the posterior distribution when we have a complete data set telling informing us of every reaction which occurs as well as perfect readings of the numbers of reactants.
+We are also interested in the two simplifying assumptions we saw in the previous section.
+
+In all three cases, we find the posterior distribution for $\mathbf{k} \in \mathcal{X}$. These four parameters are assigned Gamma prior distributions,
+\[
+	k_i \sim \text{Gamma}(\cdot; \alpha_i, \beta_i), \quad \text{for} \quad i = 1, \dots, 4.
+\]
+The hyper-parameters corresponding to each of these prior distributions are given in Table~\ref{tab:multiscale_priors}. These priors are the same for each of the three posterior distributions.
+
+\begin{table}
+\centering
+\begin{tabular}{|l|R{2em}R{2em}R{2em}R{2em}|}
+	\hline
+	parameter/dimension & 1 & 2 & 3 & 4 \\ \hline
+	$\alpha_i$ & 150 & 5 & 5 & 3 \\
+	$\beta_i$ & 15/9 & 5/12 & 5/12 & 1 \\ \hline
+\end{tabular}
+\caption{Hyper-parameters in the prior distributions for the multiscale problem described in Section~\ref{sec:multiscale_posterior}.}
+\label{tab:multiscale_priors}
+\end{table}
+
+Under the QEA and CMA simplifications, we reduce the information contained in the data, as defined in Table~\ref{tab:chem_data}. We now only have one species, $S$, and two reactions,
+\[
+	R_1\colon \emptyset \xrightarrow{k_1} S, \quad \text{and} \quad R_2\colon S \xrightarrow{\hat{k}_4} \emptyset,
+\]
+hence the state space of our system is much smaller.
+This means it is much faster to evaluate the posterior given in Equation~\eqref{eqn:chem_posterior}.
+For the full data scenario, the dimension of the parameter $\mathbf{k}$ is $d$ and we have $d$ reactions, however when using the QEA or CMA simplifications, the number of reactions is reduced to $n=2$ so that $d \neq n$.
+i.e.
+We slightly modify the posterior from Equation~\eqref{eqn:chem_posterior} to
+\begin{align}
+	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
+				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k})
+				+\sum_{j=1}^{n} \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
+			\Bigg]  \notag\\
+		&	\qquad\qquad\qquad{} + \sum_{i=1}^d ((a_i-1)\log k_i - b_ik_i)\Bigg\}, \label{eqn:chem_QEA_posterior}
+\end{align}
+where $n=4$ for the full data model, and $n=2$ for the simplifications.
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.9\textwidth]{"images/CMA_posterior"}
+\caption{Posterior for the constrained multiscale problem outlined in Section~\ref{sec:chem_multiscale}.}
+\label{fig:chem_CMA_posterior}
+\end{figure}
+
+Figure~\ref{fig:chem_CMA_posterior} shows how this posterior looks when we use the constrained approach in Equation~\eqref{eqn:chem_CMA_rate} to model the effective degradation rate $\hat{k}_4$.
+
+\subsubsection{Implementation}\label{sec:chem_implementation}
+
+For the posterior distribution in \eqref{eqn:chem_QEA_posterior}, we consider several proposal methods.
+First we implement both the PAIS and MH algorithms with a Gaussian proposal distribution.
+In the case of the PAIS algorithm, this is a Gaussian mixture proposal distribution.
+The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard RWMH algorithm.
+We also compare the PAIS and MH algorithms when using a transport map proposal distribution.
+This proposal method was discussed in detail in Chapter~\ref{sec:TransportPAIS}.
+
+\begin{figure}
+	\centering
+	\begin{tikzpicture}
+		\pgfmathsetseed{1}
+
+		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
+		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$R$};
+
+		\draw (A) \irregularcircle{1cm}{2.5mm};
+		\draw (B) \irregularcircle{1cm}{0.5mm};
+
+		\draw [->] (A) -- (B);
+
+		\node at (2.5,1) {$T'\colon\mathcal{X}\rightarrow R$};
+
+		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
+		\node at (5,-1.5) {$r\in R\subset\mathbb{R}^d$};
+	\end{tikzpicture}
+	\caption{Couplings between parameter space and reference when using the transport map proposal method to propose moves on parameter space.}
+	\label{fig:chem_coupling}
+\end{figure}
+
+Figure~\ref{fig:chem_coupling} shows how we define a bijective map, $T'$, between parameter space $\mathcal{X}$ and a reference space $R$.
+In practice we cannot ensure that the approximate map $\tilde{T}'$ is uniquely invertible over the whole of $R$ and so $\tilde{T}'$ is not truly bijective.
+This leads to problems for our strictly positive state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in $R$ do not necessarily map back on to $\mathcal{X}$.
+This motivates the use of a third intermediate space.
+When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
+MH-RW (RWMH) and MH-TRW, as well as PAIS-RW and PAIS-TRW.
+
+We also consider how these algorithms perform when they are applied to the log of the reaction rates.
+We choose to use this log transformation since it converts our strictly positive parameter space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset \mathbb{R}^d$.
+This allows us to define $T$ between two subsets of $\mathbb{R}^d$, which means that even if $\tilde{T}$ is not uniquely invertible in some region of $R$, all possibilities are valid proposals.
+As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
+MH-logRW and MH-logTRW, with, PAIS-logRW and PAIS-logTRW.
+
+\begin{figure}
+	\centering
+	\begin{tikzpicture}
+		\pgfmathsetseed{1}
+
+		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
+		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
+		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};
+
+		\draw (A) \irregularcircle{1cm}{2.5mm};
+		\draw (B) \irregularcircle{1cm}{1.1mm};
+		\draw (C) \irregularcircle{1cm}{0.5mm};
+
+		\draw [->] (A) -- (B);
+		\draw [->] (B) -- (C);
+
+		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
+		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};
+
+		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
+		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
+		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
+	\end{tikzpicture}
+	\caption{Couplings between parameter space, an intermediate space and reference space when using the transport map proposal method to propose moves on a log space.}
+	\label{fig:chem_log_coupling}
+\end{figure}
+
+Figure~\ref{fig:chem_log_coupling} displays the composition of maps when we include this intermediate space $\mathcal{Y}$.
+Every proposal in $R$ results in a valid proposal on $\mathcal{X}$.
+The inclusion of this additional map means that we must again alter our importance weight definition to reflect the pullback from $R$ through $\mathcal{Y}$.
+The weight is now
+\[
+	w_i(\theta') = \frac{\pi(\theta'|\mathbf{R},\mathbf{T})}{Q(\tilde{T}\circ\log(\theta')|\tilde{T}\circ\log(\theta^{(i-1)}))|J_{\tilde{T}\circ\log}(\theta')|},
+\]
+where $\theta'$ is a proposal on $\mathcal{X}$, $\theta^{(i-1)}$ is the ensemble of states from the previous iteration, and $J_{\tilde{T}\circ\log}(\theta')$ is the Jacobian of the composition of the two maps.
+This Jacobian is straightforward to calculate,
+\[
+	|J_{\tilde{T}\circ\log}(\theta')| = |J_{\tilde{T}}(\log(\theta'))||J_{\log}(\theta')|,
+\]
+where the first determinant is as we saw in the previous chapter, and the second is
+\[
+	|J_{\log}(\theta')| = \prod\limits_{i=1}^d \frac{1}{\theta'_i}.
+\]
+
+% choices made in transport map
+For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=4$ as the basis functions, i.e.
+\[
+	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
+\]
+and
+\[
+	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
+\]
+Since $\mathcal{X}$ is four-dimensional, this yeilds a total of 125 map coefficients across the four index sets.
+This number can be reduced by using smaller index sets as discussed in~\cite{parno2014transport}.
+
+% choice of resampler?
+As with the mixture model in Section~\ref{sec:app_mixture}, we will use the AMR resampler with an ensemble size of $M=500$.
+This increase in ensemble size from Chapter~\ref{sec:PAIS} compensates for the increase in parameter dimension.
+
+% analysis method
+To measure the convergence of the sampling methods in this section, we will compare the convergence of the mean of each parameter.
+We approximate $\mathbb{E}(\mathbf{k})$ using 2.4 billion samples from the MH-RW algorithm.
+
+%*** numerics.
+\subsubsection{Convergence Analysis for the Constrained Approach to a Multiscale Chemical Reaction Model}\label{sec:chem_conv}
+
+In this section, we demonstrate the convergence of the sample mean $\hat{\mathbf{k}}$ by approximating the relative error between the sample mean and the true mean $\mathbb{E}(\mathbf{k})$.
+We do this for each of the eight algorithms introduced in the previous section.
+Convergence is shown for the constrained approach to the multiscale system, i.e.
+to the posterior in Equation~\eqref{eqn:chem_QEA_posterior} with the effective degredation rate, $\hat{k}_4^{\text{CMA}}$, from Equation~\eqref{eqn:chem_CMA_rate}.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{rrrrrr}
+\toprule
+	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{PAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
+	& & $\delta_\%$ & & $\delta_{\text{ESS}}$ & ESS \\ \midrule
+	RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
+	TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
+	logRW & & 1.7e-1 & & 1.2e-0 & 6.0e-2 \\
+	logTRW & & 2.7e-2 & & 1.5e-1 & 3.5e-1 \\
+\bottomrule
+\end{tabular}
+\caption{Optimal scaling parameters for the MH and PAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}.
+MH parameters optimised by acceptance rate, and PAIS parameters optimised using effective sample size.}
+\label{tab:chem_multiscale_scaling}
+\end{table}
+
+The optimal scaling parameters are given in Table~\ref{tab:chem_multiscale_scaling}.
+We note that for MH-TRW and PAIS-TRW the scaling parameter is near to 1, particularly for MH-RW, this is what we should expect since the proposal is made on a reference space which should be near to $\mathcal{N}_d(0, \text{I})$.
+The same should be true for the MH-logTRW and PAIS-logTRW algorithms since these have the same target reference space, however the optimal scaling parameters here are much smaller.
+We see that the ESS is higher for the algorithms which sample on $\mathcal{Y}$, and we expect that convergence will be fastest for the PAIS-logTRW algorithm.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Sampling algorithms on $\mathcal{X}$.]{\includegraphics[width=0.495\textwidth]{"images/CMA_L2_param_space"}}
+\subfigure[Sampling algorithms on $\mathcal{Y}$.]{\includegraphics[width=0.495\textwidth]{"images/CMA_L2_log_space"}}
+\caption{Convergence of the constrained multiscale example described in Section~\ref{sec:chem_multiscale}.}
+\label{fig:chem_multiscale_L2}
+\end{figure}
+
+Convergence of the eight algorithms for this example is shown in Figure~\ref{fig:chem_multiscale_L2}.
+We first note the poor performance of the MH based algorithms, each of them taking roughly 10,000 samples to begin converging.
+Only the MH-TRW is at all competitive with the PAIS algorithms.
+During the simulation interval, the MH-TRW algorithm has not settled down to the expected $\mathcal{O}(1/\sqrt{N})$ rate which means that the estimate is still biased by the long burn-in time.
+As we have seen in previous examples, the burn-in time for the PAIS algorithm is negligible.
+
+The PAIS variants with RW and TRW proposals perform similarly on both sample spaces.
+When sampling on $\mathcal{X}$ the transport map is not quite as efficient, largely due to the difficulties discussed in the previous section i.e.
+many proposals are made which do result in negative reaction rates.
+Sampling on $\mathcal{Y}$ leads to more comparable Monte Carlo errors, with the logTRW being apparently slightly less stable.
+This proposal method becomes more stable as we increase either the ensemble size, or the number of iterations between updates of the transport map, $T$.
+Overall we see the smallest Monte Carlo errors for a given amount of computational effort coming from the PAIS-logTRW algorithm.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Reference space mapped to by $T'$.]{\includegraphics[width=0.495\textwidth]{"images/CMA_ref_space_MHTRW"}}
+\subfigure[Reference space mapped to by $T\circ\log$.]{\includegraphics[width=0.495\textwidth]{"images/CMA_ref_space_MHlogTRW"}}
+\caption{Reference space for the TRW and logTRW proposal distributions. Components are linked by the relation $r_i = T_i'(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
+\label{fig:chem_reference_spaces}
+\end{figure}
+
+We now look at the proposal distributions of the transport map accelerated algorithms.
+In Figure~\ref{fig:chem_reference_spaces}, we see the reference spaces found by; in (a) mapping the posterior through the map $T'$, and in (b) by mapping the posterior through $T\circ\log$.
+For the most part, each of these marginal distributions can be recognised as a Gaussian.
+However, with the exception of $\mathbb{P}(r_2,r_3)$, we would not consider them to be close to a standardised $\mathcal{N}(0, \text{I})$ distribution.
+Before thinking that the transport map has not helped us to find a `nicer' space on which to propose new values we should consider that the dimensions are now (1) largely uncorrelated, and (2) the variances in each dimension are much more similar than they are in Figure~\ref{fig:chem_CMA_posterior}.
+
+Particularly in Figure~\ref{fig:chem_reference_spaces} (b) we see that $\text{var}(r_1)$ and $\text{var}(r_4)$ are much smaller than $\text{var}(r_2)$ and $\text{var}(r_3)$.
+To combat this we have a number of choices, we might wish to use two different scaling parameters to match these scales, which would require knowledge of the reference space before beginning sampling.
+We could alternatively use a proposal distribution such as in the LPAIS algorithm to adaptively learn this at each iteration.
+Another option is to increase the total order of our index set.
+For these numerics we have chosen $p=4$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
+
+\subsubsection{Comparison of the Constrained and QEA approaches}
+
+The convergence analysis has been performed for the constrained approach to this multiscale system.
+We now look at the differences between the constrained and QEA posterior distributions.
+Recall that the approaches differed only in the form of the effective degradation rate $\hat{k}_4$,
+\[
+	\hat{k}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{k}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
+\]
+This difference in the denominator causes a shift in the parameters as can be seen in Figure~\ref{fig:chem_diff}. The figure shows the difference in posteriors,
+\begin{equation}\label{eqn:chem_diff}
+	\text{diff}(\mu^{\text{CMA}}, \mu^{\text{QEA}}) = \pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
+\end{equation}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.9\textwidth]{"images/Diff_CMA_QEA"}
+\caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
+\label{fig:chem_diff}
+\end{figure}
+
+Since the two posteriors have been approximated using an MCMC sample, there is a significant amount of noise, particularly in the tails of the distributions We can see that the differences in the marginals for $k_1, k_2,$ and $k_3$ are relatively small and the differences are largely positive.
+This means that the constrained approach is slightly more informative about the values of these first three parameters.
+However the maginal for $k_4$ varies by a significant amount and is more negative.
+This means that the mean estimates for $k_4$ vary by a large amount, and we should be more confident in the solution given by the QEA.
+
+% compare solution with full data.
+We now consider how we should interpret the information given by these models.
+The QEA assumption tells us that we cannot observe the parameters $k_2$, $k_3$ and $k_4$ indendently from the reduced data set, but we can observe the quantity $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$.
+Similarly, the constrained approach tells us that we are able to observe the quantity $\hat{k}_4^{\text{CMA}} = k_2k_4/(k_2+k_3+k_4)$.
+To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model.
+
+A conventional way to compare two models under a Bayesian framework is to calculate the Bayes factors~\cite{chen2012monte}.
+The Bayes factor, $B_{1,2}$, between two models, $\mathcal{M}_1$ and $\mathcal{M}_2$, can be interpreted as a ratio of the normalisation constants of the posterior distributions given each model,
+\[
+	B_{1,2} = \frac{\mathbb{P}(D|\mathcal{M}_1)}{\mathbb{P}(D|\mathcal{M}_2)}, \quad \text{where} \quad \mathbb{P}(D|\mathcal{M}_k) = \int_\mathcal{X} \mathbb{P}(D|\theta_k,\mathcal{M}_k)\mathbb{P}(\theta_k|\mathcal{M}_k) \, \text{d}\theta_k.
+\]
+Under the PAIS framework, it is straightforward to calculate these factors using the Monte Carlo estimator for the normalisation constants.
+From~\cite{robert2013monte} these normalisation constants take the form
+\[
+	Z_k \approx \frac{1}{NM}\sum_{i=1}^N\sum_{j=1}^M w_j^{(i)}(\mathcal{M}_k),
+\]
+where $w_j^{(i)}(\mathcal{M}_k)$ is the weight under model $\mathcal{M}_k$ corresponding to the $j$-th ensemble member on the $i$-th iteration.
+Hence, $B_{1,2} = Z_1/Z_2$.
+We can compare more than two models in this way by selecting the model with the largest marginal distribution as the best model.
+
+We now label the constrained model as $\mathcal{M}_1$, the model arising from the QEA as $\mathcal{M}_2$, and the full data model as $\mathcal{M}_0$.
+Under $\mathcal{M}_1$, the parameter $\theta_1 = (k_1, \hat{k}_4^{\text{CMA}})^\top$, and under $\mathcal{M}_2$, the parameter $\theta_2 = (k_1, \hat{k}_4^{\text{QEA}})^\top$.
+The full model $\mathcal{M}_0$ observes all four parameters $\theta_0 = (k_1, k_2, k_3, k_4)^\top$.
+The marginal densities for the data, evaluated at the observed data, given the models are displayed in Table~\ref{tab:chem_Bayes_marginals}.
+
+\begin{table}[!htb]
+\centering
+\begin{tabular}{rr}
+	\toprule
+	$k$ & \quad $\mathbb{P}(D = (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$ \\ \cmidrule(lr){1-2}
+	0 & 6.8e-3 \\
+	1 & 3.2e-3 \\
+	2 & 1.7e-3 \\ \bottomrule
+\end{tabular}
+\caption{Marginal distributions for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:multiscale_posterior}.}
+\label{tab:chem_Bayes_marginals}
+\end{table}
+
+Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals} we see that we should of course prefer the model in which we observe all reactions perfectly, however this model is not significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$, ~\cite{kass1995bayes}).
+Again the constrained model is not substantially more attractive than the QEA model when we consider the Bayes factor $B_{1,2} = 1.96 < 3.2$.
+The Bayes factor $B_{0,2} = 4.1 > 3.2$ does tell us that we should significantly prefer the full model to the QEA model.
+These Bayes factors present a weak argument that that the constrained model provides us with a better description of the data than the QEA model.
+We can also present this information graphically, which might provide us with a greater justification for preferring the constrained model.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/QEA_approx_marginal"}}
+\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/CMA_approx_marginal"}}
+\caption{Comparison of the approximate marginal densities for the `observable parameter' $\hat{k}_4$ under models $\mathcal{M}_1$ and $\mathcal{M}_2$.
+Marginals for these two parameters are approximated using samples which have been produced by targeting the three densities $\pi_i = \mathbb{P}(\theta_i|D,\mathcal{M}_i)$ for $i = 0, 1, 2$.}
+\label{fig:chem_model_comp}
+\end{figure}
+
+Figure~\ref{fig:chem_model_comp} displays the marginal distributions for the parameters $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$.
+For each of these two observable parameters we obtain approximations for three marginal distributions, $\pi_i(\cdot|D, \mathcal{M}_i)$, $i = 0, 1, 2$.
+Each of these approximations has been produced by marginalising a sample drawn from the full posterior $\pi_i(\mathbf{k}|D, \mathcal{M}_i)$ defined in terms of the respective model $\mathcal{M}_i$.
+In other words, to calculate $\hat{\pi}_1(\hat{k}_4^{\text{QEA}}|D)$, we first draw a sample using the PAIS algorithm targeting the posterior density $\pi_1(\mathbf{k}|D,\mathcal{M}_1)$.
+We then calculate $\theta^{(j)} = k_2^{(j)}k_4^{(j)}/(k_2^{(j)}+k_3^{(j)})$ for each sample produced.
+Finally we produce a histogram from this sample $\theta^{(j)}$.
+
+The first thing to note is that in subfigure (a) when we approximate the marginals for $\hat{k}_4^{\text{QEA}}$ the density $\hat{\pi}_2(\cdot|D,\mathcal{M}_2)$ is much more peaked than $\hat{\pi}_1(\cdot|D,\mathcal{M}_1)$.
+As we might expect the same is true in reverse in subfigure (b).
+What is interesting here is that for both parameters, the marginal distribution which assumes $\mathcal{M}_2$ assigns a small density value to the true value of the parameter, while the marginals which assume $\mathcal{M}_1$ assign a relatively high density to the truth.
+We also draw attention to the similarities between the constrained model and the full data model.
+In subfigure (a) these two marginals are peaked around a similar value, and in subfigure (b) the full data and constrained models almost exactly coincide.
+This supports the hypothesis that the constrained approach provides a more accurate approximation to the dynamics of the full system.
 

### 2017-07-11 TPAIS.tex
Added some more stuff to the beginning of Section 2

@@
 
-In this section we give details on how \cite{parno2014transport} propose to couple the target,
-$\mu_{\theta}$, with the reference distribution, $\mu_r$.
-When sampling, ideally we would like to draw a sample directly from the target distribution.
-When we have the exact map, $T\in\mathcal{T}$, $T\colon \mathcal{X}\rightarrow\mathbb{R}^d$, we are able to map samples from the posterior distribution onto a reference density, $\pi_r$ which is a standard Gaussian.
-We can then draw a proposal sample from $\pi_r$, before mapping these samples back onto target space using $T^{-1}$.
+In this section, we give details on how we have adapted the method
+proposed in \cite{parno2014transport} to couple the target,
+$\mu_{\theta}$, with the reference distribution, $\mu_r$. In many
+applications ***REFS*** inverse problems arise which lead to posterior
+distributions from which it is challenging to sample efficiently. This
+can be due to multimodality, strong correlations between parameters,
+or vastly different scales of sensitivity in different directions, for
+example the concept of sloppy models. Standard random walks become
+inefficient in this context since the proposal distributions do not
+take into account these types of challenging structures in the
+density, leading to a large number of proposals being made in low
+density regions.
+
+One way that this can be tackled is to implement more complex MCMC
+methods which are able to explore the state space more
+efficiently. For example the Metropolis-Adjusted Langevin Algorithm (MALA) or Hamiltonian Monte
+Carlo (HMC) methods use gradient information about the target density
+in order to make proposals in better directions that RWMH. Riemannian
+manifold MCMC methods can use additionally approximations of the
+Hessian of the target in order to make very large moves with a good
+probability of being accepted.
+
+Another approach is the simplification of the posterior density
+itself. In \cite{parno2014transport}, the authors proposed the
+approximation of an invertible map which exactly couples the postieror distribution
+$\mu_{\theta}$ to a reference measure $\mu_r$. Given such a map,
+$T\in\mathcal{T}$, $T\colon \mathcal{X}\rightarrow\mathbb{R}^d$, we can then draw a proposal sample from $\pi_r$, before mapping these samples back onto target space using $T^{-1}$.
 These proposed samples are distributed according to the target distribution.
@@
 \end{equation}
-which is equal to the target density when $T$ is exact.
-Clearly we require $T$ to be monotonic and have continuous first derivatives, which means that our space of feasible transference plans, $\mathcal{T}^\uparrow$, may not include the exact map.
-Since it is unlikely that $T\in\mathcal{T}^\uparrow$, we are motivated to formulate an optimisation problem as discussed in Section~\ref{sec:intro_optim}.
+which is equal to the target density when $T$ is exact. Clearly we
+require $T$ to be monotonic and have continuous first derivatives,
+which means that our space of feasible transference plans,
+$\mathcal{T}^\uparrow$, may not include the exact map. Since it is
+unlikely that $T\in\mathcal{T}^\uparrow$, we are motivated to
+formulate an optimisation problem as discussed in
+Section~\ref{sec:intro_optim}.  ****PAUL, what is
+$\mathcal{T}^\uparrow$, what is in Section~\ref{sec:intro_optim}*****
 

### 2018-01-17 TPAIS_2018.tex
Starting new manuscript file
Starting new file, and using the final version of Paul’s thesis work

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks,soul}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{tikz,hhline}
\usepackage{tabularx}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\setlength{\parindent}{0in}
\usepackage{amsmath,amsfonts,amsbsy,amssymb}
\newcommand{\RARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longrightarrow}^{#3}\; #2}
\newcommand{\RARRlong}[3]{#1
  \;\displaystyle\mathop{-\!\!\!-\!\!\!-\!\!\!-\!\!\!-\!\!\!\!\displaystyle
  \longrightarrow}^{#3}\; #2}
\newcommand{\LARR}[3]{#1
  \;\displaystyle\mathop{\displaystyle\longleftarrow}^{#3}\; #2}
\newcommand{\LRARR}[4]{{\mbox{ \raise 0.4 mm \hbox{$#1$}}} \;
  \mathop{\stackrel{\displaystyle\longrightarrow}\longleftarrow}^{#3}_{#4}
  \; {\mbox{\raise 0.4 mm\hbox{$#2$}}}}
\newcommand{\bX}{{\bf X}}
\newcommand{\vecx}{{\mathbf x}}
\newcommand{\vecy}{{\mathbf y}}
\newcommand{\vecz}{{\mathbf z}}
\newcommand{\vecq}{{\mathbf q}}
\newcommand{\bs}{{\mathbf s}}
\newcommand{\vecr}{{\mathbf r}}
\newcommand{\vecX}{{\mathbf X}}
\newcommand{\vecv}{{\mathbf v}}
\newcommand{\tick}{\ding{52}}
\newcommand{\cross}{\ding{54}}
\newcommand{\vecn}{{\mathbf n}}
\newcommand{\vecp}{{\mathbf p}}
\newcommand{\cT}{{\mathcal T}}
\newcommand{\dt}{{\mbox{d}t}}
\newcommand{\dx}{{\mbox{d} \vecx}}
\newcommand{\boldnu}{{\boldsymbol \nu}}
\newcommand{\er}{{\mathbb R}}
\renewcommand{\div}{{\rm div}}
\newcommand{\bnu}{{\bf \nu}}
\newcommand{\divergence}{\mathop{\mbox{div}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\FP}{P_{\rm{FP}}}
\newcommand{\ME}{P_{\rm{ME}}}
\newcommand{\MEs}{P_{\rm{ME}_{S}}}
\newcommand{\D}{\mathcal{D}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\X}{{\mathbf X}}
\newcommand{\Y}{{\mathbf Y}}
\newcommand{\W}{{\mathbf W}}
\newcommand{\data}{D}
\newcommand{\neff}{n_{\text{eff}}}
\newcommand{\E}{{\mathbb E}}
\renewcommand{\b}[1]{{\bf #1}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\picturesAB}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}
\newcommand{\picturesCD}[6]{
\centerline{
\hskip #4
\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
\hskip #5
\epsfig{file=#1,height=#3}
\hskip #6
\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
\hskip #5
\epsfig{file=#2,height=#3}
}}

\makeatletter  
\newcommand{\xleftrightarrows}[2][]{\mathrel{%  
 \raise.40ex\hbox{$  
       \ext@arrow 3095\leftarrowfill@{\phantom{#1}}{#2}$}%  
 \setbox0=\hbox{$\ext@arrow 0359\rightarrowfill@{#1}{\phantom{#2}}$}%  
 \kern-\wd0 \lower.4ex\box0}}

\newcommand{\xrightleftarrows}[2][]{\mathrel{%  
 \raise.40ex\hbox{$\ext@arrow 3095\rightarrowfill@{\phantom{#1}}{#2}$}%  
 \setbox0=\hbox{$\ext@arrow 0359\leftarrowfill@{#1}{\phantom{#2}}$}%  
 \kern-\wd0 \lower.4ex\box0}}

\def\leftrightarrowfill@{%
 \arrowfill@\leftarrow\relbar\rightarrow%
 }
\newcommand*{\centerfloat}{%
  \parindent \z@
  \leftskip \z@ \@plus 1fil \@minus \textwidth
  \rightskip\leftskip
  \parfillskip \z@skip}
\makeatother
\makeatother

\newcommand\irregularcircle[2]{% radius, irregularity
  \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
  +(0:\len pt)
  \foreach \a in {10,20,...,350}{
    \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
    -- +(\a:\len pt)
  } -- cycle
}

\newcommand{\into}{\operatornamewithlimits{\longrightarrow}}
\newtheorem{dfn}{Definition}[section]

\newcommand{\ltri}{%
\,\resizebox{!}{0.25\baselineskip}{%
\begin{tikzpicture}%
\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
\end{tikzpicture}%  
}\xspace%
}%

\newcommand{\smallltri}{%
\,\resizebox{!}{0.15\baselineskip}{%
\begin{tikzpicture}%
\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
\end{tikzpicture}%  
}\xspace%
}%

\title{Transport map accelerated-PAIS, and application to inverse problems arising from
  multiscale stochastic reaction networks}
\author{Simon Cotter, Yannis Kevrekidis, Paul Russell}
\begin{document}
\maketitle
\begin{abstract}
In many applications, inverse problems arise where where there are
complex correlations between the different parameters which we wish to
infer from data. The correlations often manifest themselves as lower
dimensional manifolds on which the likelihood function is
invariant, or varies very little. This can be due to trying to infer
unobservable parameters, or due to sloppiness in the model which is
being used to describe the data. In such a situation, standard
sampling methods for characterising the posterior distribution which
do not incorporate information about this structure will be highly
inefficient. Moreover, most methods are inherently serial in nature,
and as such are not expoiting the parallelised  nature of modern
computer infrastructure. In this paper, we seek to develop a method to
tackle this problem, using optimal transport maps to simplify
posterior distributions which are concentrated on lower dimensional
manifolds.

We demonstrate the approach by considering inverse problems arising
from partially observed stochastic reaction networks. In particular,
we consider systems which exhibit multiscale behaviour, but for which
only the slow variables in the system are observable. We demonstrate
that certain multiscale approximations lead to more consistent
approximations of the posterior than others.
\end{abstract}

\section{Introduction}
%Topics to cover: parallel MCMC, pMC, PAIS
%Transport maps, transport map MCMC
%Stochastic reaction networks
%Multiscale approximations, QSSA/QEA, CMA

In Section \ref{sec:map} we show how an appropriate transport map can
be constructed from importance samples which maps the posterior close
to a reference Gaussian measure. In Section \ref{sec:TPAIS} we show
how such a map can be incorporated into a sophisticated parallel MCMC
infrastructure in order to accelerate mixing. In Section
\ref{sec:multi} we consider how likelihoods can be approximated using
 multiscale methodologies in order to carry out inference for
multiscale and/or partially observed stochastic reaction networks. In
Section \ref{sec:num} we present some numerical examples, which serve
to demonstrate the increased efficiency of the described sampling
methodologies, as well as investigating the posterior approximations
discussed in the previous section. We conclude with a discussion in
Section \ref{sec:conc}.

\section{Construction of transport maps in importance sampling} \label{sec:map}
%Follow the other papers, show adaptation(s)

In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
distribution to the posterior distribution, the idea being that one could draw a moderately sized
sample from the prior distribution and use this sample to approximate a map onto the target space.
Once this map was known to the desired accuracy a larger sample from the prior could be used to
investigate the posterior distribution. This
methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
distribution, the map transforms a sample from the posterior onto a reference space.
The reference density is chosen to allow efficient proposals using a simple proposal
distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.

Proposing new states in this way allows us to make large steps around complex probability distributions.
It is also feasible in this framework to assume that the reference density is close enough to a standard Gaussian that we can efficiently propose moves using a proposal distribution which is independent of the current state, e.g. choose $q(\theta) = \mathcal{N}(0,I_n)$.

In this Section we outline the methodology in \cite{parno2014transport} for coupling the target,
$\mu_{\theta}$, with the reference distribution, $\mu_r$, and
show how the map can be constructed using a weighted sample
and hence how we can incorporate the map into importance sampling schemes.

\begin{dfn}[(Exact) Transport Map $T$]
	An (exact) transport map $T$ is a function $T\colon
        \mathcal{X}\rightarrow\mathbb{R}^d$ such that the {\it
          pullback} of the reference measure with density $\phi(\cdot)$,
	\begin{equation}\label{eq:pullback}
		\tilde{\pi}(\theta) = \phi(T(\theta))|J_T(\theta)|,
	\end{equation}
	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
	\[
		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
	\]
\end{dfn}

There are infinitely many such maps, a subset of which will be
invertible. In the case that we have an exact invertible map,
$T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible
maps, we are able to draw a sample from $\pi_r = \phi$, the density of
the reference distribution, which could be picked to be the
standardised Gaussian distribution for example, and map these samples back
onto target space using $T^{-1}$. These proposed samples are then distributed according to the target distribution.

\begin{dfn}[Target and Reference Space]
	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target measure $\mu_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference measure $\mu_r$.
\end{dfn}

On target space, the proposal density is induced by the pullback of $\phi$ through $T^{-1}$.
Clearly the pullback only exists when $T$ is monotonic, i.e. has a positive definite Jacobian, and has continuous first derivatives.
Not all maps satisfy these conditions, so we define a smaller space of maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all feasible maps.
An exact map $T$ is not necessarily in $\mathcal{T}^\uparrow$, so we
are motivated to formulate an optimisation problem to find the map $T
\in \mathcal{T}^\uparrow$ which most closely maps the target density
to the reference density.

We attempt to find the deterministic coupling of two continuous
probability distributions, $(\mu_\theta, \hat{\mu_r})$, such that
$\hat{\mu_r} = T\mu_\theta$ where the distance between $\mu_r$ (the
desired reference measure) and $\hat{\mu_r}$ (the achieved reference
measure) is minimised, for $T \in \mathcal{T}^\uparrow$. As in
\cite{parno2014transport} we aim to minimise the Kullback-Liebler (KL)
divergence between the density of $\mu_{\theta}$ and the pullback of
the density of $\mu_{r}$, i.e. the distance between $\pi(\theta)$ and
$\tilde{\pi}(\theta)$. For two absolutely continuous measures with
densities $\pi_1$ and $\pi_2$ respectively, the KL
divergence is given by
\[D_\text{KL}(\pi_1\|\pi_2) = 
		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\]
The KL divergence is not itself a norm, since
it is not symmetric, i.e. $D_\text{KL}(\pi_1\|\pi_2) \neq
D_\text{KL}(\pi_2\|\pi_1)$ in general. However, it is still a useful
measure of the similarity of two probability distributions, not least since the
square root of the KL-divergence is an upperbound to the Hellinger
distance metric.

As in previous work in \cite{parno2014transport}, when we optimise the cost function
\[
	C(T) = D_\text{KL}(\pi\|\tilde{\pi}),
\]
to ensure invertibility we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
\[
	T(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
\]
where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We assume that the target and reference probability densities are absolutely continuous on
$\mathbb{R}^d$. Under this formulation we are guaranteed a unique invertible map $\tilde{T}$ with the property that
\[
	\tilde{T}\mu_{\theta} \approx \mu_r.
\]
Relaxing the equality constraint to finding the approximate map
$\tilde{T} \in \mathcal{T}^{\ltri}$ which minimises $C(T)$, gives us a
practical route to finding a good candidate map.

\subsection{The optimisation problem}

With these constraints in mind we formulate the optimisation problem explicitly. The cost function
is chosen to be the Kullback-Leibler divergence between the posterior density and the pullback density,
\[
	D_\text{KL}(\pi\|\tilde{\pi}) =
		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
\]
This divergence results in some nice properties which we will explore in the following derivation. The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to it's relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
\[
	D_{KL}(p\|q) \geq D_H^2(p,q),
\]
which is a true metric between probability distributions $p$ and $q$.
Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
\[
	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
		\log\left|J_{\tilde{T}}(\theta)\right|\right].
\]
We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function. This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown. However this is exactly the scenario for which we would turn to MCMC methods for a solution.

To find the best coupling, $\tilde{T} \in \mathcal{T}^{\ltri}$, we solve the optimisation problem,
\[
	\tilde{T} = \arg\min_{T \in \mathcal{T}^{\smallltri}} \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right]
\]
which has a unique solution since the cost function is convex.

We also include a regularisation term which is required for reasons which will become clear later. The optimisation problem now takes the form
\begin{equation}\label{eq:gen_map_optim}
	\tilde{T} = \arg\min_{T\in\mathcal{T}^{\smallltri}} \left[
		 \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
		\log\left|J_T(\theta)\right|\right] + \beta\mathbb{E}(T(\theta)- \theta)^2 \right].
\end{equation}
This parameter $\beta$ does not need to be tuned, experimentation has shown that the choice
$\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
closer to the identity.

\subsection{The structure of the map}

Before we continue with the derivation of the optimisation problem, we consider the structure
of the map in more detail. The lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density, as well as the inverse of the map, $\tilde{T}^{-1}$. The Jacobian of $\tilde{T}$ is a lower triangular matrix,
\[
	J_T(\theta) = \begin{bmatrix}
		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & \partial_{\theta_d} \tilde{T}_1(\theta)\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
	\end{bmatrix} = \begin{bmatrix}
		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & 0\\
		\vdots & \ddots & \vdots \\
		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
	\end{bmatrix}
\]
since $\partial_{\theta_n} \tilde{T}_k(\theta) = 0$ for all $n > k$. This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
\begin{equation}\label{eqn:separable_jacobian}
	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta).
\end{equation}
Here we note that this term is separable in terms of the dimension $i$.

%Inverting $T$ is simplified by the lower triangular structure in the sense that in each dimension, $i$, we need only invert a univariate polynomial. This simplification comes about because we invert each dimension of the map independently, beginning with $T_1$ which is a function of only $\theta_1$. Finding the root of $T_1(\theta_1) = r_1$, gives us the value of $\theta_1$ which we can then use to find the root of $T_2(\theta_2;\theta_1) = r_2$, and so on. These roots, $r_1,\dots,r_d$, are restricted to be real since $\mathcal{X} \subset \mathbb{R}^d$, which allows us to find the unique point, $\theta$, which corresponds to the point $r$.

Inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map. The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, so we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$. This inversion tells us the value of $\theta_1$, which means the next component is again a univariate polynomial, $T_2(\theta_2; \theta_1)=r_2$. We can then perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.

We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions. Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not necessarily orthogonal with respect to the target distribution. Each component of the map is defined as a multivariate polynomial expansion,
\begin{equation}\label{eq:map_defn}
	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
\end{equation}
The parameter $\gamma_i$ is a vector of coefficients in $\mathbb{R}^{M_i}$. Each component of $\gamma_i$ corresponds to a basis function
$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$. These multi-indices are elements of the multi-index set $\mathcal{J}_i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
\[
	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
\]
and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$. Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.

The cardinalities of the multi-index sets, $M_i = \text{card}(\mathcal{J}_i)$, give the number of unknowns in our
optimisation problem, and so we would like to keep this number as small as possible. One option is
to use polynomials of total order $p$,
\[
	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
\]
which is optimal in terms of the amount of information captured by the map about the target. The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = {i+p \choose p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$. Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$. These index sets are discussed
in~\cite{parno2014transport}. Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}. Here we stick with polynomials of total order $p$ since we work with low dimensional problems with the PAIS algorithm.

\subsection{Implementation of the optimisation problem}\label{sec:transport_implementation}

We now discuss how we can evaluate Equation~\eqref{eq:gen_map_optim} using a sample from the target distribution. We first reformulate the expectation in the cost functional in terms of a MC estimator,
\begin{align}
	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
			\beta\mathbb{E}(T(\theta)-\theta)^2 \notag \\
		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. \label{eqn:TM_full_cost}
\end{align}
Optimisation of this cost function results in a map from $\pi$ to some reference density $\pi_r$. By choosing the reference density to be a Gaussian density, we can simplify this expression greatly. Substitution of the Gaussian density into Equation~\eqref{eqn:TM_full_cost} leads to
\begin{equation}\label{eq:gauss_map_optim}
	C(T) = \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[\frac{1}{2}
		T_i^2(\theta^{(k)}) - \log\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}) +
		\beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right].
\end{equation}
Note that since we assume that the map is monotonic, the derivatives of each component are
positive and so this functional is always finite. In practice it is infeasible to enforce this condition across the whole parameter space. We instead enforce this condition by ensuring that the derivatives are positive at each sample point. This means that when we sample away from these support points while in reference space, it is possible to enter a region of space where the map is not monotonic.

We now return to the structure of the map components given in Equation~\eqref{eq:map_defn}. Since the basis functions are
fixed, the optimisation problem in \eqref{eq:gen_map_optim} is really over the map components $\bar{\gamma} = (\gamma_1, \dots,
\gamma_d)$ where $\gamma_i \in \mathbb{R}^{M_i}$. Note that $C(T)$ is the sum of $d$ expectations, and these expectations each only concern one dimension. Therefore we can rewrite \eqref{eq:gen_map_optim} as $d$ separable optimisation problems.
\begin{align}\label{eq:gamma_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{K}\sum\limits_{k=1}^K
		\left[\frac{1}{2}T_i^2(\theta^{(k)}; \gamma_i) - \log\frac{\partial T_i}{\partial\theta_i}(\theta^{(k)}; \gamma_i) + \beta(T_i(\theta^{(k)};
		\gamma_i)-\theta^{(k)})^2\right], \\
	&\text{subject to} \quad \frac{\partial T_i}{\partial\theta_i}(\theta^{(k)};
		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,d.
		\notag
\end{align}
The sum in Equation~\eqref{eq:map_defn} is an inner
product between the vector of map coefficients, and the evaluations of the basis function at a
particular $\theta^{(k)}$. If we organise our basis evaluations into two matrices,
\[
	(F_i)_{k,\mathbf{j}} = \psi_\mathbf{j}(\theta^{(k)}), \quad \text{and} \quad (G_i)_{k,\mathbf{j}} =
\frac{\partial\psi_\mathbf{j}}{\partial\theta_i}(\theta^{(k)}),
\]
for all $\mathbf{j}
\in \mathcal{J}_i^\text{TO}$, and $k = 1,\dots,K$, then we have that
\[
	T_i(\theta^{(k)}) = (F_i)_{k\cdot}\gamma_i \quad \text{and} \quad \frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}; \gamma_i) = (G_i)_{k\cdot}\gamma_i,
\]
so \eqref{eq:gamma_map_optim} becomes
\begin{align}\label{eq:blas_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2}(F_i\gamma_i)^\top(F_i\gamma_i) -
		c^\top\log(G_i\gamma_i) + \beta\sum\limits_{k=1}^K \!
		(F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}), \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
In this expression, the vector $c$ is a $K\times 1$ vector of ones, and $\log(G_i\gamma_i)$ is to be
evaluated element-wise. As the Monte Carlo simulations advance, new rows can be appended to the
$F_i$ and $G_i$ matrices, and $F_i^\top F_i$ can be efficiently updated via the addition of rank-1 matrices.

The regularisation term in Equation~\eqref{eq:blas_map_optim} can be approximated using Parseval's identity,
\[
	\sum\limits_{k=1}^K \! (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}) \approx
		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\mu_\theta =
		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
\]
where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\mu_\theta$; however this
approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples. If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.

These simplifications result in the efficiently implementable, regularised optimisation problem for
computing the map coefficients,
\begin{align}\label{eq:final_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2}\gamma_i^\top F_i^\top F_i\gamma_i -
		c^\top\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
This optimisation problem can be efficiently solved using Newton iterations. It is suggested
in~\cite{parno2014transport} that this method usually converges in around 10-15 iterations, and we
have seen no evidence that this is not a reasonable estimate. When calculating the map several times
during a Monte Carlo run, using previous guesses of the optimal map to seed the Newton algorithm
results in much faster convergence, usually taking only a couple of iterations to satisfy the stopping
criteria.

\subsection{Implementation of the optimisation problem in PAIS}

In the PAIS algorithm, we use weighted samples to approximate the posterior rather than equally weighted samples. Fortunately, the majority of the derivation of this cost function follows unchanged. We look at the importance sampling Monte Carlo estimate of $C(T)$, compared with
Equation~\eqref{eq:gauss_map_optim},
\begin{equation}\label{eqn:TPAIS_objective}
	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
\end{equation}
here $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
all these weights. This necessitates a minor alteration to the optimisation problem in
Equation~\eqref{eq:final_map_optim},
\begin{align}\label{eq:weighted_map_optim}
	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \frac{\beta}{K}\|\gamma_i-\iota\|^2, \\
	&\text{subject to} \quad G_i\gamma_i > 0. \notag
\end{align}
We introduce a diagonal matrix $W = \text{diag}(w)$, where $w = (w_1,\dots,w_K)^\top$ into the log of the reference density, and replace the ones vector, $c$, with $w$. Analogously to the unweighted case, we include a $1/K$ to the regularisation term so that it has less influence as we obtain more samples. If we allowed this regularisation term to exert influence at all updates we would converge to a suboptimal map. We should converge to the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation. The weights are strictly
positive resulting in a positive definite matrix $W$. This means that the Hessian of the objective
function is still positive definite, and so the optimisation problem remains convex. It is important in implementation to ensure that any weights which are numerically zero are dealt with to ensure this positive definiteness.

This optimisation problem can be efficiently solved using the Newton optimisation algorithm. The Hessian takes the
form
\begin{equation}\label{eqn:TPAIS_hessian}
	HC_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i + G_i^\top
		W\text{diag}([G_i\gamma_i]^{-2})G_i\right] + \beta I,
\end{equation}
where $[G_i\gamma_i]^{-2}$ is to be taken element-wise, and $I$ is the $M_i\times M_i$
identity matrix. The first derivative of $C_i(T)$ is
\[
	\nabla C_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i\gamma_i - G_i^\top
		W[G_i\gamma_i]^{-1}\right] + \beta(\gamma_i - \iota),
\]
again $[G_i\gamma_i]^{-1}$ is taken element-wise.

\section[Transport map MCMC]{Using the transport map to propose states in MCMC algorithms}\label{sec:transport_algs}

Given samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
target measure to a reference measure. We now consider how to implement an MCMC algorithm which uses
these maps to propose new states. These algorithms fit into the adaptive framework described in
Section~\ref{sec:adapt} since we use the full history of the chain when updating the map, and this update changes the shape of our proposal
distribution. Convergence of this adaptation is shown in~\cite{parno2014transport}.

\subsection{Transport map MH algorithms}

As introduced in~\cite{parno2014transport}, a MH algorithm using the transport map to propose new
states can be designed, as given in Algorithm~\ref{alg:TransportMH}. These algorithms
map a sample from the target onto the reference space, a new state is then efficiently
proposed using a Gaussian kernel and is mapped back onto target space. This can
result in rapid mixing, where step size and direction are now a function of the current particles location in state space, similar to the ideas of Riemann manifold Hamiltonian Monte Carlo~\cite{girolami2011riemann}.

\begin{table}
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Initialise state $\theta^{(1)} = \theta_0$.\;
Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
\For{$k \leftarrow 1, \dots, L-1$}{
	Compute $r = \tilde{T}(\theta^{(k)}; \bar{\gamma}^{(k)})$.\;
	Sample $r' \sim q_r(\cdot; r)$.\;
	Invert $\theta' = \tilde{T}^{-1}(r'; \bar{\gamma}^{(k)})$.\;
	Calculate:
	\[
		\alpha = 1 \wedge
			\frac{\pi(\theta')}{\pi(\theta^{(k)})}\frac{q_r(r|r')|J_{\tilde{T}}(\theta^{(k)};
			\bar{\gamma}^{(k)})|}{q_r(r'|r)|J_{\tilde{T}}(\theta';\bar{\gamma}^{(k)})|}.
	\]

	Sample $u \sim U[0,1]$.\;
	Set $\theta^{(k+1)}$ to $\theta'$ with probability $\alpha$, otherwise $\theta^{(k+1)} =
		\theta^{(k)}$.

	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
		\For{$i \leftarrow 1, \dots, n$}{
			Solve \eqref{eq:final_map_optim} with $\{\theta^{(1)}, \dots, \theta^{(k+1)}\}$ and
				update $\gamma_i^{(k+1)}$.\;
		}
	}{
		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
	}
}
\caption{MH algorithm with adaptive transport map~\cite{parno2014transport}\label{alg:TransportMH}}
\end{algorithm}
\end{table}

Initially we begin with the identity map, $\iota$, while we build enough samples to produce an approximate
map. For simplicity, we update the map at uniform intervals of $K_U$ iterations, given an initial
burn-in phase. In order to satisfy the conditions for adaptive algorithms to be ergodic, discussed
in Section~\ref{sec:adapt}, this adaptive phase must come to an end. It is sensible to stop
adaptation once the map has sufficiently converged, e.g. when updates do not change the coefficients, $\gamma$,
above some tolerance, or as suggested by~\cite{parno2014transport} when the variance of $D_\text{KL}(\pi\|\tilde{\pi})$ is sufficiently close to zero.

Once the map is sufficiently converged, we can decide whether we would like to continue using a local proposal distribution, or whether to use a distribution which is independent of the current state. An independent proposal is more efficient
if the reference space is close to Gaussian; however, if it is not then certain regions of the target
space could be less likely to be visited than in a standard MH algorithm. A compromise can be found by randomly selecting which kernel to sample from, i.e. the proposal distribution
\[
	q(x, \cdot) = p\mathcal{N}(x, \Sigma)+(1-p)\mathcal{N}(0,\text{I}), \quad p \in [0, 1].
\]

\subsection{Transport map PAIS algorithms}

In a similar way, we can use the Transport map derived in Equation~\eqref{eq:weighted_map_optim} to
design a proposal scheme for the PAIS algorithm. In this case we have a choice in how to proceed; we
propose new samples on reference space and resample on target space, or we both propose and resample on reference space, mapping onto target space to output the samples. The first option allows us to reuse much of the framework
from the standard PAIS algorithm and in the numerics later we see that this performs better than
both the Transport MH algorithm, and the standard PAIS algorithm. The second option requires some
restructuring but results in improved performance from the resampler.

\begin{table}
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
\For{$k \leftarrow 1, \dots, L-1$}{
	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
	Calculate:
	\[
		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
	\]

	Resample $\theta^{(k+1)} \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{\hat{\theta}^{(k)}_j}(\cdot)$.\;

	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
		\For{$i \leftarrow 1, \dots, n$}{
			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
				and update $\gamma_i^{(k+1)}$.\;
		}
	}{
		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
	}
}
\caption{PAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportPAIS1}}
\end{algorithm}
\end{table}

The first option is given in Algorithm~\ref{alg:TransportPAIS1}. We denote the ensembles of states in target space $\theta^{(k)} = \{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble size. Similarly, the proposal states are denoted $r' = \{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) = \{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M, \hat{\theta}^{(k)}_M)\}$, where these pairs are the states which we consider to be our sample from the target distribution. As in the standard version of the PAIS algorithm we use the deterministic mixture weights.

\begin{table}
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
\For{$k \leftarrow 1, \dots, N-1$}{
	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
	Calculate:
	\[
		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
	\]

	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\label{algline:TPAIS_resample}\;
	Invert $\theta^{(k+1)}_i = \tilde{T}^{-1}(r^*_i)$.\;
	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
		\For{$i \leftarrow 1, \dots, n$}{
			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
				and update $\gamma_i^{(k+1)}$.\;
		}
	}{
		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
	}
}
\caption{PAIS algorithm with adaptive transport map. Option 2.\label{alg:TransportPAIS2}}
\end{algorithm}
\end{table}

The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to the first except on Line~\ref{algline:TPAIS_resample} where rather than resampling in target space we resample in reference space. In reference space the dimensions are roughly uncorrelated, and the Gaussian marginals are easy to approximate with fewer ensemble members. This means that the resampling step will be more efficient in higher dimensions, which we discuss in Section~\ref{sec:TPAIS_higher_dim}.

\section[Convergence of transport MCMC]{Convergence of the transport proposal based MCMC algorithms}

In this section we study the convergence of the transport based proposal distributions which we have described in Section~\ref{sec:transport_algs}. We follow a similar strategy to that in Chapter~\ref{sec:PAIS}. We begin by finding optimal values for the scaling parameters by performing a series of simulations with differing scaling parameters. We then produce 1 million samples from the target distribution with the optimal scaling parameter and an ensemble size of $M=150$. We take 32 repeats of this optimal simulation and present the geometric average of the convergence rates. This process is then repeated for each of the algorithms we have discussed, Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2}.

In this section we return to the Rosenbrock banana-shaped density which in Chapter~\ref{sec:LPAIS} we labelled $R_1$. This target density is
\begin{equation}\label{eqn:R2_repeat}
	\pi(\theta) = \frac{\sqrt{10}}{\pi}\exp\left\{ -(1 - \theta_1)^2 - 10(\theta_2 - \theta_1^2)^2 \right\}.
\end{equation}
A contour plot of the target density is given in Figure~\ref{fig:R2_posterior}.
\begin{figure}
\centering
\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_1"}}
\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_2"}}
\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_posterior}}
\caption{Visualisation of the density of example $R_1$, as given in Equation~\eqref{eqn:R2_repeat}.}
\label{fig:R2_posterior}
\end{figure}

\subsection{Implementation details}

We now demonstrate some properties of the transport maps we will be using in our MCMC algorithms. We draw 1 million samples from the density in \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:transport_implementation} to build a transport map. We use this map to push forward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian. We then pull the sample back on to target space using the inverse map to check that our map is invertible and well behaved.

For this example, we use an index set of total order 3 with monomial basis functions. This results in a map of the form
\[
	T(\theta_1, \theta_2) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \end{bmatrix},
\]
where
\begin{align*}
		T_1(\theta_1) &= \gamma_{1,1} + \gamma_{1,2}\theta_1 + \gamma_{1,3}\theta_1^2 + \gamma_{1,4}\theta_1^3, \\
		T_2(\theta_1, \theta_2) &= \gamma_{2,1} + \gamma_{2,2}\theta_1 + \gamma_{2,3}\theta_1^2 + \gamma_{2,4}\theta_1^3
					+ \gamma_{2,5}\theta_2 + \gamma_{2,6}\theta_1\theta_2 \\
				 & \qquad \quad + \gamma_{2,7}\theta_1^2\theta_2 + \gamma_{2,8}\theta_2^2 + \gamma_{2,9}\theta_1\theta_2^2 +
					 \gamma_{2,10}\theta_2^3.
\end{align*}
Clearly even with only basis functions of total order 3, we have a large number of unknowns in our optimisation problem, $\bar{\gamma} \in \mathbb{R}^{14}$. If we were to increase the dimension of $\theta$ further we would need to reduce the number of terms we include in the expansion by, for example, removing all the cross terms. This reduces the quality of our map but since we only require an approximate map we can afford to reduce the accuracy.

\begin{figure}[htpb]
\centering
\subfigure[Original sample $\theta$ from RWMH algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
\caption{Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
\label{fig:R2_transport}
\end{figure}

Figure~\ref{fig:R2_transport} shows the efficiency of the transport map. Even though we have truncated the infinite expansion in the monomial basis down to 4 and 10 terms in respective dimensions, the push forward of the sample is still a unimodal distribution centred at the origin with standard deviation 1. As you move out into the tails of the reference density more non-Gaussian features form, but these are not too much of a problem when using a suitable proposal distribution. The pullback from reference space, in Figure~\ref{fig:R2_transport}, is an exact match of the original sample since we have not perturbed the sample in reference space. This inversion is well defined in the sampling region, although not necessarily outside.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{images/TPAIS/zero_jacobian}
\caption{An example transport map which does not allow exploration of the full target distribution, and has a non-invertible point.}
\label{fig:TM_zero_jacobian}
\end{figure}

If we consider a map built from polynomials of total order $p=2$, it is clear that the map $\tilde{T}$ does not map onto $\mathbb{R}^d$, but on to a subset of it. Figure~\ref{fig:TM_zero_jacobian} illustrates this point. We see that in this dimension, the $\tilde{T}^{-1}$ will not map any proposal $r > 0.8$ to a point in the target space. The point at which no inverse exists for a point $r_1$ can be made arbitrarily far out in the tails of the posterior distribution, but will be finite. The point at which the Jacobian is zero is also a non-invertible point since the logarithm in the pullback becomes infinite. This means that (1) we will never sample from the tail of the target distribution which lies beyond this turning point, and (2) all points in reference space $r < 0.8$ have the possibility of being mapped to two distinct points in target space. We can choose to truncate our parameter space at this turning point which allows our map to be bijective, however we will not be able to sample from the true posterior. For this reason it is best to chose an index set where the maximum order is an odd number.

\subsection[Numerical results]{Numerical results for convergence of transport map based algorithms}

We first find the optimal scaling parameters for the individual algorithms. This is done as before by optimising the effective sample size in the PAIS algorithm, and by tuning the relative $L^2$ error in the MH algorithm. There is currently no guidance on the best way of tuning the MH algorithm with transport map proposals although one might expect results similar to the standard MH results. As in the PAIS algorithm, the effective sample size might be the best option.

\begin{table}[!h]
\centering
\begin{tabular}{lrrr}
\toprule
	Statistic \quad / \quad Algorithm & \ref{alg:TransportMH} & \ref{alg:TransportPAIS1} & \ref{alg:TransportPAIS2}  \\ \cmidrule(lr){1-4}
	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
	$\delta_{\text{ESS}}$				 & - & 1.0e-1 & 5.2e-1 \\ \cmidrule(lr){1-4}
	Acc. rate							 & 0.23 & - & - \\
	ESS ratio							 & - & 0.62 & 0.71 \\
\bottomrule
\end{tabular}
\caption{Optimal scaling parameters for the transport map based algorithms applied to $R_1$.}
\label{tab:R2_opt_scaling}
\end{table}

The optimal scaling parameters are given in Table~\ref{tab:R2_opt_scaling}. Here we see that the effective sample size is much lower than we see in the one-dimensional examples with the PAIS algorithms. However, in $R_1$ we are dealing with a much more complicated correlation structure, as well as a very slowly decaying tail in $\theta_2$. We have seen in Section~\ref{sec:LPAIS_R1} that the standard PAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.

\begin{figure}[!h]
\centering
\subfigure[Comparison of the two Transport PAIS options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
\subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
\caption{Convergence of Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2} for $R_1$. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
\label{fig:R2_l2_convergence}
\end{figure}

The convergence of the three algorithms is displayed in Figure~\ref{fig:R2_l2_convergence}. Figure (a) shows that the two variations of the transport based PAIS algorithms converge with similar rates. The second version, which performs the resampling stage in reference space rather than target space, has a slightly higher ESS, and is more stable than option (1). This version also has a property that we can exploit in Section~\ref{sec:TPAIS_higher_dim}.

\section{Sampling in higher dimensions}\label{sec:TPAIS_higher_dim}

Algorithm~\ref{alg:TransportPAIS2} allows us to decorrelate the dimensions of our random parameter on reference space, where we then can resample and map the resulting ensemble back onto target space. Since, on reference space, the dimensions are uncorrelated, we are able to resample in each dimension separately. Resampling in a single dimension allows for optimisations in resampling code, and also means that the resampler is not affected by the curse of dimensionality.

If we can approximate the posterior well with our mixture and with the transport map, we should not be affected by the increase in dimension to the extent we have been with the standard PAIS-RW algorithm. In one dimension the ETPF algorithm can be implemented very efficiently. As described in~\cite{reich2013nonparametric}, the coupling matrix has all non-zero entries in a staircase pattern when the state space is ordered. We can exploit this knowledge to produce Algorithm~\ref{alg:ETPF_1d}. Which is much faster than using the simplex algorithm to minimise the associated cost function, and faster than the AMR algorithm.

\begin{table}[!htpb]
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
Sort the states, $\{(w_i, x_i)\}_{i=1}^M$, into ascending order.\;
Normalise the weights $p_i = w_i/\|w\|_1$.\;
Set $y_i \leftarrow 0$ for all $i=1,\dots,M$.\;
Set $c \leftarrow 0$\;
\For{$i \leftarrow 1, \dots, M$}{
	Set $t \leftarrow p_i$\;
	\While{$j \leq M$ and $t > 0$}
	{
		Set $s \leftarrow \left(M^{-1}-c\right) \wedge t$\;
		Increase $y_j$ by $M\times s\times x_i$.\;
		Decrease $t$ by $s$.\;
		Increase $c$ by $s$.\;
		\If{$t>0$}
		{
			Increase $j$ by 1.\;
			Set $c \leftarrow 0$.
		}
	}
}
Return $y$.\;
\caption{ETPF algorithm in one dimension.\label{alg:ETPF_1d}}
\end{algorithm}
\end{table}

\section{Transport map-accelerated Parallel Adaptive Importance
  Sampling (TPAIS)}\label{sec:TPAIS}

\section{Multiscale Methods for Stochastic Chemical Reaction
  Networks}\label{sec:multi}

\section{Numerical Examples}\label{sec:num}

\section{Discussion}\label{sec:conc}

\bibliographystyle{siam}
\bibliography{bibliography}

\end{document}

### 2018-01-18 TPAIS_2018.tex
Transport map construction chapter edited

@@
 
-In this Section we outline the methodology in \cite{parno2014transport} for coupling the target,
+In this Section we outline the methodology in
+\cite{parno2014transport} for approximately coupling the target,
 $\mu_{\theta}$, with the reference distribution, $\mu_r$, and
@@
 \begin{dfn}[(Exact) Transport Map $T$]
-	An (exact) transport map $T$ is a function $T\colon
+	A transport map $T$ is a function $T\colon
         \mathcal{X}\rightarrow\mathbb{R}^d$ such that the {\it
@@
 
-There are infinitely many such maps, a subset of which will be
-invertible. In the case that we have an exact invertible map,
-$T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible
-maps, we are able to draw a sample from $\pi_r = \phi$, the density of
-the reference distribution, which could be picked to be the
-standardised Gaussian distribution for example, and map these samples back
-onto target space using $T^{-1}$. These proposed samples are then distributed according to the target distribution.
+% In the case that we have an invertible map,
+% $T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible
+% maps, we are able to draw a sample from $\pi_r = \phi$, the density of
+% the reference distribution, which could be picked to be the
+% standardised Gaussian distribution for example, and map these samples back
+% onto target space using $T^{-1}$. These proposed samples are then distributed according to the target distribution.
 
@@
 
-On target space, the proposal density is induced by the pullback of $\phi$ through $T^{-1}$.
+Armed with such a map, independent samples can be made of the target
+measure, using the pullback of the reference density $\phi$ through $T^{-1}$.
 Clearly the pullback only exists when $T$ is monotonic, i.e. has a positive definite Jacobian, and has continuous first derivatives.
-Not all maps satisfy these conditions, so we define a smaller space of maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all feasible maps.
-An exact map $T$ is not necessarily in $\mathcal{T}^\uparrow$, so we
-are motivated to formulate an optimisation problem to find the map $T
+Not all maps satisfy these conditions, so we define a smaller space of
+maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all
+feasible maps. This space does not necessarily contain an exact
+coupling between target and reference space, and so we are motivated to formulate an optimisation problem to find the map $\tilde{T}
 \in \mathcal{T}^\uparrow$ which most closely maps the target density
@@
 
-We attempt to find the deterministic coupling of two continuous
-probability distributions, $(\mu_\theta, \hat{\mu_r})$, such that
-$\hat{\mu_r} = T\mu_\theta$ where the distance between $\mu_r$ (the
-desired reference measure) and $\hat{\mu_r}$ (the achieved reference
-measure) is minimised, for $T \in \mathcal{T}^\uparrow$. As in
-\cite{parno2014transport} we aim to minimise the Kullback-Liebler (KL)
-divergence between the density of $\mu_{\theta}$ and the pullback of
-the density of $\mu_{r}$, i.e. the distance between $\pi(\theta)$ and
-$\tilde{\pi}(\theta)$. For two absolutely continuous measures with
-densities $\pi_1$ and $\pi_2$ respectively, the KL
-divergence is given by
-\[D_\text{KL}(\pi_1\|\pi_2) = 
-		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\]
-The KL divergence is not itself a norm, since
-it is not symmetric, i.e. $D_\text{KL}(\pi_1\|\pi_2) \neq
-D_\text{KL}(\pi_2\|\pi_1)$ in general. However, it is still a useful
-measure of the similarity of two probability distributions, not least since the
-square root of the KL-divergence is an upperbound to the Hellinger
-distance metric.
+% We attempt to find the deterministic coupling of two continuous
+% probability distributions, $(\mu_\theta, \hat{\mu_r})$, such that
+% $\hat{\mu_r} = T\mu_\theta$ where the distance between $\mu_r$ (the
+% desired reference measure) and $\hat{\mu_r}$ (the achieved reference
+% measure) is minimised, for $T \in \mathcal{T}^\uparrow$. As in
+% \cite{parno2014transport} we aim to minimise the Kullback-Liebler (KL)
+% divergence between the density of $\mu_{\theta}$ and the pullback of
+% the density of $\mu_{r}$, i.e. the distance between $\pi(\theta)$ and
+% $\tilde{\pi}(\theta)$. For two absolutely continuous measures with
+% densities $\pi_1$ and $\pi_2$ respectively, the KL
+% divergence is given by
+% \[D_\text{KL}(\pi_1\|\pi_2) = 
+% 		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\]
+% The KL divergence is not itself a norm, since
+% it is not symmetric, i.e. $D_\text{KL}(\pi_1\|\pi_2) \neq
+% D_\text{KL}(\pi_2\|\pi_1)$ in general. However, it is still a useful
+% measure of the similarity of two probability distributions, not least since the
+% square root of the KL-divergence is an upperbound to the Hellinger
+% distance metric.
 
-As in previous work in \cite{parno2014transport}, when we optimise the cost function
-\[
-	C(T) = D_\text{KL}(\pi\|\tilde{\pi}),
-\]
-to ensure invertibility we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
+As in previous work in \cite{parno2014transport}, we can ensure
+invertibility if we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
 \[
@@
 \]
-where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We assume that the target and reference probability densities are absolutely continuous on
-$\mathbb{R}^d$. Under this formulation we are guaranteed a unique invertible map $\tilde{T}$ with the property that
-\[
-	\tilde{T}\mu_{\theta} \approx \mu_r.
-\]
-Relaxing the equality constraint to finding the approximate map
-$\tilde{T} \in \mathcal{T}^{\ltri}$ which minimises $C(T)$, gives us a
-practical route to finding a good candidate map.
+where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. % We assume that the target and reference probability densities are absolutely continuous on
+% $\mathbb{R}^d$. Under this formulation, with appropriate
+% regularisation (which will be detailed later), we are guaranteed a unique invertible map $\tilde{T}$ with the property that
+% \[
+% 	\tilde{T}\mu_{\theta} \approx \mu_r.
+% \]
+% % Relaxing the equality constraint to finding the approximate map
+% % $\tilde{T} \in \mathcal{T}^{\ltri}$ which minimises $C(T)$, gives us a
+% % practical route to finding a good candidate map.
 
 \subsection{The optimisation problem}
-
-With these constraints in mind we formulate the optimisation problem explicitly. The cost function
-is chosen to be the Kullback-Leibler divergence between the posterior density and the pullback density,
+Our aim is now to find the lower triangular map $\tilde{T} \in
+\mathcal{T}^{\ltri}$ such that the difference between target
+density and the pullback of the reference density is minimised. As in
+\cite{parno2014transport}, we choose the cost function
+to be the Kullback-Leibler (KL) divergence between the posterior density and the pullback density,
 \[
@@
 which has a unique solution since the cost function is convex.
-
-We also include a regularisation term which is required for reasons which will become clear later. The optimisation problem now takes the form
+We also include a regularisation term, which is required for reasons which will become clear later.
+The optimisation problem now takes the form
 \begin{equation}\label{eq:gen_map_optim}
@@
 \end{equation}
-This parameter $\beta$ does not need to be tuned, experimentation has shown that the choice
-$\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
-closer to the identity.
+The parameter $\beta>0$ does not need to be tuned, as experimentation has shown that the choice
+$\beta=1$ is sufficient for most problems. This expectation can be
+approximated by using an MCMC approximation. The form of the penalisation term promotes maps which are
+closer to the identity, and so prevents overfitting when the quality
+or size of the current sample from the posterior is not sufficient.
 
@@
 \begin{equation}\label{eqn:separable_jacobian}
-	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta).
+	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta),
 \end{equation}
-Here we note that this term is separable in terms of the dimension $i$.
-
-%Inverting $T$ is simplified by the lower triangular structure in the sense that in each dimension, $i$, we need only invert a univariate polynomial.
-This simplification comes about because we invert each dimension of the map independently, beginning with $T_1$ which is a function of only $\theta_1$.
-Finding the root of $T_1(\theta_1) = r_1$, gives us the value of $\theta_1$ which we can then use to find the root of $T_2(\theta_2;\theta_1) = r_2$, and so on.
-These roots, $r_1,\dots,r_d$, are restricted to be real since $\mathcal{X} \subset \mathbb{R}^d$, which allows us to find the unique point, $\theta$, which corresponds to the point $r$.
+where we note that this term is separable in terms of the dimension $i$.
 
@@
 \end{equation}
-The parameter $\gamma_i$ is a vector of coefficients in $\mathbb{R}^{M_i}$. Each component of $\gamma_i$ corresponds to a basis function
+The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
 $\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$.
@@
 
-We now discuss how we can evaluate Equation~\eqref{eq:gen_map_optim} using a sample from the target distribution.
-We first reformulate the expectation in the cost functional in terms of a MC estimator,
+We now discuss how we can evaluate the cost function in
+Equation~\eqref{eq:gen_map_optim}. In \cite{parno2014transport}, this
+expectation is approximated using an MCMC estimator, such that
 \begin{align}
@@
 		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
-			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. \label{eqn:TM_full_cost}
+			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. 
 \end{align}
+Here we diverge from previous work, as we aim to build a map from
+samples from an importance sampling scheme. Such samples no longer
+carry equal weight, and as such the Monte Carlo estimator becomes
+\begin{equation}\label{eqn:TM_full_cost}
+	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
+		w_k \left[-\log\pi_r(T_i(\theta^{(k)})) -
+			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
+\end{equation}
+where $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
+all these weights.
 Optimisation of this cost function results in a map from $\pi$ to some reference density $\pi_r$.
@@
 Substitution of the Gaussian density into Equation~\eqref{eqn:TM_full_cost} leads to
-\begin{equation}\label{eq:gauss_map_optim}
-	C(T) = \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[\frac{1}{2}
-		T_i^2(\theta^{(k)}) - \log\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}) +
-		\beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right].
+\begin{equation}\label{eqn:TPAIS_objective}
+	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
+		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
+		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
 \end{equation}
+
 Note that since we assume that the map is monotonic, the derivatives of each component are
@@
 \begin{align}\label{eq:gamma_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{K}\sum\limits_{k=1}^K
-		\left[\frac{1}{2}T_i^2(\theta^{(k)}; \gamma_i) - \log\frac{\partial T_i}{\partial\theta_i}(\theta^{(k)}; \gamma_i) + \beta(T_i(\theta^{(k)};
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{\bar{w}}\sum\limits_{k=1}^K
+		w_k \left[\frac{1}{2}T_i^2(\theta^{(k)}; \gamma_i) - \log\frac{\partial T_i}{\partial\theta_i}(\theta^{(k)}; \gamma_i) + \beta(T_i(\theta^{(k)};
 		\gamma_i)-\theta^{(k)})^2\right], \\
@@
 \begin{align}\label{eq:blas_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2}(F_i\gamma_i)^\top(F_i\gamma_i) -
-		c^\top\log(G_i\gamma_i) + \beta\sum\limits_{k=1}^K \!
-		(F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}), \\
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}}
+          \frac{1}{2}(F_i\gamma_i)^\top W (F_i\gamma_i) -
+		{\bf w}^\top\log(G_i\gamma_i) + \frac{\beta}{\bar{w}}\sum\limits_{k=1}^K \!
+		w_k(F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}), \\
 	&\text{subject to} \quad G_i\gamma_i > 0. \notag
 \end{align}
-In this expression, the vector $c$ is a $K\times 1$ vector of ones, and $\log(G_i\gamma_i)$ is to be
-evaluated element-wise. As the Monte Carlo simulations advance, new rows can be appended to the
-$F_i$ and $G_i$ matrices, and $F_i^\top F_i$ can be efficiently updated via the addition of rank-1 matrices.
+In this expression, the vector ${\bf w} = [w_1, w_2, \ldots,
+w_K]^\top$ is the vector of the weights, $W$ is the diagonal matrix $W
+= \text{diag}(w)$ and $\log(G_i\gamma_i)$ is to be
+evaluated element-wise. As more importance samples are made, new rows can be appended to the
+$F_i$ and $G_i$ matrices, and $F_i^\top W F_i$ can be efficiently updated via the addition of rank-1 matrices.
 
@@
 \[
-	\sum\limits_{k=1}^K \! (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}) \approx
+	\frac{1}{\bar{w}}\sum\limits_{k=1}^K \! w_k
+        (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)})
+        \xrightarrow[K \to \infty]{}
 		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\mu_\theta =
@@
 the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\mu_\theta$; however this
-approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples.
-If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.
+approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples.% If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.
 
@@
 computing the map coefficients, 
-\begin{align}\label{eq:final_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2}\gamma_i^\top F_i^\top F_i\gamma_i -
-		c^\top\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
-	&\text{subject to} \quad G_i\gamma_i > 0. \notag
+\begin{align}\label{eq:weighted_map_optim}
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
+		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
+	&\text{subject to} \quad G_i\gamma_i > 0, \notag
 \end{align}
+%where additionally we have introduced a factor of $\frac{1}{K}$ to the
+%regularisation term to prevent over-smoothing as the number of samples increases.
 This optimisation problem can be efficiently solved using Newton iterations. It is suggested
@@
 
-\subsection{Implementation of the optimisation problem in PAIS}
-
-In the PAIS algorithm, we use weighted samples to approximate the posterior rather than equally weighted samples.
-Fortunately, the majority of the derivation of this cost function follows unchanged.
-We look at the importance sampling Monte Carlo estimate of $C(T)$, compared with
-Equation~\eqref{eq:gauss_map_optim},
-\begin{equation}\label{eqn:TPAIS_objective}
-	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
-		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
-		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
-\end{equation}
-here $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
-all these weights. This necessitates a minor alteration to the optimisation problem in
-Equation~\eqref{eq:final_map_optim},
-\begin{align}\label{eq:weighted_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
-		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \frac{\beta}{K}\|\gamma_i-\iota\|^2, \\
-	&\text{subject to} \quad G_i\gamma_i > 0. \notag
-\end{align}
-We introduce a diagonal matrix $W = \text{diag}(w)$, where $w = (w_1,\dots,w_K)^\top$ into the log of the reference density, and replace the ones vector, $c$, with $w$.
-Analogously to the unweighted case, we include a $1/K$ to the regularisation term so that it has less influence as we obtain more samples.
-If we allowed this regularisation term to exert influence at all updates we would converge to a suboptimal map.
-We should converge to the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation.
-The weights are strictly
-positive resulting in a positive definite matrix $W$. This means that the Hessian of the objective
-function is still positive definite, and so the optimisation problem remains convex.
-It is important in implementation to ensure that any weights which are numerically zero are dealt with to ensure this positive definiteness.
-
-This optimisation problem can be efficiently solved using the Newton optimisation algorithm. The Hessian takes the
+The Hessian takes the
 form
@@
 
-\section[Transport map MCMC]{Using the transport map to propose states in MCMC algorithms}\label{sec:transport_algs}
+\section[Transport map MCMC]{Transport map usage in PAIS
+  and other PMC algorithms}\label{sec:TPAIS}
 
-Given samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
-target measure to a reference measure. We now consider how to implement an MCMC algorithm which uses
-these maps to propose new states. These algorithms fit into the adaptive framework described in
-Section~\ref{sec:adapt} since we use the full history of the chain when updating the map, and this update changes the shape of our proposal
-distribution. Convergence of this adaptation is shown in~\cite{parno2014transport}.
+Given importance samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
+target measure to a reference measure. We now consider how to
+implement an importance sampling-based MCMC algorithm which uses
+these maps to propose new states. In \cite{parno2014transport} it was
+shown how approximate transport maps can be used to accelerate
+Metropolis-Hastings methods, with the map being periodically updated
+with the samples produced from the target measure. Convergence of this
+adaptation is shown in~\cite{parno2014transport}. In this section, we
+will discuss how similarly, these maps can be used to construct highly
+efficient importance sampling schemes.
 
-\subsection{Transport map MH algorithms}
-
-As introduced in~\cite{parno2014transport}, a MH algorithm using the transport map to propose new
-states can be designed, as given in Algorithm~\ref{alg:TransportMH}. These algorithms
-map a sample from the target onto the reference space, a new state is then efficiently
-proposed using a Gaussian kernel and is mapped back onto target space. This can
-result in rapid mixing, where step size and direction are now a function of the current particles location in state space, similar to the ideas of Riemann manifold Hamiltonian Monte Carlo~\cite{girolami2011riemann}.
-
-\begin{table}
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\BlankLine
-Initialise state $\theta^{(1)} = \theta_0$.\;
-Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
-\For{$k \leftarrow 1, \dots, L-1$}{
-	Compute $r = \tilde{T}(\theta^{(k)}; \bar{\gamma}^{(k)})$.\;
-	Sample $r' \sim q_r(\cdot; r)$.\;
-	Invert $\theta' = \tilde{T}^{-1}(r'; \bar{\gamma}^{(k)})$.\;
-	Calculate:
-	\[
-		\alpha = 1 \wedge
-			\frac{\pi(\theta')}{\pi(\theta^{(k)})}\frac{q_r(r|r')|J_{\tilde{T}}(\theta^{(k)};
-			\bar{\gamma}^{(k)})|}{q_r(r'|r)|J_{\tilde{T}}(\theta';\bar{\gamma}^{(k)})|}.
-	\]
-
-	Sample $u \sim U[0,1]$.\;
-	Set $\theta^{(k+1)}$ to $\theta'$ with probability $\alpha$, otherwise $\theta^{(k+1)} =
-		\theta^{(k)}$.
-
-	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
-		\For{$i \leftarrow 1, \dots, n$}{
-			Solve \eqref{eq:final_map_optim} with $\{\theta^{(1)}, \dots, \theta^{(k+1)}\}$ and
-				update $\gamma_i^{(k+1)}$.\;
-		}
-	}{
-		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
-	}
-}
-\caption{MH algorithm with adaptive transport map~\cite{parno2014transport}\label{alg:TransportMH}}
-\end{algorithm}
-\end{table}
-
-Initially we begin with the identity map, $\iota$, while we build enough samples to produce an approximate
-map. For simplicity, we update the map at uniform intervals of $K_U$ iterations, given an initial
-burn-in phase. In order to satisfy the conditions for adaptive algorithms to be ergodic, discussed
-in Section~\ref{sec:adapt}, this adaptive phase must come to an end. It is sensible to stop
-adaptation once the map has sufficiently converged, e.g. when updates do not change the coefficients, $\gamma$,
-above some tolerance, or as suggested by~\cite{parno2014transport} when the variance of $D_\text{KL}(\pi\|\tilde{\pi})$ is sufficiently close to zero.
-
-Once the map is sufficiently converged, we can decide whether we would like to continue using a local proposal distribution, or whether to use a distribution which is independent of the current state.
-An independent proposal is more efficient
-if the reference space is close to Gaussian; however, if it is not then certain regions of the target
-space could be less likely to be visited than in a standard MH algorithm. A compromise can be found by randomly selecting which kernel to sample from, i.e. the proposal distribution
-\[
-	q(x, \cdot) = p\mathcal{N}(x, \Sigma)+(1-p)\mathcal{N}(0,\text{I}), \quad p \in [0, 1].
-\]
-
-\subsection{Transport map PAIS algorithms}
-
-In a similar way, we can use the Transport map derived in Equation~\eqref{eq:weighted_map_optim} to
+In particular, we will show how we can use the Transport map derived in Equation~\eqref{eq:weighted_map_optim} to
 design a proposal scheme for the PAIS algorithm. In this case we have a choice in how to proceed; we
@@
 
-The first option is given in Algorithm~\ref{alg:TransportPAIS1}.
-We denote the ensembles of states in target space $\theta^{(k)} = \{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble size.
-Similarly, the proposal states are denoted $r' = \{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) = \{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M, \hat{\theta}^{(k)}_M)\}$, where these pairs are the states which we consider to be our sample from the target distribution.
-As in the standard version of the PAIS algorithm we use the deterministic mixture weights.
+The first option is given in Algorithm~\ref{alg:TransportPAIS1}. We
+denote the ensembles of states in target space $\theta^{(k)} =
+\{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the
+reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble
+size. Similarly, the proposal states are denoted $r' =
+\{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) =
+\{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M,
+\hat{\theta}^{(k)}_M)\}$, where these pairs are the states which
+together form our sample from the target distribution. As in the standard version of the PAIS algorithm we use the deterministic mixture weights.
 
@@
 
-\section{Transport map-accelerated Parallel Adaptive Importance
-  Sampling (TPAIS)}\label{sec:TPAIS}
+%\section{Transport map-accelerated Parallel Adaptive Importance
+  %Sampling (TPAIS)}\label{sec:TPAIS}
 

### 2018-01-18 TPAIS_2018.tex
Finished for the day
Lots of small and big changes from after the transport map section

@@
 infrastructure in order to accelerate mixing. In Section
+\ref{sec:conv} we seek to show the advantages of this approach through
+the analysis of test problems. In Section
 \ref{sec:multi} we consider how likelihoods can be approximated using
@@
 which is optimal in terms of the amount of information captured by the map about the target.
-The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = {i+p \choose p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$.
+The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = \binom{i+p}{p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$.
 Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$.
@@
 with the samples produced from the target measure. Convergence of this
-adaptation is shown in~\cite{parno2014transport}. In this section, we
-will discuss how similarly, these maps can be used to construct highly
+adaptation is shown in~\cite{parno2014transport}. In this Section, we
+will show how similarly, these maps can be used to construct highly
 efficient importance sampling schemes.
 
-In particular, we will show how we can use the Transport map derived in Equation~\eqref{eq:weighted_map_optim} to
+In particular, we will show how we can use the transport map derived in Equation~\eqref{eq:weighted_map_optim} to
 design a proposal scheme for the PAIS algorithm. In this case we have a choice in how to proceed; we
@@
 
-The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to the first except on Line~\ref{algline:TPAIS_resample} where rather than resampling in target space we resample in reference space.
-In reference space the dimensions are roughly uncorrelated, and the Gaussian marginals are easy to approximate with fewer ensemble members.
-This means that the resampling step will be more efficient in higher dimensions, which we discuss in Section~\ref{sec:TPAIS_higher_dim}.
+The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to
+the first except on Line~\ref{algline:TPAIS_resample} where rather
+than resampling in target space we resample in reference space. In
+reference space the dimensions are roughly uncorrelated, and the
+Gaussian marginals are easy to approximate with fewer ensemble
+members. This means that the resampling step will be more efficient in
+moderately higher dimensions, which we discuss in Section~\ref{sec:TPAIS_higher_dim}.
 
-\section[Convergence of transport MCMC]{Convergence of the transport proposal based MCMC algorithms}
+\section[Convergence of transport MCMC]{Convergence of the transport proposal based MCMC algorithms}\label{sec:conv}
 
-In this section we study the convergence of the transport based proposal distributions which we have described in Section~\ref{sec:transport_algs}.
-We follow a similar strategy to that in Chapter~\ref{sec:PAIS}.
-We begin by finding optimal values for the scaling parameters by performing a series of simulations with differing scaling parameters.
-We then produce 1 million samples from the target distribution with the optimal scaling parameter and an ensemble size of $M=150$.
-We take 32 repeats of this optimal simulation and present the geometric average of the convergence rates.
-This process is then repeated for each of the algorithms we have discussed, Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2}.
-
-In this section we return to the Rosenbrock banana-shaped density which in Chapter~\ref{sec:LPAIS} we labelled $R_1$. This target density is
+In this Section we study the convergence of the transport based
+proposal distributions which we have described in
+Section~\ref{sec:TPAIS}.
+We take as a test problem the ubiquitous Rosenbrock banana-shaped
+density. This target density is given by
 \begin{equation}\label{eqn:R2_repeat}
@@
 \subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_posterior}}
-\caption{Visualisation of the density of example $R_1$, as given in Equation~\eqref{eqn:R2_repeat}.}
+\caption{Visualisation of the Rosenbrock density as given in Equation~\eqref{eqn:R2_repeat}.}
 \label{fig:R2_posterior}
 \end{figure}
+This problem is challenging to sample from since it has a highly peaked and curved ridge, and is often used
+as a test problem in optimisation and MCMC communities.
 
@@
 
-We now demonstrate some properties of the transport maps we will be using in our MCMC algorithms.
-We draw 1 million samples from the density in \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:transport_implementation} to build a transport map.
+Before looking at the performance of the MCMC algorithms, we demonstrate some properties of the transport maps we will be using in our MCMC algorithms.
+We draw 1 million samples from the density in \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:map} to build a transport map.
 We use this map to push forward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian.
@@
 
-For this example, we use an index set of total order 3 with monomial basis functions. This results in a map of the form
+For this example, we use an index set of total order 3 with monomial
+basis functions. It is important that total order is an odd number,
+since otherwise the map will not be surjective. This results in a map of the form
 \[
@@
 Clearly even with only basis functions of total order 3, we have a large number of unknowns in our optimisation problem, $\bar{\gamma} \in \mathbb{R}^{14}$.
-If we were to increase the dimension of $\theta$ further we would need to reduce the number of terms we include in the expansion by, for example, removing all the cross terms.
+If we were to increase the dimension of $\theta$ further we would need to reduce the number of terms we include in the expansion by, for example, removing all the ``cross'' terms.
 This reduces the quality of our map but since we only require an approximate map we can afford to reduce the accuracy.
@@
 
-Figure~\ref{fig:R2_transport} shows the efficiency of the transport map.
-Even though we have truncated the infinite expansion in the monomial basis down to 4 and 10 terms in respective dimensions, the push forward of the sample is still a unimodal distribution centred at the origin with standard deviation 1.
-As you move out into the tails of the reference density more non-Gaussian features form, but these are not too much of a problem when using a suitable proposal distribution.
-The pullback from reference space, in Figure~\ref{fig:R2_transport}, is an exact match of the original sample since we have not perturbed the sample in reference space.
-This inversion is well defined in the sampling region, although not necessarily outside.
+Figure~\ref{fig:R2_transport} shows the output of the approximate
+transport map. Even though we have truncated the infinite expansion in
+the monomial basis down to 4 and 10 terms in respective dimensions,
+the push forward of the sample is still a unimodal distribution
+centred at the origin with standard deviation 1. As you move out into
+the tails of the reference density more non-Gaussian features are
+clearly visible. However, overall, the push forward of the target
+density does not look a challenging one to sample from, with even
+relatively simple MCMC methods such as RWMH. The pullback from
+reference space, in Figure~\ref{fig:R2_transport}, is an exact match
+of the original sample since we have not perturbed the sample in
+reference space. This inversion is well defined in the sampling
+region, although not necessarily outside\cite{parno2014transport}.
 
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{images/TPAIS/zero_jacobian}
-\caption{An example transport map which does not allow exploration of the full target distribution, and has a non-invertible point.}
-\label{fig:TM_zero_jacobian}
-\end{figure}
+\subsection[Numerical results]{Numerical results for convergence of
+  transport map based algorithms on the Rosenbrock density}
 
-If we consider a map built from polynomials of total order $p=2$, it is clear that the map $\tilde{T}$ does not map onto $\mathbb{R}^d$, but on to a subset of it.
-Figure~\ref{fig:TM_zero_jacobian} illustrates this point.
-We see that in this dimension, the $\tilde{T}^{-1}$ will not map any proposal $r > 0.8$ to a point in the target space.
-The point at which no inverse exists for a point $r_1$ can be made arbitrarily far out in the tails of the posterior distribution, but will be finite.
-The point at which the Jacobian is zero is also a non-invertible point since the logarithm in the pullback becomes infinite.
-This means that (1) we will never sample from the tail of the target distribution which lies beyond this turning point, and (2) all points in reference space $r < 0.8$ have the possibility of being mapped to two distinct points in target space.
-We can choose to truncate our parameter space at this turning point which allows our map to be bijective, however we will not be able to sample from the true posterior.
-For this reason it is best to chose an index set where the maximum order is an odd number.
+We first find the optimal scaling parameters for the individual
+algorithms. This is done, as in ****CITE US**** by optimising for the effective sample size in
+the PAIS algorithm, and by tuning the relative $L^2$ error in the MH
+algorithm. There is currently no
+guidance on the best way of tuning the MH algorithm with transport map
+proposals although one might expect results similar to the standard MH
+results, especially if adaptation of the map is stopped after a given
+point. As in the PAIS algorithm, optimising for the effective sample size might be the best option.
 
-\subsection[Numerical results]{Numerical results for convergence of transport map based algorithms}
-
-We first find the optimal scaling parameters for the individual algorithms.
-This is done as before by optimising the effective sample size in the PAIS algorithm, and by tuning the relative $L^2$ error in the MH algorithm.
-There is currently no guidance on the best way of tuning the MH algorithm with transport map proposals although one might expect results similar to the standard MH results.
-As in the PAIS algorithm, the effective sample size might be the best option.
-
-\begin{table}[!h]
+\begin{table}[!ht]
 \centering
@@
 \toprule
-	Statistic \quad / \quad Algorithm & \ref{alg:TransportMH} & \ref{alg:TransportPAIS1} & \ref{alg:TransportPAIS2}  \\ \cmidrule(lr){1-4}
+	Statistic \quad / \quad Algorithm & Transport M-H &
+                                                            Alg. \ref{alg:TransportPAIS1} & Alg. \ref{alg:TransportPAIS2}  \\ \cmidrule(lr){1-4}
 	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
@@
 
-The optimal scaling parameters are given in Table~\ref{tab:R2_opt_scaling}.
-Here we see that the effective sample size is much lower than we see in the one-dimensional examples with the PAIS algorithms.
-However, in $R_1$ we are dealing with a much more complicated correlation structure, as well as a very slowly decaying tail in $\theta_2$.
-We have seen in Section~\ref{sec:LPAIS_R1} that the standard PAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.
+The optimal scaling parameters are given in
+Table~\ref{tab:R2_opt_scaling}. Here we see that the effective sample
+size is much lower than we see in the one-dimensional examples with
+the PAIS algorithms. However, in the Rosenbrock density
+\eqref{eqn:R2_repeat} we are dealing with a much more complicated
+correlation structure, as well as a very slowly decaying tail in
+$\theta_2$.
+From our experiments, we have observed that the standard PAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 \centering
@@
 \subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
-\caption{Convergence of Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2} for $R_1$. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
+\caption{Convergence of Algorithms Transport M-H,
+  \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2} for density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
 \label{fig:R2_l2_convergence}
@@
 
-\section{Sampling in higher dimensions}\label{sec:TPAIS_higher_dim}
+%\section{Transport map-accelerated Parallel Adaptive Importance
+  %Sampling (TPAIS)}\label{sec:TPAIS}
+
+\section{Multiscale Methods for Stochastic Chemical Reaction
+  Networks}\label{sec:multi}
+In this Section, we discuss some recent advances in multiscale methods
+for stochastic reaction networks. Inverse problems arising in this
+area often lead to highly correlated and complex posterior
+distributions, which traditional MCMC methods can struggle to sample
+from. We will then go on to solve some inverse problems related to
+this in Section \ref{sec:num}.
+
+\section{Numerical Examples}\label{sec:num}
+
+\section{Discussion}\label{sec:conc}
+
+\subsection{Sampling in (moderately) higher
+  dimensions}\label{sec:TPAIS_higher_dim}
+One major problem with importance sampling schemes is the curse of
+dimensionality, which means that methods such as PAIS, and other PMC
+methods, can only be used for relatively low dimensional
+problems. Here, we will briefly discuss how transport maps could aid
+with making moderately higher dimensional problems accessible to this
+family of methods.
 
@@
 
-If we can approximate the posterior well with our mixture and with the transport map, we should not be affected by the increase in dimension to the extent we have been with the standard PAIS-RW algorithm.
-In one dimension the ETPF algorithm can be implemented very efficiently.
-As described in~\cite{reich2013nonparametric}, the coupling matrix has all non-zero entries in a staircase pattern when the state space is ordered.
-We can exploit this knowledge to produce Algorithm~\ref{alg:ETPF_1d}.
-Which is much faster than using the simplex algorithm to minimise the associated cost function, and faster than the AMR algorithm.
+If we can approximate the posterior well with our mixture and with the
+transport map, we should not be affected by the increase in dimension
+to the extent we have been with the standard PAIS-RW algorithm. In one
+dimension the ETPF algorithm can be implemented very efficiently. As
+described in~\cite{reich2013nonparametric}, the coupling matrix has
+all non-zero entries in a staircase pattern when the state space is
+ordered. We can exploit this knowledge to produce
+Algorithm~\ref{alg:ETPF_1d}. Which is much faster than using the
+simplex algorithm to minimise the associated cost function, and faster
+than the AMR algorithm****CITE OUR OTHER PAPER****.
 
@@
 
-%\section{Transport map-accelerated Parallel Adaptive Importance
-  %Sampling (TPAIS)}\label{sec:TPAIS}
-
-\section{Multiscale Methods for Stochastic Chemical Reaction
-  Networks}\label{sec:multi}
-
-\section{Numerical Examples}\label{sec:num}
-
-\section{Discussion}\label{sec:conc}
-
 \bibliographystyle{siam}

### 2018-01-19 TPAIS_2018.tex
Wrote PAIS section

@@
 
-In Section \ref{sec:map} we show how an appropriate transport map can
+In Section \ref{sec:PAIS}, we will briefly reintroduce Parallel
+Adaptive Importance Sampling (PAIS), a variant of Population Monte
+Carlo (PMC) which incorporates state of the art resamplers. In Section \ref{sec:map} we show how an appropriate transport map can
 be constructed from importance samples which maps the posterior close
@@
 Section \ref{sec:conc}.
+
+\section{Parallel Adaptive Importance Sampling}\label{sec:PAIS}
+PAIS****CITE**** is a variant of PMC****CITE****, which is a family of methods which are based
+on importance sampling. In importance sampling, we attempt to
+characterise a target density through sampling from that
+density. However, the target density $\pi$ itself is often too complex to sample
+from directly, so we instead sample from a
+proposal density $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
+$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
+sampling from a different distribution to $\pi$. Monte Carlo estimates
+using a sample of size $N$
+of a function $f$ with respect to $\pi$ can then be made through the
+formula
+\[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
+  w_kf(\theta^(k)).\]
+This method works well when $\pi$ and $\chi$ are close, but can be
+excruciatingly slow when they are not. The idea behind PMC methods is
+to construct a good proposal distribution, either from the entire
+history of the algorithm up to the current point, or to use the
+current state of a whole ensemble of $M$ particles in the system.
+
+In PAIS, the proposal distribution $\chi$ is chosen to be the equally
+weighted mixture of any choice of MCMC proposal kernel, evaluated at
+each of the current particles in the system. If $\theta^{(k)} = [\theta_1^{(k)},
+\theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
+ensemble, and we wish to use an MCMC proposal density $q(\cdot ;
+\cdot, \beta)$ then 
+\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}.\]
+Often the variance of the MCMC proposal kernels can be tuned using
+their respective algorithmic parameters $\beta$. Good values for these
+algorithmic parameters can be
+found by optimising for the effective sample size of the importance
+sample that is produced (see ***CITE*** for more details).
+
+If the ensemble is large enough, and the chain has entered
+probabilistic stationarity, then the current state of the ensemble is
+a good rough discrete approximation of the target density, and in turn
+$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
+sample $\{\hat{\theta}^{(k)},w^{(k)}\}$. It can be advantageous to use stratified sampling of the
+mixture in order to
+ensure that the sample made is as representative as possible of the
+target density, i.e.
+\[ \hat{\theta}_i^{(k)} \sim q(\cdot ; \theta_i^{(k)}\]
+for each $i = 1,2,\ldots,M$. We now have a weighted sample, and it
+would be inadvisable to use an equally weighted mixture of proposal
+distributions from each of these points. Therefore, before starting
+the next iteration, the importance sample
+$\{\hat{\theta}^{(k)},w^{(k)}\}$ is resampled to produce an equally
+weighted sample, ready for the next iteration of the algorithm. In
+PAIS, a state-of-the-art resampler is used, which uses optimal
+transport methods to find the equally weighted discrete sample which
+best represents the statistics of the importance
+sample***CITE****. For larger ensemble sizes $M$ this can become
+expensive, in which case a greedy approximation of this algorithm, the
+Approximate Multinomial Resampler (AMR) can be
+implemented***CITE****. The output of the resampler is then denoted
+$\theta^{(k+1)}$ and the algorithm is ready for the next
+iteration. The importance samples $\{\hat{\theta}^{(k)},w^{(k)}\}$ The PAIS is summarised in Algorithm \ref{alg:PAIS}.
+
+\begin{table}[!h]
+\centering
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+	Initialise $\theta^{(1)} \sim \mu_0$.\;
+	\For{$k=1,\ldots,N$}{
+		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_j^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:PAIS_propose}\;
+		Calculate $w^{(k)} = (w_1^{(k)}, \ldots, w_M^{(k)})^\top$, where
+			\[
+				w^{(k)}_i = \frac{\mu(\hat{\theta}^{(k)}_i)}{\chi^{(k)}(\hat{\theta}^{(k)}_i; \theta^{(k)}, \beta)}.
+			\]
+		\label{algline:PAIS_weights}
+
+		Resample $\theta^{(k+1)} \leftarrow \| w^{(k)}\|_1^{-1}\sum_{j=1}^M w_j^{(k)}\delta_{\hat{\theta}_j^{(k)}}(\cdot)$.\label{algline:PAIS_resample}\;
+	}
+	Output $\{(w^{(n)}, \hat{\theta}^{(n)}\}_{n=1}^N$.
+\caption{The PAIS Algorithm.\label{alg:PAIS}}
+\end{algorithm}
+\end{table}
+
+One problem with the PAIS and other PMC methods can become apparent if
+the target density has very strong correlations in its structure, in
+particular if that correlation is not global but only local. In this
+case, unless the proposal densities $q$ are informed by this local
+structure, the mixture distribution proposal may not well approximate
+$\pi$ without a very large large ensemble size $M$, which can become
+inhibitively expensive. Some methods have been proposed ****CITE****
+which use samples local to each particle to inform local covariance
+structure.
+
+In this paper, we investigate the use of transport maps to learn local
+covariances across the whole of the domain, in order to stabilise
+PMC-type methods, and make these methods more applicable to a wider
+range of more challenging inference problems.
 

### 2018-01-19 TPAIS_2018.tex
Added huge amounts + edits
Paper now a monster.

@@
 %Follow the other papers, show adaptation(s)
-
+In this Section, we describe the construction of transport maps which
+allow for the simplification of complex posterior distributions in
+order to allow for improved sampling, in particular for methods based
+on importance sampling.
 In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
@@
 from. We will then go on to solve some inverse problems related to
-this in Section \ref{sec:num}.
+this in Section \ref{sec:num}, using the transport map versions of the
+PAIS algorithm, as described in Section \ref{sec:TPAIS}.
+
+We consider chemical reaction networks of $N_s$ chemical species $\{S_j\}_{j=1}^{N_s}$,
+with population numbers given by $X(t) = [X_1(t), X_2(t), \ldots, X_{N_s}(t)]^\top \in
+\mathbb{N}_0^{N_s}$ reacting in a small reactor, through $N_r$ different reaction
+channels. When population numbers of one or more of the chemical
+species in the reactor is small, as is the case with chemical
+reactions occurring within living cells, the sporadic and discrete way
+in which reactions occur can not be well modelled by deterministic
+continuous models, such as ODEs. In such a situation, we may wish to
+model the dynamics through a discrete stochastic model, such as a
+continuous time Markov chain.
+
+For each reaction $R_j$, for $j = 1,2,\ldots N_r$, there is a
+propensity or hazard function $\alpha_j(X(t)$ which indicates how
+likely that reaction is to fire, defined by
+\[\alpha_j(X(t)) = \lim_{dt \to 0} \mathbb{P}(\text{Reaction $R_j$ in
+    the time interval  } \quad s \in [t, t+ dt] ).\] If a system
+satsifies what is called \emph{mass action kinetics}, then the form of
+the function $\alpha_j$ is determined, up to a rate constant, by the
+reactants involved in that reaction:
+\begin{equation}\label{eq:MAK}
+\alpha_j(X) = k_j \prod_{m=1}^{N_s} \prod_{n=0}^{\nu_{j,m} -1} (X_m - n),
+\end{equation}
+where $\nu_{j,m}$ is the $m$th component of the stoichiometric vector
+$\nu_j$, $X_m$ is the $m$th component of the state vector $X$, and the
+$k_j$ are rate constants.
+
+Following each reaction $R_j$ there is an instantaneous change in the
+current state, as the reactants of the reaction are consumed, and the
+products produced. This is modelled by a change in the state vector
+$X(t) = X(t) + \nu_j$ where $\nu_j$ is that stochiometric vector for
+reaction $R_j$.
+
+The model can be represented as the following expression involving
+$N_r$ different unit rate Poisson
+processes****CITE ANDERSON ET AL**** $Y_j$, given by:
+\begin{equation}\label{eq:RTC}
+X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j \int_0^t \alpha_j(X(s)) ds.
+\end{equation}
+
+The master equation for this model is only solvable in certain
+situations, for example monomolecular networks****CITE****, or for
+steady-state solutions certain deficiency
+zero networks****CITE*****. Trajectories for this system can be
+sampled exactly, for instance using the Gillespie SSA***CITE***, or
+its variants\cite{gillespie2007stochastic}****CITE****. However, if the system is stiff, i.e. there
+are some reactions which are firing many times on a timescale for
+which others are unlikely to fire at all, then trajectories can become
+prohibitively expensive to simulate, since these methods simulate
+every single reaction event with the same cost. In such a system, one
+might employ multiscale methods in order to approximate trajectories
+at a lower cost than the exact algorithms.
+
+One common approach is to induce the quasi-steady-state approximation
+QSSA. This approximation makes the assumption that fast reactions
+enter quasi-equillibrium on a timescale which is negligible with
+respect to the timescale on which the slow dynamics in the system are
+evolving. This amounts to taking the assymptotic limit that the
+propensities of the ``slow'' reaction channels are equal to zero. This
+allows us to sample approximate trajectories of the slow reactions
+without the need to compute many fast reaction events.
+
+This approach can work well when there is a large timescale gap
+between the fast and slow reactions in a system, but where the
+timescale gap is less pronounced, it can lead to significant
+errors***cite cotter erban errors paper***. Another approach is the
+constrained multiscale method (CMA)***CITE***, based in part on the
+equation-free approach to multiscale computations***CITE
+YANNIS****. This approach also assumes quasi-equillibrium in the
+system, but takes into account the effect that the slow reactions can
+have on the invariant distribution of the fast variables in the
+system. For a more detailed description of this method, please refer
+to the literature***CITATIONS***.
+
+Multiscale methods are not only of use when forward evaluations of the
+dynamics are costly, but can also be of use where we attempt to solve
+an inverse problem where the fast variables are unobservable. One
+approach in this situation would be to integrate over all possible
+trajectories of the fast variables, but this would almost always be
+prohibitively expensive. Another approach would be to use an
+appropriate multiscale approximation, so that the effective dynamics
+of the slow variables can be approximate without the need to consider
+the rapid fluctuations of the fast variables in the system.
+
+\subsection{Likelihoods arising from stochastic reaction networks}
+Suppose that we are able to exactly observe the number of molecules of
+each chemical species in a system which satisfies mass action
+kinetics, and which can be well approximated by
+\eqref{eq:RTC}. Suppose that we wish to be able to infer the value of
+the rate constants of each reaction from these observations. Even with no observational noise, since the dynamics
+of the system are stochastic, this still leads to a Bayesian inverse
+problem where we can only infer a joint probability distributions on the
+reaction parameters.
+
+Suppose that we are in state $X(t) = X_0$. There are two independent
+univariate random variables which decide when and what the next
+reaction in the system is. There is the $j$th waiting time $\tau_j$ to the
+next reaction, which is given by an exponential random variable
+\[\tau_j \sim \exp \left ( \frac{1}{\alpha_0(X(t))} \right ),\]
+where $\alpha_0(X(t)) =\sum_{i}^{N_r} \alpha_i(X(t))$ is the total
+propensity in the system. The second is a multinomial random variable $r_j$
+which dictates which reaction has occurred during the $j$th reaction,
+which takes the value
+$r \in \{1,2,\ldots,N_r\}$ with probability
+\[\mathbb{P}(r = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\]
+As such, in order to compute the likelihood of a particular trajectory
+given a possible realisation of the reaction parameters, it is
+sufficient to have access to the total time spent in each state, and
+the frequency of each reaction which led to leaving each state.
+
+From this formulation, we see that the random variables $(\tau_j, r_j)$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are Markovian.
+This conditional independence means that we can group events together by what state the system was in when the event happened.
+We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
+\[
+	T(\mathbf{Y}) = \sum\limits_{j=1}^M \tau_j\; \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y}).
+\]
+This random variable, $T(\mathbf{Y})$, is a sum of exponentially
+distributed random variable, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
+\begin{equation}\label{eqn:chem_time_dist}
+	T(\mathbf{Y}) \sim \text{Gamma}\left(\alpha=K(\mathbf{Y}),~\beta = \alpha_0(\mathbf{Y}; \mathbf{k})\right),
+\end{equation}
+where $K(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y})$.
+
+Similarly, we can define the reactions which occurred when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_{N_r}(\mathbf{Y})]^\top$ where
+\[
+	r_i(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(r_j = i \textbf{ and }\mathbf{X}(t_{j-1}) = \mathbf{Y}).
+\]
+Here all the random variables $r_j$ follow the same multinomial distribution, and so
+\begin{equation}\label{eqn:chem_react_dist}
+	\mathbf{r}(\mathbf{Y}) \sim \text{Multinomial}(K(\mathbf{Y}),~\mathbf{p}(\mathbf{Y})). 
+\end{equation}
+The random variables defined in Equations~\eqref{eqn:chem_time_dist} and \eqref{eqn:chem_react_dist} are sufficient statistics for the posterior distribution $\pi(\mathbf{k}|D)$.
+With these definitions we define two new structures
+\[
+	\mathbf{T} = [T(\mathbf{Y}_1), \dots, T(\mathbf{Y}_K)]^\top, \quad \text{and} \quad \mathbf{R} = [\mathbf{r}(\mathbf{Y}_1), \dots, \mathbf{r}(\mathbf{Y}_K)]^\top,
+\]
+where $K = |\mathcal{S}|$, the number of states in $\mathcal{S}$, and each state $\mathbf{Y}_i \in \mathcal{S}$ has been enumerated. We use these structures to define shorter notation,
+\[
+	\mathbf{T}_i = T(\mathbf{Y}_i), \quad \mathbf{R}_{ij} = r_j(\mathbf{Y}_i), \quad \text{and} \quad \mathbf{K}_i = K(\mathbf{Y}_i).
+\]
+
+To construct the posterior distribution for the reaction rates, $\mathbf{k}$, in the chemical system, we formulate the likelihood using the sufficient statistics derived in the previous section.
+Due to the non-negativity of these reaction rates, we assign a Gamma prior distribution to each rate.
+Given the distributions in Equations~\eqref{eqn:chem_time_dist} and~\eqref{eqn:chem_react_dist} for our data, the likelihood of observing the data $\mathbf{R}$ and $\mathbf{T}$ is
+\[
+	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
+\]
+where again $\mathbf{Y}_i \in \mathcal{S}$ and the propensities
+$\alpha_i$ and probabilities $p_i$ depend on the reaction rates
+$\mathbf{k} = [k_1, k_2, \ldots k_{N_r}]^\top$ through the concept of
+mass action kinetics given in \eqref{eq:MAK}.
+
+Our choice of Gamma prior distributions with hyper-parameters $(a_i, b_i)$ results in the posterior distribution of the form
+\begin{align}
+	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
+	\prod_{i=1}^{N_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
+		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
+				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
+%		& \qquad\qquad\qquad
+				+\sum_{j=1}^{N_r} \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
+			\Bigg]  \notag\\
+		&	\qquad\qquad\qquad{} + \sum_{i=1}^{N_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}. \label{eqn:chem_posterior}
+\end{align}
+
+\subsection{Approximation of likelihoods in multiscale chemical
+  networks}
+Suppose now that we are only observing the slower variables in a
+multiscale chemical system of this type. Suppose that $n_r < N_r$ is the number of slow
+reactions in the system, and the slow reactions are given by $\{R_{s_1},
+R_{s_2}, \ldots R_{s_{n_r}} \} \subset \{R_1, R_2, \ldots , R_{N_r}
+\}$. The propensities of the slow
+reactions $\alpha_{s_i}$ may depend on the value of the fast variables which is
+unknown. However, the invariant distribution of the fast variables
+conditioned on the slow variables in the system can be approximated
+using a multiscale method, such as the QSSA or the CMA, as described
+briefly in Section \ref{sec:multi}. Once the approximations have been
+made, we arrive at approximate \emph{effective} propensities $\bar{\alpha}_{s_i}$, which
+have been averaged over the computed invariant distributions of the
+fast variables. Then the approximate effective dynamics on the slow variable $S(t)$ are given by 
+\begin{equation}\label{eq:RTCS}
+S(t) = S(0) + \sum_{j=1}^{n_r} \nu_{s_j} \int_0^t \bar{\alpha}_{s_j}(S(s)) ds.
+\end{equation}
+
+In turn, the approximate posterior distribution on the reaction
+parameters $\mathbf{k}$ is then given by
+\begin{align}
+	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
+	\prod_{i=1}^{N_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
+		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
+				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
+%		& \qquad\qquad\qquad
+				+\sum_{j=1}^{N_r} \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
+			\Bigg]  \notag\\
+		&	\qquad\qquad\qquad{} + \sum_{i=1}^{N_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}. \label{eq:Apos}
+\end{align}
 
 \section{Numerical Examples}\label{sec:num}
+In this section we look at two examples of chemical systems to
+demonstrate the effectiveness of the Bayesian approach we have
+proposed, along with the transport PAIS algorithms for sampling the
+challenging posterior distribution on the reaction parameters that
+arise.
+
+% The first example will only contain at most mono-molecular reactions,
+% i.e. reactions which convert one species into another species at a
+% particular rate. As shown in~\cite{jahnke2007solving} these systems
+% have a Poisson invariant distributions on the populations, given by
+% \begin{equation}\label{eqn:mono_stat_dist}
+% 	\mathbb{P}(X_1=x_1, \ldots, X_n=x_n) = \left(\prod_{i=1}^n \! \frac{\bar{c}_i^{x_i}}{x_i!}\right)\exp\left(-\sum_i^n \! \bar{c}_i\right),
+% \end{equation}
+% where $x_i \in \mathbb{N}$ and $\bar{c}_i$ are the complex balances or the concentration of chemical $X_i$ according to the solution of a steady state approximation of the system.
+We make use of this distribution to formulate posteriors for the chemical reaction rates using the constrained multiscale approach, as in~\cite{cotter2016constrained}.
+
+%%%%%
+
+\subsection{A Multiscale Chemical System}\label{sec:chem_multiscale}
+
+First we consider a simple multiscale example involving two chemical
+species $S_1$ and $S_2$ involved in only zeroth and first order reactions:
+\begin{equation}\label{eqn:full_system}
+	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
+\end{equation}
+Each arrow represents a reaction from a reactant to a product, with
+some rate constant $k_i$, and where mass action kinetics is
+assumed. The parameters $k_i$ are non-negative, and $\mathbf{k} =
+[k_1,\dots,k_4]^\top \in \mathbb{R}_+^4 = \mathcal{X}$. We denote the
+population count of
+species $S_i$ by $X_i \in \mathbb{N}_0$. We assume that we are in a
+parameter regime such that the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
+more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
+chemical species are involved in fast reactions. However, the quantity
+$S = X_1 + X_2$, is conserved by both of the fast reactions,
+and as such, this is the slowly changing quantity in this system.
+
+We assume that we are only able to observe
+the slow variable $S(t)$. 
+The effective dynamics of $S$ can be approximated by a system of only
+two reactions:
+\begin{equation}\label{eqn:QSSA_system}
+	\emptyset \xrightarrow{\alpha_1(S)} S \xrightarrow{\bar{\alpha}_4(S)} \emptyset.
+\end{equation}
+The effective propensity $\bar{\alpha}_4(S)$ can be  approximated through
+application of a multiscale approximation, as detailed in Section
+\ref{sec:multi}, and in more detail in ****CITE****. Under the QEA,
+this is given by
+\[
+	\bar{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3}.
+\]
+Similarly, the analysis of the constrained system as discussed in~\cite{cotter2016constrained} yields the effective propensity
+\begin{equation}\label{eqn:chem_CMA_rate}
+	\bar{\alpha}_4^{\text{CMA}}(s) = \mathbb{E}_{\text{CMA}}\left[k_4X_2|S=s\right] = \frac{k_2k_4s}{k_2+k_3+k_4}.
+\end{equation}
+
+Our observations are uninformative about the reaction rates $k_2, k_3$, and $k_4$, as there
+is a sub-manifold $\mathcal{M} \subset \mathcal{X}$ along which the effective rate
+$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
+problem. This type of problem is notoriously difficult to
+sample from using standard MH algorithms, as the algorithms quickly
+gravitate towards the manifold $\mathcal{M}$ but then
+exploration around $\mathcal{M}$ is slow.
+
+We aim to recover three different posterior distributions. In the
+first, which is of the form \eqref{eqn:chem_posterior} we assume that we can fully and exactly observe the whole state
+vector for the whole of the observation time window. In the second and
+third, the observations are restricted to observations of the slow
+variable $S = X_1 + X_2$. In these two examples, we approximate the
+posterior using \eqref{eq:Apos} with QEA and CMA approximations of the
+effective dynamics respectively.
+
+In all three cases, we find the posterior distribution for $\mathbf{k} \in \mathcal{X}$. These four parameters are assigned Gamma prior distributions,
+\[
+	k_i \sim \text{Gamma}(\cdot; \alpha_i, \beta_i), \quad \text{for} \quad i = 1, \dots, 4.
+\]
+The hyper-parameters corresponding to each of these prior distributions are given in Table~\ref{tab:multiscale_priors}. These priors are the same for each of the three posterior distributions.
+
+\begin{table}
+\centering
+\begin{tabular}{|l|R{2em}R{2em}R{2em}R{2em}|}
+	\hline
+	parameter/dimension & 1 & 2 & 3 & 4 \\ \hline
+	$\alpha_i$ & 150 & 5 & 5 & 3 \\
+	$\beta_i$ & 15/9 & 5/12 & 5/12 & 1 \\ \hline
+\end{tabular}
+\caption{Hyper-parameters in the prior distributions for the multiscale problem described in Section~\ref{sec:chem_multiscale}.}
+\label{tab:multiscale_priors}
+\end{table}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
+\caption{Posterior for the constrained multiscale problem outlined in Section~\ref{sec:chem_multiscale}.}
+\label{fig:chem_CMA_posterior}
+\end{figure}
+
+Figure~\ref{fig:chem_CMA_posterior} shows how this posterior looks
+when we use the CMA to model the effective degradation propensity $\bar{\alpha}_4$.
+
+We consider several different algorithms for sampling from these
+distributions. First we implement both the PAIS and MH algorithms with
+a Gaussian proposal distribution. In the case of the PAIS algorithm,
+this is a Gaussian mixture proposal distribution, and in the MH
+algorithm this is equivalent to a vanilla random walk.
+The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard RWMH algorithm.
+We also compare the PAIS and MH algorithms when using a transport map proposal distribution.
+This proposal method was discussed in detail in Chapter~\ref{sec:TPAIS}.
+
+\begin{figure}
+	\centering
+	\begin{tikzpicture}
+		\pgfmathsetseed{1}
+
+		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
+		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$R$};
+
+		\draw (A) \irregularcircle{1cm}{2.5mm};
+		\draw (B) \irregularcircle{1cm}{0.5mm};
+
+		\draw [->] (A) -- (B);
+
+		\node at (2.5,1) {$T\colon\mathcal{X}\rightarrow R$};
+
+		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
+		\node at (5,-1.5) {$r\in R\subset\mathbb{R}^d$};
+	\end{tikzpicture}
+	\caption{Couplings between parameter space and reference when using the transport map proposal method to propose moves on parameter space.}
+	\label{fig:chem_coupling}
+\end{figure}
+
+Figure~\ref{fig:chem_coupling} shows how we define a bijective map,
+$\hat{T}$, between parameter space $\mathcal{X}$ and a reference space
+$R$.
+When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
+MH-RW (RWMH) and MH-TRW, as well as PAIS-RW and PAIS-TRW.
+In practice we cannot ensure that the approximate map $\tilde{T}$
+is uniquely invertible over the whole of $R$ and so $\tilde{T}$ is not
+truly bijective. This leads to problems for our strictly positive
+state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in
+$R$ do not necessarily map back onto $\mathcal{X}$. This motivates the
+use of an intermediate space.
+
+We also consider how these algorithms perform when they are applied to
+the log of the reaction rates. We choose to use this log
+transformation since it converts our strictly positive parameter
+space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset
+\mathbb{R}^d$. This allows us to define $T$ between two subsets of
+$\mathbb{R}^d$, which means that even if $\tilde{T}$ is not uniquely
+invertible in some region of $R$, all possibilities are valid
+proposals. It also reduces the work required of the map, which since
+we use only a finite dimensional approximation, can improve
+performance and stability.
+As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
+MH-logRW and MH-logTRW, with, PAIS-logRW and PAIS-logTRW.
+
+\begin{figure}
+	\centering
+	\begin{tikzpicture}
+		\pgfmathsetseed{1}
+
+		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
+		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
+		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};
+
+		\draw (A) \irregularcircle{1cm}{2.5mm};
+		\draw (B) \irregularcircle{1cm}{1.1mm};
+		\draw (C) \irregularcircle{1cm}{0.5mm};
+
+		\draw [->] (A) -- (B);
+		\draw [->] (B) -- (C);
+
+		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
+		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};
+
+		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
+		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
+		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
+	\end{tikzpicture}
+	\caption{Couplings between parameter space, an intermediate space and reference space when using the transport map proposal method to propose moves on a log space.}
+	\label{fig:chem_log_coupling}
+\end{figure}
+
+Figure~\ref{fig:chem_log_coupling} displays the composition of maps when we include this intermediate space $\mathcal{Y}$.
+Every proposal in $R$ results in a valid proposal on $\mathcal{X}$.
+The inclusion of this additional map means that we must again alter our importance weight definition to reflect the pullback from $R$ through $\mathcal{Y}$.
+The weight is now
+\[
+	w_i(\theta) = \frac{\pi(\theta|\mathbf{R},\mathbf{T})}{\chi(\tilde{T}\circ\log(\theta)|\tilde{T}\circ\log(\theta^{(i-1)}))|J_{\tilde{T}\circ\log}(\theta)|},
+\]
+where $\theta$ is a proposed new state, $\theta^{(i-1)}$ is the ensemble of states from the previous iteration, and $J_{\tilde{T}\circ\log}(\theta)$ is the Jacobian of the composition of the two maps.
+This Jacobian is straightforward to calculate,
+\[
+	|J_{\tilde{T}\circ\log}(\theta')| = |J_{\tilde{T}}(\log(\theta'))||J_{\log}(\theta')|,
+\]
+where the first determinant is as we saw in the previous chapter, and the second is
+\[
+	|J_{\log}(\theta')| = \prod\limits_{i=1}^d \frac{1}{\theta'_i}.
+\]
+
+% choices made in transport map
+For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=3$ as the basis functions, i.e.
+\[
+	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
+\]
+and
+\[
+	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
+\]
+
+% choice of resampler?
+We will use the AMR***CITE*** resampler with an ensemble size of
+$M=500$. This increase in ensemble size in comparison with previous
+example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.
+
+% analysis method
+To measure the convergence of the sampling methods in this section, we
+will compare the convergence of the mean of each parameter. We
+approximate $\mathbb{E}(\mathbf{k})$ using 2.4 billion samples from
+the MH-RW algorithm. We do this for each of the eight algorithms we
+have so far discussed. Convergence is shown only for the CMA
+approximation of the posterior \eqref{eq:Apos} with effective rate for
+the degradation of $S$ given by \eqref{eqn:chem_CMA_rate}, but we
+expect very similar results for the other posterior distributions discussed.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{rrrrrr}
+\toprule
+	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{PAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
+	& & $\delta_\%$ & & $\delta_{\text{ESS}}$ & ESS \\ \midrule
+	RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
+	TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
+	logRW & & 1.7e-1 & & 1.2e-0 & 6.0e-2 \\
+	logTRW & & 2.7e-2 & & 1.5e-1 & 3.5e-1 \\
+\bottomrule
+\end{tabular}
+\caption{Optimal scaling parameters for the MH and PAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}.
+MH parameters optimised by acceptance rate, and PAIS parameters optimised using effective sample size.}
+\label{tab:chem_multiscale_scaling}
+\end{table}
+
+The optimal scaling parameters for the proposal distributions are given in Table~\ref{tab:chem_multiscale_scaling}.
+We note that for MH-TRW and PAIS-TRW the scaling parameter is near to 1, particularly for MH-RW, this is what we should expect since the proposal is made on a reference space which should be near to $\mathcal{N}_d(0, \text{I})$.
+The same should be true for the MH-logTRW and PAIS-logTRW algorithms since these have the same target reference space, however the optimal scaling parameters here are much smaller.
+We see that the ESS is higher for the algorithms which sample on $\mathcal{Y}$, and we expect that convergence will be fastest for the PAIS-logTRW algorithm.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Sampling algorithms on $\mathcal{X}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
+\subfigure[Sampling algorithms on $\mathcal{Y}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
+\caption{Convergence of the constrained multiscale example described in Section~\ref{sec:chem_multiscale}.}
+\label{fig:chem_multiscale_L2}
+\end{figure}
+
+Convergence of the eight algorithms for this example is shown in Figure~\ref{fig:chem_multiscale_L2}.
+We first note the poor performance of the MH based algorithms, each of them taking roughly 10,000 samples to begin converging.
+Only the MH-TRW is at all competitive with the PAIS algorithms.
+During the simulation interval, the MH-TRW algorithm has not settled down to the expected $\mathcal{O}(1/\sqrt{N})$ rate which means that the estimate is still biased by the long burn-in time.
+As we have seen in previous examples, the burn-in time for the PAIS algorithm is negligible.
+
+The PAIS variants with RW and TRW proposals perform similarly on both sample spaces.
+When sampling on $\mathcal{X}$ the transport map is not quite as efficient, largely due to the difficulties discussed in the previous section i.e.
+many proposals are made which do result in negative reaction rates.
+Sampling on $\mathcal{Y}$ leads to more comparable Monte Carlo errors, with the logTRW being apparently slightly less stable.
+This proposal method becomes more stable as we increase either the ensemble size, or the number of iterations between updates of the transport map, $T$.
+Overall we see the smallest Monte Carlo errors for a given amount of computational effort coming from the PAIS-logTRW algorithm.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Reference space mapped to by $\hat{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
+\subfigure[Reference space mapped to by $T\circ\log$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
+\caption{Reference space for the problem described in Section~\ref{sec:multiscale_posterior} found using the TRW and logTRW proposal distributions.
+Components are linked by the relation $r_i = T_i'(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
+\label{fig:chem_reference_spaces}
+\end{figure}
+
+We now look at the proposal distributions of the transport map accelerated algorithms.
+In Figure~\ref{fig:chem_reference_spaces}, we see the reference spaces found by; in (a) mapping the posterior through the map $\hat{T}$, and in (b) by mapping the posterior through $T\circ\log$.
+For the most part, each of these marginal distributions can be recognised as a Gaussian.
+However, with the exception of $\mathbb{P}(r_2,r_3)$, we would not consider them to be close to a standardised $\mathcal{N}(0, \text{I})$ distribution.
+Before thinking that the transport map has not helped us to find a `nicer' space on which to propose new values we should consider that the dimensions are now (1) largely uncorrelated, and (2) the variances in each dimension are much more similar than they are in Figure~\ref{fig:chem_CMA_posterior}.
+
+Particularly in Figure~\ref{fig:chem_reference_spaces} (b) we see that $\text{var}(r_1)$ and $\text{var}(r_4)$ are much smaller than $\text{var}(r_2)$ and $\text{var}(r_3)$.
+To combat this we have a number of choices, we might wish to use two different scaling parameters to match these scales, which would require knowledge of the reference space before beginning sampling.
+We could alternatively use a proposal distribution such as in the LPAIS algorithm to adaptively learn this at each iteration.
+Another option is to increase the total order of our index set.
+For these numerics we have chosen $p=3$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
+
+\subsubsection{Comparison of the Constrained and QEA approaches}
+
+The convergence analysis has been performed for the constrained approach to this multiscale system.
+We now look at the differences between the constrained and QEA posterior distributions.
+Recall that the approaches differed only in the form of the effective degradation rate $\hat{k}_4$,
+\[
+	\hat{k}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{k}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
+\]
+This difference in the denominator causes a shift in the parameters as can be seen in Figure~\ref{fig:chem_diff}. The figure shows the difference in posteriors,
+\begin{equation}\label{eqn:chem_diff}
+	\text{diff}(\mu^{\text{CMA}}, \mu^{\text{QEA}}) = \pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
+\end{equation}
+
+\begin{figure}[htb]
+\centering
+\includegraphics[width=0.9\textwidth]{"images/Applications/Diff_CMA_QEA"}
+\caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
+\label{fig:chem_diff}
+\end{figure}
+
+Since the two posteriors have been approximated using an MCMC sample, there is a significant amount of noise, particularly in the tails of the distributions We can see that the differences in the marginals for $k_1, k_2,$ and $k_3$ are relatively small and the differences are largely positive.
+These positive differences tell us that the constrained approach claims to be more informative about the values of these first three parameters.
+However the marginal for $k_4$ varies by a significant amount and is more negative.
+This means that the mean estimates for $k_4$ differ by a large amount, the posterior for the QEA is most peaked.
+
+% compare solution with full data.
+We now consider how we should interpret the information given by these models.
+The QEA assumption tells us that we cannot observe the parameters $k_2$, $k_3$ and $k_4$ independently from the reduced data set, but we can observe the quantity $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$.
+Similarly, the constrained approach tells us that we are able to observe the quantity $\hat{k}_4^{\text{CMA}} = k_2k_4/(k_2+k_3+k_4)$.
+To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most accurate and informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model, with all reactions observed.
+
+A conventional way to compare two models under a Bayesian framework is to calculate the Bayes factors~\cite{chen2012monte}.
+The Bayes factor, $B_{1,2}$, between two models, $\mathcal{M}_1$ and $\mathcal{M}_2$, can be interpreted as a ratio of the normalisation constants of the posterior distributions given each model,
+\[
+	B_{1,2} = \frac{\mathbb{P}(D|\mathcal{M}_1)}{\mathbb{P}(D|\mathcal{M}_2)}, \quad \text{where} \quad \mathbb{P}(D|\mathcal{M}_k) = \int_\mathcal{X} \mathbb{P}(D|\theta_k,\mathcal{M}_k)\mathbb{P}(\theta_k|\mathcal{M}_k) \, \text{d}\theta_k.
+\]
+Under the PAIS framework, it is straightforward to calculate these factors using the Monte Carlo estimator for the normalisation constants.
+From~\cite{robert2013monte} these normalisation constants take the form
+\[
+	Z_k \approx \hat{Z}_k\frac{1}{NM}\sum_{i=1}^N\sum_{j=1}^M w_j^{(i)}(\mathcal{M}_k),
+\]
+where $w_j^{(i)}(\mathcal{M}_k)$ is the weight under model $\mathcal{M}_k$ corresponding to the $j$-th ensemble member on the $i$-th iteration.
+Hence, $B_{1,2} \approx \hat{Z}_1/\hat{Z}_2$.
+We can compare more than two models in this way by selecting the model with the largest marginal distribution as the best model.
+
+We now label the constrained model as $\mathcal{M}_1$, the model arising from the QEA as $\mathcal{M}_2$, and the full data model as $\mathcal{M}_0$.
+Under $\mathcal{M}_1$, the parameter $\theta_1 = (k_1, \hat{k}_4^{\text{CMA}})^\top$, and under $\mathcal{M}_2$, the parameter $\theta_2 = (k_1, \hat{k}_4^{\text{QEA}})^\top$.
+The full model $\mathcal{M}_0$ observes all four parameters $\theta_0 = (k_1, k_2, k_3, k_4)^\top$.
+The marginal densities for the data, evaluated at the observed data, given the models are displayed in Table~\ref{tab:chem_Bayes_marginals}.
+
+\begin{table}[!htb]
+\centering
+\begin{tabular}{rr}
+	\toprule
+	$k$ & \quad $\mathbb{P}(D = (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$ \\ \cmidrule(lr){1-2}
+	0 & 6.8e-3 \\
+	1 & 3.2e-3 \\
+	2 & 1.7e-3 \\ \bottomrule
+\end{tabular}
+\caption{Marginal distributions for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:multiscale_posterior}.}
+\label{tab:chem_Bayes_marginals}
+\end{table}
+
+Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals} we see that we should of course prefer the model in which we observe all reactions perfectly, however this model is not significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$, ~\cite{kass1995bayes}).
+Again the constrained model is not substantially more attractive than the QEA model when we consider the Bayes factor $B_{1,2} = 1.96 < 3.2$.
+The Bayes factor $B_{0,2} = 4.1 > 3.2$ does tell us that we should significantly prefer the full model to the QEA approximation of the posterior.
+These Bayes factors present a weak argument that that the constrained model provides us with a better description of the data than the QEA model.
+We can also present this information graphically, which might provide us with a greater justification for preferring the constrained approximation of the posterior.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/QEA_approx_marginal"}}
+\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_approx_marginal"}}
+\caption{Comparison of the approximate marginal densities for the `observable parameter' $\hat{k}_4$ under models $\mathcal{M}_1$ and $\mathcal{M}_2$.
+Marginals for these two parameters are approximated using samples which have been produced by targeting the three densities $\pi_i = \mathbb{P}(\theta_i|D,\mathcal{M}_i)$ for $i = 0, 1, 2$.}
+\label{fig:chem_model_comp}
+\end{figure}
+
+Figure~\ref{fig:chem_model_comp} displays the marginal distributions for the parameters $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$.
+For each of these two observable parameters we obtain approximations for three marginal distributions, $\pi_i(\cdot|D, \mathcal{M}_i)$, $i = 0, 1, 2$.
+Each of these approximations has been produced by marginalising a sample drawn from the full posterior $\pi_i(\mathbf{k}|D, \mathcal{M}_i)$ defined in terms of the respective model $\mathcal{M}_i$.
+In other words, to calculate $\hat{\pi}_1(\hat{k}_4^{\text{QEA}}|D)$, we first draw a sample using the PAIS algorithm targeting the posterior density $\pi_1(\mathbf{k}|D,\mathcal{M}_1)$.
+We then calculate $\theta^{(j)} = k_2^{(j)}k_4^{(j)}/(k_2^{(j)}+k_3^{(j)})$ for each sample produced.
+Finally we produce a histogram from this sample $\theta^{(j)}$.
+
+The first thing to note is that in subfigure (a) when we approximate the marginals for $\hat{k}_4^{\text{QEA}}$ the density $\hat{\pi}_2(\cdot|D,\mathcal{M}_2)$ is much more peaked than $\hat{\pi}_1(\cdot|D,\mathcal{M}_1)$.
+As we might expect the same is true in reverse in subfigure (b).
+What is interesting here is that for both parameters, the marginal distribution which assumes $\mathcal{M}_2$ assigns a small density value to the true value of the parameter, while the marginals which assume $\mathcal{M}_1$ assign a relatively high density to the truth.
+We also draw attention to the similarities between the constrained model and the full data model.
+In subfigure (a) these two marginals are peaked around a similar value, and in subfigure (b) the full data and constrained models almost exactly coincide.
+This supports the hypothesis that the constrained approach provides a more accurate approximation to the dynamics of the full system.
+
+\subsection{Gene Regulatory Networks}\label{sec:grn}
+In this example, we look at a model of a gene regulatory network (GRN) mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering} used by a particular cell to regulate the amount of a protein, $P$, present.
+In this model, we again have reactions occurring at particular rates, $k_i$, between four different species.
+Here there is a special species, $G$, representing the gene responsible for building the protein, of which there is only one at a time, and this species can either be turned on or off.
+When the gene is off we denote it as $G'$.
+\begin{align}
+	P+P \xleftrightarrows[k_2]{k_1} D \quad & \quad G+D \xleftrightarrows[k_4]{k_3} G' \nonumber\\
+	G \xrightarrow{k_5} G+M \quad & \quad M\xrightarrow{k_6} M+P  \label{eqn:GRN_reactions} \\
+	P \xrightarrow{k_7} \emptyset \quad & \quad M \xrightarrow{k_8} \emptyset \nonumber
+\end{align}
+The eight reactions are given in Equation~\eqref{eqn:GRN_reactions}.
+In addition to the gene, $G$, and protein, $P$, the system also involves molecules of mRNA, $M$, and dimers, $D$, which are two proteins joined together.
+
+In this example protein is created by a molecule of mRNA at a rate $k_6$.
+The amount of protein in the cell is then governed by the rate of decay of $P$, and the number of mRNA molecules in the cell creating $P$.
+In turn the number of mRNA molecules is controlled by its own decay rate and whether or not the gene $G$, which produces the mRNA, is active.
+The gene can be turned on and off depending on how many dimers are present in the system, which is related to how many protein molecules there are.
+This completes a feedback cycle whereby if the protein population gets too high then the gene will turn off and the rate of new protein formation will decrease, causing the level to drop and the gene to turn back on.
+Figure~\ref{fig:GRN_cycle} shows what happens to the molecules in the system as the gene turns on and off.
+
+\begin{figure}
+NEW FIGURE HERE
+\end{figure}
+
+%As in Section~\ref{sec:chem_multiscale} we use the constrained multiscale approach to model the system.
+Due to the small scale of the system and the speed of the reactions, we are only able to measure the total amount of protein, $T$, and mRNA, $M$, while the number of dimers and whether or not the gene is switched on must be considered to be missing data.
+For this scenario, the likelihood function is intractable and so we must use a multiscale approximation of the system.
+Here we choose the constrained multiscale approximation due to its success in the previous example.
+
+The formulation here is slightly more complex.
+We define a fast subsystem which acts on the fast variables $\mathcal{F} = [P, D, G]$ and incorporates reactions $R_1$-$R_4$.
+For the true parameter values we have chosen, 90\% of reactions which occur in a particular time frame are one of these four reactions.
+\[
+	\text{Fast subsystem:} \quad P+P\xleftrightarrows[k_2]{k_1} D, \quad G+D\xleftrightarrows[k_4]{k_3} G'.
+\]
+We also define the slow variables, $T = P+2D+2(1-G)$ and $M$ which are held constant by these fast reactions.
+
+When we observe a cell, we cannot distinguish between the fast species and so don't know which of the first four reactions are happening.
+Due to this, we observe how many of reactions $R_5$-$R_8$ occur for each combination of the two slow variables $\mathcal{S} = [T, M]$.
+Since we don't observe $P$ and $G$ for every reaction, the propensities must be altered to take this into account,
+\begin{align}
+	\hat{\alpha}_5 = k_5\mathbb{E}[G|T=t],\quad&\quad \alpha_6 = k_6M, \notag\\
+	\hat{\alpha}_7 = k_7\mathbb{E}[P|T=t], \quad&\quad \alpha_8 = k_8M.\label{eqn:grn_propensities}
+\end{align}
+
+To calculate these expectations, we note that the fast subsystem is a deficiency zero system~\cite{anderson2010product}, and is also reversible. i.e. The system has four complexes, 
+\[
+	C_1 = P+P, \quad C_2 = D, \quad C_3 = G+D \quad \text{and} \quad  C_4 = G',
+\]
+there are two linkage classes (or two sets of reversible reactions), and the space spanned by the stoichiometries is two dimensional,
+\[
+	\text{span}\left\{\pm[-2, 1, 0, 0]^\top, \pm[0, -1, 0, -1]^\top\right\},
+\]
+hence the deficiency is $d = 4 - 2 - 2 = 0$. This means that, by \cite{anderson2010product}, the conditional probabilities have the closed form equation,
+\[
+	\mathbb{P}[\mathcal{F}|T=t] \propto \text{Poisson}(P_t;\bar{p})\text{Poisson}(D_t;\bar{d})\text{Poisson}(G_t;\bar{g})\text{Poisson}(1-G_t;1-\bar{g}),
+\]
+from which we can calculate the expected values of $P_t$, $D_t$, and $G_t$.
+The fast variables $P_t$, $D_t$ and $G_t$ are constrained to the line defined by $T=P+2D+2(1-G)=t$.
+The concentrations $\bar{p}$, $\bar{d}$ and $\bar{g}$, must be calculated as the solution to the nonlinear system produced by examining the steady state behaviour of the fast subsystem,
+\begin{align}
+	T &= p + 2d + 2(1-g), \notag\\
+	\frac{\text{d}p}{\text{d}t} &= -2k_1p^2+2k_2d = 0, \notag\\
+	\frac{\text{d}d}{\text{d}t} &= k_1p^2 - k_2d - k_3gd + k_4(1-g) = 0, \notag\\
+	\frac{\text{d}g}{\text{d}t} &= -k_3gd + k_4(1-g) = 0, \label{eqn:grn_nonlinear_system}
+\end{align}
+subject to $p \in [0, T]$, $g \in [0, 1]$, and $0 \leq d \leq \text{floor}(T/2)$.
+
+With these effective propensities, given in Equation~\eqref{eqn:grn_propensities}, we can calculate the posterior distribution as given in Equation~\eqref{eqn:chem_posterior}, again assigning Gamma priors to the eight reaction rates.
+We note that the reaction rates for the fast variables do not appear in the posterior density explicitly which only uses the propensities $\alpha_5$-$\alpha_8$, however they appear through the solution to the nonlinear system in Equation~\eqref{eqn:grn_nonlinear_system}.
+
+\subsubsection{Target Distribution}
+
+As in the previous chemical system example we use the target distribution as defined in Equation~\eqref{eqn:chem_posterior}.
+The application here is slightly more involved as discussed previously in Section~\ref{sec:grn}.
+We have an eight dimensional parameter space, with one reaction rate corresponding to one of eight reactions.
+The first four reactions, $R_1$-$R_4$, are combined into a fast subsystem for which we do not observe the time or identifier for which reaction has fired.
+We instead use the constrained multiscale approach to approximate what has happened in this subsystem between occurrences of the slow reactions firing (reactions $R_5$-$R_8$).
+For this reason, we apply the form in Equation~\eqref{eqn:chem_posterior} to reactions $R_5$-$R_8$, and allow the rates for reactions $R_1$-$R_4$ to implicitly influence the posterior density through the effective propensities, $\hat{\alpha}_5$ and $\hat{\alpha}_7$.
+
+As we mentioned earlier, for this problem we only consider the CMA's approximation to the system dynamics.
+
+The priors for this problem were chosen to be fairly uninformative with respect to the likelihood function for each reaction rate.
+A list of the hyper-parameters corresponding to each Gamma prior can be found in Table~\ref{tab:grn_priors}.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{|l|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|}
+	\hline
+	Dimension $i$ & 1 & 2& 3& 4 & 5 & 6 & 7 & 8 \\ \hline
+	$\alpha_i$ & 2 & 100 & 100 & 3 & 3 & 3 & 2 & 2\\ \hline
+	$\beta_i$ &50 & 0.02 & 1 & 1 & 0.6 & 1 & 50 & 50 \\ \hline
+\end{tabular}
+\caption{Hyper parameters for the Gamma priors on each of the reaction rates in the GRN example in Section~\ref{sec:grn}.}
+\label{tab:grn_priors}
+\end{table}
+
+\begin{figure}[htb]
+\centering
+\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/Applications/GRN_posterior"}}%
+\caption{Marginal densities of the posterior distribution for the GRN example in Section~\ref{sec:grn}.}
+\label{fig:GRN_posterior}
+\end{figure}
+
+The one- and two-dimensional marginal distributions for the full posterior distribution are displayed in Figure~\ref{fig:GRN_posterior}.
+The posterior does not contain many interesting correlation structures, however several dimensions have leptokurtic tails which are difficult for the standard PAIS algorithm to sample from.
+The marginal densities also vary over very different scales, which might require us to use a variant of the LPAIS algorithm.
+
+\subsubsection{Implementation}
+
+%% proposal methods
+%For the posterior distribution in \eqref{eqn:chem_QEA_posterior}, we consider several proposal methods.
+First we implement both the PAIS and MH algorithms with a Gaussian proposal distribution.
+In the case of the PAIS algorithm, this is a Gaussian mixture proposal distribution.
+The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard RWMH algorithm.
+We also compare the PAIS and MH algorithms when using a transport map proposal distribution.
+This proposal method was discussed in detail in Chapter~\ref{sec:TransportPAIS}.
+%
+%
+%Figure~\ref{fig:chem_coupling} shows how we define a bijective map, $\hat{T}$, between parameter space $\mathcal{X}$ and a reference space $R$.
+In practice we cannot ensure that the approximate map $\tilde{T}'$ is uniquely invertible over the whole of $R$ and so $\tilde{T}'$ is not truly bijective.
+This leads to problems for our strictly positive state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in $R$ do not necessarily map back on to $\mathcal{X}$.
+This motivates the use of a third intermediate space.
+When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
+MH-RW (RWMH) and MH-TRW, as well as PAIS-RW and PAIS-TRW.
+%
+%We also consider how these algorithms perform when they are applied to the log of the reaction rates.
+We choose to use this log transformation since it converts our strictly positive parameter space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset \mathbb{R}^d$.
+This allows us to define $T$ between two subsets of $\mathbb{R}^d$, which means that even if $\tilde{T}$ is not uniquely invertible in some region of $R$, all possibilities are valid proposals.
+As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
+MH-logRW and MH-logTRW, with, PAIS-logRW and PAIS-logTRW.
+We apply the eight RW and logRW proposal distributions, with and without transport map acceleration, which were discussed in Section~\ref{sec:chem_multiscale} to this posterior.
+Again the intermediate log space is required so that our transport map is a function $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$, rather than a map from $\mathbb{R}_+^d$ to $\mathbb{R}^d$.
+
+%% choices made in transport map
+%For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=4$ as the basis functions, i.e.
+%\[
+%	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
+%\]
+%and
+%\[
+%	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
+%\]
+%Since $\mathcal{X}$ is four-dimensional, this yeilds a total of 125 map coefficients across the four index sets.
+This number can be reduced by using smaller index sets as discussed in~\cite{parno2014transport}.
+The transport map setup has not changed from the previous example.
+However, due to the scaling of the weight we have had to be more careful about which samples we use to build our map.
+We require that the objective function $C(T)$, from Equation~\eqref{eqn:TPAIS_objective} is convex, which requires that the second derivative, given in Equation~\eqref{eqn:TPAIS_hessian}, is positive definite.
+This positive definite property depends on all weights in our sample being strictly positive, which is not always possible on a computer.
+For this reason we filter out any samples from the optimisation problem where the weight is a numeric zero.
+This does not affect the validity of the map since a weight zero would not contribute to the expectation, and we do not require (and could not enforce) an exact map.
+
+In all eight algorithms we adapt the proposal variances during the burn-in phase of the algorithms to avoid behaviour like that seen in Example~\ref{sec:ex_W1} where we needed a huge ensemble size to approximate the posterior with our mixture of isotropic Gaussian kernels.
+
+%% choice of resampler?
+%As with the mixture model in Section~\ref{sec:app_mixture}, we will use the AMR resampler with an ensemble size of $M=500$.
+This increase in ensemble size from Chapter~\ref{sec:PAIS} compensates for the increase in parameter dimension.
+We again use the AMR resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size again compensates for the increase in dimension (from four to eight parameters.)
+
+%% analysis method
+%To measure the convergence of the sampling methods in this section, we will compare the convergence of the mean of each parameter.
+We approximate $\mathbb{E}(\mathbf{k})$ using 2.4 billion samples from the MH-RW algorithm.
+As in the previous chemical reaction example, we measure convergence of the algorithms only through the convergence of the mean, for computational ease.
+We use the sample Mahalanobis distance, with `true' covariance and `true' mean built from a sample of size 2.4 billion using the MH-RW algorithm.
+
+\subsubsection{Convergence results}
+
+The scaling parameter selection is done by optimising the acceptance rate for the MH algorithms, and optimising the effective sample size for the PAIS algorithms.
+The optimal parameters are given in Table~\ref{tab:grn_scaling_parameters}.
+
+\begin{table}[!h]
+	\centering
+	\begin{tabular}{lrrrrrrrr}
+	\toprule
+		 & \multicolumn{4}{c}{MH} & \multicolumn{4}{c}{PAIS} \\ \cmidrule(lr){2-5} \cmidrule(lr){6-9}
+		& RW & logRW & TRW & log-TRW & RW & logRW & TRW & log-TRW \\ \midrule
+		$\beta_{\%}^*$	 	& 1.0e-1 & 3.9e-1 & 1.0e-0 &6.0e-1 & - & - & - & - \\
+		$\beta_{\text{ESS}}^*$	& - 	        & -           & -            & -            & 1.0e-0 & 1.3e-0 & 8.0e-1 & 5.0e-1 \\
+		ESS/$M$		 		& - 	        & -           & -            & -            & 3.5e-4 & 4.7e-2 & 4.8e-2 & 1.6e-1  \\
+	\bottomrule
+	\end{tabular}
+	\caption{Scaling parameters for the sampling algorithms applied to the GRN example in Section~\ref{sec:grn}.}
+	\label{tab:grn_scaling_parameters}
+\end{table}
+
+From the effective sample sizes shown in Table~\ref{tab:grn_scaling_parameters} we can see the improvement to the efficiency of the algorithm both by proposing on the log space, and by transforming the parameter space into something closer to Gaussian.
+
+Due to the numerical cost of calculating the full relative $L^2$ errors for this posterior, we quantify the convergence speeds of these algorithms using the sample Mahalanobis distance between the sample mean and the `true' mean.
+This convergence analysis is shown in Figure~\ref{fig:grn_L2}.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Sampling algorithms on $\mathcal{X}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
+\subfigure[Sampling algorithms on $\mathcal{Y}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
+\caption{Convergence of the GRN example described in Section~\ref{sec:grn}.}
+\label{fig:grn_L2}
+\end{figure}
+
+The first thing to note is that the first PAIS measurement happens after 2500 samples, whereas the first MH measurement occurs after 50 samples.
+This is due to the much larger ensemble size required to sample in higher dimensions.
+We believe that an ensemble size of 2500 is over-cautious in this example and we could have used a smaller ensemble size.
+We also think that the required ensemble size to sample from these posteriors is reduced by the use of the transport map.
+This is due to the way the tail behaviour is improved by the transformation.
+
+The second obvious feature of these convergence plots is that the PAIS algorithms outperform the MH algorithms by a large margin - roughly a reduction of two orders of magnitude in the relative error over the same number of samples.
+A less positive result is that for this example the PAIS-TRW algorithm has found a map which has prevented further convergence of the mean.
+This is likely because one dimension of the transport map is sending proposals onto the negative real line and so all samples are being assigned a zero weight.
+This is behaviour which was anticipated, and it is always suggested that the transport map is constructed as a map $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$ rather than $T\colon\mathcal{X}\rightarrow\mathbb{R}^d$ where $\mathcal{X}\subset\mathbb{R}^d$.
+The large increase in the effective sample size observed between the PAIS-logRW and PAIS-logTRW algorithms is converted into an estimate which is twice as accurate after 1 million samples.
+
+A similar pattern is seen in the MH algorithms, where the MH-TRW performs the worst, and the MH-logTRW algorithm performs the best, although only marginally.
 

### 2018-01-22 TPAIS_2018.tex
a typo
Funding bodies

@@
   multiscale stochastic reaction networks}
-\author{Simon Cotter, Yannis Kevrekidis, Paul Russell}
+\author{Simon Cotter (EPSRC and INI), Yannis Kevrekidis (?), Paul
+  Russell (EPSRC)}
 \begin{document}
@@
 \ref{sec:conv} we seek to show the advantages of this approach through
-the analysis of test problems. In Section
+the analysis of a challenging test problem. In Section
 \ref{sec:multi} we consider how likelihoods can be approximated using

### 2018-01-23 TPAIS_2018.tex
Completed first run through of numerics sections

@@
 \centering
-\subfigure[Original sample $\theta$ from RWMH algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
+\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
 \subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
@@
 density does not look a challenging one to sample from, with even
-relatively simple MCMC methods such as RWMH. The pullback from
+relatively simple MCMC methods such as MH-RW. The pullback from
 reference space, in Figure~\ref{fig:R2_transport}, is an exact match
@@
 \end{tabular}
-\caption{Optimal scaling parameters for the transport map based algorithms applied to $R_1$.}
+\caption{Optimal scaling parameters for the transport map based
+  algorithms applied to $R_1$, optimising for the $L^2$ error and
+  the effective sample size (ESS) for PAIS algorithms, and for average
+  acceptance rate for MH algorithms.}
 \label{tab:R2_opt_scaling}
@@
 	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
-	\prod_{i=1}^{N_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
+	\prod_{i=1}^{n_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
 		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
-				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
+				\mathbf{K}_i\log
+                  \bar{\alpha}_0(\mathbf{Y}_i; \mathbf{k}) -
+                  \mathbf{T}_i \bar{\alpha}_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
 %		& \qquad\qquad\qquad
-				+\sum_{j=1}^{N_r} \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
+				+\sum_{j=1}^{n_r} \mathbf{R}_{ij}\log \bar{\mathbf{p}}_j(\mathbf{Y}_i; \mathbf{k})
 			\Bigg]  \notag\\
-		&	\qquad\qquad\qquad{} + \sum_{i=1}^{N_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}. \label{eq:Apos}
+		&	\qquad\qquad\qquad{} + \sum_{i=1}^{n_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}, \label{eq:Apos}
 \end{align}
-
+where $\bar{\alpha}_0(\mathbf{Y}_i; \mathbf{k}) = \sum_{j=1}^{n_r}
+\bar{\alpha}_j(\mathbf{Y}_i; \mathbf{k})$ is the total of the
+effective propensities, $n_r$ is the number of slow reactions,
+$\bar{\mathbf{p}}(\mathbf{Y}_i; \mathbf{k})$ is the multiscale
+approximation of the expected proportion of each slow reaction at
+state $\mathbf{Y}_i$ and with reaction rates $\mathbf{k}$ and the
+data is limited only to changes in the slow variable due to the
+occurrence of slow reactions.
 \section{Numerical Examples}\label{sec:num}
@@
 
+We create trajectories of this system with the following parameters:****
+\begin{equation}
+k_1 = ? etc
+\end{equation}
+over the time period ****$t \in [0,?]$.
+
 We assume that we are only able to observe
@@
 application of a multiscale approximation, as detailed in Section
-\ref{sec:multi}, and in more detail in ****CITE****. Under the QEA,
-this is given by
+\ref{sec:multi}, and in more detail in ****CITE****. These effective
+propensities can often be computed without the need for expensive
+simulations, in particular when the fast subsystem is deficiency zero,
+which is often the case****CITE DIF ZERO PAPERS****. For this very
+simple example, the dependance of these effective
+propensities on the original rate parameters can be explicitly
+understood. Under the QEA,
+the effective rate of degradation of the slow variable $S$ is given by
 \[
@@
 
-Our observations are uninformative about the reaction rates $k_2, k_3$, and $k_4$, as there
-is a sub-manifold $\mathcal{M} \subset \mathcal{X}$ along which the effective rate
-$\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
+Our observations are uninformative about the reaction rates $k_2,
+k_3$, and $k_4$, as for each multiscale approximation there
+is a manifold $\mathcal{M} \subset \mathcal{X}$ along which the effective propensity
+$\hat{\alpha}_4$ is invariant, leading to a highly ill-posed inverse
 problem. This type of problem is notoriously difficult to
@@
 We aim to recover three different posterior distributions. In the
-first, which is of the form \eqref{eqn:chem_posterior} we assume that we can fully and exactly observe the whole state
-vector for the whole of the observation time window. In the second and
-third, the observations are restricted to observations of the slow
-variable $S = X_1 + X_2$. In these two examples, we approximate the
+first, which is of the form \eqref{eqn:chem_posterior}, we assume that we can fully and exactly observe the whole state
+vector for the whole of the observation time window $t \in [0,T]$. In
+the second and
+third, the same data are used, but restricted to observations of the slow
+variable $S = X_1 + X_2$. In these last two examples, we approximate the
 posterior using \eqref{eq:Apos} with QEA and CMA approximations of the
@@
 	\hline
-	parameter/dimension & 1 & 2 & 3 & 4 \\ \hline
+	Parameter$\backslash$Dimension $i$ & 1 & 2 & 3 & 4 \\ \hline
 	$\alpha_i$ & 150 & 5 & 5 & 3 \\
-	$\beta_i$ & 15/9 & 5/12 & 5/12 & 1 \\ \hline
+	$\beta_i$ & $\frac{15}{9}$ & $\frac{5}{12}$ & $\frac{5}{12}$ & $1$ \\ \hline 
 \end{tabular}
@@
 algorithm this is equivalent to a vanilla random walk.
-The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard RWMH algorithm.
+The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard MH-RW algorithm.
 We also compare the PAIS and MH algorithms when using a transport map proposal distribution.
@@
 	\end{tikzpicture}
-	\caption{Couplings between parameter space and reference when using the transport map proposal method to propose moves on parameter space.}
+
+	\begin{tikzpicture}
+		\pgfmathsetseed{1}
+
+		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
+		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
+		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};
+
+		\draw (A) \irregularcircle{1cm}{2.5mm};
+		\draw (B) \irregularcircle{1cm}{1.1mm};
+		\draw (C) \irregularcircle{1cm}{0.5mm};
+
+		\draw [->] (A) -- (B);
+		\draw [->] (B) -- (C);
+
+		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
+		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};
+
+		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
+		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
+		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
+	\end{tikzpicture}
+	\caption{Top: Couplings between parameter space and reference
+          when using the transport map. Bottom: Couplings between parameter space and reference space
+          when using the transport map proposal with a $\log$ preconditioner.}
 	\label{fig:chem_coupling}
@@
 
-Figure~\ref{fig:chem_coupling} shows how we define a bijective map,
+Figure~\ref{fig:chem_coupling} (top) shows how we define a bijective map,
 $\hat{T}$, between parameter space $\mathcal{X}$ and a reference space
@@
 When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
-MH-RW (RWMH) and MH-TRW, as well as PAIS-RW and PAIS-TRW.
+MH-RW  and MH-TRW, as well as PAIS-RW and PAIS-TRW.
 In practice we cannot ensure that the approximate map $\tilde{T}$
@@
 state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in
-$R$ do not necessarily map back onto $\mathcal{X}$. This motivates the
-use of an intermediate space.
-
-We also consider how these algorithms perform when they are applied to
-the log of the reaction rates. We choose to use this log
-transformation since it converts our strictly positive parameter
+$R$ do not necessarily map back onto $\mathcal{X}$.
+
+Therefore we also consider how these algorithms perform when the transport map
+is preconditioned by the logarithmic function applied to the parameters. We choose to use this log
+preconditioner since it converts our strictly positive parameter
 space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset
 \mathbb{R}^d$. This allows us to define $T$ between two subsets of
-$\mathbb{R}^d$, which means that even if $\tilde{T}$ is not uniquely
-invertible in some region of $R$, all possibilities are valid
-proposals. It also reduces the work required of the map, which since
+$\mathbb{R}^d$, which means that all proposals made on the reference
+space are mapped to valid
+proposals on the true parameter space. It also reduces the work required of the map, which since
 we use only a finite dimensional approximation, can improve
@@
 
-\begin{figure}
-	\centering
-	\begin{tikzpicture}
-		\pgfmathsetseed{1}
-
-		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
-		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
-		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};
-
-		\draw (A) \irregularcircle{1cm}{2.5mm};
-		\draw (B) \irregularcircle{1cm}{1.1mm};
-		\draw (C) \irregularcircle{1cm}{0.5mm};
-
-		\draw [->] (A) -- (B);
-		\draw [->] (B) -- (C);
-
-		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
-		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};
-
-		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
-		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
-		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
-	\end{tikzpicture}
-	\caption{Couplings between parameter space, an intermediate space and reference space when using the transport map proposal method to propose moves on a log space.}
-	\label{fig:chem_log_coupling}
-\end{figure}
-
-Figure~\ref{fig:chem_log_coupling} displays the composition of maps when we include this intermediate space $\mathcal{Y}$.
-Every proposal in $R$ results in a valid proposal on $\mathcal{X}$.
+Figure~\ref{fig:chem_coupling} (bottom) displays the composition of maps
+when we include this $\log$ preconditioner leading to an intermediate space $\mathcal{Y}$.
 The inclusion of this additional map means that we must again alter our importance weight definition to reflect the pullback from $R$ through $\mathcal{Y}$.
@@
 \]
-where the first determinant is as we saw in the previous chapter, and the second is
+where the first determinant is computed as in
+\eqref{eqn:separable_jacobian}, and the second is given by
 \[
@@
 	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{PAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
-	& & $\delta_\%$ & & $\delta_{\text{ESS}}$ & ESS \\ \midrule
+	& & $\delta$ & & $\delta$ & ESS \\ \midrule
 	RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
@@
 
-The optimal scaling parameters for the proposal distributions are given in Table~\ref{tab:chem_multiscale_scaling}.
-We note that for MH-TRW and PAIS-TRW the scaling parameter is near to 1, particularly for MH-RW, this is what we should expect since the proposal is made on a reference space which should be near to $\mathcal{N}_d(0, \text{I})$.
-The same should be true for the MH-logTRW and PAIS-logTRW algorithms since these have the same target reference space, however the optimal scaling parameters here are much smaller.
-We see that the ESS is higher for the algorithms which sample on $\mathcal{Y}$, and we expect that convergence will be fastest for the PAIS-logTRW algorithm.
+The optimal scaling parameters for the proposal distributions are
+given in Table~\ref{tab:chem_multiscale_scaling}. These are
+precomputed by running the algorithms for different parameter
+values. MH-RW based algorithms are tuned to an acceptance rate close
+to the optimal $23.4\%$, and PAIS algorithms are tuned to maximise the
+effective sample size (ESS). The optimal parameter values for the MH
+algorithms are not particularly informative about their performance,
+since they relate to algorithms operating on different
+transformations. The ESS is highest for the logTRW variant
+of the PAIS algorithm, as we would hope.
 
@@
 \centering
-\subfigure[Sampling algorithms on $\mathcal{X}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
-\subfigure[Sampling algorithms on $\mathcal{Y}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
+\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
+\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
 \caption{Convergence of the constrained multiscale example described in Section~\ref{sec:chem_multiscale}.}
@@
 
-The PAIS variants with RW and TRW proposals perform similarly on both sample spaces.
-When sampling on $\mathcal{X}$ the transport map is not quite as efficient, largely due to the difficulties discussed in the previous section i.e.
-many proposals are made which do result in negative reaction rates.
-Sampling on $\mathcal{Y}$ leads to more comparable Monte Carlo errors, with the logTRW being apparently slightly less stable.
+The PAIS variants with RW and TRW proposals perform similarly on both
+sample spaces. When sampling without a preconditioner, the transport map is not
+quite as efficient, largely due to the difficulties discussed in the
+previous section i.e. many proposals are made which do result in
+negative reaction rates.
+Sampling with transport maps with a $\log$ preconditioner leads to more comparable Monte Carlo errors, with the logTRW being apparently slightly less stable.
 This proposal method becomes more stable as we increase either the ensemble size, or the number of iterations between updates of the transport map, $T$.
@@
 \subfigure[Reference space mapped to by $T\circ\log$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
-\caption{Reference space for the problem described in Section~\ref{sec:multiscale_posterior} found using the TRW and logTRW proposal distributions.
+\caption{Push forward of the target measure on reference space for the problem described in Section~\ref{sec:multiscale_posterior} found using the TRW and logTRW proposal distributions.
 Components are linked by the relation $r_i = T_i'(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
@@
 
-Particularly in Figure~\ref{fig:chem_reference_spaces} (b) we see that $\text{var}(r_1)$ and $\text{var}(r_4)$ are much smaller than $\text{var}(r_2)$ and $\text{var}(r_3)$.
-To combat this we have a number of choices, we might wish to use two different scaling parameters to match these scales, which would require knowledge of the reference space before beginning sampling.
-We could alternatively use a proposal distribution such as in the LPAIS algorithm to adaptively learn this at each iteration.
+In Figure~\ref{fig:chem_reference_spaces} (b) we see that
+$\text{var}(r_1)$ and $\text{var}(r_4)$ are much smaller than
+$\text{var}(r_2)$ and $\text{var}(r_3)$. To combat this we have a
+number of choices. We might wish to learn the covariance structure of
+the push forward of the target onto the reference space, and
+incorporate this information into the covariances of the mixture
+components in an adaptive scheme.
 Another option is to increase the total order of our index set.
-For these numerics we have chosen $p=3$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
+For these numerics we have chosen $p=3$, but we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
 
@@
 
-The convergence analysis has been performed for the constrained approach to this multiscale system.
-We now look at the differences between the constrained and QEA posterior distributions.
-Recall that the approaches differed only in the form of the effective degradation rate $\hat{k}_4$,
-\[
-	\hat{k}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{k}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
-\]
-This difference in the denominator causes a shift in the parameters as can be seen in Figure~\ref{fig:chem_diff}. The figure shows the difference in posteriors,
+The convergence analysis has been performed for the constrained
+approximation of the posterior distribution. We now look at the
+differences in the approximations of the posterior distribution
+arising in this problem between the CMA and QEA. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
+\[
+	\hat{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{\alpha}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
+\]
+This difference in the denominator causes a shift in the manifold on
+which the posterior is concentrated, as can be seen in
+Figure~\ref{fig:chem_diff}. The figure shows the difference in
+posteriors densities
 \begin{equation}\label{eqn:chem_diff}
@@
 
-Since the two posteriors have been approximated using an MCMC sample, there is a significant amount of noise, particularly in the tails of the distributions We can see that the differences in the marginals for $k_1, k_2,$ and $k_3$ are relatively small and the differences are largely positive.
-These positive differences tell us that the constrained approach claims to be more informative about the values of these first three parameters.
-However the marginal for $k_4$ varies by a significant amount and is more negative.
-This means that the mean estimates for $k_4$ differ by a large amount, the posterior for the QEA is most peaked.
-
-% compare solution with full data.
-We now consider how we should interpret the information given by these models.
-The QEA assumption tells us that we cannot observe the parameters $k_2$, $k_3$ and $k_4$ independently from the reduced data set, but we can observe the quantity $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$.
-Similarly, the constrained approach tells us that we are able to observe the quantity $\hat{k}_4^{\text{CMA}} = k_2k_4/(k_2+k_3+k_4)$.
-To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most accurate and informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model, with all reactions observed.
+Since the two posteriors have been approximated using an MCMC sample,
+there is a significant amount of noise, particularly in the tails of
+the distributions. We can see that the differences in the marginals
+for $k_1, k_2,$ and $k_3$ are relatively small, but the marginal for
+$k_4$ varies by a significant amount.
+
+In the QEA approximation, the effective rate of degradation of $S$ is
+given by $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$, and as such, our
+observations should be informative about this quantity. Using the CMA,
+this effective rate is approximated by $\hat{k}_4^{\text{CMA}} =
+k_2k_4/(k_2+k_3+k_4)$. To validate our inferences on $k_2$, $k_3$ and
+$k_4$ we would like to discover which model is most accurate and
+informative about these quantities, and which model gives us results
+which are most similar to that which can be obtained from the full
+model, with all reactions (including fast reactions) observed.
 
@@
 
-We now label the constrained model as $\mathcal{M}_1$, the model arising from the QEA as $\mathcal{M}_2$, and the full data model as $\mathcal{M}_0$.
-Under $\mathcal{M}_1$, the parameter $\theta_1 = (k_1, \hat{k}_4^{\text{CMA}})^\top$, and under $\mathcal{M}_2$, the parameter $\theta_2 = (k_1, \hat{k}_4^{\text{QEA}})^\top$.
-The full model $\mathcal{M}_0$ observes all four parameters $\theta_0 = (k_1, k_2, k_3, k_4)^\top$.
+We now label the constrained model as $\mathcal{M}_1$, the model
+arising from the QEA as $\mathcal{M}_2$, and the full data model as
+$\mathcal{M}_0$. Model $\mathcal{M}_1$ is dependent on the parameter
+$\theta_1 = (k_1, \hat{k}_4^{\text{CMA}})^\top$, and model
+$\mathcal{M}_2$ is dependent on the parameters $\theta_2 = (k_1,
+\hat{k}_4^{\text{QEA}})^\top$. The full model $\mathcal{M}_0$ is
+dependent on all four parameters $\theta_0 = (k_1, k_2, k_3, k_4)^\top$.
 The marginal densities for the data, evaluated at the observed data, given the models are displayed in Table~\ref{tab:chem_Bayes_marginals}.
@@
 
-Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals} we see that we should of course prefer the model in which we observe all reactions perfectly, however this model is not significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$, ~\cite{kass1995bayes}).
-Again the constrained model is not substantially more attractive than the QEA model when we consider the Bayes factor $B_{1,2} = 1.96 < 3.2$.
-The Bayes factor $B_{0,2} = 4.1 > 3.2$ does tell us that we should significantly prefer the full model to the QEA approximation of the posterior.
-These Bayes factors present a weak argument that that the constrained model provides us with a better description of the data than the QEA model.
-We can also present this information graphically, which might provide us with a greater justification for preferring the constrained approximation of the posterior.
+Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals}
+we see that we should of course prefer the model $\mathcal{M}_0$ in
+which we observe all reactions perfectly. However this model is not
+significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$,
+~\cite{kass1995bayes}). The Bayes factor $B_{0,2} = 4.1 > 3.2$ tells
+us that we should significantly prefer the full model to the QEA
+approximation of the posterior. These Bayes factors present a
+relatively weak
+argument that the constrained approximation of the posterior is
+superior to the QEA approximation.
 
@@
 
-Figure~\ref{fig:chem_model_comp} displays the marginal distributions for the parameters $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$.
-For each of these two observable parameters we obtain approximations for three marginal distributions, $\pi_i(\cdot|D, \mathcal{M}_i)$, $i = 0, 1, 2$.
-Each of these approximations has been produced by marginalising a sample drawn from the full posterior $\pi_i(\mathbf{k}|D, \mathcal{M}_i)$ defined in terms of the respective model $\mathcal{M}_i$.
-In other words, to calculate $\hat{\pi}_1(\hat{k}_4^{\text{QEA}}|D)$, we first draw a sample using the PAIS algorithm targeting the posterior density $\pi_1(\mathbf{k}|D,\mathcal{M}_1)$.
-We then calculate $\theta^{(j)} = k_2^{(j)}k_4^{(j)}/(k_2^{(j)}+k_3^{(j)})$ for each sample produced.
-Finally we produce a histogram from this sample $\theta^{(j)}$.
-
-The first thing to note is that in subfigure (a) when we approximate the marginals for $\hat{k}_4^{\text{QEA}}$ the density $\hat{\pi}_2(\cdot|D,\mathcal{M}_2)$ is much more peaked than $\hat{\pi}_1(\cdot|D,\mathcal{M}_1)$.
-As we might expect the same is true in reverse in subfigure (b).
-What is interesting here is that for both parameters, the marginal distribution which assumes $\mathcal{M}_2$ assigns a small density value to the true value of the parameter, while the marginals which assume $\mathcal{M}_1$ assign a relatively high density to the truth.
-We also draw attention to the similarities between the constrained model and the full data model.
-In subfigure (a) these two marginals are peaked around a similar value, and in subfigure (b) the full data and constrained models almost exactly coincide.
-This supports the hypothesis that the constrained approach provides a more accurate approximation to the dynamics of the full system.
-
-\subsection{Gene Regulatory Networks}\label{sec:grn}
-In this example, we look at a model of a gene regulatory network (GRN) mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering} used by a particular cell to regulate the amount of a protein, $P$, present.
-In this model, we again have reactions occurring at particular rates, $k_i$, between four different species.
-Here there is a special species, $G$, representing the gene responsible for building the protein, of which there is only one at a time, and this species can either be turned on or off.
-When the gene is off we denote it as $G'$.
+Figure~\ref{fig:chem_model_comp} displays more compelling evidence in
+this direction. Here we show the marginal distributions for the
+quantities $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$ with
+respect to each of the three posterior distributions, $\pi_i(\cdot|D,
+\mathcal{M}_i)$, $i = 0, 1, 2$.
+
+Firstly, in
+Figure~\ref{fig:chem_model_comp} (a), displaying the marginals for
+$\hat{k}_4^{\text{QEA}}$ we see that $\pi_0$, the density arising from
+the full model with all reactions observed, is peaked a distance away
+from $\pi_2$, QEA approximation. The CMA approximation $\pi_1$, in contrast, is
+peaked close to peak of $\pi_0$, but there is more uncertainty here
+since we do not observe the fast reactions.
+
+In contrast, in Figure~\ref{fig:chem_model_comp} (b), the marginal
+densities for $\hat{k}_4^{\text{CMA}}$ for $\pi_0$ and $\pi_1$ are
+barely distinguishable, even with the vastly reduced data available in
+the posterior $\pi_1$ with CMA approximation. The QEA approximation
+$\pi_2$ is again peaked far away from both of the other distributions.
+This appears to demonstrate that the data regarding the slow variable
+$S$ is only informative about $k_1 $ and $\hat{k}_4^{\text{CMA}}$, as
+predicted by the CMA multiscale approximation.
+
+\subsection{Gene Regulatory Network}\label{sec:grn}
+In this example, we look at a model of a gene regulatory network (GRN)
+mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering}
+used by a particular cell to regulate the amount of a protein, $P$,
+present. Proteins can bind in pairs to form a dimer $D$. The dimer can
+bind to the gene $G$ to a ``switched off'' state $G'$, in which the
+production of MRNA molecules $M$ is inhibited. The MRNA encodes for
+production of the protein $P$, and both $P$ and $M$ can degrade. 
+The eight reactions are given in Equation~\eqref{eqn:GRN_reactions}.
 \begin{align}
@@
 \end{align}
-The eight reactions are given in Equation~\eqref{eqn:GRN_reactions}.
-In addition to the gene, $G$, and protein, $P$, the system also involves molecules of mRNA, $M$, and dimers, $D$, which are two proteins joined together.
-
-In this example protein is created by a molecule of mRNA at a rate $k_6$.
-The amount of protein in the cell is then governed by the rate of decay of $P$, and the number of mRNA molecules in the cell creating $P$.
-In turn the number of mRNA molecules is controlled by its own decay rate and whether or not the gene $G$, which produces the mRNA, is active.
-The gene can be turned on and off depending on how many dimers are present in the system, which is related to how many protein molecules there are.
-This completes a feedback cycle whereby if the protein population gets too high then the gene will turn off and the rate of new protein formation will decrease, causing the level to drop and the gene to turn back on.
-Figure~\ref{fig:GRN_cycle} shows what happens to the molecules in the system as the gene turns on and off.
-
-\begin{figure}
-NEW FIGURE HERE
-\end{figure}
-
-%As in Section~\ref{sec:chem_multiscale} we use the constrained multiscale approach to model the system.
-Due to the small scale of the system and the speed of the reactions, we are only able to measure the total amount of protein, $T$, and mRNA, $M$, while the number of dimers and whether or not the gene is switched on must be considered to be missing data.
-For this scenario, the likelihood function is intractable and so we must use a multiscale approximation of the system.
-Here we choose the constrained multiscale approximation due to its success in the previous example.
-
-The formulation here is slightly more complex.
-We define a fast subsystem which acts on the fast variables $\mathcal{F} = [P, D, G]$ and incorporates reactions $R_1$-$R_4$.
-For the true parameter values we have chosen, 90\% of reactions which occur in a particular time frame are one of these four reactions.
-\[
-	\text{Fast subsystem:} \quad P+P\xleftrightarrows[k_2]{k_1} D, \quad G+D\xleftrightarrows[k_4]{k_3} G'.
-\]
-We also define the slow variables, $T = P+2D+2(1-G)$ and $M$ which are held constant by these fast reactions.
-
-When we observe a cell, we cannot distinguish between the fast species and so don't know which of the first four reactions are happening.
-Due to this, we observe how many of reactions $R_5$-$R_8$ occur for each combination of the two slow variables $\mathcal{S} = [T, M]$.
-Since we don't observe $P$ and $G$ for every reaction, the propensities must be altered to take this into account,
+
+We create trajectories of this system with the following parameters:****
+\begin{equation}
+k_1 = ? etc
+\end{equation}
+over the time period ****$t \in [0,?]$.
+
+T-cells can be manufactured which bind to
+proteins of a certain type, and cause them to phosphoresce, allowing
+for approximate observations of the concentration levels of that
+protein. However, the T-cells will not necessarily differentiate
+between monomers and dimers of the protein. We replicate such a
+scenario, in a zero-observational noise setting, by assuming that we
+can only observe $T=P + 2D + 2G'$, the total number of protein
+molecules in the system, alongside $M$, the concentration of MRNA. The
+number of dimers, and the status of the genetic switch, are assumed
+unobservable. In
+order to sample from the posterior distribution on the reaction
+parameters in the system given these observations, we can either integrate over all possible
+trajectories of the fast variables (which is computationally
+intractable), or as in Section \ref{sec:chem_multiscale} we can use a
+multiscale method to approximate the posterior.
+
+The effective dynamics of $T$ and $M$ are then given by:
 \begin{align}
-	\hat{\alpha}_5 = k_5\mathbb{E}[G|T=t],\quad&\quad \alpha_6 = k_6M, \notag\\
-	\hat{\alpha}_7 = k_7\mathbb{E}[P|T=t], \quad&\quad \alpha_8 = k_8M.\label{eqn:grn_propensities}
+	\emptyset \xrightarrow{\hat{\alpha}_5} M \quad & \quad M\xrightarrow{\alpha_6} M+T  \label{eqn:GRN_eff} \\
+	T \xrightarrow{\hat{\alpha}_7} \emptyset \quad & \quad M \xrightarrow{\alpha_8} \emptyset \nonumber
 \end{align}
-
-To calculate these expectations, we note that the fast subsystem is a deficiency zero system~\cite{anderson2010product}, and is also reversible. i.e. The system has four complexes, 
-\[
-	C_1 = P+P, \quad C_2 = D, \quad C_3 = G+D \quad \text{and} \quad  C_4 = G',
-\]
-there are two linkage classes (or two sets of reversible reactions), and the space spanned by the stoichiometries is two dimensional,
-\[
-	\text{span}\left\{\pm[-2, 1, 0, 0]^\top, \pm[0, -1, 0, -1]^\top\right\},
-\]
-hence the deficiency is $d = 4 - 2 - 2 = 0$. This means that, by \cite{anderson2010product}, the conditional probabilities have the closed form equation,
-\[
-	\mathbb{P}[\mathcal{F}|T=t] \propto \text{Poisson}(P_t;\bar{p})\text{Poisson}(D_t;\bar{d})\text{Poisson}(G_t;\bar{g})\text{Poisson}(1-G_t;1-\bar{g}),
-\]
-from which we can calculate the expected values of $P_t$, $D_t$, and $G_t$.
-The fast variables $P_t$, $D_t$ and $G_t$ are constrained to the line defined by $T=P+2D+2(1-G)=t$.
-The concentrations $\bar{p}$, $\bar{d}$ and $\bar{g}$, must be calculated as the solution to the nonlinear system produced by examining the steady state behaviour of the fast subsystem,
-\begin{align}
-	T &= p + 2d + 2(1-g), \notag\\
-	\frac{\text{d}p}{\text{d}t} &= -2k_1p^2+2k_2d = 0, \notag\\
-	\frac{\text{d}d}{\text{d}t} &= k_1p^2 - k_2d - k_3gd + k_4(1-g) = 0, \notag\\
-	\frac{\text{d}g}{\text{d}t} &= -k_3gd + k_4(1-g) = 0, \label{eqn:grn_nonlinear_system}
-\end{align}
-subject to $p \in [0, T]$, $g \in [0, 1]$, and $0 \leq d \leq \text{floor}(T/2)$.
-
-With these effective propensities, given in Equation~\eqref{eqn:grn_propensities}, we can calculate the posterior distribution as given in Equation~\eqref{eqn:chem_posterior}, again assigning Gamma priors to the eight reaction rates.
-We note that the reaction rates for the fast variables do not appear in the posterior density explicitly which only uses the propensities $\alpha_5$-$\alpha_8$, however they appear through the solution to the nonlinear system in Equation~\eqref{eqn:grn_nonlinear_system}.
+Here, the propensities are displayed, as opposed to the rates, with
+approximations of the propensities $\hat{\alpha}_5$ and
+$\hat{\alpha}_7$ to be computed. We omit the derivation of these
+effective propensities, using the CMA methodology, for brevity, but
+interested readers can see more details in ****CITE THESIS****.
 
 \subsubsection{Target Distribution}
-
-As in the previous chemical system example we use the target distribution as defined in Equation~\eqref{eqn:chem_posterior}.
-The application here is slightly more involved as discussed previously in Section~\ref{sec:grn}.
-We have an eight dimensional parameter space, with one reaction rate corresponding to one of eight reactions.
-The first four reactions, $R_1$-$R_4$, are combined into a fast subsystem for which we do not observe the time or identifier for which reaction has fired.
-We instead use the constrained multiscale approach to approximate what has happened in this subsystem between occurrences of the slow reactions firing (reactions $R_5$-$R_8$).
-For this reason, we apply the form in Equation~\eqref{eqn:chem_posterior} to reactions $R_5$-$R_8$, and allow the rates for reactions $R_1$-$R_4$ to implicitly influence the posterior density through the effective propensities, $\hat{\alpha}_5$ and $\hat{\alpha}_7$.
-
-As we mentioned earlier, for this problem we only consider the CMA's approximation to the system dynamics.
+As alluded to in the previous section, we aim to sample from the CMA
+approximation of the posterior distribution as defined in
+Equation~\eqref{eqn:chem_posterior}. This example is more involved
+than the previous one, since we have an eight dimensional parameter
+space corresponding to the eight reaction rates in the system
+\eqref{eqn:GRN_reactions}. However half of those reactions are
+unobservable, and their effect on the posterior is only felt through
+their effect on the effective propensities $\hat{\alpha}_5$ and $\hat{\alpha}_7$.
 
@@
 
-The one- and two-dimensional marginal distributions for the full posterior distribution are displayed in Figure~\ref{fig:GRN_posterior}.
-The posterior does not contain many interesting correlation structures, however several dimensions have leptokurtic tails which are difficult for the standard PAIS algorithm to sample from.
-The marginal densities also vary over very different scales, which might require us to use a variant of the LPAIS algorithm.
+The one- and two-dimensional marginal distributions for the full
+posterior distribution are displayed in
+Figure~\ref{fig:GRN_posterior}. The posterior does not contain many
+interesting correlation structures, however several dimensions have
+leptokurtic tails which are difficult for the standard PAIS algorithm
+to sample from. The marginal densities also vary over very different
+scales, which could lead to problems with the algorithms which do not
+benefit from the transport map.
 
@@
 
-%% proposal methods
-%For the posterior distribution in \eqref{eqn:chem_QEA_posterior}, we consider several proposal methods.
-First we implement both the PAIS and MH algorithms with a Gaussian proposal distribution.
-In the case of the PAIS algorithm, this is a Gaussian mixture proposal distribution.
-The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard RWMH algorithm.
-We also compare the PAIS and MH algorithms when using a transport map proposal distribution.
-This proposal method was discussed in detail in Chapter~\ref{sec:TransportPAIS}.
-%
-%
-%Figure~\ref{fig:chem_coupling} shows how we define a bijective map, $\hat{T}$, between parameter space $\mathcal{X}$ and a reference space $R$.
-In practice we cannot ensure that the approximate map $\tilde{T}'$ is uniquely invertible over the whole of $R$ and so $\tilde{T}'$ is not truly bijective.
-This leads to problems for our strictly positive state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in $R$ do not necessarily map back on to $\mathcal{X}$.
-This motivates the use of a third intermediate space.
-When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
-MH-RW (RWMH) and MH-TRW, as well as PAIS-RW and PAIS-TRW.
-%
-%We also consider how these algorithms perform when they are applied to the log of the reaction rates.
-We choose to use this log transformation since it converts our strictly positive parameter space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset \mathbb{R}^d$.
-This allows us to define $T$ between two subsets of $\mathbb{R}^d$, which means that even if $\tilde{T}$ is not uniquely invertible in some region of $R$, all possibilities are valid proposals.
-As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
-MH-logRW and MH-logTRW, with, PAIS-logRW and PAIS-logTRW.
-We apply the eight RW and logRW proposal distributions, with and without transport map acceleration, which were discussed in Section~\ref{sec:chem_multiscale} to this posterior.
-Again the intermediate log space is required so that our transport map is a function $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$, rather than a map from $\mathbb{R}_+^d$ to $\mathbb{R}^d$.
-
-%% choices made in transport map
-%For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=4$ as the basis functions, i.e.
-%\[
-%	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
-%\]
-%and
-%\[
-%	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
-%\]
-%Since $\mathcal{X}$ is four-dimensional, this yeilds a total of 125 map coefficients across the four index sets.
-This number can be reduced by using smaller index sets as discussed in~\cite{parno2014transport}.
+We apply the eight RW and logRW proposal distributions, with and
+without transport map acceleration, which were discussed in
+Section~\ref{sec:chem_multiscale} to this posterior. As with this
+example, the log conditioner is required so that our transport map is a function $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$, rather than a map from $\mathbb{R}_+^d$ to $\mathbb{R}^d$.
+
 The transport map setup has not changed from the previous example.
@@
 For this reason we filter out any samples from the optimisation problem where the weight is a numeric zero.
-This does not affect the validity of the map since a weight zero would not contribute to the expectation, and we do not require (and could not enforce) an exact map.
-
-In all eight algorithms we adapt the proposal variances during the burn-in phase of the algorithms to avoid behaviour like that seen in Example~\ref{sec:ex_W1} where we needed a huge ensemble size to approximate the posterior with our mixture of isotropic Gaussian kernels.
-
-%% choice of resampler?
-%As with the mixture model in Section~\ref{sec:app_mixture}, we will use the AMR resampler with an ensemble size of $M=500$.
-This increase in ensemble size from Chapter~\ref{sec:PAIS} compensates for the increase in parameter dimension.
-We again use the AMR resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size again compensates for the increase in dimension (from four to eight parameters.)
-
-%% analysis method
-%To measure the convergence of the sampling methods in this section, we will compare the convergence of the mean of each parameter.
-We approximate $\mathbb{E}(\mathbf{k})$ using 2.4 billion samples from the MH-RW algorithm.
-As in the previous chemical reaction example, we measure convergence of the algorithms only through the convergence of the mean, for computational ease.
+This does not affect the validity of the map since a weight zero would not contribute to the expectation.
+
+In all eight algorithms we adapt the proposal variances during the
+burn-in phase of the algorithms to avoid the need for a huge ensemble size to approximate the posterior with a mixture of isotropic Gaussian kernels.
+We use the AMR resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size compensates for the increase in dimension from four to eight.
+
+As in the previous chemical reaction example, we measure convergence of the algorithms only through the convergence of the mean.
 We use the sample Mahalanobis distance, with `true' covariance and `true' mean built from a sample of size 2.4 billion using the MH-RW algorithm.
@@
 
-The scaling parameter selection is done by optimising the acceptance rate for the MH algorithms, and optimising the effective sample size for the PAIS algorithms.
+The scaling parameters used are selected by optimising the acceptance rate for the MH algorithms, and optimising the effective sample size for the PAIS algorithms.
 The optimal parameters are given in Table~\ref{tab:grn_scaling_parameters}.
@@
 
-From the effective sample sizes shown in Table~\ref{tab:grn_scaling_parameters} we can see the improvement to the efficiency of the algorithm both by proposing on the log space, and by transforming the parameter space into something closer to Gaussian.
+From the effective sample sizes shown in
+Table~\ref{tab:grn_scaling_parameters} we can see the improvement to
+the efficiency of the algorithm both by transforming the parameter
+space into something closer to Gaussian, and additionally by utilising
+a log preconditioner.
 
@@
 \centering
-\subfigure[Sampling algorithms on $\mathcal{X}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
-\subfigure[Sampling algorithms on $\mathcal{Y}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
+\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
+\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
 \caption{Convergence of the GRN example described in Section~\ref{sec:grn}.}
@@
 
-The first thing to note is that the first PAIS measurement happens after 2500 samples, whereas the first MH measurement occurs after 50 samples.
-This is due to the much larger ensemble size required to sample in higher dimensions.
-We believe that an ensemble size of 2500 is over-cautious in this example and we could have used a smaller ensemble size.
-We also think that the required ensemble size to sample from these posteriors is reduced by the use of the transport map.
-This is due to the way the tail behaviour is improved by the transformation.
-
-The second obvious feature of these convergence plots is that the PAIS algorithms outperform the MH algorithms by a large margin - roughly a reduction of two orders of magnitude in the relative error over the same number of samples.
-A less positive result is that for this example the PAIS-TRW algorithm has found a map which has prevented further convergence of the mean.
-This is likely because one dimension of the transport map is sending proposals onto the negative real line and so all samples are being assigned a zero weight.
-This is behaviour which was anticipated, and it is always suggested that the transport map is constructed as a map $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$ rather than $T\colon\mathcal{X}\rightarrow\mathbb{R}^d$ where $\mathcal{X}\subset\mathbb{R}^d$.
-The large increase in the effective sample size observed between the PAIS-logRW and PAIS-logTRW algorithms is converted into an estimate which is twice as accurate after 1 million samples.
-
-A similar pattern is seen in the MH algorithms, where the MH-TRW performs the worst, and the MH-logTRW algorithm performs the best, although only marginally.
+The first thing to note is that the first PAIS measurement happens
+after 2500 samples, whereas the first MH measurement occurs after 50
+samples. This is due to the much larger ensemble size required to
+sample in higher dimensions. We believe that an ensemble size of 2500
+is over-cautious in this example and we could have used a smaller
+ensemble size. We also believe that the required ensemble size to sample
+from these posteriors is reduced by the use of the transport map.
+
+The second obvious feature of these convergence plots is that the PAIS
+algorithms outperform the MH algorithms by a large margin - roughly a
+reduction of two orders of magnitude in the relative error over the
+same number of samples. A less positive result is that for this
+example the PAIS-TRW algorithm has found a map which has prevented
+further convergence of the mean. This is likely because one dimension
+of the transport map is sending proposals onto the negative real line
+and so all samples are being assigned a zero weight. This is behaviour
+which was anticipated, and it is always suggested that appropriate
+preconditioners are used where the posterior is not absolutely
+continuous with respect to Lebesgue measure, as is the case when the
+posterior is only non-zero on the positive quadrant of $\mathbb{R}^d$
+as in these examples. The large increase in the effective sample size
+observed between the PAIS-logRW and PAIS-logTRW algorithms is
+converted into an estimate which is twice as accurate after 1 million
+samples.
+
+A similar pattern is seen in the MH algorithms, where the MH-TRW
+performs the worst, and the MH-logTRW algorithm performs the best,
+although only marginally.
 
@@
 
-Algorithm~\ref{alg:TransportPAIS2} allows us to decorrelate the dimensions of our random parameter on reference space, where we then can resample and map the resulting ensemble back onto target space.
-Since, on reference space, the dimensions are uncorrelated, we are able to resample in each dimension separately.
-Resampling in a single dimension allows for optimisations in resampling code, and also means that the resampler is not affected by the curse of dimensionality.
+Algorithm~\ref{alg:TransportPAIS2} allows us to decorrelate the
+dimensions of our random parameter on reference space, where we then
+can resample and map the resulting ensemble back onto target
+space. Since, on reference space, the dimensions are uncorrelated, we
+are able to resample in each dimension separately. Resampling in a
+single dimension allows for optimisations in resampling code, and also
+means that the resampler is not affected by the curse of
+dimensionality.
 

### 2018-01-24 TPAIS_2018.tex
First draft nearly done
Apart from intro and discussion

@@
 
-\title{Transport map accelerated-PAIS, and application to inverse problems arising from
+\author{Simon L. Cotter\thanks{School of
+    Mathematics, University of Manchester, Manchester, UK. e:
+    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
+    grant award EP/L023393/1. SLC would like to thank the Isaac Newton
+    Institute for Mathematical Sciences for support and hospitality
+    during the programme ``Uncertainty quantification for complex systems: theory and methodologies'' when work on this paper was undertaken. This work was supported by:
+EPSRC grant number EP/K032208/1.} \and Ioannis G. Kevrekidis\thanks{} \and Paul
+  Russell\thanks{School of
+    Mathematics, University of Manchester, Manchester, UK.}}
+
+\title{Transport map accelerated adaptive importance sampling, and application to inverse problems arising from
   multiscale stochastic reaction networks}
-\author{Simon Cotter (EPSRC and INI), Yannis Kevrekidis (?), Paul
-  Russell (EPSRC)}
 \begin{document}
@@
 \section{Parallel Adaptive Importance Sampling}\label{sec:PAIS}
-PAIS****CITE**** is a variant of PMC****CITE****, which is a family of methods which are based
+Parallel adaptive importance sampling (PAIS)\cite{cotter2015parallel} is a variant of PMC\cite{cappe2012population}, which is a family of methods which are based
 on importance sampling. In importance sampling, we attempt to
@@
 found by optimising for the effective sample size of the importance
-sample that is produced (see ***CITE*** for more details).
+sample that is produced (see \cite{cotter2015parallel, russ2017parallel} for more details).
 
@@
 best represents the statistics of the importance
-sample***CITE****. For larger ensemble sizes $M$ this can become
+sample\cite{reich2013nonparametric}. For larger ensemble sizes $M$ this can become
 expensive, in which case a greedy approximation of this algorithm, the
 Approximate Multinomial Resampler (AMR) can be
-implemented***CITE****. The output of the resampler is then denoted
+implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
 $\theta^{(k+1)}$ and the algorithm is ready for the next
@@
 $\pi$ without a very large large ensemble size $M$, which can become
-inhibitively expensive. Some methods have been proposed ****CITE****
+inhibitively expensive. Some methods have been
+proposed\cite{douc2007minimum}
 which use samples local to each particle to inform local covariance
@@
 
-\begin{dfn}[(Exact) Transport Map $T$]
+\begin{dfn}[Transport Map $T$]
 	A transport map $T$ is a function $T\colon
@@
 \subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
-\caption{Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
+\caption{The effect of the approximate transport map $\tilde{T}$ on a
+  sample from the Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
 \label{fig:R2_transport}
@@
 We first find the optimal scaling parameters for the individual
-algorithms. This is done, as in ****CITE US**** by optimising for the effective sample size in
+algorithms. This is done, as in \cite{cotter2015parallel}, by optimising for the effective sample size in
 the PAIS algorithm, and by tuning the relative $L^2$ error in the MH
@@
 \subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
-\caption{Convergence of Algorithms Transport M-H,
-  \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2} for density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
+\caption{Convergence of algorithms; Transport M-H,
+  algorithm \ref{alg:TransportPAIS1}, algorithm \ref{alg:TransportPAIS2} for
+  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
 \label{fig:R2_l2_convergence}
@@
 $N_r$ different unit rate Poisson
-processes****CITE ANDERSON ET AL**** $Y_j$, given by:
+processes\cite{anderson2011continuous} $Y_j$, given by:
 \begin{equation}\label{eq:RTC}
@@
 The master equation for this model is only solvable in certain
-situations, for example monomolecular networks****CITE****, or for
+situations, for example monomolecular networks\cite{jahnke2007solving}, or for
 steady-state solutions certain deficiency
-zero networks****CITE*****. Trajectories for this system can be
-sampled exactly, for instance using the Gillespie SSA***CITE***, or
-its variants\cite{gillespie2007stochastic}****CITE****. However, if the system is stiff, i.e. there
+zero networks\cite{anderson2010product,anderson2016product}. Trajectories for this system can be
+sampled exactly, for instance using the Gillespie SSA\cite{gillespie1977exact}, or
+its variants\cite{gillespie2007stochastic,gibson1998efficient,cao2004efficient,anderson2007modified}. However, if the system is stiff, i.e. there
 are some reactions which are firing many times on a timescale for
@@
 timescale gap is less pronounced, it can lead to significant
-errors***cite cotter erban errors paper***. Another approach is the
-constrained multiscale method (CMA)***CITE***, based in part on the
-equation-free approach to multiscale computations***CITE
-YANNIS****. This approach also assumes quasi-equillibrium in the
+errors\cite{cotter2016error}. Another approach is the
+constrained multiscale method (CMA)\cite{cotter2011constrained,cotter2016constrained}, based in part on the
+equation-free approach to multiscale computations\cite{kevrekidis2003equation,erban2006gene}. This approach also assumes quasi-equillibrium in the
 system, but takes into account the effect that the slow reactions can
@@
 system. For a more detailed description of this method, please refer
-to the literature***CITATIONS***.
+to the literature\cite{cotter2016constrained}.
 
@@
 population count of
-species $S_i$ by $X_i \in \mathbb{N}_0$. We assume that we are in a
-parameter regime such that the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
-more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
+species $S_i$ by $X_i \in \mathbb{N}_0$. 
+We consider this system with the following parameters:
+\begin{equation}\label{eq:params1}
+k_1 = 100, \qquad k_2 = k_3 = 10, \qquad k_4 = 1,
+\end{equation}
+
+ In this 
+parameter regime the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
+more frequently than the other reactions. Notice that both
 chemical species are involved in fast reactions. However, the quantity
@@
 
-We create trajectories of this system with the following parameters:****
-\begin{equation}
-k_1 = ? etc
-\end{equation}
-over the time period ****$t \in [0,?]$.
-
-We assume that we are only able to observe
+We observe this system over the time period $t \in [0,500]$ with
+initial condition ******, but we assume that we are only able to observe
 the slow variable $S(t)$. 
@@
 \begin{equation}\label{eqn:QSSA_system}
-	\emptyset \xrightarrow{\alpha_1(S)} S \xrightarrow{\bar{\alpha}_4(S)} \emptyset.
-\end{equation}
+	\emptyset \xrightarrow{\alpha_1(S)} S \xrightarrow{\bar{\alpha}_4(S)} \emptyset,
+\end{equation}
+where here the propensities are shown as opposed to the rates.
+
 The effective propensity $\bar{\alpha}_4(S)$ can be  approximated through
 application of a multiscale approximation, as detailed in Section
-\ref{sec:multi}, and in more detail in ****CITE****. These effective
+\ref{sec:multi}, and in more detail in \cite{cotter2016constrained}. These effective
 propensities can often be computed without the need for expensive
 simulations, in particular when the fast subsystem is deficiency zero,
-which is often the case****CITE DIF ZERO PAPERS****. For this very
+which is often the case\cite{anderson2010product,anderson2016product}. For this very
 simple example, the dependance of these effective
@@
 \includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
-\caption{Posterior for the constrained multiscale problem outlined in Section~\ref{sec:chem_multiscale}.}
+\caption{CMA approximation \eqref{eq:Apos} of the posterior arising from observations of
+  the slow variable $S = X_1 + X_2$ for system
+  \eqref{eqn:full_system}  with true rates given by \eqref{eq:params1}.}
 \label{fig:chem_CMA_posterior}
@@
 	\end{tikzpicture}
-	\caption{Top: Couplings between parameter space and reference
-          when using the transport map. Bottom: Couplings between parameter space and reference space
-          when using the transport map proposal with a $\log$ preconditioner.}
+	\caption{Couplings between parameter space and reference
+          when using the transport map with (bottom) and without (top)
+        a log preconditioner.}
 	\label{fig:chem_coupling}
@@
 % choice of resampler?
-We will use the AMR***CITE*** resampler with an ensemble size of
+We will use the AMR\cite{russ2017parallel} resampler with an ensemble size of
 $M=500$. This increase in ensemble size in comparison with previous
@@
 \subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
-\caption{Convergence of the constrained multiscale example described in Section~\ref{sec:chem_multiscale}.}
+\caption{Convergence of eight different MCMC algorithms for the CMA
+  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}.}
 \label{fig:chem_multiscale_L2}
@@
 \centering
-\subfigure[Reference space mapped to by $\hat{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
-\subfigure[Reference space mapped to by $T\circ\log$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
-\caption{Push forward of the target measure on reference space for the problem described in Section~\ref{sec:multiscale_posterior} found using the TRW and logTRW proposal distributions.
-Components are linked by the relation $r_i = T_i'(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
+\subfigure[Push forward of the posterior sample through$\hat{T}$ with no log preconditioner.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
+\subfigure[Push forward of the posterior sample through$\hat{T}$ with log preconditioner.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
+\caption{Push forward of the target measure on reference space for the the CMA
+  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}, found using the TRW and logTRW proposal distributions.
+Components are linked by the relation $r_i = T_i(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
 \label{fig:chem_reference_spaces}
@@
 \end{tabular}
-\caption{Marginal distributions for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:multiscale_posterior}.}
+\caption{Marginal distributions for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:chem_multiscale}.}
 \label{tab:chem_Bayes_marginals}
@@
 \subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_approx_marginal"}}
-\caption{Comparison of the approximate marginal densities for the `observable parameter' $\hat{k}_4$ under models $\mathcal{M}_1$ and $\mathcal{M}_2$.
-Marginals for these two parameters are approximated using samples which have been produced by targeting the three densities $\pi_i = \mathbb{P}(\theta_i|D,\mathcal{M}_i)$ for $i = 0, 1, 2$.}
+\caption{Comparison of the approximate marginal densities for the
+  quantities $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$
+  across the three posterior densities $\pi_i = \mathbb{P}(\theta_i|D,\mathcal{M}_i)$ for $i = 0, 1, 2$.}
 \label{fig:chem_model_comp}
@@
 effective propensities, using the CMA methodology, for brevity, but
-interested readers can see more details in ****CITE THESIS****.
+interested readers can see more details in \cite{russ2017parallel}.
 
@@
 \makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/Applications/GRN_posterior"}}%
-\caption{Marginal densities of the posterior distribution for the GRN example in Section~\ref{sec:grn}.}
+\caption{Marginal densities of the CMA approximation of the posterior
+  distribution for the GRN example in Section~\ref{sec:grn}.}
 \label{fig:GRN_posterior}
@@
 \subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
-\caption{Convergence of the GRN example described in Section~\ref{sec:grn}.}
+\caption{Convergence of eight different MCMC algorithms for the CMA
+  approximation of the  posterior arising from the example in Section \ref{sec:grn}.}
 \label{fig:grn_L2}
@@
 
+One major issue with PMC-type algorithms is the curse of
+dimensionality. This curse is two-fold, since as the target
+distribution's dimension is increased, a larger ensemble size is often
+required in order to capture the addition complexity of the density
+that the higher dimension enables. Furthermore, the cost of the
+resampling step within the algorithm grows with the ensemble size. For
+example, the ETPF algorithm is $\mathcal{O}(n^2)$. However, the
+simiplification that the transport map makes has some potential to go
+some way to tackling this problem.
+
 \subsection{Sampling in (moderately) higher
   dimensions}\label{sec:TPAIS_higher_dim}
-One major problem with importance sampling schemes is the curse of
-dimensionality, which means that methods such as PAIS, and other PMC
-methods, can only be used for relatively low dimensional
-problems. Here, we will briefly discuss how transport maps could aid
+Here, we will briefly discuss how transport maps could aid
 with making moderately higher dimensional problems accessible to this
@@
 can resample and map the resulting ensemble back onto target
-space. Since, on reference space, the dimensions are uncorrelated, we
+space. Since, on reference space, the dimensions are approximately uncorrelated, we
 are able to resample in each dimension separately. Resampling in a
@@
 
-If we can approximate the posterior well with our mixture and with the
-transport map, we should not be affected by the increase in dimension
-to the extent we have been with the standard PAIS-RW algorithm. In one
+In one
 dimension the ETPF algorithm can be implemented very efficiently. As
@@
 ordered. We can exploit this knowledge to produce
-Algorithm~\ref{alg:ETPF_1d}. Which is much faster than using the
+Algorithm~\ref{alg:ETPF_1d}, which is much faster than using the
 simplex algorithm to minimise the associated cost function, and faster
-than the AMR algorithm****CITE OUR OTHER PAPER****.
+than the AMR algorithm\cite{cotter2015parallel}. Moreover, the task of
+running this resampler can then be parallelised over $d$
+processors. This could be an area for further study.
 

### 2018-06-12 TPAIS_2018.tex
Updating

@@
     during the programme ``Uncertainty quantification for complex systems: theory and methodologies'' when work on this paper was undertaken. This work was supported by:
-EPSRC grant number EP/K032208/1.} \and Ioannis G. Kevrekidis\thanks{} \and Paul
+EPSRC grant number EP/K032208/1.} \and Ioannis
+G. Kevrekidis\thanks{Chemical and Biomolecular Engineering \& Applied
+  Mathematics and Statistics, Johns Hopkins University, Baltimore, MD
+  21218. The work of IGK was partially supported by the US National Science Foundation and by DARPA.} \and Paul
   Russell\thanks{School of
@@
 \section{Introduction}
+The importance of Markov chain Monte Carlo (MCMC) methods is becoming
+increasingly apparent, in a world replete with datasets which need to be
+combined with complex dynamical systems in order for us to be able
+make progress in a range of scientific fields. Different MCMC methods
+have been designed with different challenges in mind; high dimensional
+targets****CITE FASTER PAPER*****, large data sets ****zigzag****,
+other methods which exploit piecewise deterministic Markov processes
+***bouncey particle***,
+methods which are able to make very large moves informed by the
+data****stochastic mala, HMC****, methods which are able to exploit
+the geometry of the target***riemannian****. Population Monte Carlo
+(pMC) methods uses efficient importance sampling proposals informed by an
+ensemble of particles which have already learned about the structure
+of the target***cite standard papers including AIS etc****. These methods are particularly useful in the context of
+low dimensional target distributions which have complex structure, for
+which traditional Metropolis-Hastings methods tend to struggle, for
+example when the target is multimodal or if the density is
+concentrated on a lower-dimensional manifold. Moreover, they are an
+excellent candidate for parallelisation, since the task of computing
+the likelihood for each particle in the ensemble can be distributed
+across a set of processors. This is of utmost importance in an age
+where increases in computational power are most likely to come from
+large-scale parallelisation as opposed to the large speed-ups in
+individual processors that we saw throughout the 20th Century.
+
+In ****cite PAIS***, we explored ideas from adaptive MCMC within a
+pMC-type structure, in order to construct an adaptive importance
+sampler which could automatically adapt to optimise
+efficiency. Alongside this, we explored the use of state-of-the-art
+resamplers****cite sebastian***, and a greedy approximation (the
+approximate multinomial resampler) in order to improve the quality of
+the importance sampler proposals.
+
+Other approaches to challenging posterior structure have been
+proposed, including the work of Marzouk and Parnod***cite***, who
+demonstrated how invertible transport maps can be constructed which
+map the target distribution close to an easily-explored reference
+density, such as a d-dimensional Gaussian. Since the map is
+invertible, Gaussian proposal densities on the reference space can be
+mapped to highly informed moves on the parameter space.
+
+In this paper, we particularly concern ourselves with the problem of
+inverse problems in stochastic reaction networks. This type of model
+is used in situations where ODE or diffusion****Chem langevin****
+approximations of chemical reaction kinetics are not viable, usually
+because of the low concentration of one or more chemical
+species****Gillespie****. Biochemical networks typically fall into
+this category, because of the small volume of the reactors involved (a
+cell).
+
+One challenge with networks arising from these applications is that
+they often have some reactions which are occurring many times on a
+timescale in which other reactions in the system are unlikely to
+occur. This multiscale behaviour throws up two main challenges, both
+in observability in an experimental setting, but also in the context
+of modelling, where multiscale systems can make the simulation of the
+continuous time Markov chain computationally intractable.
+
+A range of numerical methods have been suggested to tackle this latter
+problem, including tau-leaping***cite****, diffusion approximations
+***radek and Yannis, cotter and Yannis****, quasi-steady state
+approximations ****slow scale and nested****, and other averaging
+arguments ****cotter comp. phys****. Some of these approaches are more
+computationally intensive than others, but this can be mitigated with the
+use of analytical results about the invariant densities of certain
+types of fast subsystems***product-form papers****.
+
+We would not expect the fast processes in such a system to be
+experimentally observable. In such a situation where only the slow
+variables in a system are identifiable, in an ideal world, we would be
+be able to integrate over the space of all possible trajectories of
+the fast processes. Since this is an insurmountable computational
+task, we might instead wish to approximate the likelihoods arising
+from such a set of observations by the use of an appropriate
+multiscale methodology.
+
+The partial observability of these systems leads to only
+partial-observability of the unknown parameters in the system, for
+example the reaction rates of the fast reactions. The resulting
+multiscale approximations, and the resulting approximations of the
+likelihoods arising from the inverse problem for such a network, then
+often exhibit 
+insensitivity to changes in certain directions of
+parameter space. This causes problems when aiming to sample from such
+a target, since the density is concentrated on a manifold whose
+tangential vectors point in directions of insensitivity. This
+\emph{sloppiness}***cite*** of the likelihood is a challenging problem
+that appears in many applications in science and engineering***active
+subspaces literature here****.
+
+In this work, we aim to improve the PAIS machinery, which is limited,
+as the same proposal kernel is used for each particle in the
+ensemble. This homogeniety of proposal kernels means that complex
+structures in the density can only be well represented with a large
+increase in the ensemble size, increasing computational cost. If
+computational resources do not allow for such an increase, this can
+lead to stability issues in the algorithm, due to the proposal
+distribution in the importance sampler not representing well the
+target density, leading to high variance in the weights, causing slow
+convergence.
+
+As shown in ***cite marzouk****, transport maps can be used to create
+proposal densities which are highly informed by the structure of the
+target density. In this paper, we demonstrate how such maps can also
+be constructed using importance weights arising from a pMC-type
+algorithm in order to accelerate and stabilise convergence. We will
+then apply the methodology to some complex posterior distributions
+arising from Bayesian inversion of multiscale (bio-)chemical networks.
+
 %Topics to cover: parallel MCMC, pMC, PAIS
@@
 
-In Section \ref{sec:PAIS}, we will briefly reintroduce Parallel
-Adaptive Importance Sampling (PAIS), a variant of Population Monte
-Carlo (PMC) which incorporates state of the art resamplers. In Section \ref{sec:map} we show how an appropriate transport map can
+In Section \ref{sec:PAIS}, we will briefly reintroduce PAIS. In Section \ref{sec:map} we show how an appropriate transport map can
 be constructed from importance samples which maps the posterior close
@@
 how such a map can be incorporated into a sophisticated parallel MCMC
-infrastructure in order to accelerate mixing. In Section
+infrastructure in order to accelerate mixing, and improve stability. In Section
 \ref{sec:conv} we seek to show the advantages of this approach through

### 2018-06-12 TPAIS_2018.tex
Intro and corrections/amendments nearly complete
Still need rates and time frame of GRN example.

@@
 \begin{abstract}
-In many applications, inverse problems arise where where there are
-complex correlations between the different parameters which we wish to
-infer from data. The correlations often manifest themselves as lower
+In many applications, Bayesian inverse problems can give rise to
+probability distributions which contain complexities due to the
+Hessian varying greatly across parameter space. This complexity often manifests itself as lower
 dimensional manifolds on which the likelihood function is
@@
 being used to describe the data. In such a situation, standard
-sampling methods for characterising the posterior distribution which
-do not incorporate information about this structure will be highly
+sampling methods for characterising the posterior distribution, which
+do not incorporate information about this structure, will be highly
 inefficient. Moreover, most methods are inherently serial in nature,
@@
 have been designed with different challenges in mind; high dimensional
-targets****CITE FASTER PAPER*****, large data sets ****zigzag****,
+targets\cite{cotter2013mcmc}, large data sets\cite{bierkens2016zig},
 other methods which exploit piecewise deterministic Markov processes
-***bouncey particle***,
+\cite{bouchard2018bouncy},
 methods which are able to make very large moves informed by the
-data****stochastic mala, HMC****, methods which are able to exploit
-the geometry of the target***riemannian****. Population Monte Carlo
+data\cite{teh2016consistency,duane1987hybrid}, methods which are able to exploit
+the geometry of the target\cite{girolami2011riemann}. Population Monte Carlo
 (pMC) methods uses efficient importance sampling proposals informed by an
 ensemble of particles which have already learned about the structure
-of the target***cite standard papers including AIS etc****. These methods are particularly useful in the context of
+of the target\cite{cappe2012population,cappe2008adaptive,douc2007convergence,douc2007minimum, martino2017layered, cornuet2012adaptive, martino2015adaptive}.
+These methods are particularly useful in the context of
 low dimensional target distributions which have complex structure, for
@@
 
-In ****cite PAIS***, we explored ideas from adaptive MCMC within a
+In \cite{cotter2015parallel}, we explored ideas from adaptive MCMC within a
 pMC-type structure, in order to construct an adaptive importance
@@
 efficiency. Alongside this, we explored the use of state-of-the-art
-resamplers****cite sebastian***, and a greedy approximation (the
+resamplers\cite{reich2013nonparametric}, and a greedy approximation (the
 approximate multinomial resampler) in order to improve the quality of
@@
 Other approaches to challenging posterior structure have been
-proposed, including the work of Marzouk and Parnod***cite***, who
+proposed, including the work of Marzouk and Parnod\cite{parno2014transport}, who
 demonstrated how invertible transport maps can be constructed which
@@
 invertible, Gaussian proposal densities on the reference space can be
-mapped to highly informed moves on the parameter space.
+mapped to highly informed moves on the parameter space. A related idea
+is the use of the anisotropic diffusion maps\cite{singer2008non}, in
+which a map which describes the slow manifold is constructed using
+principle component analysis. These ideas have also been applied to
+stochastic reaction
+networks\cite{singer2009detecting,dsilva2016data}. In
+\cite{chiavazzo2017intrinsic}, these maps were used to propogate
+data-mining into unexplored regions which were predicted to be close
+to the lower-dimensional manifold. The idea is similar to that of the
+Lamperti or Girsanov transformations for SDEs\cite{flandrin2003stationarity,gardiner2009stochastic},  the aim being to map complex
+processes onto a standard one which is well understood, through a
+change of measure.
+
+Bayesian inverse problems are not the only area in which probability
+densities arise which are concentrated on, or close, lower dimensional
+manifolds. This is also a common feature of molecular dynamics
+problems, where
 
@@
 inverse problems in stochastic reaction networks. This type of model
-is used in situations where ODE or diffusion****Chem langevin****
+is used in situations where ODE or diffusion\cite{gillespie2000chemical}
 approximations of chemical reaction kinetics are not viable, usually
 because of the low concentration of one or more chemical
-species****Gillespie****. Biochemical networks typically fall into
+species\cite{gillespie1977exact}. Biochemical networks typically fall into
 this category, because of the small volume of the reactors involved (a
@@
 
-A range of numerical methods have been suggested to tackle this latter
-problem, including tau-leaping***cite****, diffusion approximations
-***radek and Yannis, cotter and Yannis****, quasi-steady state
-approximations ****slow scale and nested****, and other averaging
-arguments ****cotter comp. phys****. Some of these approaches are more
+A range of numerical methods have been suggested to tackle this second
+problem, including tau-leaping\cite{rathinam2003stiffness,cao2006efficient}, diffusion approximations\cite{erban2006gene,cotter2011constrained,cotter2016error}, quasi-steady state
+approximations\cite{ea2005nested,cao2005slow}, and other averaging
+arguments\cite{cotter2016constrained}. Some of these approaches are more
 computationally intensive than others, but this can be mitigated with the
 use of analytical results about the invariant densities of certain
-types of fast subsystems***product-form papers****.
+types of fast subsystems\cite{jahnke2007solving,anderson2010product,anderson2016product}.
 
@@
 tangential vectors point in directions of insensitivity. This
-\emph{sloppiness}***cite*** of the likelihood is a challenging problem
-that appears in many applications in science and engineering***active
-subspaces literature here****.
+\emph{sloppiness}\cite{gutenkunst2007universally,apgar2010sloppy} of the likelihood is a challenging problem
+that appears in many applications in science and engineering\cite{constantine2014active}.
 
 In this work, we aim to improve the PAIS machinery, which is limited,
-as the same proposal kernel is used for each particle in the
+since in the current presentation the same proposal kernel is used for each particle in the
 ensemble. This homogeniety of proposal kernels means that complex
@@
 
-As shown in ***cite marzouk****, transport maps can be used to create
+As shown in\cite{parno2014transport}, transport maps can be used to create
 proposal densities which are highly informed by the structure of the
@@
 \cdot, \beta)$ then 
-\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}.\]
+\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}).\]
 Often the variance of the MCMC proposal kernels can be tuned using
@@
 target density, i.e.
-\[ \hat{\theta}_i^{(k)} \sim q(\cdot ; \theta_i^{(k)}\]
+\[ \hat{\theta}_i^{(k)} \sim q(\cdot ; \theta_i^{(k)})\]
 for each $i = 1,2,\ldots,M$. We now have a weighted sample, and it
@@
 	}
-	Output $\{(w^{(n)}, \hat{\theta}^{(n)}\}_{n=1}^N$.
+	Output $\{(w^{(n)}, \hat{\theta}^{(n)})\}_{n=1}^N$.
 \caption{The PAIS Algorithm.\label{alg:PAIS}}
@@
 One problem with the PAIS and other PMC methods can become apparent if
-the target density has very strong correlations in its structure, in
-particular if that correlation is not global but only local. In this
+the target density has highly curved manifolds on which the density is
+concentrated. In this
 case, unless the proposal densities $q$ are informed by this local
@@
 sample from the prior distribution and use this sample to approximate a map onto the target space.
-Once this map was known to the desired accuracy a larger sample from the prior could be used to
+Once this map was known to the desired accuracy, a larger sample from the prior could be used to
 investigate the posterior distribution. This
@@
 \begin{dfn}[Transport Map $T$]
-	A transport map $T$ is a function $T\colon
+	A transport map $T$ is a smooth function $T\colon
         \mathcal{X}\rightarrow\mathbb{R}^d$ such that the {\it
@@
 This divergence results in some nice properties which we will explore in the following derivation.
-The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to it's relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
+The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to its relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
 \[
@@
 		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
-	&\text{subject to} \quad G_i\gamma_i > 0, \notag
+	&\text{subject to} \quad G_i\gamma_i > 0. \notag
 \end{align}
@@
 both the Transport MH algorithm, and the standard PAIS algorithm. The second option requires some
-restructuring but results in improved performance from the resampler.
+restructuring, but can result in improved performance from the resampler.
 
@@
 density does not look a challenging one to sample from, with even
-relatively simple MCMC methods such as MH-RW. The pullback from
+relatively simple MCMC methods such as Metropolis-Hasting random walk (MH-RW). The pullback from
 reference space, in Figure~\ref{fig:R2_transport}, is an exact match
@@
 \eqref{eqn:R2_repeat} we are dealing with a much more complicated
-correlation structure, as well as a very slowly decaying tail in
+curved structure, as well as a very slowly decaying tail in
 $\theta_2$.
@@
 for stochastic reaction networks. Inverse problems arising in this
-area often lead to highly correlated and complex posterior
+area often lead to complex posterior
 distributions, which traditional MCMC methods can struggle to sample
@@
 
-In all three cases, we find the posterior distribution for $\mathbf{k} \in \mathcal{X}$. These four parameters are assigned Gamma prior distributions,
+In all three cases, we find the posterior distribution for $\mathbf{k}
+\in \mathcal{X} = \mathbb{R}_+^4$. These four parameters are assigned Gamma prior distributions,
 \[
@@
 \centering
-\subfigure[Push forward of the posterior sample through$\hat{T}$ with no log preconditioner.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
-\subfigure[Push forward of the posterior sample through$\hat{T}$ with log preconditioner.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
+\subfigure[Push forward of the posterior sample through $\hat{T}$ with
+no log
+preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
+\\
+\subfigure[Push forward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
 \caption{Push forward of the target measure on reference space for the the CMA
@@
 
-A conventional way to compare two models under a Bayesian framework is to calculate the Bayes factors~\cite{chen2012monte}.
-The Bayes factor, $B_{1,2}$, between two models, $\mathcal{M}_1$ and $\mathcal{M}_2$, can be interpreted as a ratio of the normalisation constants of the posterior distributions given each model,
+A conventional way to compare two models under a Bayesian framework is
+to calculate the Bayes factors~\cite{chen2012monte}. The Bayes factor,
+$B_{1,2}$, between two models, $\mathcal{M}_1$ and $\mathcal{M}_2$,
+given data $D$, can be interpreted as a ratio of the normalisation constants of the posterior distributions given each model,
 \[
@@
 
-The scaling parameters used are selected by optimising the acceptance rate for the MH algorithms, and optimising the effective sample size for the PAIS algorithms.
-The optimal parameters are given in Table~\ref{tab:grn_scaling_parameters}.
+The scaling parameters used are selected by optimising the acceptance
+rate for the MH algorithms, and optimising the effective sample size
+for the PAIS algorithms. The optimal parameters are given in
+Table~\ref{tab:grn_scaling_parameters}. Here, for Metropolis-Hastings
+variants, $\beta_{\%}^*$ is the variance of the proposal, optimised
+for an acceptance rate of $50\%$. For the PAIS variants,
+$\beta_{\text{ESS}}^*$ is the variance of each of the proposal kernels
+around each particle, optimised numerically for the maximal effective
+sample size (ESS). ESS/$M$ denotes the ratio of ESS to the number of
+particles, with values closer to one denoting a more efficient algorithm.
 
@@
 \section{Discussion}\label{sec:conc}
+In this paper, we have investigated the use of transport maps to
+accelerate and stabilise ensemble importance sampling schemes. In
+particular we looked at the application of these methods to problems
+arising in inverse problems for multiscale stochastic chemical
+networks, where the posterior distributions which arise can be highly
+concentrated along curved manifolds in parameter space.
+Below we highlight some areas for future work.
+
+\subsection{Sampling in higher
+  dimensions}\label{sec:TPAIS_higher_dim}
 
@@
 example, the ETPF algorithm is $\mathcal{O}(n^2)$. However, the
-simiplification that the transport map makes has some potential to go
+simplification that the transport map makes has some potential to go
 some way to tackling this problem.
 
-\subsection{Sampling in (moderately) higher
-  dimensions}\label{sec:TPAIS_higher_dim}
 Here, we will briefly discuss how transport maps could aid
 with making moderately higher dimensional problems accessible to this
-family of methods.
+family of methods. It is important to emphasise here that these
+methods are unlikely to be suitable in anything other than low
+dimensional problems, but the idea that we present here, which is the
+subject of future investigation, could make them applicable to more
+problems which have more than a handful of parameters.
 
@@
 
+\subsection{The Chicken and Egg Paradox}
+Another issue with the use of transport maps of this ilk, is that in
+order to construct such a map, we require a representative sample from
+the posterior distribution. Since these methods are designed to sample
+from distributions which are inherently difficult to sample from, be
+it due to multimodality, as looked at in \cite{cotter2015parallel}, or
+as in this paper, due to the density being highly concentrated close
+to a curved manifold in parameter space, it could be argued that accessing this initial set of
+samples remains a hard problem. This remains an open problem, which we
+are aiming to address in future work.
+
 \bibliographystyle{siam}

### 2018-06-12 TPAIS_2018.tex
Standardised bibliography

@@
 problem, including tau-leaping\cite{rathinam2003stiffness,cao2006efficient}, diffusion approximations\cite{erban2006gene,cotter2011constrained,cotter2016error}, quasi-steady state
-approximations\cite{ea2005nested,cao2005slow}, and other averaging
+approximations\cite{weinan2007nested,cao2005slow}, and other averaging
 arguments\cite{cotter2016constrained}. Some of these approaches are more
@@
 sampled exactly, for instance using the Gillespie SSA\cite{gillespie1977exact}, or
-its variants\cite{gillespie2007stochastic,gibson1998efficient,cao2004efficient,anderson2007modified}. However, if the system is stiff, i.e. there
+its variants\cite{gillespie2007stochastic,cao2004efficient,anderson2007modified}. However, if the system is stiff, i.e. there
 are some reactions which are firing many times on a timescale for

### 2018-06-12 TPAIS_2018.tex
Spellcheck

@@
 inefficient. Moreover, most methods are inherently serial in nature,
-and as such are not expoiting the parallelised  nature of modern
+and as such are not exploiting the parallelised  nature of modern
 computer infrastructure. In this paper, we seek to develop a method to
@@
 networks\cite{singer2009detecting,dsilva2016data}. In
-\cite{chiavazzo2017intrinsic}, these maps were used to propogate
+\cite{chiavazzo2017intrinsic}, these maps were used to propagate
 data-mining into unexplored regions which were predicted to be close
@@
 since in the current presentation the same proposal kernel is used for each particle in the
-ensemble. This homogeniety of proposal kernels means that complex
+ensemble. This homogeneity of proposal kernels means that complex
 structures in the density can only be well represented with a large
@@
     the time interval  } \quad s \in [t, t+ dt] ).\] If a system
-satsifies what is called \emph{mass action kinetics}, then the form of
+satisfies what is called \emph{mass action kinetics}, then the form of
 the function $\alpha_j$ is determined, up to a rate constant, by the
@@
 products produced. This is modelled by a change in the state vector
-$X(t) = X(t) + \nu_j$ where $\nu_j$ is that stochiometric vector for
+$X(t) = X(t) + \nu_j$ where $\nu_j$ is that stoichiometric vector for
 reaction $R_j$.
@@
 QSSA. This approximation makes the assumption that fast reactions
-enter quasi-equillibrium on a timescale which is negligible with
+enter quasi-equilibrium on a timescale which is negligible with
 respect to the timescale on which the slow dynamics in the system are
-evolving. This amounts to taking the assymptotic limit that the
+evolving. This amounts to taking the asymptotic limit that the
 propensities of the ``slow'' reaction channels are equal to zero. This
@@
 constrained multiscale method (CMA)\cite{cotter2011constrained,cotter2016constrained}, based in part on the
-equation-free approach to multiscale computations\cite{kevrekidis2003equation,erban2006gene}. This approach also assumes quasi-equillibrium in the
+equation-free approach to multiscale computations\cite{kevrekidis2003equation,erban2006gene}. This approach also assumes quasi-equilibrium in the
 system, but takes into account the effect that the slow reactions can

### 2018-06-15 TPAIS_2018.tex
Added another section to discussion

@@
 
-\subsection{The Chicken and Egg Paradox}
+\subsection{Invertibility of the transport map}
+In this paper, we have considered an approach very similar to that
+deseminated in \cite{parno2014transport}, where we aim to find an
+invertible map which approximates the true transport map between the
+reference measure and the target measure, however this true map is
+invariably itself not invertible. However, by considering embedding
+the map in higher dimensions\cite{whitney_self-intersections_1994,nash1956imbedding,takens_detecting_1981}, these issues of non-invertibility may be
+avoided, and higher quality maps may be found which better aid
+exploration and sampling of the whole state space.
+
+\subsection{The chicken and egg paradox}
 Another issue with the use of transport maps of this ilk, is that in

### 2018-06-15 TPAIS_2018.tex
Draft complete

@@
 We observe this system over the time period $t \in [0,500]$ with
-initial condition ******, but we assume that we are only able to observe
-the slow variable $S(t)$. 
+initial condition $S_1(0) = S_2(0) = 0$, but we assume that we are only able to observe
+the slow variable $S(t) = S_1(t) + S_2(t)$. 
 The effective dynamics of $S$ can be approximated by a system of only
@@
 
-We create trajectories of this system with the following parameters:****
-\begin{equation}
-k_1 = ? etc
-\end{equation}
-over the time period ****$t \in [0,?]$.
+We create trajectories of this system with the following parameters:
+\begin{eqnarray*}
+k_1 = 0.04, \qquad
+k_2 = 5000, \qquad
+k_3 = 100, \qquad
+k_4 = 1, \\ 
+k_5 = 0.5, \qquad
+k_6 = 2, \qquad
+k_7 = 0.2, \qquad
+k_8 = 0.05,
+\end{eqnarray*}
+over the time period $t \in [0,500]$.
 

### 2018-06-22 TPAIS_2018.tex
Paul's changes made

@@
 make progress in a range of scientific fields. Different MCMC methods
-have been designed with different challenges in mind; high dimensional
-targets\cite{cotter2013mcmc}, large data sets\cite{bierkens2016zig},
-other methods which exploit piecewise deterministic Markov processes
+have been designed with different challenges in mind, for example high dimensional
+targets\cite{cotter2013mcmc}, or large data
+sets\cite{bierkens2016zig}. 
+Other methods which exploit piecewise deterministic Markov processes
 \cite{bouchard2018bouncy},
@@
 Other approaches to challenging posterior structure have been
-proposed, including the work of Marzouk and Parnod\cite{parno2014transport}, who
+proposed, including the work of Marzouk and Parno\cite{parno2014transport}, who
 demonstrated how invertible transport maps can be constructed which
 map the target distribution close to an easily-explored reference
-density, such as a d-dimensional Gaussian. Since the map is
+density, such as a $d$-dimensional Gaussian. Since the map is
 invertible, Gaussian proposal densities on the reference space can be
@@
 Bayesian inverse problems are not the only area in which probability
-densities arise which are concentrated on, or close, lower dimensional
+densities arise which are concentrated on, or close to, lower dimensional
 manifolds. This is also a common feature of molecular dynamics
-problems, where
+problems, where ******
 
@@
 
-In this work, we aim to improve the PAIS machinery, which is limited,
-since in the current presentation the same proposal kernel is used for each particle in the
-ensemble. This homogeneity of proposal kernels means that complex
+In this paper, we aim to incorporate transport maps into an ensemble
+importance sampling setting, in order to develop stable algorithms which are
+capable of sampling from probability distributions with complex
+structures. These maps are constructed using an importance sample of
+the posterior distribution, and are used to simplify the posterior
+distribution, mapping them as close as possible to a reference
+Gaussian measure. Through this map, straightforward Gaussian proposal
+densities are transformed into complex densities, which match as well
+as possible the target density. Less well-informed proposal kernels means that complex
 structures in the density can only be well represented with a large
@@
 target density, leading to high variance in the weights, causing slow
-convergence.
-
-As shown in\cite{parno2014transport}, transport maps can be used to create
-proposal densities which are highly informed by the structure of the
-target density. In this paper, we demonstrate how such maps can also
-be constructed using importance weights arising from a pMC-type
-algorithm in order to accelerate and stabilise convergence. We will
-then apply the methodology to some complex posterior distributions
-arising from Bayesian inversion of multiscale (bio-)chemical networks.
+convergence. Better informed proposal densities
+leads to more stable and faster converging importance sampling
+schemes, and we will demonstrate this through examples
+arising from Bayesian inverse problems on multiscale stochastic
+reaction networks.
 
@@
 \section{Parallel Adaptive Importance Sampling}\label{sec:PAIS}
-Parallel adaptive importance sampling (PAIS)\cite{cotter2015parallel} is a variant of PMC\cite{cappe2012population}, which is a family of methods which are based
+Parallel adaptive importance sampling (PAIS)\cite{cotter2015parallel} is a variant of pMC\cite{cappe2012population}, which is a family of methods which are based
 on importance sampling. In importance sampling, we attempt to
@@
 \[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
-  w_kf(\theta^(k)).\]
+  w_kf(\theta^{(k)}),\]
+where $\bar{w} = \sum_{k=1}^N w_k$ is the total weight.
 This method works well when $\pi$ and $\chi$ are close, but can be
-excruciatingly slow when they are not. The idea behind PMC methods is
+excruciatingly slow when they are not. The idea behind pMC methods is
 to construct a good proposal distribution, either from the entire
@@
 \cdot, \beta)$ then 
-\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}).\]
+\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}, \beta).\]
 Often the variance of the MCMC proposal kernels can be tuned using
@@
 $\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
-sample $\{\hat{\theta}^{(k)},w^{(k)}\}$. It can be advantageous to use stratified sampling of the
+sample $\{(\hat{\theta}^{(k)},w^{(k)})\}$,  where $\theta^{(k)} =
+[\theta_1^{(k)}, \ldots, \theta_M^{(k)}]^\top$ and $w^{(k)} =
+[w_1^{(k)}, \ldots, w_M^{(k)}]^\top$. It can be advantageous to use stratified sampling of the
 mixture in order to
@@
 target density, i.e.
-\[ \hat{\theta}_i^{(k)} \sim q(\cdot ; \theta_i^{(k)})\]
+\[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
 for each $i = 1,2,\ldots,M$. We now have a weighted sample, and it
@@
 the next iteration, the importance sample
-$\{\hat{\theta}^{(k)},w^{(k)}\}$ is resampled to produce an equally
+$\{(\theta^{(k)},w^{(k)})\}$ is resampled to produce an equally
 weighted sample, ready for the next iteration of the algorithm. In
@@
 $\theta^{(k+1)}$ and the algorithm is ready for the next
-iteration. The importance samples $\{\hat{\theta}^{(k)},w^{(k)}\}$ The PAIS is summarised in Algorithm \ref{alg:PAIS}.
+iteration. The PAIS is summarised in Algorithm \ref{alg:PAIS}, where
+the importance samples $\{(\hat{\theta}^{(k)},w^{(k)})\}$ are stored as
+the output.
 
@@
 	}
-	Output $\{(w^{(n)}, \hat{\theta}^{(n)})\}_{n=1}^N$.
+	Output $\{(\hat{\theta}^{(n)},w^{(n)})\}_{n=1}^N$.
 \caption{The PAIS Algorithm.\label{alg:PAIS}}
@@
 
-One problem with the PAIS and other PMC methods can become apparent if
+One problem with the PAIS and other pMC methods can become apparent if
 the target density has highly curved manifolds on which the density is
@@
 covariances across the whole of the domain, in order to stabilise
-PMC-type methods, and make these methods more applicable to a wider
+pMC-type methods, and make these methods more applicable to a wider
 range of more challenging inference problems.
@@
 Proposing new states in this way allows us to make large steps around complex probability distributions.
-It is also feasible in this framework to assume that the reference density is close enough to a standard Gaussian that we can efficiently propose moves using a proposal distribution which is independent of the current state, e.g.
-choose $q(\theta) = \mathcal{N}(0,I_n)$.
+It is also feasible in this framework to assume that the reference
+density is close enough to a standard Gaussian that we can efficiently
+propose moves using a proposal distribution which is independent of
+the current state, for example sampling from the reference Gaussian
+itself, or a slightly more diffuse distribution.
 
@@
 maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all
-feasible maps. This space does not necessarily contain an exact
+invertible maps. This space does not necessarily contain an exact
 coupling between target and reference space, and so we are motivated to formulate an optimisation problem to find the map $\tilde{T}
@@
 \[
-	T(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
+	\tilde{T}(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
 		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
@@
 The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
-$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$.
-These multi-indices are elements of the multi-index set $\mathcal{J}_i$.
-A multi-index defines a product of univariate polynomials in $\theta_k$,
+$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in
+\mathcal{J}_i \subset \mathbb{N}_0^i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
 \[
@@
 \section[Transport map MCMC]{Transport map usage in PAIS
-  and other PMC algorithms}\label{sec:TPAIS}
+  and other pMC algorithms}\label{sec:TPAIS}
 
@@
 \subfigure[Comparison of the two Transport PAIS options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
-\subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
+\subfigure[Comparison of Transport PAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
 \caption{Convergence of algorithms; Transport M-H,
@@
 For each reaction $R_j$, for $j = 1,2,\ldots N_r$, there is a
-propensity or hazard function $\alpha_j(X(t)$ which indicates how
+propensity or hazard function $\alpha_j(X(t))$ which indicates how
 likely that reaction is to fire, defined by
@@
 
-One common approach is to induce the quasi-steady-state approximation
-QSSA. This approximation makes the assumption that fast reactions
+One common approach is to induce the quasi-equillibrium approximation
+QEA. This approximation makes the assumption that fast reactions
 enter quasi-equilibrium on a timescale which is negligible with
@@
 which takes the value
-$r \in \{1,2,\ldots,N_r\}$ with probability
-\[\mathbb{P}(r = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\]
+$i \in \{1,2,\ldots,N_r\}$ with probability
+\[\mathbb{P}(r_j = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\]
 As such, in order to compute the likelihood of a particular trajectory
@@
 \[
-	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
+	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{R}_{i\cdot}; \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
 \]
@@
 conditioned on the slow variables in the system can be approximated
-using a multiscale method, such as the QSSA or the CMA, as described
+using a multiscale method, such as the QEA or the CMA, as described
 briefly in Section \ref{sec:multi}. Once the approximations have been
@@
 \centering
-\begin{tabular}{rr}
+\begin{tabular}{cccc}
 	\toprule
-	$k$ & \quad $\mathbb{P}(D = (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$ \\ \cmidrule(lr){1-2}
-	0 & 6.8e-3 \\
-	1 & 3.2e-3 \\
-	2 & 1.7e-3 \\ \bottomrule
+	Model description & $k$ & \quad $\mathbb{P}(D =
+                                  (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$
+  & $B_{0,k}$ \\ \cmidrule(lr){1-4}
+	Full & 0 & 6.8e-3 & 1 \\
+	CMA & 1 & 3.2e-3 & 2.09 \\
+	QEA & 2 & 1.7e-3 & 4.10 \\ \bottomrule
 \end{tabular}
-\caption{Marginal distributions for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:chem_multiscale}.}
+\caption{Marginal distributions and Bayes factors for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:chem_multiscale}.}
 \label{tab:chem_Bayes_marginals}
@@
 bind to the gene $G$ to a ``switched off'' state $G'$, in which the
-production of MRNA molecules $M$ is inhibited. The MRNA encodes for
+production of mRNA molecules $M$ is inhibited. The mRNA encodes for
 production of the protein $P$, and both $P$ and $M$ can degrade. 
@@
 can only observe $T=P + 2D + 2G'$, the total number of protein
-molecules in the system, alongside $M$, the concentration of MRNA. The
+molecules in the system, alongside $M$, the concentration of mRNA. The
 number of dimers, and the status of the genetic switch, are assumed
@@
 
-One major issue with PMC-type algorithms is the curse of
+One major issue with pMC-type algorithms is the curse of
 dimensionality. This curse is two-fold, since as the target

### 2019-01-09 TPAIS_2018.tex
Yannis's comments incorporated, PAIS -> ETAIS everywhere

@@
   Mathematics and Statistics, Johns Hopkins University, Baltimore, MD
-  21218. The work of IGK was partially supported by the US National Science Foundation and by DARPA.} \and Paul
+  21218. The work of IGK was partially supported by the US National
+  Science Foundation, by the US AFOSR, and by DARPA.} \and Paul
   Russell\thanks{School of
@@
 and as such are not exploiting the parallelised  nature of modern
-computer infrastructure. In this paper, we seek to develop a method to
+computer infrastructure. In this paper, we seek to develop an approach to
 tackle this problem, using optimal transport maps to simplify
@@
 sets\cite{bierkens2016zig}. 
-Other methods which exploit piecewise deterministic Markov processes
+Other methods include piecewise deterministic Markov processes
 \cite{bouchard2018bouncy},
@@
 the geometry of the target\cite{girolami2011riemann}. Population Monte Carlo
-(pMC) methods uses efficient importance sampling proposals informed by an
+(pMC) methods use efficient importance sampling proposals informed by an
 ensemble of particles which have already learned about the structure
@@
 These methods are particularly useful in the context of
-low dimensional target distributions which have complex structure, for
+low dimensional target distributions which have complex\footnote{We
+  use ``complex'' with the meaning ``complicated'' rather than
+  containing imaginary numbers.} structure, for
 which traditional Metropolis-Hastings methods tend to struggle, for
@@
 large-scale parallelisation as opposed to the large speed-ups in
-individual processors that we saw throughout the 20th Century.
+individual processors that we saw throughout the 20th century.
 
@@
 
-Bayesian inverse problems are not the only area in which probability
-densities arise which are concentrated on, or close to, lower dimensional
-manifolds. This is also a common feature of molecular dynamics
-problems, where ******
-
-In this paper, we particularly concern ourselves with the problem of
+In this paper, we particularly concern ourselves with the subject of
 inverse problems in stochastic reaction networks. This type of model
@@
 this category, because of the small volume of the reactors involved (a
-cell).
+single cell).
 
@@
 
-The partial observability of these systems leads to only
+The partial observability of these systems lead to only
 partial-observability of the unknown parameters in the system, for
@@
 
-In Section \ref{sec:PAIS}, we will briefly reintroduce PAIS. In Section \ref{sec:map} we show how an appropriate transport map can
+In Section \ref{sec:ETAIS}, we will briefly reintroduce ETAIS. In Section \ref{sec:map} we show how an appropriate transport map can
 be constructed from importance samples which maps the posterior close
-to a reference Gaussian measure. In Section \ref{sec:TPAIS} we show
+to a reference Gaussian measure. In Section \ref{sec:TETAIS} we show
 how such a map can be incorporated into a sophisticated parallel MCMC
@@
 
-\section{Parallel Adaptive Importance Sampling}\label{sec:PAIS}
-Parallel adaptive importance sampling (PAIS)\cite{cotter2015parallel} is a variant of pMC\cite{cappe2012population}, which is a family of methods which are based
+\section{Ensemble Transport Adaptive Importance Sampling}\label{sec:ETAIS}
+Ensemble transport adaptive importance sampling (ETAIS)\cite{cotter2015parallel} is a variant of pMC\cite{cappe2012population}, which is a family of methods which are based
 on importance sampling. In importance sampling, we attempt to
 characterise a target density through sampling from that
-density. However, the target density $\pi$ itself is often too complex to sample
+density. However, the target density $\pi$ itself is often too complicated to sample
 from directly, so we instead sample from a
@@
 
-In PAIS, the proposal distribution $\chi$ is chosen to be the equally
+In ETAIS, the proposal distribution $\chi$ is chosen to be the equally
 weighted mixture of any choice of MCMC proposal kernel, evaluated at
@@
 \[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
-for each $i = 1,2,\ldots,M$. We now have a weighted sample, and it
+for each $i = 1,2,\ldots,M$. Usually the kernel $q(\cdot ;
+x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
+is scaled with $\beta$. We now have a weighted sample, and it
 would be inadvisable to use an equally weighted mixture of proposal
@@
 weighted sample, ready for the next iteration of the algorithm. In
-PAIS, a state-of-the-art resampler is used, which uses optimal
+ETAIS, a state-of-the-art resampler is used, which uses optimal
 transport methods to find the equally weighted discrete sample which
@@
 $\theta^{(k+1)}$ and the algorithm is ready for the next
-iteration. The PAIS is summarised in Algorithm \ref{alg:PAIS}, where
+iteration. The ETAIS is summarised in Algorithm \ref{alg:ETAIS}, where
 the importance samples $\{(\hat{\theta}^{(k)},w^{(k)})\}$ are stored as
@@
 	\For{$k=1,\ldots,N$}{
-		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_j^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:PAIS_propose}\;
+		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_j^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:ETAIS_propose}\;
 		Calculate $w^{(k)} = (w_1^{(k)}, \ldots, w_M^{(k)})^\top$, where
@@
 			\]
-		\label{algline:PAIS_weights}
-
-		Resample $\theta^{(k+1)} \leftarrow \| w^{(k)}\|_1^{-1}\sum_{j=1}^M w_j^{(k)}\delta_{\hat{\theta}_j^{(k)}}(\cdot)$.\label{algline:PAIS_resample}\;
+		\label{algline:ETAIS_weights}
+
+		Resample $\theta^{(k+1)} \leftarrow \| w^{(k)}\|_1^{-1}\sum_{j=1}^M w_j^{(k)}\delta_{\hat{\theta}_j^{(k)}}(\cdot)$.\label{algline:ETAIS_resample}\;
 	}
 	Output $\{(\hat{\theta}^{(n)},w^{(n)})\}_{n=1}^N$.
-\caption{The PAIS Algorithm.\label{alg:PAIS}}
+\caption{The ETAIS Algorithm.\label{alg:ETAIS}}
 \end{algorithm}
@@
 
-One problem with the PAIS and other pMC methods can become apparent if
+One problem with the ETAIS and other pMC methods can become apparent if
 the target density has highly curved manifolds on which the density is
@@
 structure, the mixture distribution proposal may not well approximate
-$\pi$ without a very large large ensemble size $M$, which can become
-inhibitively expensive. Some methods have been
+$\pi$ without a very large ensemble size $M$, which can become
+prohibitively expensive. Some methods have been
 proposed\cite{douc2007minimum}
@@
 Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}.
-Here we stick with polynomials of total order $p$ since we work with low dimensional problems with the PAIS algorithm.
+Here we stick with polynomials of total order $p$ since we work with low dimensional problems with the ETAIS algorithm.
 
@@
 Substitution of the Gaussian density into Equation~\eqref{eqn:TM_full_cost} leads to
-\begin{equation}\label{eqn:TPAIS_objective}
+\begin{equation}\label{eqn:TETAIS_objective}
 	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
@@
 form
-\begin{equation}\label{eqn:TPAIS_hessian}
+\begin{equation}\label{eqn:TETAIS_hessian}
 	HC_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i + G_i^\top
@@
 
-\section[Transport map MCMC]{Transport map usage in PAIS
-  and other pMC algorithms}\label{sec:TPAIS}
+\section[Transport map MCMC]{Transport map usage in ETAIS
+  and other pMC algorithms}\label{sec:TETAIS}
 
@@
 In particular, we will show how we can use the transport map derived in Equation~\eqref{eq:weighted_map_optim} to
-design a proposal scheme for the PAIS algorithm. In this case we have a choice in how to proceed; we
+design a proposal scheme for the ETAIS algorithm. In this case we have a choice in how to proceed; we
 propose new samples on reference space and resample on target space, or we both propose and resample on reference space, mapping onto target space to output the samples.
 The first option allows us to reuse much of the framework
-from the standard PAIS algorithm and in the numerics later we see that this performs better than
-both the Transport MH algorithm, and the standard PAIS algorithm. The second option requires some
+from the standard ETAIS algorithm and in the numerics later we see that this performs better than
+both the Transport MH algorithm, and the standard ETAIS algorithm. The second option requires some
 restructuring, but can result in improved performance from the resampler.
@@
 }
-\caption{PAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportPAIS1}}
+\caption{ETAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportETAIS1}}
 \end{algorithm}
@@
 
-The first option is given in Algorithm~\ref{alg:TransportPAIS1}. We
+The first option is given in Algorithm~\ref{alg:TransportETAIS1}. We
 denote the ensembles of states in target space $\theta^{(k)} =
@@
 \hat{\theta}^{(k)}_M)\}$, where these pairs are the states which
-together form our sample from the target distribution. As in the standard version of the PAIS algorithm we use the deterministic mixture weights.
+together form our sample from the target distribution. As in the standard version of the ETAIS algorithm we use the deterministic mixture weights.
 
@@
 
-	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\label{algline:TPAIS_resample}\;
+	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\label{algline:TETAIS_resample}\;
 	Invert $\theta^{(k+1)}_i = \tilde{T}^{-1}(r^*_i)$.\;
@@
 }
-\caption{PAIS algorithm with adaptive transport map. Option 2.\label{alg:TransportPAIS2}}
+\caption{ETAIS algorithm with adaptive transport map. Option 2.\label{alg:TransportETAIS2}}
 \end{algorithm}
@@
 
-The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to
-the first except on Line~\ref{algline:TPAIS_resample} where rather
+The second option, Algorithm~\ref{alg:TransportETAIS2}, is similar to
+the first except on Line~\ref{algline:TETAIS_resample} where rather
 than resampling in target space we resample in reference space. In
@@
 members. This means that the resampling step will be more efficient in
-moderately higher dimensions, which we discuss in Section~\ref{sec:TPAIS_higher_dim}.
+moderately higher dimensions, which we discuss in Section~\ref{sec:TETAIS_higher_dim}.
 
@@
 proposal distributions which we have described in
-Section~\ref{sec:TPAIS}.
+Section~\ref{sec:TETAIS}.
 We take as a test problem the ubiquitous Rosenbrock banana-shaped
@@
 the push forward of the sample is still a unimodal distribution
-centred at the origin with standard deviation 1. As you move out into
+centred at the origin with standard deviation 1. As one moves out into
 the tails of the reference density more non-Gaussian features are
@@
 algorithms. This is done, as in \cite{cotter2015parallel}, by optimising for the effective sample size in
-the PAIS algorithm, and by tuning the relative $L^2$ error in the MH
+the ETAIS algorithm, and by tuning the relative $L^2$ error in the MH
 algorithm. There is currently no
@@
 results, especially if adaptation of the map is stopped after a given
-point. As in the PAIS algorithm, optimising for the effective sample size might be the best option.
+point. As in the ETAIS algorithm, optimising for the effective sample size might be the best option.
 
@@
 	Statistic \quad / \quad Algorithm & Transport M-H &
-                                                            Alg. \ref{alg:TransportPAIS1} & Alg. \ref{alg:TransportPAIS2}  \\ \cmidrule(lr){1-4}
+                                                            Alg. \ref{alg:TransportETAIS1} & Alg. \ref{alg:TransportETAIS2}  \\ \cmidrule(lr){1-4}
 	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
@@
   algorithms applied to $R_1$, optimising for the $L^2$ error and
-  the effective sample size (ESS) for PAIS algorithms, and for average
+  the effective sample size (ESS) for ETAIS algorithms, and for average
   acceptance rate for MH algorithms.}
@@
 size is much lower than we see in the one-dimensional examples with
-the PAIS algorithms. However, in the Rosenbrock density
+the ETAIS algorithms. However, in the Rosenbrock density
 \eqref{eqn:R2_repeat} we are dealing with a much more complicated
@@
 $\theta_2$.
-From our experiments, we have observed that the standard PAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.
+From our experiments, we have observed that the standard ETAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.
 
@@
 \centering
-\subfigure[Comparison of the two Transport PAIS options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
-\subfigure[Comparison of Transport PAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
+\subfigure[Comparison of the two Transport ETAIS
+options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
+\subfigure[Comparison of Transport ETAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
 \caption{Convergence of algorithms; Transport M-H,
-  algorithm \ref{alg:TransportPAIS1}, algorithm \ref{alg:TransportPAIS2} for
+  algorithm \ref{alg:TransportETAIS1}, algorithm \ref{alg:TransportETAIS2} for
   density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
@@
 The convergence of the three algorithms is displayed in Figure~\ref{fig:R2_l2_convergence}.
-Figure (a) shows that the two variations of the transport based PAIS algorithms converge with similar rates.
+Figure (a) shows that the two variations of the transport based ETAIS algorithms converge with similar rates.
 The second version, which performs the resampling stage in reference space rather than target space, has a slightly higher ESS, and is more stable than option (1).
-This version also has a property that we can exploit in Section~\ref{sec:TPAIS_higher_dim}.
+This version also has a property that we can exploit in Section~\ref{sec:TETAIS_higher_dim}.
 
 %\section{Transport map-accelerated Parallel Adaptive Importance
-  %Sampling (TPAIS)}\label{sec:TPAIS}
+  %Sampling (TETAIS)}\label{sec:TETAIS}
 
@@
 this in Section \ref{sec:num}, using the transport map versions of the
-PAIS algorithm, as described in Section \ref{sec:TPAIS}.
+ETAIS algorithm, as described in Section \ref{sec:TETAIS}.
 
@@
 demonstrate the effectiveness of the Bayesian approach we have
-proposed, along with the transport PAIS algorithms for sampling the
+proposed, along with the transport ETAIS algorithms for sampling the
 challenging posterior distribution on the reaction parameters that
@@
 We consider several different algorithms for sampling from these
-distributions. First we implement both the PAIS and MH algorithms with
-a Gaussian proposal distribution. In the case of the PAIS algorithm,
+distributions. First we implement both the ETAIS and MH algorithms with
+a Gaussian proposal distribution. In the case of the ETAIS algorithm,
 this is a Gaussian mixture proposal distribution, and in the MH
@@
 The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard MH-RW algorithm.
-We also compare the PAIS and MH algorithms when using a transport map proposal distribution.
-This proposal method was discussed in detail in Chapter~\ref{sec:TPAIS}.
+We also compare the ETAIS and MH algorithms when using a transport map proposal distribution.
+This proposal method was discussed in detail in Chapter~\ref{sec:TETAIS}.
 
@@
 When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
-MH-RW  and MH-TRW, as well as PAIS-RW and PAIS-TRW.
+MH-RW  and MH-TRW, as well as ETAIS-RW and ETAIS-TRW.
 In practice we cannot ensure that the approximate map $\tilde{T}$
@@
 As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
-MH-logRW and MH-logTRW, with, PAIS-logRW and PAIS-logTRW.
+MH-logRW and MH-logTRW, with, ETAIS-logRW and ETAIS-logTRW.
 
@@
 \toprule
-	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{PAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
+	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{ETAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
 	& & $\delta$ & & $\delta$ & ESS \\ \midrule
@@
 \end{tabular}
-\caption{Optimal scaling parameters for the MH and PAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}.
-MH parameters optimised by acceptance rate, and PAIS parameters optimised using effective sample size.}
+\caption{Optimal scaling parameters for the MH and ETAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}.
+MH parameters optimised by acceptance rate, and ETAIS parameters optimised using effective sample size.}
 \label{tab:chem_multiscale_scaling}
@@
 values. MH-RW based algorithms are tuned to an acceptance rate close
-to the optimal $23.4\%$, and PAIS algorithms are tuned to maximise the
+to the optimal $23.4\%$, and ETAIS algorithms are tuned to maximise the
 effective sample size (ESS). The optimal parameter values for the MH
@@
 transformations. The ESS is highest for the logTRW variant
-of the PAIS algorithm, as we would hope.
+of the ETAIS algorithm, as we would hope.
 
@@
 We first note the poor performance of the MH based algorithms, each of them taking roughly 10,000 samples to begin converging.
-Only the MH-TRW is at all competitive with the PAIS algorithms.
+Only the MH-TRW is at all competitive with the ETAIS algorithms.
 During the simulation interval, the MH-TRW algorithm has not settled down to the expected $\mathcal{O}(1/\sqrt{N})$ rate which means that the estimate is still biased by the long burn-in time.
-As we have seen in previous examples, the burn-in time for the PAIS algorithm is negligible.
-
-The PAIS variants with RW and TRW proposals perform similarly on both
+As we have seen in previous examples, the burn-in time for the ETAIS algorithm is negligible.
+
+The ETAIS variants with RW and TRW proposals perform similarly on both
 sample spaces. When sampling without a preconditioner, the transport map is not
@@
 This proposal method becomes more stable as we increase either the ensemble size, or the number of iterations between updates of the transport map, $T$.
-Overall we see the smallest Monte Carlo errors for a given amount of computational effort coming from the PAIS-logTRW algorithm.
+Overall we see the smallest Monte Carlo errors for a given amount of computational effort coming from the ETAIS-logTRW algorithm.
 
@@
 \]
-Under the PAIS framework, it is straightforward to calculate these factors using the Monte Carlo estimator for the normalisation constants.
+Under the ETAIS framework, it is straightforward to calculate these factors using the Monte Carlo estimator for the normalisation constants.
 From~\cite{robert2013monte} these normalisation constants take the form
@@
 interesting correlation structures, however several dimensions have
-leptokurtic tails which are difficult for the standard PAIS algorithm
+leptokurtic tails which are difficult for the standard ETAIS algorithm
 to sample from. The marginal densities also vary over very different
@@
 However, due to the scaling of the weight we have had to be more careful about which samples we use to build our map.
-We require that the objective function $C(T)$, from Equation~\eqref{eqn:TPAIS_objective} is convex, which requires that the second derivative, given in Equation~\eqref{eqn:TPAIS_hessian}, is positive definite.
+We require that the objective function $C(T)$, from Equation~\eqref{eqn:TETAIS_objective} is convex, which requires that the second derivative, given in Equation~\eqref{eqn:TETAIS_hessian}, is positive definite.
 This positive definite property depends on all weights in our sample being strictly positive, which is not always possible on a computer.
@@
 rate for the MH algorithms, and optimising the effective sample size
-for the PAIS algorithms. The optimal parameters are given in
+for the ETAIS algorithms. The optimal parameters are given in
 Table~\ref{tab:grn_scaling_parameters}. Here, for Metropolis-Hastings
 variants, $\beta_{\%}^*$ is the variance of the proposal, optimised
-for an acceptance rate of $50\%$. For the PAIS variants,
+for an acceptance rate of $50\%$. For the ETAIS variants,
 $\beta_{\text{ESS}}^*$ is the variance of each of the proposal kernels
@@
 	\toprule
-		 & \multicolumn{4}{c}{MH} & \multicolumn{4}{c}{PAIS} \\ \cmidrule(lr){2-5} \cmidrule(lr){6-9}
+		 & \multicolumn{4}{c}{MH} & \multicolumn{4}{c}{ETAIS} \\ \cmidrule(lr){2-5} \cmidrule(lr){6-9}
 		& RW & logRW & TRW & log-TRW & RW & logRW & TRW & log-TRW \\ \midrule
@@
 
-The first thing to note is that the first PAIS measurement happens
+The first thing to note is that the first ETAIS measurement happens
 after 2500 samples, whereas the first MH measurement occurs after 50
@@
 
-The second obvious feature of these convergence plots is that the PAIS
+The second obvious feature of these convergence plots is that the ETAIS
 algorithms outperform the MH algorithms by a large margin - roughly a
@@
 same number of samples. A less positive result is that for this
-example the PAIS-TRW algorithm has found a map which has prevented
+example the ETAIS-TRW algorithm has found a map which has prevented
 further convergence of the mean. This is likely because one dimension
@@
 as in these examples. The large increase in the effective sample size
-observed between the PAIS-logRW and PAIS-logTRW algorithms is
+observed between the ETAIS-logRW and ETAIS-logTRW algorithms is
 converted into an estimate which is twice as accurate after 1 million
@@
 \subsection{Sampling in higher
-  dimensions}\label{sec:TPAIS_higher_dim}
+  dimensions}\label{sec:TETAIS_higher_dim}
 
@@
 
-Algorithm~\ref{alg:TransportPAIS2} allows us to decorrelate the
+Algorithm~\ref{alg:TransportETAIS2} allows us to decorrelate the
 dimensions of our random parameter on reference space, where we then
@@
 \subsection{Invertibility of the transport map}
-In this paper, we have considered an approach very similar to that
-deseminated in \cite{parno2014transport}, where we aim to find an
+In this paper, we have considered an approach similar to that
+presented in \cite{parno2014transport}, where we aim to find an
 invertible map which approximates the true transport map between the

### 2019-01-10 TETAIS.tex
New file to reflect new title
Also changed AMR -> MT throughout

--- 2019-01-09 TPAIS_2018.tex
@@
 expensive, in which case a greedy approximation of this algorithm, the
-Approximate Multinomial Resampler (AMR) can be
+Multinomial Transformation (MT) can be
 implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
@@
   algorithm \ref{alg:TransportETAIS1}, algorithm \ref{alg:TransportETAIS2} for
-  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
+  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the MT algorithm.}
 \label{fig:R2_l2_convergence}
@@
 % choice of resampler?
-We will use the AMR\cite{russ2017parallel} resampler with an ensemble size of
+We will use the MT\cite{russ2017parallel} resampler with an ensemble size of
 $M=500$. This increase in ensemble size in comparison with previous
@@
 burn-in phase of the algorithms to avoid the need for a huge ensemble size to approximate the posterior with a mixture of isotropic Gaussian kernels.
-We use the AMR resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size compensates for the increase in dimension from four to eight.
+We use the MT resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size compensates for the increase in dimension from four to eight.
 
@@
 simplex algorithm to minimise the associated cost function, and faster
-than the AMR algorithm\cite{cotter2015parallel}. Moreover, the task of
+than the MT algorithm\cite{cotter2015parallel}. Moreover, the task of
 running this resampler can then be parallelised over $d$

### 2019-01-10 TETAIS.tex
Updated bib

@@
 
-\subsection{The chicken and egg paradox}
-Another issue with the use of transport maps of this ilk, is that in
-order to construct such a map, we require a representative sample from
-the posterior distribution. Since these methods are designed to sample
-from distributions which are inherently difficult to sample from, be
-it due to multimodality, as looked at in \cite{cotter2015parallel}, or
-as in this paper, due to the density being highly concentrated close
-to a curved manifold in parameter space, it could be argued that accessing this initial set of
-samples remains a hard problem. This remains an open problem, which we
-are aiming to address in future work.
+% \subsection{The chicken and egg paradox}
+% Another issue with the use of transport maps of this ilk, is that in
+% order to construct such a map, we require a representative sample from
+% the posterior distribution. Since these methods are designed to sample
+% from distributions which are inherently difficult to sample from, be
+% it due to multimodality, as looked at in \cite{cotter2015parallel}, or
+% as in this paper, due to the density being highly concentrated close
+% to a curved manifold in parameter space, it could be argued that accessing this initial set of
+% samples remains a hard problem. This remains an open problem, which we
+% are aiming to address in future work.
 

### 2019-01-15 TETAIS.tex
Cover letter added plus minor amendments

@@
 do not incorporate information about this structure, will be highly
-inefficient. Moreover, most methods are inherently serial in nature,
-and as such are not exploiting the parallelised  nature of modern
-computer infrastructure. In this paper, we seek to develop an approach to
-tackle this problem, using optimal transport maps to simplify
+inefficient.
+
+In this paper, we seek to develop an approach to
+tackle this problem when using adaptive importance sampling methods,
+by using optimal transport maps to simplify
 posterior distributions which are concentrated on lower dimensional
-manifolds.
+manifolds. This approach is applicable to a whole range of
+problems for which Monte Carlo Markoc chain (MCMC) methods mix slowly.
 
@@
 that certain multiscale approximations lead to more consistent
-approximations of the posterior than others.
+approximations of the posterior than others. The use of optimal
+transport maps stabilises the ensemble transform adaptive importance
+sampling (ETAIS) method, and allows for efficient sampling with
+smaller ensemble sizes.
 \end{abstract}
@@
 targets\cite{cotter2013mcmc}, or large data
-sets\cite{bierkens2016zig}. 
-Other methods include piecewise deterministic Markov processes
-\cite{bouchard2018bouncy},
-methods which are able to make very large moves informed by the
-data\cite{teh2016consistency,duane1987hybrid}, methods which are able to exploit
+sets\cite{bierkens2016zig}. New types of stochastic processes
+have been considered, such as piecewise deterministic Markov processes
+\cite{bouchard2018bouncy}.
+Other methods are able to make very large moves informed by the
+data\cite{teh2016consistency,duane1987hybrid}, and others are able to exploit
 the geometry of the target\cite{girolami2011riemann}. Population Monte Carlo
@@
 
-In \cite{cotter2015parallel}, we explored ideas from adaptive MCMC within a
-pMC-type structure, in order to construct an adaptive importance
-sampler which could automatically adapt to optimise
+In \cite{cotter2015parallel}, we explored ideas from adaptive MCMC
+within adaptive importance samplers, in order to construct a method which could automatically adapt to optimise
 efficiency. Alongside this, we explored the use of state-of-the-art
-resamplers\cite{reich2013nonparametric}, and a greedy approximation (the
-approximate multinomial resampler) in order to improve the quality of
-the importance sampler proposals.
+resamplers\cite{reich2013nonparametric}, and the multinomial transform
+(MT), a greedy approximation of the ensemble transform method, in order to improve the quality of
+the importance sampler proposal distribution.
 
@@
 is the use of the anisotropic diffusion maps\cite{singer2008non}, in
-which a map which describes the slow manifold is constructed using
+which a map which describes the manifold is constructed using
 principle component analysis. These ideas have also been applied to
@@
 multiscale approximations, and the resulting approximations of the
-likelihoods arising from the inverse problem for such a network, then
+likelihoods arising from the inverse problem for such a network, 
 often exhibit 
@@
 
-In Section \ref{sec:ETAIS}, we will briefly reintroduce ETAIS. In Section \ref{sec:map} we show how an appropriate transport map can
+In Section \ref{sec:ETAIS}, we will briefly reintroduce the ensemble
+transport adaptive importance sampling (ETAIS) method. In Section \ref{sec:map} we show how an appropriate transport map can
 be constructed from importance samples which maps the posterior close
@@
 \section{Ensemble Transport Adaptive Importance Sampling}\label{sec:ETAIS}
-Ensemble transport adaptive importance sampling (ETAIS)\cite{cotter2015parallel} is a variant of pMC\cite{cappe2012population}, which is a family of methods which are based
-on importance sampling. In importance sampling, we attempt to
-characterise a target density through sampling from that
-density. However, the target density $\pi$ itself is often too complicated to sample
-from directly, so we instead sample from a
-proposal density $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
+Ensemble transport adaptive importance sampling
+(ETAIS)\cite{cotter2015parallel} is an adaptive importance sampling framework
+which uses an ensemble of particles and state-of-the-art resampling
+methods to construct a mixture proposal distribution which closely
+matches the target distribution.
+
+In importance sampling, we aim to sample from a target $\pi$, by
+sampling from a chosen proposal distribution $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
 $w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
@@
 This method works well when $\pi$ and $\chi$ are close, but can be
-excruciatingly slow when they are not. The idea behind pMC methods is
+excruciatingly slow when they are not. The idea behind adaptive
+importance sampling methods is
 to construct a good proposal distribution, either from the entire
@@
 In ETAIS, the proposal distribution $\chi$ is chosen to be the equally
-weighted mixture of any choice of MCMC proposal kernel, evaluated at
-each of the current particles in the system. If $\theta^{(k)} = [\theta_1^{(k)},
+weighted mixture of set of proposal kernels. If $\theta^{(k)} = [\theta_1^{(k)},
 \theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
-ensemble, and we wish to use an MCMC proposal density $q(\cdot ;
+ensemble, and we wish to use a mixture of the proposal density $q(\cdot ;
 \cdot, \beta)$ then 
-\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)}, \beta).\]
-Often the variance of the MCMC proposal kernels can be tuned using
-their respective algorithmic parameters $\beta$. Good values for these
-algorithmic parameters can be
+\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)},
+\beta).\]
+Usually the kernel $q(\cdot ;
+x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
+scales with $\beta$. Good values for the
+algorithmic parameter(s) $\beta$ can be
 found by optimising for the effective sample size of the importance
@@
 \[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
-for each $i = 1,2,\ldots,M$. Usually the kernel $q(\cdot ;
-x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
-is scaled with $\beta$. We now have a weighted sample, and it
+for each $i = 1,2,\ldots,M$.  We now have a weighted sample, and it
 would be inadvisable to use an equally weighted mixture of proposal
@@
 \BlankLine
-	Initialise $\theta^{(1)} \sim \mu_0$.\;
+	Initialise $\theta^{(1)} \sim \pi_0$.\;
 	\For{$k=1,\ldots,N$}{
@@
 			\[
-				w^{(k)}_i = \frac{\mu(\hat{\theta}^{(k)}_i)}{\chi^{(k)}(\hat{\theta}^{(k)}_i; \theta^{(k)}, \beta)}.
+				w^{(k)}_i = \frac{\pi(\hat{\theta}^{(k)}_i)}{\chi^{(k)}(\hat{\theta}^{(k)}_i; \theta^{(k)}, \beta)}.
 			\]
@@
 
-One problem with the ETAIS and other pMC methods can become apparent if
+One problem with the ETAIS and other similar methods can become apparent if
 the target density has highly curved manifolds on which the density is
@@
 covariances across the whole of the domain, in order to stabilise
-pMC-type methods, and make these methods more applicable to a wider
+adaptive importance sampling methods, and make these methods more applicable to a wider
 range of more challenging inference problems.
@@
 
-Proposing new states in this way allows us to make large steps around complex probability distributions.
-It is also feasible in this framework to assume that the reference
-density is close enough to a standard Gaussian that we can efficiently
+Proposing new states in this way allows us to make large, informed
+moves, even when sampling from probability distributions with complex structure.
+If a very good such map can be found, it is also possible that the
+push forward of the posterior density through the map is close enough
+to the reference Gaussian that we can efficiently
 propose moves using a proposal distribution which is independent of
@@
 In this Section we outline the methodology in
-\cite{parno2014transport} for approximately coupling the target,
-$\mu_{\theta}$, with the reference distribution, $\mu_r$, and
+\cite{parno2014transport} for approximately coupling the target
+measure with density
+$\pi_{\theta}$, with the reference distribution with density $\pi_r$, and
 show how the map can be constructed using a weighted sample
@@
 \begin{dfn}[Target and Reference Space]
-	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target measure $\mu_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference measure $\mu_r$.
+	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target density $\pi_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference density $\pi_r$.
 \end{dfn}
@@
         \xrightarrow[K \to \infty]{}
-		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\mu_\theta =
+		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\pi_\theta =
 		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
@@
 where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
-the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\mu_\theta$; however this
+the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\pi_\theta$; however this
 approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples.% If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.
@@
 \section[Transport map MCMC]{Transport map usage in ETAIS
-  and other pMC algorithms}\label{sec:TETAIS}
+  and other importance sampling algorithms}\label{sec:TETAIS}
 
@@
 posteriors densities
-\begin{equation}\label{eqn:chem_diff}
-	\text{diff}(\mu^{\text{CMA}}, \mu^{\text{QEA}}) = \pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
+\begin{equation}\label{eqn:chem_diff}\pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
 \end{equation}
@@
 
-One major issue with pMC-type algorithms is the curse of
+One major issue with importance sampling-based algorithms is the curse of
 dimensionality. This curse is two-fold, since as the target

### 2019-01-15 cover_letter.tex
Cover letter added plus minor amendments

\documentclass{letter}
\usepackage{txfonts}

\setlength{\textheight}{10.2in} 
\setlength{\topmargin}{-0.9in}

\signature{Dr Simon Cotter (corresponding author), Prof. Ioannis
  Kevrekidis, and Dr Paul Russell}
\address{School of Mathematics \\ Alan Turing Building \\ Oxford
  Road\\Manchester\\ M13 9PL \\ UK \\\textit{email:} simon.cotter@manchester.ac.uk}

\begin{document}

\date{}
\begin{letter}{ }

\opening{Dear Prof. Higdon,}

Please find attached our new manuscript entitled ``Transport map accelerated adaptive importance sampling, and application to inverse problems arising from
  multiscale stochastic reaction networks''. In this paper, we expand
  on recent work adaptive importance samplers, by incorporating the
  use of transport maps to simplify challenging posterior
  distributions, giving rise to efficient and stable importance
  sampling. The efficacy of the approach is demonstrated through
  application to inverse problems for multiscale biochemical networks,
  for which the posterior is highly concentrated close to curved lower
  dimensional manifolds.

We feel that SIAM Journal of Uncertainty
  Quantification would be the best place for this work, since the main
  two references, one by Youssef Marzouk's group and the other by my
  own group, are published and in review respectively, with this journal.
We very much look forward to hearing from you in due course regarding this paper.

\closing{Yours sincerely,}
\end{letter}

\end{document}

### 2019-01-15 TETAIS.tex
Got to 7.1.1
Final proof read and edits complete up to 7.1.1.

@@
 \begin{dfn}[Target and Reference Space]
-	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target density $\pi_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference density $\pi_r$.
+	The transport map pushes a sample from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target density $\pi_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference density $\pi_r$.
 \end{dfn}
@@
 \]
-which is a true metric between probability distributions $p$ and $q$.
+which is a true distance metric on probability distributions.
 Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
@@
 target measure to a reference measure. We now consider how to
-implement an importance sampling-based MCMC algorithm which uses
+implement an ensemble importance sampling algorithm which uses
 these maps to propose new states. In \cite{parno2014transport} it was
@@
 the first except on Line~\ref{algline:TETAIS_resample} where rather
-than resampling in target space we resample in reference space. In
-reference space the dimensions are roughly uncorrelated, and the
+than resampling in target space we resample in reference
+space. Assuming a good transport map, the push forward of the target
+measure onto 
+reference space has no strong correlations, and the
 Gaussian marginals are easy to approximate with fewer ensemble
@@
 
-Before looking at the performance of the MCMC algorithms, we demonstrate some properties of the transport maps we will be using in our MCMC algorithms.
-We draw 1 million samples from the density in \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:map} to build a transport map.
+Before looking at the performance of the algorithms, we demonstrate some properties of the transport maps we will be using.
+We draw 1 million samples from the density \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:map} to build a transport map.
 We use this map to push forward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian.
@@
 
-The convergence of the three algorithms is displayed in Figure~\ref{fig:R2_l2_convergence}.
-Figure (a) shows that the two variations of the transport based ETAIS algorithms converge with similar rates.
-The second version, which performs the resampling stage in reference space rather than target space, has a slightly higher ESS, and is more stable than option (1).
-This version also has a property that we can exploit in Section~\ref{sec:TETAIS_higher_dim}.
+The convergence of the three algorithms is displayed in
+Figure~\ref{fig:R2_l2_convergence}. Figure (a) shows that the two
+variations of the transport based ETAIS algorithms converge with
+similar rates. The second version, which performs the resampling stage
+in reference space rather than target space, has a slightly higher
+ESS, and is more stable than option (1). This version also has a
+property that we can exploit in Section~\ref{sec:TETAIS_higher_dim}.
+
+Figure~\ref{fig:R2_l2_convergence} (b) compares the performance of
+RWMH, transport map-accelerated RWMH, ETAIS-RW and transport
+map-accelerated ETAIS-RW. The difference in performance between RWMH
+and its accelerated version is relatively negligible, in part because
+this version of the Rosenbrock is not overtly pathological. The story
+is different for the ETAIS algorithms. The slowly decaying tails of
+the Rosenbrock cause a problem for ETAIS since these are not
+well-represented in the proposal distributions, leading to high weight
+variance and poor convergence. A much larger
+ensemble would be required to ensure a stable ETAIS implementation for this
+problem, which leads to high overhead costs for the resampling step
+and evaluation of the importance weights. The transport map-accelerated
+ETAIS, on the other hand, is stabilised by the map, which transforms the proposal
+distributions in order to adequately approximate the slowly decaying
+tails of the target. This stabilisation allows us to benefit from the
+improved convergence of the ETAIS methodology, with a smaller ensemble
+size.
 
@@
 situations, for example monomolecular networks\cite{jahnke2007solving}, or for
-steady-state solutions certain deficiency
+steady-state solutions for weakly reversible deficiency
 zero networks\cite{anderson2010product,anderson2016product}. Trajectories for this system can be
@@
 errors\cite{cotter2016error}. Another approach is the
-constrained multiscale method (CMA)\cite{cotter2011constrained,cotter2016constrained}, based in part on the
+constrained multiscale algorithm (CMA)\cite{cotter2011constrained,cotter2016constrained}, based in part on the
 equation-free approach to multiscale computations\cite{kevrekidis2003equation,erban2006gene}. This approach also assumes quasi-equilibrium in the
@@
   networks}
-Suppose now that we are only observing the slower variables in a
+Suppose now that we only observe the slower variables in a
 multiscale chemical system of this type. Suppose that $n_r < N_r$ is the number of slow
@@
 In this section we look at two examples of chemical systems to
-demonstrate the effectiveness of the Bayesian approach we have
-proposed, along with the transport ETAIS algorithms for sampling the
+demonstrate the effectiveness of transport ETAIS algorithms for
+sampling from the
 challenging posterior distribution on the reaction parameters that
@@
 First we consider a simple multiscale example involving two chemical
-species $S_1$ and $S_2$ involved in only zeroth and first order reactions:
+species $S_1$ and $S_2$,
 \begin{equation}\label{eqn:full_system}
@@
 \begin{equation}\label{eq:params1}
-k_1 = 100, \qquad k_2 = k_3 = 10, \qquad k_4 = 1,
+k_1V = 100, \qquad k_2 = k_3 = 10, \qquad k_4 = 1,
 \end{equation} 
-
+where $V$ is the volume of the reactor.
  In this 
@@
 the second and
-third, the same data are used, but restricted to observations of the slow
+third, the same data are used, but projected down to one dimensional observations of the slow
 variable $S = X_1 + X_2$. In these last two examples, we approximate the
@@
 Figure~\ref{fig:chem_CMA_posterior} shows how this posterior looks
-when we use the CMA to model the effective degradation propensity $\bar{\alpha}_4$.
+when we use the CMA to model the effective degradation propensity
+$\bar{\alpha}_4$. Although this posterior does not appear to be
+pathologically difficult to sample using MCMC from these plots, it most certainly is. The posterior
+density is concentrated close to a curved manifold. Since the manifold
+is curved, the joint marginals do not look unreasonable, where in fact
+the density has very differing derivatives in different directions
+when evaluated close to this manifold.
 
@@
 this is a Gaussian mixture proposal distribution, and in the MH
-algorithm this is equivalent to a vanilla random walk.
-The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard MH-RW algorithm.
-We also compare the ETAIS and MH algorithms when using a transport map proposal distribution.
-This proposal method was discussed in detail in Chapter~\ref{sec:TETAIS}.
+algorithm this is equivalent to a vanilla random walk. The proposal
+distribution uses a covariance matrix which has been constructed using
+the sample covariance of a sample produced using a standard MH-RW
+algorithm. This is necessary since these algorithms otherwise fail to
+converge in a reasonable time for
+this problem. We also compare the ETAIS and MH algorithms when using a
+transport map proposal distributionm, as was discussed in detail in Chapter~\ref{sec:TETAIS}.
 
@@
 We will use the MT\cite{russ2017parallel} resampler with an ensemble size of
-$M=500$. This increase in ensemble size in comparison with previous
+$M=500$. This increase in ensemble size in comparison with the previous
 example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.
@@
 the degradation of $S$ given by \eqref{eqn:chem_CMA_rate}, but we
-expect very similar results for the other posterior distributions discussed.
+expect very similar results for the other multiscale approximation of
+posterior distributions
+discussed.
 
@@
 transformations. The ESS is highest for the logTRW variant
-of the ETAIS algorithm, as we would hope.
+of the ETAIS algorithm, as we would hope. We note here that as
+outlined in \cite{cotter2015parallel}, it is possible to construct
+adaptive versions of ETAIS which automatically tune these algorithmic
+parameters, but we compare against the static optimal parameter values
+for all algorithms to ensure fairness.
 
@@
 
-Convergence of the eight algorithms for this example is shown in Figure~\ref{fig:chem_multiscale_L2}.
-We first note the poor performance of the MH based algorithms, each of them taking roughly 10,000 samples to begin converging.
+Convergence of the eight algorithms for this example are shown in
+Figure~\ref{fig:chem_multiscale_L2}. We first note the poor
+performance of the MH based algorithms, despite being preconditioned
+with a pre-computed sample covariance.
 Only the MH-TRW is at all competitive with the ETAIS algorithms.
@@
 previous section i.e. many proposals are made which do result in
-negative reaction rates.
-Sampling with transport maps with a $\log$ preconditioner leads to more comparable Monte Carlo errors, with the logTRW being apparently slightly less stable.
-This proposal method becomes more stable as we increase either the ensemble size, or the number of iterations between updates of the transport map, $T$.
-Overall we see the smallest Monte Carlo errors for a given amount of computational effort coming from the ETAIS-logTRW algorithm.
+negative reaction rates. Sampling with transport maps with a $\log$
+preconditioner leads to more comparable Monte Carlo errors, with the
+logTRW being apparently slightly less stable. This proposal method
+becomes more stable as we increase either the ensemble size, or the
+number of iterations between updates of the transport map,
+$T$. Overall we see the smallest Monte Carlo errors for a given number
+of likelihood evaluations coming from the ETAIS-logTRW algorithm.
 
@@
 We now look at the proposal distributions of the transport map accelerated algorithms.
-In Figure~\ref{fig:chem_reference_spaces}, we see the reference spaces found by; in (a) mapping the posterior through the map $\hat{T}$, and in (b) by mapping the posterior through $T\circ\log$.
+In Figure~\ref{fig:chem_reference_spaces}, we see the reference spaces found by; (a) mapping the posterior through the map $\hat{T}$, (b) by mapping the posterior through $T\circ\log$.
 For the most part, each of these marginal distributions can be recognised as a Gaussian.

### 2019-01-15 images/Applications/createfigure.m

@@ -1,0 +1,52 @@
+function createfigure(X1, YMatrix1, X2, YMatrix2)
+%CREATEFIGURE(X1, YMATRIX1, X2, YMATRIX2)
+%  X1:  vector of x data
+%  YMATRIX1:  matrix of y data
+%  X2:  vector of x data
+%  YMATRIX2:  matrix of y data
+
+%  Auto-generated by MATLAB on 15-Jan-2019 15:50:42
+
+% Create figure
+figure1 = figure('PaperSize',[20.99999864 29.69999902]);
+
+% Create axes
+axes1 = axes('Parent',figure1,'YMinorTick','on','YScale','log',...
+    'XMinorTick','on',...
+    'XScale','log',...
+    'XTick',[10 100 1000 10000 100000 1000000],...
+    'FontSize',20);
+%% Uncomment the following line to preserve the X-limits of the axes
+% xlim(axes1,[10 1000000]);
+box(axes1,'on');
+hold(axes1,'on');
+
+% Create multiple lines using matrix input to loglog
+loglog1 = loglog(X1,YMatrix1(1:2,:),'LineWidth',2);
+set(loglog1(1),'DisplayName','MH-logRW','Marker','o','Color',[0 0 1]);
+set(loglog1(2),'DisplayName','MH-logTRW','Marker','^','Color',[1 0 0]);
+
+% Create multiple lines using matrix input to loglog
+loglog2 = loglog(X2,YMatrix2,'LineWidth',2,'LineStyle','--');
+set(loglog2(1),'DisplayName','PAIS-logRW','Marker','o',...
+    'Color',[0 0.749019622802734 0.749019622802734]);
+set(loglog2(2),'DisplayName','PAIS-logTRW','Marker','^',...
+    'Color',[0.87058824300766 0.490196079015732 0]);
+
+
+
+loglog3 = loglog(X1,YMatrix1(3,:),'LineWidth',2);
+set(loglog3,'DisplayName','$\mathcal{O}(1/\sqrt{N})$','LineStyle','--',...
+    'Color',[0 0 0]);
+
+% Create xlabel
+xlabel('Number of samples','Margin',2,'FontSize',20,'Interpreter','latex');
+
+% Create ylabel
+ylabel('relative error','Margin',2,'FontSize',20,'Interpreter','latex');
+
+% Create legend
+legend1 = legend(axes1,'show');
+set(legend1,'Location','southwest','Interpreter','latex','FontSize',20,...
+    'EdgeColor',[0 0 0]);
+

### 2019-01-17 TETAIS.tex
Final edits complete, including spell check

@@
 manifolds. This approach is applicable to a whole range of
-problems for which Monte Carlo Markoc chain (MCMC) methods mix slowly.
+problems for which Monte Carlo Markov chain (MCMC) methods mix slowly.
 
@@
 
-One common approach is to induce the quasi-equillibrium approximation
+One common approach is to induce the quasi-equilibrium approximation
 QEA. This approximation makes the assumption that fast reactions
@@
 this problem. We also compare the ETAIS and MH algorithms when using a
-transport map proposal distributionm, as was discussed in detail in Chapter~\ref{sec:TETAIS}.
+transport map proposal distribution, as was discussed in detail in Chapter~\ref{sec:TETAIS}.
 
@@
 
-The convergence analysis has been performed for the constrained
+The convergence analysis in the previous section was performed for the constrained
 approximation of the posterior distribution. We now look at the
-differences in the approximations of the posterior distribution
-arising in this problem between the CMA and QEA. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
+differences in the CMA and QEA approximations of the posterior distribution
+arising in this problem. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
 \[
@@
 which the posterior is concentrated, as can be seen in
-Figure~\ref{fig:chem_diff}. The figure shows the difference in
+Figure~\ref{fig:chem_diff}, in which we plot the difference in the 
 posteriors densities
@@
 Since the two posteriors have been approximated using an MCMC sample,
-there is a significant amount of noise, particularly in the tails of
+there is a visible amount of Monte Carlo static, particularly in the tails of
 the distributions. We can see that the differences in the marginals
@@
 us that we should significantly prefer the full model to the QEA
-approximation of the posterior. These Bayes factors present a
-relatively weak
+approximation of the posterior. These Bayes factors provide evidence
+for our
 argument that the constrained approximation of the posterior is
@@
 respect to each of the three posterior distributions, $\pi_i(\cdot|D,
-\mathcal{M}_i)$, $i = 0, 1, 2$.
-
-Firstly, in
+\mathcal{M}_i)$, $i = 0, 1, 2$. Firstly, in
 Figure~\ref{fig:chem_model_comp} (a), displaying the marginals for
@@
 $\pi_2$ is again peaked far away from both of the other distributions.
-This appears to demonstrate that the data regarding the slow variable
+This demonstrates that the data regarding the slow variable
 $S$ is only informative about $k_1 $ and $\hat{k}_4^{\text{CMA}}$, as
-predicted by the CMA multiscale approximation.
+predicted by the constrained multiscale approximation.
 
@@
 mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering}
-used by a particular cell to regulate the amount of a protein, $P$,
+used by a cell to regulate the amount of a protein, $P$,
 present. Proteins can bind in pairs to form a dimer $D$. The dimer can
-bind to the gene $G$ to a ``switched off'' state $G'$, in which the
+bind to the gene $G$ and convert it into a ``switched off'' state $G'$, in which the
 production of mRNA molecules $M$ is inhibited. The mRNA encodes for
@@
 
-We create trajectories of this system with the following parameters:
+We create trajectories of this system, using the Gillespie SSA, with the following parameters:
 \begin{eqnarray*}
@@
 protein. However, the T-cells will not necessarily differentiate
-between monomers and dimers of the protein. We replicate such a
+between monomers and dimers of the protein. We approximately replicate such a
 scenario, in a zero-observational noise setting, by assuming that we
@@
 space corresponding to the eight reaction rates in the system
-\eqref{eqn:GRN_reactions}. However half of those reactions are
+\eqref{eqn:GRN_reactions}. Half of those reactions are
 unobservable, and their effect on the posterior is only felt through
-their effect on the effective propensities $\hat{\alpha}_5$ and $\hat{\alpha}_7$.
-
-The priors for this problem were chosen to be fairly uninformative with respect to the likelihood function for each reaction rate.
-A list of the hyper-parameters corresponding to each Gamma prior can be found in Table~\ref{tab:grn_priors}.
+their effect on the effective propensities $\hat{\alpha}_5$ and
+$\hat{\alpha}_7$, and observations of the frequencies of these reactions.
+
+The priors for this problem were chosen to be fairly uninformative,
+and a list of the hyper-parameters corresponding to each Gamma prior can be found in Table~\ref{tab:grn_priors}.
 
@@
 posterior distribution are displayed in
-Figure~\ref{fig:GRN_posterior}. The posterior does not contain many
-interesting correlation structures, however several dimensions have
+Figure~\ref{fig:GRN_posterior}. The posterior does not appear from
+these joint marginals to contain many
+interesting correlation structures, although several dimensions have
 leptokurtic tails which are difficult for the standard ETAIS algorithm
-to sample from. The marginal densities also vary over very different
+to sample from. However as with the previous example, the density is
+concentrated close to a lower-dimensional manifold. The marginal densities also vary over very different
 scales, which could lead to problems with the algorithms which do not
@@
 
-The transport map setup has not changed from the previous example.
-However, due to the scaling of the weight we have had to be more careful about which samples we use to build our map.
-We require that the objective function $C(T)$, from Equation~\eqref{eqn:TETAIS_objective} is convex, which requires that the second derivative, given in Equation~\eqref{eqn:TETAIS_hessian}, is positive definite.
-This positive definite property depends on all weights in our sample being strictly positive, which is not always possible on a computer.
+The transport map setup has not changed from the previous
+example. However, due to the scaling of the weight we have had to be
+more careful about which samples we use to build our map. We require
+that the objective function $C(T)$, from
+Equation~\eqref{eqn:TETAIS_objective} is convex, which requires that
+the second derivative, given in Equation~\eqref{eqn:TETAIS_hessian},
+is positive definite. This positive definite property depends on all
+weights in our sample being strictly positive, which is not always
+possible on a computer, where very small positive numbers can be
+rounded down to absolute zero.
 For this reason we filter out any samples from the optimisation problem where the weight is a numeric zero.
@@
 
-The first thing to note is that the first ETAIS measurement happens
-after 2500 samples, whereas the first MH measurement occurs after 50
-samples. This is due to the much larger ensemble size required to
-sample in higher dimensions. We believe that an ensemble size of 2500
-is over-cautious in this example and we could have used a smaller
-ensemble size. We also believe that the required ensemble size to sample
+We believe that an ensemble size of 2500
+is over-cautious, despite the larger number of dimensions in this
+example, and that the required ensemble size to sample
 from these posteriors is reduced by the use of the transport map.
@@
 reduction of two orders of magnitude in the relative error over the
-same number of samples. A less positive result is that for this
-example the ETAIS-TRW algorithm has found a map which has prevented
-further convergence of the mean. This is likely because one dimension
-of the transport map is sending proposals onto the negative real line
-and so all samples are being assigned a zero weight. This is behaviour
-which was anticipated, and it is always suggested that appropriate
-preconditioners are used where the posterior is not absolutely
-continuous with respect to Lebesgue measure, as is the case when the
-posterior is only non-zero on the positive quadrant of $\mathbb{R}^d$
-as in these examples. The large increase in the effective sample size
+same number of samples.
+
+We note that in this 
+example the ETAIS-TRW algorithm does not perform as well as the ETAIS-RW
+algorithm. This is due to the lack of log preconditioner. The
+posterior has support only on the positive quadrant, and without the
+preconditioner, the transport map has to work hard to enforce
+this. With a relatively low-dimensional approximation being used for
+the transport map, this makes it more likely that the map produced
+may hinder sampling rather than improving it. This highlights the
+importance of using preconditioners with transport maps in order to
+ensure absolute continuity of the posterior distribution with respect
+to the pullback of the proposal distribution on reference space.
+
+In contrast, when we look at the convergence of these two algorithms
+with the log preconditioner, the transport map version exhibits more
+consistent and stable convergence, with some of the 32 repeats for the
+ETAIS-logRW algorithm producing rare large weights which hinder convergence.
+The large increase in the effective sample size
 observed between the ETAIS-logRW and ETAIS-logTRW algorithms is
 converted into an estimate which is twice as accurate after 1 million
-samples.
+samples. This demonstrates our claim that the transport map can
+produce stable ETAIS implementations with smaller ensemble sizes than without.
 
@@
 In this paper, we have investigated the use of transport maps to
-accelerate and stabilise ensemble importance sampling schemes. In
+accelerate and stabilise ensemble importance sampling schemes. We
+considered the ETAIS methodology here, but the approach used is
+applicable to all such methods. The idea is powerful since it allows
+adaptive importance sampling schemes to be efficient and stable for
+smaller ensemble sizes. This in turn suggests that the improvements in
+performance that adaptive importance sampling schemes can deliver
+could be made applicable to a bigger class of problems.
+
+In
 particular we looked at the application of these methods to problems
@@
 networks, where the posterior distributions which arise can be highly
-concentrated along curved manifolds in parameter space.
-Below we highlight some areas for future work.
+concentrated along curved manifolds in parameter space. Through these
+results we also demonstrated the improvement in accuracy of inference
+for multiscale problems when more sophisticated multiscale
+approximations are employed.
+Below we highlight two areas for future work.
 
@@
 
-Algorithm~\ref{alg:TransportETAIS2} allows us to decorrelate the
-dimensions of our random parameter on reference space, where we then
-can resample and map the resulting ensemble back onto target
-space. Since, on reference space, the dimensions are approximately uncorrelated, we
-are able to resample in each dimension separately. Resampling in a
+Armed with a high quality transport map, we can transform the target
+distribution close to a Gaussian reference measure, which is
+uncorrelated in all dimensions. This lack of correlations allows us to
+resample in each dimension separately. Resampling in a
 single dimension allows for optimisations in resampling code, and also
@@
 In one
-dimension the ETPF algorithm can be implemented very efficiently. As
+dimension, the ETPF algorithm can be implemented very efficiently. As
 described in~\cite{reich2013nonparametric}, the coupling matrix has
@@
 simplex algorithm to minimise the associated cost function, and faster
-than the MT algorithm\cite{cotter2015parallel}. Moreover, the task of
+than the MT algorithm\cite{cotter2015parallel}, which is itself a
+greedy approximation to the output of the ETPF resampler. Moreover, the task of
 running this resampler can then be parallelised over $d$
@@
 invertible map which approximates the true transport map between the
-reference measure and the target measure, however this true map is
+reference measure and the target measure. However, this true map is
 invariably itself not invertible. However, by considering embedding
@@
 exploration and sampling of the whole state space.
+
+\subsection{Adaptivity of ensemble sizes}
+At initialisation, our transport map approximation is equal to the
+identity, and the algorithm is essentially the standard ETAIS
+algorithm. In order to build a good transport map approximation, we
+require a good sample from the posterior distribution. Since our aim
+is to apply this methodology to challenging problems, this is itself a
+challenge. However, in the limit of the number of ensemble members
+going to infinity, this approach is consistent and stable. However,
+large ensemble sizes come with large computational costs.
+
+We propose that the transport map acceleration could be used in
+conjunction with an adaptive ensemble size mechanism. First we can
+initialise the algorithm with a large ensemble size, and the identity transport map. As
+we explore the density, the transport map quickly improves, making it
+possible to sample efficiently and stably with a smaller ensemble
+size. The problem then amounts to finding a reliable method to
+approximate the lowest stable ensemble size.
 

### 2019-10-29 arxiv/TETAIS.tex
arXiV

--- 2019-01-17 TETAIS.tex
@@
-\documentclass[final]{siamltex}
+\documentclass[12pt]{article}
+\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
 %test change
@@
 sampling (ETAIS) method, and allows for efficient sampling with
-smaller ensemble sizes.
+smaller ensemble sizes.\\
+{\bf Keywords: Importance sampling, ensemble, transport map,
+  multiscale, stochastic reaction networks, Bayesian inverse problems}
 \end{abstract}
@@
 \centering
-\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_1"}}
-\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_2"}}
-\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_posterior}}
+\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"R2_marginal_1"}}
+\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"R2_marginal_2"}}
+\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{R2_posterior}}
 \caption{Visualisation of the Rosenbrock density as given in Equation~\eqref{eqn:R2_repeat}.}
@@
 \centering
-\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
-\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
-\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
+\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"R2_transport_orig"}}\quad
+\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"R2_transport_ref"}}\quad
+\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{R2_transport_pullback}}
 \caption{The effect of the approximate transport map $\tilde{T}$ on a
@@
 \subfigure[Comparison of the two Transport ETAIS
-options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
-\subfigure[Comparison of Transport ETAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
+options.]{\includegraphics[width=0.45\textwidth]{R1_L2}}
+\subfigure[Comparison of Transport ETAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{R1_L2_all}}
 \caption{Convergence of algorithms; Transport M-H,
@@
 \centering
-\includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
+\includegraphics[width=0.9\textwidth]{"CMA_posterior"}
 \caption{CMA approximation \eqref{eq:Apos} of the posterior arising from observations of
@@
 \centering
-\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
-\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
+\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"CMA_L2_param_space"}}
+\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"CMA_L2_log_space"}}
 \caption{Convergence of eight different MCMC algorithms for the CMA
@@
 no log
-preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
+preconditioner.]{\includegraphics[width=0.8\textwidth]{"CMA_ref_space_MHTRW"}}
 \\
-\subfigure[Push forward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
+\subfigure[Push forward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"CMA_ref_space_MHlogTRW"}}
 \caption{Push forward of the target measure on reference space for the the CMA
@@
 \centering
-\includegraphics[width=0.9\textwidth]{"images/Applications/Diff_CMA_QEA"}
+\includegraphics[width=0.9\textwidth]{"Diff_CMA_QEA"}
 \caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
@@
 \centering
-\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/QEA_approx_marginal"}}
-\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_approx_marginal"}}
+\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"QEA_approx_marginal"}}
+\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"CMA_approx_marginal"}}
 \caption{Comparison of the approximate marginal densities for the
@@
 \centering
-\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/Applications/GRN_posterior"}}%
+\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"GRN_posterior"}}%
 \caption{Marginal densities of the CMA approximation of the posterior
@@
 \centering
-\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
-\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
+\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"GRN_convergence"}}
+\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"GRN_convergence_log"}}
 \caption{Convergence of eight different MCMC algorithms for the CMA

### 2019-10-29 cover_letter.tex
arXiV

@@
 
-\opening{Dear Prof. Higdon,}
+\opening{Dear Prof. Stuart and Prof. Challenor,}
 
@@
   multiscale stochastic reaction networks''. In this paper, we expand
-  on recent work adaptive importance samplers, by incorporating the
+  on recent work on adaptive importance samplers, by incorporating the
   use of transport maps to simplify challenging posterior

### 2019-11-01 Resubmission/response.tex
Resubmission folder and response.tex

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{amsmath,amssymb}
\usepackage{enumerate}
\usepackage{color}
\usepackage{amsthm,amsmath}
\usepackage{bbm}
\usepackage{mhchem}

%for stacking below a sum or max...
\usepackage{mathtools}

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Transport map-accelerated adaptive importance sampling}
\maketitle

Firstly, we would like to thank the referees, associate editor and editor for their careful consideration of our manuscript. We largely agree with the comments made, and have adjusted the paper to reflect the extremely helpful suggestions that have been given. In particular, both referees commented on the long length of the paper and the somewhat disjoint feel between the two main topics discussed within it. In order to address this primary point, we have made substantial reductions to the paper in the areas suggested by the referees. We have also reduced the emphasis on the multiscale stochastic reaction network material, aiming to use this as an example of a nice application for the methods outlined, as opposed to an equally-weighted component of the work. We hope that this has gone a long way to knitting the two parts of the work together in a way which tells a more coherent story.

Below we list the comments made by the referees and respond to them in turn, highlighting where we have made changes in the manuscript to address them. We also attach a version of the paper with the changes in text highlighted in red. Since a lot of the larger changes involved removal of sections of text and material, we have also attached a copy of the originally submitted paper with the removed sections highlighted in blue.

We would like to thank the referees and editors for their time in considering our manuscript for publication in SIAM JUQ.

\section*{Referee 1 comments}

\comment{Length and scope. Not including references, this manuscript is 32 pages long and seems to have two somewhat disjoint parts. The first part develops the transport map accelerated importance sampler and provides details of the algorithm. The second part seems to focus more on multiscale approximations of stochastic reaction networks, and the impact that these approximations have on the posterior distribution. While interesting, especially within a model discrepancy context, I feel that this part of the paper is tangential to the paper's main algorithmic contributions. With the length of this paper in mind, I would suggest shortening Sections 6 and 7 in order to focus more on numerical comparisons of the new algorithm and less on stochastic reaction networks. Details of the multiscale material could be moved into supplementary material or even into another manuscript.}{}

\comment{Regularization. In regards to the regularization terms in equation (3.2), the authors claim that "experimentation has shown that the choice $\beta=1$ is sufficient for most problems." I would like to see more discussion on this choice. With a sufficiently large number of samples, decreasing $\beta$ should result in a more accurate transport map, so why does a smaller value of $\beta$ not yield a more efficient algorithm? With $\beta=1$, is the map somehow able to capture enough of the posterior structure to enable effective importance sampling?}{}

\comment{Proposal distribution. The proposed transport map ETAIS algorithm is tested using local random walk proposals for each ensemble member. However, for accurate maps, an independent Gaussian proposal or some form of mixture proposal could potentially generate nearly iid samples of the target distribution. Is there a reason for sticking with the random walk proposals? I think adding a comparison of proposal types, even if only on the Rosenbrock toy problem, would help readers understand the importance of each part in the transport map ETAIS algorithm: the proposal, map, resampler, and ensemble size.}

\comment{Ensemble Size. How is the ensemble size chosen? The authors mention that a smaller ensemble can be used when the transport map is accurate, but are there any guidelines on how to choose the ensemble size given the map?}{}

\comment{Closing Remarks. I found the discussion (Section 8) ended rather abruptly. When reading the manuscript, I felt like I walked off the end of the manuscript. I think a sentence or two that ties things together would provide a better closing statement.}

\comment{I would prefer numbering all equations, but leave that to the author's discretion.}{}

\comment{Subscript in Algorithm 1 on line 3. Should the proposal be centered at $\theta_j$ or $\theta_i$}{}

\comment{Second equation on page 4. Should the sample be labeled with a $\hat{ }$?}{}

\comment{When discussing monotonicity of the map, it might be worth pointing out the monotone parameterizations introduced in "An Introduction to Sampling via Measure Transport" by Marzouk et al.}{}

\comment{In Section 4, the notation for proposed points on the reference space (marked with an apostrophe) and proposed points on the target space (marked with a $\hat{ }$) are different. It might be easier to follow if these were consistent, but that is up to the authors.}{}

\comment{Equation (6.2) $Y_j$ is used in the text but $X$ is used in the equation. Is that a typo?}{}

\comment{Sentence before equation (6.3). "variable" should be plural "variables"}{}

\section*{Referee 2 comments}

\comment{I do think that having these realistic applications is a strength—but I think the paper would be much easier to digest if the pieces were better woven together at the outset. On that note, the introduction takes a rather long time to get to the point of combining transport-ETIAS with partially observed stochastic reaction networks. Perhaps the necessary link here is “sloppiness,” mentioned on page 3?}{}

\comment{Also, it should be made more clear what exactly is new within the stochastic reaction networks portion of the paper}{}

\comment{In the numerical examples, what exactly is the “relative L2 error,” or $\delta_{L^2}$? I don’t see where this is defined, and I don’t know it as a standard MCMC or importance sampling diagnostic. Is it the relative $L_2$ error in the estimate of some posterior moment? If so, which moment? And relative to what? Similarly, can you define $\delta_{ESS}$ as a function of the ESS?}

\comment{Do the error versus \# samples plots (as in Figure 5.3, Figure 7.3, Figure 7.8) correspond to a single runs of MCMC? Or are they the error averaged over many independent chains (in
1
which case I think plotting all the runs in the same figure would be more illustrative). These are Monte Carlo schemes, so I expect the error in any given running estimate to be noisy.}{}

\comment{By \# of samples, in the plots for MCMC error, do you simply mean the number of MCMC iterations? Similarly, what is meant by \# of samples in the ETIAS error plots? Using “samples” without context is confusing, as there are many intermediate populations of samples at hand.}{}

\comment{To complement these plots, I think that a very useful and widely used diagnostic, applicable both to MCMC and ETIAS and their transport-accelerated versions, would be ESS per posterior density evaluation, or simply ESS per unit of computational time.}{}

\comment{I think there is some confusion about invertibility in various places.
In the discussion following Definition 3.2, it is stated that the space of all invertible maps might not contain an exact coupling. For absolutely continuous distributions with strictly positive densities, which seems to correspond to the cases here, I do not believe that this statement is correct. There is always a bijective map from one distribution to the other (and vice versa); indeed, there is an infinity of such maps. If one only considers maps with triangular structure, then there is a unique (up to sets of measure zero) invertible transformation. This is more or less the Knothe–Rosenblatt rearrangement; see one of Villani’s books, or [Bogachev, Kolesnikov, Medvedev 2005] for a more formal treatment.
Similarly, the statement in Section 8.2 about the true map being “invariably itself not in- vertible” seemed odd to me. If by the true map one means the KR rearrangement, then it is invertible. Maybe I am confused about what is meant by “true map?”}{}

\comment{ In comparing the transport MH and the transport-ETIAS algorithms, I think it is somehow not surprising that in many problems the latter performs better. But I also have a hunch that the difference may be due to “global” versus local proposals. Transport MH, as I understand the implementation here, is restricted to using local random-walk proposals in the reference space. But some of the larger efficiency gains in [46] resulted from the use of a Metropolis independence proposal on the reference space, in particular draws from a standard Gaussian, i.e., the entire reference measure. These global moves yielded excellent mixing when the map was sufficiently accurate, and were combined with local moves using a delayed rejection scheme. ETIAS is not exactly the same, but it does have something in common with a global proposal. More broadly, this reflects the close link between independence Metropolis algorithms and importance sampling. Some comments (please agree or disagree) on this front would be illuminating.}{}

\comment{What part of Section 6 is newly developed in this paper? Is the derivation of the exact likelihood in Section 6.1 new? Or is the approximation in Section 6.2 new? This relates to my broader comment above. It would be good to have a clearer picture of where exactly this paper’s key contributions lie, and even of their relative weight.}{}

\comment{Here are several suggestions for content that could be removed entirely or shortened, along with a few areas that could be expanded:
  \begin{itemize}
    \item The use of log-transformations to map strictly positive variables to the real line is a standard trick in MCMC, and has also been employed in previous work on transport. One could simply mention this and move forward. Thus I don’t see the need of including any results without the log preconditioner in problems where it is warranted. Log- preconditioning is well recognized as a good thing to do.
\item A lot of Section 3 could be abridged by appropriate referencing of [46].
\item Section 4 is quite terse, but the“second option” (Algorithm 3)—both proposing and resampling on the reference space—seems to be the cleaner algorithm and, if I understand the authors, the more effective one. So perhaps the first option can be mentioned in passing, and all the results can focus on the second? Is there any good reason to use Algorithm 2 instead of Algorithm 3?
\item The description of the ETIAS algorithm in Section 2 is very nice, but here an illustrative figure might help. \end{itemize}}{}

\comment{End of Section 2, “learn local covariances across the whole of the domain”: I believe I under- stand what is meant, but this awkwardly phrased}{}

\end{document}

### 2019-11-01 Resubmission/TPAIS.tex
Resubmission folder and response.tex

--- 2017-06-15 TPAIS.tex
@@
 \usepackage{tikz,hhline}
+\usetikzlibrary{positioning,automata}
+\usetikzlibrary{arrows,fit}
+\usetikzlibrary{decorations.pathmorphing}
 \usepackage{tabularx}
@@
 \newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\let\oldnl\nl% Store \nl in \oldnl
+\newcommand{\nonl}{\renewcommand{\nl}{\let\nl\oldnl}}% Remove line number for one line
 
@@
 
-In this section we give details on how \cite{parno2014transport} propose to couple the target,
+So far we have seen that the PAIS algorithm is efficient for low dimensional posterior sampling.
+PAIS is particularly useful if the posterior is multimodal, or inconveniently shaped such as the
+Rosenbrock density discussed in Chapter~\ref{sec:LPAIS}. However, we have had trouble when
+attempting to increase the dimension of the parameter space. This was seen when we attempted to
+recover the initial condition for the wave equation in Chapter~\ref{sec:PAIS}.
+
+In this Chapter we look at a method which will allow us to improve efficiency in our proposal distributions. The ideal sampling
+situation in MCMC is to explore an $d$-dimensional Gaussian distribution (or other standard distribution) using a near proportional
+proposal distribution.
+We therefore aim to transform an arbitrary distribution, $\mu_\theta$, in to a Gaussian distribution, which we can then sample from efficiently, and finally transform this sample back into a sample from $\mu_\theta$.
+In this way we hope to be able to sample from any distribution $\mu_\theta$ with a similar efficiency as we obtain when sampling from a Gaussian distribution.
+In particular for the PAIS algorithm, resampling schemes are more efficient in higher
+dimensions if the dimensions are uncorrelated. The work of Moselhy and
+Marzouk~\cite{el2012bayesian}, Parno and Marzouk~\cite{parno2014transport} and
+Parno~\cite{parno2015transport} provides a method of constructing such a map as part of an adaptive
+Metropolis-Hastings algorithm.
+
+%Given some constraints it is computationally tractable to adaptively construct invertible maps
+%during an MCMC simulation. These maps construct a continuous coupling between a target and
+%reference distribution, where the latter is chosen to be a standardised Gaussian. How close this reference
+%distribution is to Gaussian depends on three considerations; firstly the constraints applied. These
+%constraints, discussed later, ensure that the map exists, is unique, and is invertible. The
+%necessity for the map to be invertible, and preferrably easily invertible, restricts the space of
+%possible couplings to the point where it is unlikely that the exact map is included in our space of
+%feasible transference plans. Secondly, for our purposes, the map takes the form of a polynomial
+%chaos expansion which requires infinitely many terms to exactly represent the coupling. This is of course not possible on current hardware. Finally we
+%will be constructing this map using an optimisation problem where the objective function depends on
+%the sample path of our MCMC algorithm. This sample does not represent the target exactly until $N\rightarrow\infty$, and so the expectation is only calculated approximately.
+%
+%While the restrictions on the map are essential for well-posedness, the approximate map which
+%results from these constraints can be made arbitrarily accurate by increasing the number of basis
+%functions used in the expansion, and increasing the number of samples used in the optimisation
+%problem. In practice it appears that we need not aim to achieve a high accuracy solution. Excellent
+%results have been observed using monomial basis functions of total order 3, and we need a sample of
+%size of only $\mathcal{O}(10^3)$ to obtain a well converged map.
+
+In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
+distribution to the posterior distribution, the idea being that one could draw a moderately sized
+sample from the prior distribution and use this sample to approximate a map onto the target space.
+Once this map was known to the desired accuracy a larger sample from the prior could be used to
+investigate the posterior distribution. This
+methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
+algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
+distribution, the map transforms a sample from the posterior onto a reference space.
+The reference density is chosen to allow efficient proposals using a simple proposal
+distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.
+
+Proposing new states in this way allows us to make large steps around complex probability distributions.
+It is also feasible in this framework to assume that the reference density is close enough to a standard Gaussian that we can efficiently propose moves using a proposal distribution which is independent of the current state, e.g.
+choose $q(\theta) = \mathcal{N}(0,I_n)$.
+
+In this chapter we introduce the derivation of the map discussed, show how it can be adapted to work
+with a weighted sample, and that mixture distributions as used in the PAIS algorithm can more
+efficiently approximate the reference space and so lead to higher overall effective sample sizes.
+Further we show that the algorithm arising from the combination of these methods can lead to more
+efficient sampling than either can achieve on its own.
+
+\section{The Transport Map}
+
+In this section we outline the methodology in \cite{parno2014transport} for coupling the target,
 $\mu_{\theta}$, with the reference distribution, $\mu_r$.
-When sampling, ideally we would like to draw a sample directly from the target distribution.
-When we have the exact map, $T\in\mathcal{T}$, $T\colon \mathcal{X}\rightarrow\mathbb{R}^d$, we are able to map samples from the posterior distribution onto a reference density, $\pi_r$ which is a standard Gaussian.
-We can then draw a proposal sample from $\pi_r$, before mapping these samples back onto target space using $T^{-1}$.
+We also show how the map can be constructed using a weighted sample and hence how we can incorporate the map into the PAIS algorithm.
+
+\begin{dfn}[Transport Map $T$]
+	Given a probability distribution $\mu_\theta$ with support $\mathcal{X}$.
+	A transport map is a function $T\colon \mathcal{X} \rightarrow \mathbb{R}^d$ such that
+	\[
+		\mu_\theta(A) = \mu_r(T(A)) \quad \text{for any } A \subset \mathcal{X}.
+	\]
+\end{dfn}
+
+\begin{dfn}[Exact Transport Map $T$]
+	A transport map $T$ is exact if the {\it pullback} of the standard Gaussian density, $\phi(\cdot)$,
+	\begin{equation}\label{eq:pullback}
+		\tilde{\pi}(\theta) = \phi(T(\theta))|J_T(\theta)|,
+	\end{equation}
+	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
+	\[
+		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
+	\]
+\end{dfn}
+
+In the case that we have the exact map, $T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible maps, we are able to draw a sample from $\pi_r = \phi$, the density of the standardised Gaussian distribution, and map these samples back onto target space using $T^{-1}$.
 These proposed samples are distributed according to the target distribution.
-On target space, the proposal distribution is induced by the pullback of the standard Gaussian through the map,
-\begin{equation}\label{eq:pullback}
-	\tilde{\pi}(\theta) = \pi_r(T(\theta))|J_T(\theta)|,
-\end{equation}
-which is equal to the target density when $T$ is exact.
-Clearly we require $T$ to be monotonic and have continuous first derivatives, which means that our space of feasible transference plans, $\mathcal{T}^\uparrow$, may not include the exact map.
-Since it is unlikely that $T\in\mathcal{T}^\uparrow$, we are motivated to formulate an optimisation problem as discussed in Section~\ref{sec:intro_optim}.
+
+\begin{dfn}[Target and Reference Space]
+	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target measure $\mu_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference measure $\mu_r$.
+\end{dfn}
+
+On target space, the proposal density is induced by the pullback of $\phi$ through $T^{-1}$.
+Clearly the pullback only exists when $T$ is monotonic and has continuous first derivatives.
+Not all maps satisfy these conditions, so we define a smaller space of maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all feasible maps.
+The exact map $T$ is not necessarily in $\mathcal{T}^\uparrow$, so we are motivated to formulate an optimisation problem as discussed in Section~\ref{sec:intro_optim}.
 
 We attempt to find a deterministic coupling of two continuous probability distributions, $(\mu_\theta, \mu_r)$, such that $\mu_r = T\mu_\theta$ is satisfied and also that $T \in \mathcal{T}^\uparrow$.
-There may be infinitely many such couplings, so we look to find the coupling which minimises the difference between the proposal density and the target distribution, i.e.
+There may be infinitely many such couplings, so we look to find the coupling which minimises the KL-divergence between the density of $\mu_{\theta}$ and the pullback of the density of $\mu_{r}$, i.e.
 the distance between $\pi(\theta)$ and $\tilde{\pi}(\theta)$.
-In this case, we measure this distance using the Kullback-Leibler divergence.
-
-The map does not necessarily satisfy the monotonicity condition when we optimise the cost function,
+
+When we optimise the cost function
 \[
@@
 \]
-so it must be enforced in the design of each map component.
-To guarantee these properties we carefully choose the basis functions for each component, and we restrict the map to be lower triangular, i.e.
-$T \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$.
+numerically, it is not guaranteed that the resulting map will be invertible, so it must be enforced in the design of each map component.
+To ensure invertibility we restrict the map to be lower triangular, i.e.
+$\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$.
 This lower triangular map has the form,
@@
 \]
-where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We also assume that the target and reference probability densities are absolutely continuous on
-$\mathbb{R}^d$ which guarantees the maps existence.
+where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. We assume that the target and reference probability densities are absolutely continuous on
+$\mathbb{R}^d$. Under this formulation we are guaranteed a unique invertible map $\tilde{T}$ with the property that
+\[
+	\tilde{T}\mu_{\theta} \approx \mu_r.
+\]
+Relaxing the equality constraint to finding the approximate map $\tilde{T} \in \mathcal{T}^{\ltri}$ which optimises $C(T)$, allows us to formulate an efficient sampling method, in terms of the linear algebra required, and experimentation has shown that the pushforward of the target is close enough to Gaussian to achieve efficient sampling.
 
@@
 \]
-Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map, the divergence becomes
+This divergence results in some nice properties which we will explore in the following derivation.
+The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to it's relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
+\[
+	D_{KL}(p\|q) \geq D_H^2(p,q),
+\]
+which is a true metric between probability distributions $p$ and $q$.
+Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
 \[
@@
 \]
-We note the posterior density is constant in $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
-This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown, however this is exactly the scenario for which we would turn to MCMC methods for a solution.
+We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
+This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown.
+However this is exactly the scenario for which we would turn to MCMC methods for a solution.
 
@@
 \]
-which has a unique solution. However we would like to ensure that this solution is regular and monotonic. For this
-reason we regularise the problem further. The optimisation problem now takes the form
+which has a unique solution since the cost function is convex.
+
+We also include a regularisation term which is required for reasons which will become clear later. The optimisation problem now takes the form
 \begin{equation}\label{eq:gen_map_optim}
@@
 $\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
-close to the identity, and hence are likely to be monotonic.
+closer to the identity.
 
@@
 of the map in more detail.
-The lower lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density as well as the inverse of the map, $T^{-1}$.
-The Jacobian of $T$ is a lower triangular matrix,
-\[
-	DT(\theta) = \begin{bmatrix}
-		\partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta)\\
+The lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density, as well as the inverse of the map, $\tilde{T}^{-1}$.
+The Jacobian of $\tilde{T}$ is a lower triangular matrix,
+\[
+	J_T(\theta) = \begin{bmatrix}
+		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & \partial_{\theta_d} \tilde{T}_1(\theta)\\
 		\vdots & \ddots & \vdots \\
-		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
+		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
 	\end{bmatrix} = \begin{bmatrix}
-		\partial_{\theta_1} T_1(\theta) & \dots & 0\\
+		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & 0\\
 		\vdots & \ddots & \vdots \\
-		\partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta)
+		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
 	\end{bmatrix}
 \]
-since $\partial_{\theta_n} T_k(\theta) = 0$ for all $n > k$.
+since $\partial_{\theta_n} \tilde{T}_k(\theta) = 0$ for all $n > k$.
 This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
 \begin{equation}\label{eqn:separable_jacobian}
-	\log\left|J_T(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} T_i(\theta).
+	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta).
 \end{equation}
@@
 
-Inverting $T$ is simplified by the lower triangular structure in the sense that each dimension, $i$, we need only invert a univariate polynomial.
+%Inverting $T$ is simplified by the lower triangular structure in the sense that in each dimension, $i$, we need only invert a univariate polynomial.
 This simplification comes about because we invert each dimension of the map independently, beginning with $T_1$ which is a function of only $\theta_1$.
@@
 
+Inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map.
+The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, so we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$.
+This inversion tells us the value of $\theta_1$, which means the next component is again a univariate polynomial, $T_2(\theta_2; \theta_1)=r_2$.
+We can then perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.
+
 We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions.
-Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, although not necessarily orthogonal with respect to the target distribution.
+Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not necessarily orthogonal with respect to the target distribution.
 Each component of the map is defined as a multivariate polynomial expansion,
@@
 The parameter $\gamma_i$ is a vector of coefficients in $\mathbb{R}^{M_i}$. Each component of $\gamma_i$ corresponds to a basis function
-$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$, here $d$ is the dimension of
-$\theta$. These multi-indices are elements of the multi-index set $\mathcal{J}_i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
+$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in \mathbb{N}_0^d$.
+These multi-indices are elements of the multi-index set $\mathcal{J}_i$.
+A multi-index defines a product of univariate polynomials in $\theta_k$,
 \[
@@
 
-The cardinalities of the multi-index sets $M_i = \text{card}(\mathcal{J}_i)$ give the number of unknowns in our
+The cardinalities of the multi-index sets, $M_i = \text{card}(\mathcal{J}_i)$, give the number of unknowns in our
 optimisation problem, and so we would like to keep this number as small as possible. One option is
@@
 which is optimal in terms of the amount of information captured by the map about the target.
-The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = {i+p \choose p}$ which increases rapidly in $d$ and $p$.
+The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = {i+p \choose p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$.
 Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$.
-These index sets are discussed further
-in~\cite{parno2014transport}, here we stick with polynomials of total order $p$ since we work with low dimensional problems with the PAIS algorithm.
+These index sets are discussed
+in~\cite{parno2014transport}.
+Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}.
+Here we stick with polynomials of total order $p$ since we work with low dimensional problems with the PAIS algorithm.
 
@@
 
-We now return to Equation~\eqref{eq:gen_map_optim}, and show how this can be efficiently formulated
-for level-2 BLAS routines~\cite{choi1992scalapack}. The Monte Carlo estimator of a function is the
-sample average of the function evaluated at each sample, i.e.
-\begin{align*}
+We now discuss how we can evaluate Equation~\eqref{eq:gen_map_optim} using a sample from the target distribution.
+We first reformulate the expectation in the cost functional in terms of a MC estimator,
+\begin{align}
 	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
-			\beta\mathbb{E}(T(\theta)-\theta)^2 \\
+			\beta\mathbb{E}(T(\theta)-\theta)^2 \notag \\
 		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
-			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right].
-\end{align*}
-This cost function will create a map from $\pi$ to any reference density $\pi_r$.
-We now restrict ourselves to a Gaussian reference by explicitly using a standardised Gaussian density.
-The above expression simplifies to
+			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. \label{eqn:TM_full_cost}
+\end{align}
+Optimisation of this cost function results in a map from $\pi$ to some reference density $\pi_r$.
+By choosing the reference density to be a Gaussian density, we can simplify this expression greatly.
+Substitution of the Gaussian density into Equation~\eqref{eqn:TM_full_cost} leads to
 \begin{equation}\label{eq:gauss_map_optim}
@@
 Note that since we assume that the map is monotonic, the derivatives of each component are
-non-negative and so we don't need to take the absolute values in the log.
-In practise it is infeasible to enforce this condition across the whole parameter space.
-We instead enforce this condition by ensuring that the derivatives are non-negative at each sample point.
+positive and so this functional is always finite.
+In practice it is infeasible to enforce this condition across the whole parameter space.
+We instead enforce this condition by ensuring that the derivatives are positive at each sample point.
 This means that when we sample away from these support points while in reference space, it is possible to enter a region of space where the map is not monotonic.
@@
 	&\text{subject to} \quad \frac{\partial T_i}{\partial\theta_i}(\theta^{(k)};
-		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,n.
+		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,d.
 		\notag
 \end{align}
-The sum in Equation~\eqref{eq:map_defn} is really the inner
+The sum in Equation~\eqref{eq:map_defn} is an inner
 product between the vector of map coefficients, and the evaluations of the basis function at a
-particular $\theta^{(k)}$. This means that if we organise our basis evaluations into two matrices,
-$(F_i)_{k,\mathbf{j}} = \psi_\mathbf{j}(\theta^{(k)})$, and $(G_i)_{k,\mathbf{j}} =
-\frac{\partial\psi_\mathbf{j}}{\partial\theta_i}(\theta^{(k)})$, for all $\mathbf{j}
-\in \mathcal{J}_i^\text{TO}$, and $k = 1,\dots,K$, then we have that $T_i(\theta^{(k)}) = (F_i)_{k\cdot}\gamma_i$ and $\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}; \gamma_i) = (G_i)_{k\cdot}\gamma_i$, so \eqref{eq:gen_map_optim} becomes
+particular $\theta^{(k)}$. If we organise our basis evaluations into two matrices,
+\[
+	(F_i)_{k,\mathbf{j}} = \psi_\mathbf{j}(\theta^{(k)}), \quad \text{and} \quad (G_i)_{k,\mathbf{j}} =
+\frac{\partial\psi_\mathbf{j}}{\partial\theta_i}(\theta^{(k)}),
+\]
+for all $\mathbf{j}
+\in \mathcal{J}_i^\text{TO}$, and $k = 1,\dots,K$, then we have that
+\[
+	T_i(\theta^{(k)}) = (F_i)_{k\cdot}\gamma_i \quad \text{and} \quad \frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}; \gamma_i) = (G_i)_{k\cdot}\gamma_i,
+\]
+so \eqref{eq:gamma_map_optim} becomes
 \begin{align}\label{eq:blas_map_optim}
@@
 
-The regularisation term in Equation~\ref{eq:blas_map_optim} can be approximated using Parseval's identity,
+The regularisation term in Equation~\eqref{eq:blas_map_optim} can be approximated using Parseval's identity,
 \[
@@
 the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\mu_\theta$; however this
-approximation accomplishes the goal of efficiently regularising the optimisation problem.
+approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples.
+If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.
 
@@
 \begin{align}\label{eq:final_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2K}\gamma_i^\top F_i^\top F_i\gamma_i -
-		\frac{c^\top}{K}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
+	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2}\gamma_i^\top F_i^\top F_i\gamma_i -
+		c^\top\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
 	&\text{subject to} \quad G_i\gamma_i > 0. \notag
@@
 
-In the PAIS algorithm, we use weighted samples to approximate the posterior rather than equally weighted samples. Fortunately, the majority of the derivation of a weighted optimisation problem
-follows unchanged. We look at the importance sampling Monte Carlo estimate of $C(T)$, compared with
+In the PAIS algorithm, we use weighted samples to approximate the posterior rather than equally weighted samples.
+Fortunately, the majority of the derivation of this cost function follows unchanged.
+We look at the importance sampling Monte Carlo estimate of $C(T)$, compared with
 Equation~\eqref{eq:gauss_map_optim},
-\[
-	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^n \! \sum\limits_{k=1}^K
+\begin{equation}\label{eqn:TPAIS_objective}
+	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
 		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
 		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
-\]
+\end{equation}
 here $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
@@
 	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
-		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
+		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \frac{\beta}{K}\|\gamma_i-\iota\|^2, \\
 	&\text{subject to} \quad G_i\gamma_i > 0. \notag
 \end{align}
-We introduce a diagonal matrix $W = \text{diag}(w)$, where $w = (w_1,\dots,w_K)^\top$ into the log of the reference density, and replace the ones vector, $c$, with $w$, otherwise the cost function is unchanged.
-We should obtain the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation.
+We introduce a diagonal matrix $W = \text{diag}(w)$, where $w = (w_1,\dots,w_K)^\top$ into the log of the reference density, and replace the ones vector, $c$, with $w$.
+Analogously to the unweighted case, we include a $1/K$ to the regularisation term so that it has less influence as we obtain more samples.
+If we allowed this regularisation term to exert influence at all updates we would converge to a suboptimal map.
+We should converge to the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation.
 The weights are strictly
 positive resulting in a positive definite matrix $W$. This means that the Hessian of the objective
-function is still positive definite, and so the optimisation problem remains convex. Again this
-problem can be efficiently solved using the Newton optimisation algorithm. The Hessian takes the
+function is still positive definite, and so the optimisation problem remains convex.
+It is important in implementation to ensure that any weights which are numerically zero are dealt with to ensure this positive definiteness.
+
+This optimisation problem can be efficiently solved using the Newton optimisation algorithm. The Hessian takes the
 form
-\[
+\begin{equation}\label{eqn:TPAIS_hessian}
 	HC_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i + G_i^\top
 		W\text{diag}([G_i\gamma_i]^{-2})G_i\right] + \beta I,
-\]
+\end{equation}
 where $[G_i\gamma_i]^{-2}$ is to be taken element-wise, and $I$ is the $M_i\times M_i$
@@
 
-\section{T-PAIS}\label{sec:TPAIS}
-%Plug into PAIS
+\section[Transport map MCMC]{Using the transport map to propose states in MCMC algorithms}\label{sec:transport_algs}
+
+Given samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
+target measure to a reference measure. We now consider how to implement an MCMC algorithm which uses
+these maps to propose new states. These algorithms fit into the adaptive framework described in
+Section~\ref{sec:adapt} since we use the full history of the chain when updating the map, and this update changes the shape of our proposal
+distribution. Convergence of this adaptation is shown in~\cite{parno2014transport}.
+
+\subsection{Transport map MH algorithms}
+
+As introduced in~\cite{parno2014transport}, a MH algorithm using the transport map to propose new
+states can be designed, as given in Algorithm~\ref{alg:TransportMH}. These algorithms
+map a sample from the target onto the reference space, a new state is then efficiently
+proposed using a Gaussian kernel and is mapped back onto target space. This can
+result in rapid mixing, where step size and direction are now a function of the current particles location in state space, similar to the ideas of Riemann manifold Hamiltonian Monte Carlo~\cite{girolami2011riemann}.
+
+\begin{table}
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+Initialise state $\theta^{(1)} = \theta_0$.\;
+Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
+\For{$k \leftarrow 1, \dots, L-1$}{
+	Compute $r = \tilde{T}(\theta^{(k)}; \bar{\gamma}^{(k)})$.\;
+	Sample $r' \sim q_r(\cdot; r)$.\;
+	Invert $\theta' = \tilde{T}^{-1}(r'; \bar{\gamma}^{(k)})$.\;
+	Calculate:
+	\[
+		\alpha = 1 \wedge
+			\frac{\pi(\theta')}{\pi(\theta^{(k)})}\frac{q_r(r|r')|J_{\tilde{T}}(\theta^{(k)};
+			\bar{\gamma}^{(k)})|}{q_r(r'|r)|J_{\tilde{T}}(\theta';\bar{\gamma}^{(k)})|}.
+	\]
+
+	Sample $u \sim U[0,1]$.\;
+	Set $\theta^{(k+1)}$ to $\theta'$ with probability $\alpha$, otherwise $\theta^{(k+1)} =
+		\theta^{(k)}$.
+
+	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
+		\For{$i \leftarrow 1, \dots, n$}{
+			Solve \eqref{eq:final_map_optim} with $\{\theta^{(1)}, \dots, \theta^{(k+1)}\}$ and
+				update $\gamma_i^{(k+1)}$.\;
+		}
+	}{
+		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
+	}
+}
+\caption{MH algorithm with adaptive transport map~\cite{parno2014transport}\label{alg:TransportMH}}
+\end{algorithm}
+\end{table}
+
+Initially we begin with the identity map, $\iota$, while we build enough samples to produce an approximate
+map. For simplicity, we update the map at uniform intervals of $K_U$ iterations, given an initial
+burn-in phase. In order to satisfy the conditions for adaptive algorithms to be ergodic, discussed
+in Section~\ref{sec:adapt}, this adaptive phase must come to an end. It is sensible to stop
+adaptation once the map has sufficiently converged, e.g. when updates do not change the coefficients, $\gamma$,
+above some tolerance, or as suggested by~\cite{parno2014transport} when the variance of $D_\text{KL}(\pi\|\tilde{\pi})$ is sufficiently close to zero.
+
+Once the map is sufficiently converged, we can decide whether we would like to continue using a local proposal distribution, or whether to use a distribution which is independent of the current state.
+An independent proposal is more efficient
+if the reference space is close to Gaussian; however, if it is not then certain regions of the target
+space could be less likely to be visited than in a standard MH algorithm. A compromise can be found by randomly selecting which kernel to sample from, i.e. the proposal distribution
+\[
+	q(x, \cdot) = p\mathcal{N}(x, \Sigma)+(1-p)\mathcal{N}(0,\text{I}), \quad p \in [0, 1].
+\]
+
+\subsection{Transport map PAIS algorithms}
 
@@
 design a proposal scheme for the PAIS algorithm. In this case we have a choice in how to proceed; we
-can either use the reference space only to propose new values, or we can run the algorithm on reference space, including resampling, and only map samples onto the target space to calculate expectations.
+propose new samples on reference space and resample on target space, or we both propose and resample on reference space, mapping onto target space to output the samples.
 The first option allows us to reuse much of the framework
@@
 both the Transport MH algorithm, and the standard PAIS algorithm. The second option requires some
-restructuring but results in improved resampling, especially in higher dimensions.
+restructuring but results in improved performance from the resampler.
 
@@
 
-	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\;
+	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\label{algline:TPAIS_resample}\;
 	Invert $\theta^{(k+1)}_i = \tilde{T}^{-1}(r^*_i)$.\;
@@
 
-The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to the first except that rather than resampling in target space we resample in reference space.
+The second option, Algorithm~\ref{alg:TransportPAIS2}, is similar to the first except on Line~\ref{algline:TPAIS_resample} where rather than resampling in target space we resample in reference space.
 In reference space the dimensions are roughly uncorrelated, and the Gaussian marginals are easy to approximate with fewer ensemble members.
-This means that the resampling step will be more efficient in higher dimensions without needing to increase the ensemble size as much as we have needed to in the standard PAIS algorithm.
+This means that the resampling step will be more efficient in higher dimensions, which we discuss in Section~\ref{sec:TPAIS_higher_dim}.
+
+\section[Convergence of transport MCMC]{Convergence of the transport proposal based MCMC algorithms}
+
+In this section we study the convergence of the transport based proposal distributions which we have described in Section~\ref{sec:transport_algs}.
+We follow a similar strategy to that in Chapter~\ref{sec:PAIS}.
+We begin by finding optimal values for the scaling parameters by performing a series of simulations with differing scaling parameters.
+We then produce 1 million samples from the target distribution with the optimal scaling parameter and an ensemble size of $M=150$.
+We take 32 repeats of this optimal simulation and present the geometric average of the convergence rates.
+This process is then repeated for each of the algorithms we have discussed, Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2}.
+
+In this section we return to the Rosenbrock banana-shaped density which in Chapter~\ref{sec:LPAIS} we labelled $R_1$. This target density is
+\begin{equation}\label{eqn:R2_repeat}
+	\pi(\theta) = \frac{\sqrt{10}}{\pi}\exp\left\{ -(1 - \theta_1)^2 - 10(\theta_2 - \theta_1^2)^2 \right\}.
+\end{equation}
+A contour plot of the target density is given in Figure~\ref{fig:R2_posterior}.
+\begin{figure}
+\centering
+\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"images/R2_marginal_1"}}
+\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"images/R2_marginal_2"}}
+\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/R2_posterior}}
+\caption{Visualisation of the density of example $R_1$, as given in Equation~\eqref{eqn:R2_repeat}.}
+\label{fig:R2_posterior}
+\end{figure}
+
+\subsection{Implementation details}
+
+We now demonstrate some properties of the transport maps we will be using in our MCMC algorithms.
+We draw 1 million samples from the density in \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:transport_implementation} to build a transport map.
+We use this map to push forward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian.
+We then pull the sample back on to target space using the inverse map to check that our map is invertible and well behaved.
+
+For this example, we use an index set of total order 3 with monomial basis functions. This results in a map of the form
+\[
+	T(\theta_1, \theta_2) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \end{bmatrix},
+\]
+where
+\begin{align*}
+		T_1(\theta_1) &= \gamma_{1,1} + \gamma_{1,2}\theta_1 + \gamma_{1,3}\theta_1^2 + \gamma_{1,4}\theta_1^3, \\
+		T_2(\theta_1, \theta_2) &= \gamma_{2,1} + \gamma_{2,2}\theta_1 + \gamma_{2,3}\theta_1^2 + \gamma_{2,4}\theta_1^3
+					+ \gamma_{2,5}\theta_2 + \gamma_{2,6}\theta_1\theta_2 \\
+				 & \qquad \quad + \gamma_{2,7}\theta_1^2\theta_2 + \gamma_{2,8}\theta_2^2 + \gamma_{2,9}\theta_1\theta_2^2 +
+					 \gamma_{2,10}\theta_2^3.
+\end{align*}
+Clearly even with only basis functions of total order 3, we have a large number of unknowns in our optimisation problem, $\bar{\gamma} \in \mathbb{R}^{14}$.
+If we were to increase the dimension of $\theta$ further we would need to reduce the number of terms we include in the expansion by, for example, removing all the cross terms.
+This reduces the quality of our map but since we only require an approximate map we can afford to reduce the accuracy.
+
+\begin{figure}[htpb]
+\centering
+\subfigure[Original sample $\theta$ from RWMH algorithm.]{\includegraphics[width=0.31\textwidth]{"images/R2_transport_orig"}}\quad
+\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/R2_transport_ref"}}\quad
+\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/R2_transport_pullback}}
+\caption{Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
+\label{fig:R2_transport}
+\end{figure}
+
+Figure~\ref{fig:R2_transport} shows the efficiency of the transport map.
+Even though we have truncated the infinite expansion in the monomial basis down to 4 and 10 terms in respective dimensions, the push forward of the sample is still a unimodal distribution centred at the origin with standard deviation 1.
+As you move out into the tails of the reference density more non-Gaussian features form, but these are not too much of a problem when using a suitable proposal distribution.
+The pullback from reference space, in Figure~\ref{fig:R2_transport}, is an exact match of the original sample since we have not perturbed the sample in reference space.
+This inversion is well defined in the sampling region, although not necessarily outside.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.8\textwidth]{images/zero_jacobian}
+\caption{An example transport map which does not allow exploration of the full target distribution, and has a non-invertible point.}
+\label{fig:TM_zero_jacobian}
+\end{figure}
+
+If we consider a map built from polynomials of total order $p=2$, it is clear that the map $\tilde{T}$ does not map onto $\mathbb{R}^d$, but on to a subset of it.
+Figure~\ref{fig:TM_zero_jacobian} illustrates this point.
+We see that in this dimension, the $\tilde{T}^{-1}$ will not map any proposal $r > 0.8$ to a point in the target space.
+The point at which no inverse exists for a point $r_1$ can be made arbitrarily far out in the tails of the posterior distribution, but will be finite.
+The point at which the Jacobian is zero is also a non-invertible point since the logarithm in the pullback becomes infinite.
+This means that (1) we will never sample from the tail of the target distribution which lies beyond this turning point, and (2) all points in reference space $r < 0.8$ have the possibility of being mapped to two distinct points in target space.
+We can choose to truncate our parameter space at this turning point which allows our map to be bijective, however we will not be able to sample from the true posterior.
+For this reason it is best to chose an index set where the maximum order is an odd number.
+
+\subsection[Numerical results]{Numerical results for convergence of transport map based algorithms}
+
+We first find the optimal scaling parameters for the individual algorithms.
+This is done as before by optimising the effective sample size in the PAIS algorithm, and by tuning the relative $L^2$ error in the MH algorithm.
+There is currently no guidance on the best way of tuning the MH algorithm with transport map proposals although one might expect results similar to the standard MH results.
+As in the PAIS algorithm, the effective sample size might be the best option.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{lrrr}
+\toprule
+	Statistic \quad / \quad Algorithm & \ref{alg:TransportMH} & \ref{alg:TransportPAIS1} & \ref{alg:TransportPAIS2}  \\ \cmidrule(lr){1-4}
+	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
+	$\delta_{\text{ESS}}$				 & - & 1.0e-1 & 5.2e-1 \\ \cmidrule(lr){1-4}
+	Acc. rate							 & 0.23 & - & - \\
+	ESS ratio							 & - & 0.62 & 0.71 \\
+\bottomrule
+\end{tabular}
+\caption{Optimal scaling parameters for the transport map based algorithms applied to $R_1$.}
+\label{tab:R2_opt_scaling}
+\end{table}
+
+The optimal scaling parameters are given in Table~\ref{tab:R2_opt_scaling}.
+Here we see that the effective sample size is much lower than we see in the one-dimensional examples with the PAIS algorithms.
+However, in $R_1$ we are dealing with a much more complicated correlation structure, as well as a very slowly decaying tail in $\theta_2$.
+We have seen in Section~\ref{sec:LPAIS_R1} that the standard PAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.
+
+\begin{figure}[!h]
+\centering
+\subfigure[Comparison of the two Transport PAIS options.]{\includegraphics[width=0.45\textwidth]{images/R1_L2}}
+\subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/R1_L2_all}}
+\caption{Convergence of Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2} for $R_1$. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
+\label{fig:R2_l2_convergence}
+\end{figure}
+
+The convergence of the three algorithms is displayed in Figure~\ref{fig:R2_l2_convergence}.
+Figure (a) shows that the two variations of the transport based PAIS algorithms converge with similar rates.
+The second version, which performs the resampling stage in reference space rather than target space, has a slightly higher ESS, and is more stable than option (1).
+This version also has a property that we can exploit in Section~\ref{sec:TPAIS_higher_dim}.
+
+\section{Sampling in higher dimensions}\label{sec:TPAIS_higher_dim}
+
+Algorithm~\ref{alg:TransportPAIS2} allows us to decorrelate the dimensions of our random parameter on reference space, where we then can resample and map the resulting ensemble back onto target space.
+Since, on reference space, the dimensions are uncorrelated, we are able to resample in each dimension separately.
+Resampling in a single dimension allows for optimisations in resampling code, and also means that the resampler is not affected by the curse of dimensionality.
+
+If we can approximate the posterior well with our mixture and with the transport map, we should not be affected by the increase in dimension to the extent we have been with the standard PAIS-RW algorithm.
+In one dimension the ETPF algorithm can be implemented very efficiently.
+As described in~\cite{reich2013nonparametric}, the coupling matrix has all non-zero entries in a staircase pattern when the state space is ordered.
+We can exploit this knowledge to produce Algorithm~\ref{alg:ETPF_1d}.
+Which is much faster than using the simplex algorithm to minimise the associated cost function, and faster than the AMR algorithm.
+
+\begin{table}[!htpb]
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+Sort the states, $\{(w_i, x_i)\}_{i=1}^M$, into ascending order.\;
+Normalise the weights $p_i = w_i/\|w\|_1$.\;
+Set $y_i \leftarrow 0$ for all $i=1,\dots,M$.\;
+Set $c \leftarrow 0$\;
+\For{$i \leftarrow 1, \dots, M$}{
+	Set $t \leftarrow p_i$\;
+	\While{$j \leq M$ and $t > 0$}
+	{
+		Set $s \leftarrow \left(M^{-1}-c\right) \wedge t$\;
+		Increase $y_j$ by $M\times s\times x_i$.\;
+		Decrease $t$ by $s$.\;
+		Increase $c$ by $s$.\;
+		\If{$t>0$}
+		{
+			Increase $j$ by 1.\;
+			Set $c \leftarrow 0$.
+		}
+	}
+}
+Return $y$.\;
+\caption{ETPF algorithm in one dimension.\label{alg:ETPF_1d}}
+\end{algorithm}
+\end{table}
 
@@
 These algorithms simulate reactions between different species.
-A reaction has occured whenever the populations of the different species in the system change.
+A reaction has occurred whenever the populations of the different species in the system change.
 For example, a simple reaction where two chemicals from the same species join together to form a chemical from a different species, this is denoted
@@
 \]
-the likelihood of a reaction, $R_i$, occuring is determined by the \emph{propensity function} $\alpha_i(\mathbf{X}(t); \mathbf{k})$ and depends on the current state of the system and the reaction rates $\mathbf{k}$.
+the likelihood of a reaction, $R_i$, occurring is determined by the \emph{propensity function}
+\[
+	\alpha_i(\mathbf{X}(t); \mathbf{k}) = \lim\limits_{\Delta t\rightarrow 0} \! \frac{\mathbb{P}(\text{Reaction $i$ occurs in $[t, t+\Delta t)$})}{\Delta t}
+\]
+which depends on the current state of the system and the reaction rates $\mathbf{k}$.
 \end{dfn}
@@
 There is a propensity function associated with each of the $d$ possible reactions.
-The form of these functions is dependent on the the type of reaction which occurs.
+The form of these functions is dependent on the type of reaction which occurs.
 This notation can become cumbersome and so the dependence on the current state and reaction rates may be omitted.
@@
 
-The Gillespie algorithm simulates a Markov jump process about the chemical population state space.
-The algorithm assumes that reactions occur at times which follow an exponential distribution.
+The Gillespie algorithm is a stochastically exact algorithm which simulates a Markov jump process about the chemical population state space.
+The algorithm assumes that reactions occur in intervals which follow an exponential distribution.
 The rate associated with this exponential distribution is linked to the total system propensity $\alpha_0$.
 Which reaction occurs at each of these times depends on the distribution $\mathbf{p} = [p_1, \dots, p_d]^\top$.
-This assumption is the same as that in the formulation of the chemical master equation.
 
@@
 		Set $t_{j+1} \leftarrow t_j + \Delta t$.\;
-		Update populations $\mathbf{X}(t_{j+1}) = f_{r(t_{j+1})}(\mathbf{X}(t_j))$.\;
+		Update populations $\mathbf{X}(t_{j+1}) = \mathbf{X}(t_j) + \nu_{r(t_{j+1})}$.\;
 		$j \leftarrow j+1$.\;
@@
 The second, $\Delta t$, is from the exponential distribution with rate $\alpha_0$, and tells us how long until this reaction occurs.
-Finally we update the system's state, $\mathbf{X}(t_{j+1})$, taking into account which reaction has occured, using the transition function $f_{r(t_j)}$.
-These steps result in a stochastically exact trajectory from which we can determine the probability of being in any particular state.
+Finally we update the system's state, $\mathbf{X}(t_{j+1})$, according to which reaction has occurred, using the reaction state change vector, $\nu_i$.
+
+\begin{dfn}[State change vectors and the stoichiometric matrix]
+	The {\it state change vector} of a reaction $R_i$ is the vector, $\nu_i$, where the $j$th entry corresponds to the change made to the $j$th species by the reaction $R_i$.
+e.g.
+In a system with two chemicals $\mathbf{X} = [X_1, X_2]^\top$, and a reaction $R_1\colon X_1 \rightarrow X_2$, the state change vector of $R_1$ is $v_1 = [-1, 1]^\top$.
+The columns of the {\it stoichiometric matrix} are formed by the state change vectors for each reaction.
+\end{dfn}
 
@@
 
-In real world chemical systems, reactions will occur on more than one time scale.
-It is often possible to isolate which reactions are occuring more frequently (the \textit{fast reactions}) and those which are occuring less frequently (the \textit{slow reactions}).
-In many cases we are interested in the dynamics of the slow reactions, and the full simulation of the fast reactions can be considered a waste of computational resources.
+In many chemical systems, we can find variables which change on more than one time scale.
+These variables might be a single species, or a linear combination of different species.
+It is often possible to isolate which variables are changing more frequently (the \textit{fast variables}) and those which are changing less frequently (the \textit{slow variables}).
+In many cases we are interested in the dynamics of the slow variables, and the full simulation of the fast variables can be considered a waste of computational resources.
+In real world experiments it can also be difficult to measure the quantities of the fast variables given the relatively short times the system stays in a single state.
 In this thesis we consider two separate simplifying approaches for these multiscale systems.
@@
 \begin{dfn}[Quasi-equilibrium assumption]
-	The quasi-equilibrium assumption (QEA) is the assumption that the fast reactions converge in distribution on a timescale which is negligible with respect to the rate of occurence of the slow reactions.
+	The quasi-equilibrium assumption (QEA) is the assumption that the fast variables converge in distribution on a timescale which is negligible with respect to the rate of change of the slow variables.
 \end{dfn}
 
-This assumption allows us to approximate the dynamics of the slowly changing quantities in the system by assuming that the fast quantities are in equilibrium with respect to the fast reactions in isolation.
+This assumption allows us to approximate the dynamics of the slowly changing quantities in the system by assuming that the fast quantities are in equilibrium with respect to the fast variables in isolation, i.e.
+the slow rates $\rightarrow 0$.
 We also consider the constrained multiscale approach (CMA) described in \cite{cotter2011constrained,cotter2016constrained}.
-The CMA accounts for the differences in the invariant distribution of the fast species which are due to the occurence of slow reactions.
+The CMA is similar to the QEA except that it takes into account changes in the slow variables which also change the fast variables.
+We briefly define the CMA as is given in \cite{cotter2016constrained}.
+
+\begin{dfn}[Constrained Projector]\label{dfn:constrained_projector}
+	Given a partition of the state space with $N$ variables into $N_s$ slow variables $\mathbf{S} = [S_1,\dots, S_{N_s}]$ and $N_f$ fast variables $\mathbf{F} = [F_1,\dots, F_{N_f}]$.
+A constrained projector $\mathcal{P}_\mathbf{S}\colon\mathbb{N}^N_0\rightarrow\mathbb{N}^N_0$ maps the values of the slow variables to a given value $\mathbf{S}$, while preserving the values of the fast variables.
+\[
+	\mathcal{P}_\mathbf{S}([\hat{\mathbf{S}}, \hat{\mathbf{F}}]) = [\mathbf{S}, \hat{\mathbf{F}}], \quad \forall \, [\mathbf{S},\mathbf{F}] \in \mathbb{N}_0^N.
+\]
+\end{dfn}
+
+\begin{dfn}[Constrained Stoichiometric Projector]
+	Given the partition in Definition~\ref{dfn:constrained_projector}, the constrained stoichiometric projector $\mathcal{P}\colon\mathbb{N}_0^{N}\rightarrow\mathbb{N}_0^{N}$ maps all non-zero values of the slow variables to zero.
+\[
+	\mathcal{P}([\mathbf{S}, \mathbf{F}]) = [\mathbf{0}, \mathbf{F}], \quad \forall \, [\mathbf{S}, \mathbf{F}] \in \mathbb{N}_0^{N}.
+\]
+\end{dfn}
+
+\begin{dfn}[Constrained subsystem]
+	The constrained subsystem is defined for a chemical reaction system, such as those we have been discussing so far, and is the system which arises from partitioning this chemical reaction system as in Definition~\ref{dfn:constrained_projector} and applying the constrained projector, $\mathcal{P}_\mathbf{S}$ to the state vector $\mathbf{X} = [\mathbf{S}, \mathbf{F}]$ after every new reaction, or equivalently applying the constrained stoichiometric projector, $\mathcal{P}$, to the change in state resulting from each reaction.
+When a stoichiometric vector becomes null by application of $\mathcal{P}$, that reaction can be removed from the system.
+If a reaction with one of the new stoichiometries would result in a species with a negative population, the propensity of this reaction is set to zero.
+\end{dfn}
+
+The CMA accounts for the differences in the invariant distribution of the fast species which are due to the occurrence of slow reactions.
 When the fast and slow subsystems operate on very different timescales, the CMA and QEA produce near identical results, however when the systems are not so clearly separated the CMA provides superior accuracy.
-Other simulation methods such as the $\tau$-leap method introduce significant bias into the approximations.
 
@@
 The data, $D$, is a complete trajectory of a molecular system which includes the times of each reaction as well as which reaction occurs.
-In practise it might not be obvious which reaction has occured just by looking at the changes in chemical populations, however we ignore this problem for now.
+In practice it might not be obvious which reaction has occurred just by looking at the changes in chemical populations, however we ignore this problem for now.
 
@@
 
-From this formulation, we see that the random variables $(\Delta t_j, r(t_j))$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are independent of each other.
-This independence means that we can group events together by what state the system was in when the event happened.
+From this formulation, we see that the random variables $(\Delta t_j, r(t_j))$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are Markovian.
+This conditional independence means that we can group events together by what state the system was in when the event happened.
 We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
@@
 
-Similarly, we can define the reactions which occured when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_d(\mathbf{Y})]^\top$ where
+Similarly, we can define the reactions which occurred when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_d(\mathbf{Y})]^\top$ where
 \[
@@
 \[
-	\ell(\mathbf{R},~ \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
+	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
 \]
@@
 
-\section{Numerical Examples}\label{sec:num}
+\section{Mono-molecular examples}
+
+In this section we look at two examples of chemical systems to demonstrate the effectiveness of the Bayesian approach.
+These two systems will only contain mono-molecular reactions, i.e.
+reactions which convert one species into another species at a particular rate.
+As shown in~\cite{anderson2010product} these systems have a Poisson invariant distributions on the populations.
+In the limit of infinite reactions, a mono-molecular reaction system with $n$ species will follow the distribution
+\begin{equation}\label{eqn:mono_stat_dist}
+	\mathbb{P}(X_1=x_1, \ldots, X_n=x_n) = \left(\prod_{i=1}^n \! \frac{\bar{c}_i^{x_i}}{x_i!}\right)\exp\left(-\sum_i^n \! \bar{c}_i\right),
+\end{equation}
+where $x_i \in \mathbb{N}$ and $\bar{c}_i$ are the complex balances or the concentration of chemical $X_i$ according to the solution of a steady state approximation of the system.
+We make use of this distribution to formulate posteriors for the chemical reaction rates using the constrained multiscale approach, as in~\cite{cotter2016constrained}.
+
+%%%%%
+
+\subsection{A Multiscale Chemical System}\label{sec:chem_multiscale}
 
@@
 parameter regime such that the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
-much much more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
+more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
 chemical species are involved in fast reactions. However, the quantity
-$S = X_1 + X_2$ is conserved by both of the fast reactions,
+$X = X_1 + X_2$, corresponding to an artificial species $S$, is conserved by both of the fast reactions,
 and as such, this is the slowly changing quantity in this system.
+When recording our data, we are only able to take readings of the reactions which change the slow variable $S$. 
 The effective dynamics of $S$ can be represented as follows
@@
 application of the two approaches mentioned in Section~\ref{sec:SSAs}.
-Under the QEA, the value of $\mathbb{E}\left[k_4X_2|S=s\right]$ is approximated by finding the steady state of the ODE representing the fast subsystem of
-reactions:
+Under the QEA, the value of $\mathbb{E}\left[k_4X_2|S=s\right]$ is found by computing the invariant distribution of the fast subsystem of reactions~\cite{jahnke2007solving},
 \[
@@
 \]
-we arrive at
+which tells us that $X_2$ is distributed according to Bin$(S, k_3(k_3+k_4)^{-1})$ and so
 \[
@@
 Our observations are uninformative about the reaction rates $k_2, k_3$, and $k_4$, as there
-is a submanifold $\mathcal{M} \subset \mathcal{X}$ along which the effective rate
+is a sub-manifold $\mathcal{M} \subset \mathcal{X}$ along which the effective rate
 $\hat{k}_4$ is invariant, leading to a highly ill-posed inverse
@@
 We can use the posterior distribution derived in Section~\ref{sec:chem_stoch_posterior} to specify the posterior distribution for each of the three scenarios we are interested in.
-We are interested in the posterior distribution when we have a complete data set telling informing us of every reaction which occurs as well as perfect readings of the numbers of reactants.
-We are also interested in the two simplifying assumptions we saw in the previous section.
+We would like to find out how the posterior distributions we obtain from the two models, the QEA and CMA, differ to the posterior for a complete data set.
 
@@
 % choices made in transport map
-For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=4$ as the basis functions, i.e.
+For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=3$ as the basis functions, i.e.
 \[
@@
 \]
-Since $\mathcal{X}$ is four-dimensional, this yeilds a total of 125 map coefficients across the four index sets.
-This number can be reduced by using smaller index sets as discussed in~\cite{parno2014transport}.
 
@@
 
-%*** numerics.
+% numerics.
 \subsubsection{Convergence Analysis for the Constrained Approach to a Multiscale Chemical Reaction Model}\label{sec:chem_conv}
@@
 Convergence is shown for the constrained approach to the multiscale system, i.e.
-to the posterior in Equation~\eqref{eqn:chem_QEA_posterior} with the effective degredation rate, $\hat{k}_4^{\text{CMA}}$, from Equation~\eqref{eqn:chem_CMA_rate}.
+to the posterior in Equation~\eqref{eqn:chem_QEA_posterior} with the effective degradation rate, $\hat{k}_4^{\text{CMA}}$, from Equation~\eqref{eqn:chem_CMA_rate}.
 
@@
 Another option is to increase the total order of our index set.
-For these numerics we have chosen $p=4$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
+For these numerics we have chosen $p=3$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
 
@@
 Since the two posteriors have been approximated using an MCMC sample, there is a significant amount of noise, particularly in the tails of the distributions We can see that the differences in the marginals for $k_1, k_2,$ and $k_3$ are relatively small and the differences are largely positive.
-This means that the constrained approach is slightly more informative about the values of these first three parameters.
-However the maginal for $k_4$ varies by a significant amount and is more negative.
-This means that the mean estimates for $k_4$ vary by a large amount, and we should be more confident in the solution given by the QEA.
+These positive differences tell us that the constrained approach claims to be more informative about the values of these first three parameters.
+However the marginal for $k_4$ varies by a significant amount and is more negative.
+This means that the mean estimates for $k_4$ differ by a large amount, the posterior for the QEA is most peaked.
 
@@
 We now consider how we should interpret the information given by these models.
-The QEA assumption tells us that we cannot observe the parameters $k_2$, $k_3$ and $k_4$ indendently from the reduced data set, but we can observe the quantity $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$.
+The QEA assumption tells us that we cannot observe the parameters $k_2$, $k_3$ and $k_4$ independently from the reduced data set, but we can observe the quantity $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$.
 Similarly, the constrained approach tells us that we are able to observe the quantity $\hat{k}_4^{\text{CMA}} = k_2k_4/(k_2+k_3+k_4)$.
-To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model.
+To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most accurate and informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model, with all reactions observed.
 
@@
 \[
-	Z_k \approx \frac{1}{NM}\sum_{i=1}^N\sum_{j=1}^M w_j^{(i)}(\mathcal{M}_k),
+	Z_k \approx \hat{Z}_k\frac{1}{NM}\sum_{i=1}^N\sum_{j=1}^M w_j^{(i)}(\mathcal{M}_k),
 \]
 where $w_j^{(i)}(\mathcal{M}_k)$ is the weight under model $\mathcal{M}_k$ corresponding to the $j$-th ensemble member on the $i$-th iteration.
-Hence, $B_{1,2} = Z_1/Z_2$.
+Hence, $B_{1,2} \approx \hat{Z}_1/\hat{Z}_2$.
 We can compare more than two models in this way by selecting the model with the largest marginal distribution as the best model.
@@
 Again the constrained model is not substantially more attractive than the QEA model when we consider the Bayes factor $B_{1,2} = 1.96 < 3.2$.
-The Bayes factor $B_{0,2} = 4.1 > 3.2$ does tell us that we should significantly prefer the full model to the QEA model.
+The Bayes factor $B_{0,2} = 4.1 > 3.2$ does tell us that we should significantly prefer the full model to the QEA approximation of the posterior.
 These Bayes factors present a weak argument that that the constrained model provides us with a better description of the data than the QEA model.
-We can also present this information graphically, which might provide us with a greater justification for preferring the constrained model.
+We can also present this information graphically, which might provide us with a greater justification for preferring the constrained approximation of the posterior.
 
@@
 
-\section{Conclusions}\label{sec:conc}
+\subsection{Gene Regulatory Networks}\label{sec:grn}
+In this example, we look at a model of a gene regulatory network (GRN) mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering} used by a particular cell to regulate the amount of a protein, $P$, present.
+In this model, we again have reactions occurring at particular rates, $k_i$, between four different species.
+Here there is a special species, $G$, representing the gene responsible for building the protein, of which there is only one at a time, and this species can either be turned on or off.
+When the gene is off we denote it as $G'$.
+\begin{align}
+	P+P \xleftrightarrows[k_2]{k_1} D \quad & \quad G+D \xleftrightarrows[k_4]{k_3} G' \nonumber\\
+	G \xrightarrow{k_5} G+M \quad & \quad M\xrightarrow{k_6} M+P  \label{eqn:GRN_reactions} \\
+	P \xrightarrow{k_7} \emptyset \quad & \quad M \xrightarrow{k_8} \emptyset \nonumber
+\end{align}
+The eight reactions are given in Equation~\eqref{eqn:GRN_reactions}.
+In addition to the gene, $G$, and protein, $P$, the system also involves molecules of mRNA, $M$, and dimers, $D$, which are two proteins joined together.
+
+In this example protein is created by a molecule of mRNA at a rate $k_6$.
+The amount of protein in the cell is then governed by the rate of decay of $P$, and the number of mRNA molecules in the cell creating $P$.
+In turn the number of mRNA molecules is controlled by its own decay rate and whether or not the gene $G$, which produces the mRNA, is active.
+The gene can be turned on and off depending on how many dimers are present in the system, which is related to how many protein molecules there are.
+This completes a feedback cycle whereby if the protein population gets too high then the gene will turn off and the rate of new protein formation will decrease, causing the level to drop and the gene to turn back on.
+Figure~\ref{fig:GRN_cycle} shows what happens to the molecules in the system as the gene turns on and off.
+
+\begin{figure}
+	\centering
+	\begin{tikzpicture}
+
+		\node at (0, 1.1) {Increase in molecules};
+
+		\node[inner sep=.5cm] (AA) at (-3,.6) {$M$};
+		\node[inner sep=.5cm] (BB) at (0,.6) {$P$};
+		\node[inner sep=.5cm] (CC) at (3,.6) {$D$};
+		\node[inner sep=.5cm] (DD) at (5,0) {$G'$};
+
+		\node at (0, -1.1) {Decrease in molecules};
+
+		\node[inner sep=.5cm] (EE) at (3,-.6) {$M$};
+		\node[inner sep=.5cm] (FF) at (0,-.6) {$P$};
+		\node[inner sep=.5cm] (GG) at (-3,-.6) {$D$};
+		\node[inner sep=.5cm] (HH) at (-5,0) {$G$};
+
+		\draw [arrows={-latex}] (HH) -- (AA);
+		\draw [arrows={-latex}] (AA) -- (BB);
+		\draw [arrows={-latex}] (BB) -- (CC);
+		\draw [arrows={-latex}] (CC) -- (DD);
+		\draw [arrows={-latex},decorate, decoration={zigzag, segment length=4, amplitude=.9,post=lineto, post length=2pt}] (DD) -- (EE);
+		\draw [arrows={-latex},decorate, decoration={zigzag, segment length=4, amplitude=.9,post=lineto, post length=2pt}] (EE) -- (FF);
+		\draw [arrows={-latex},decorate, decoration={zigzag, segment length=4, amplitude=.9,post=lineto, post length=2pt}] (FF) -- (GG);
+		\draw [arrows={-latex},decorate, decoration={zigzag, segment length=4, amplitude=.9,post=lineto, post length=2pt}] (GG) -- (HH);
+	\end{tikzpicture}
+	\caption{Diagram of the effects of molecule populations in a gene regulatory network.}
+	\label{fig:GRN_cycle}
+\end{figure}
+
+%As in Section~\ref{sec:chem_multiscale} we use the constrained multiscale approach to model the system.
+Due to the small scale of the system and the speed of the reactions, we are only able to measure the total amount of protein, $T$, and mRNA, $M$, while the number of dimers and whether or not the gene is switched on must be considered to be missing data.
+For this scenario, the likelihood function is intractable and so we must use a multiscale approximation of the system.
+Here we choose the constrained multiscale approximation due to its success in the previous example.
+
+The formulation here is slightly more complex.
+We define a fast subsystem which acts on the fast variables $\mathcal{F} = [P, D, G]$ and incorporates reactions $R_1$-$R_4$.
+For the true parameter values we have chosen, 90\% of reactions which occur in a particular time frame are one of these four reactions.
+\[
+	\text{Fast subsystem:} \quad P+P\xleftrightarrows[k_2]{k_1} D, \quad G+D\xleftrightarrows[k_4]{k_3} G'.
+\]
+We also define the slow variables, $T = P+2D+2(1-G)$ and $M$ which are held constant by these fast reactions.
+
+When we observe a cell, we cannot distinguish between the fast species and so don't know which of the first four reactions are happening.
+Due to this, we observe how many of reactions $R_5$-$R_8$ occur for each combination of the two slow variables $\mathcal{S} = [T, M]$.
+Since we don't observe $P$ and $G$ for every reaction, the propensities must be altered to take this into account,
+\begin{align}
+	\hat{\alpha}_5 = k_5\mathbb{E}[G|T=t],\quad&\quad \alpha_6 = k_6M, \notag\\
+	\hat{\alpha}_7 = k_7\mathbb{E}[P|T=t], \quad&\quad \alpha_8 = k_8M.\label{eqn:grn_propensities}
+\end{align}
+
+To calculate these expectations, we note that the fast subsystem is a deficiency zero system~\cite{anderson2010product}, and is also reversible. i.e. The system has four complexes, 
+\[
+	C_1 = P+P, \quad C_2 = D, \quad C_3 = G+D \quad \text{and} \quad  C_4 = G',
+\]
+there are two linkage classes (or two sets of reversible reactions), and the space spanned by the stoichiometries is two dimensional,
+\[
+	\text{span}\left\{\pm[-2, 1, 0, 0]^\top, \pm[0, -1, 0, -1]^\top\right\},
+\]
+hence the deficiency is $d = 4 - 2 - 2 = 0$. This means that, by \cite{anderson2010product}, the conditional probabilities have the closed form equation,
+\[
+	\mathbb{P}[\mathcal{F}|T=t] \propto \text{Poisson}(P_t;\bar{p})\text{Poisson}(D_t;\bar{d})\text{Poisson}(G_t;\bar{g})\text{Poisson}(1-G_t;1-\bar{g}),
+\]
+from which we can calculate the expected values of $P_t$, $D_t$, and $G_t$.
+The fast variables $P_t$, $D_t$ and $G_t$ are constrained to the line defined by $T=P+2D+2(1-G)=t$.
+The concentrations $\bar{p}$, $\bar{d}$ and $\bar{g}$, must be calculated as the solution to the nonlinear system produced by examining the steady state behaviour of the fast subsystem,
+\begin{align}
+	T &= p + 2d + 2(1-g), \notag\\
+	\frac{\text{d}p}{\text{d}t} &= -2k_1p^2+2k_2d = 0, \notag\\
+	\frac{\text{d}d}{\text{d}t} &= k_1p^2 - k_2d - k_3gd + k_4(1-g) = 0, \notag\\
+	\frac{\text{d}g}{\text{d}t} &= -k_3gd + k_4(1-g) = 0, \label{eqn:grn_nonlinear_system}
+\end{align}
+subject to $p \in [0, T]$, $g \in [0, 1]$, and $0 \leq d \leq \text{floor}(T/2)$.
+
+With these effective propensities, given in Equation~\eqref{eqn:grn_propensities}, we can calculate the posterior distribution as given in Equation~\eqref{eqn:chem_posterior}, again assigning Gamma priors to the eight reaction rates.
+We note that the reaction rates for the fast variables do not appear in the posterior density explicitly which only uses the propensities $\alpha_5$-$\alpha_8$, however they appear through the solution to the nonlinear system in Equation~\eqref{eqn:grn_nonlinear_system}.
+
+\subsubsection{Target Distribution}
+
+As in the previous chemical system example we use the target distribution as defined in Equation~\eqref{eqn:chem_posterior}.
+The application here is slightly more involved as discussed previously in Section~\ref{sec:grn}.
+We have an eight dimensional parameter space, with one reaction rate corresponding to one of eight reactions.
+The first four reactions, $R_1$-$R_4$, are combined into a fast subsystem for which we do not observe the time or identifier for which reaction has fired.
+We instead use the constrained multiscale approach to approximate what has happened in this subsystem between occurrences of the slow reactions firing (reactions $R_5$-$R_8$).
+For this reason, we apply the form in Equation~\eqref{eqn:chem_posterior} to reactions $R_5$-$R_8$, and allow the rates for reactions $R_1$-$R_4$ to implicitly influence the posterior density through the effective propensities, $\hat{\alpha}_5$ and $\hat{\alpha}_7$.
+
+As we mentioned earlier, for this problem we only consider the CMA's approximation to the system dynamics.
+
+The priors for this problem were chosen to be fairly uninformative with respect to the likelihood function for each reaction rate.
+A list of the hyper-parameters corresponding to each Gamma prior can be found in Table~\ref{tab:grn_priors}.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{|l|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|}
+	\hline
+	Dimension $i$ & 1 & 2& 3& 4 & 5 & 6 & 7 & 8 \\ \hline
+	$\alpha_i$ & 2 & 100 & 100 & 3 & 3 & 3 & 2 & 2\\ \hline
+	$\beta_i$ &50 & 0.02 & 1 & 1 & 0.6 & 1 & 50 & 50 \\ \hline
+\end{tabular}
+\caption{Hyper parameters for the Gamma priors on each of the reaction rates in the GRN example in Section~\ref{sec:grn}.}
+\label{tab:grn_priors}
+\end{table}
+
+\begin{figure}[htb]
+\centering
+\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/GRN_posterior"}}%
+\caption{Marginal densities of the posterior distribution for the GRN example in Section~\ref{sec:grn}.}
+\label{fig:GRN_posterior}
+\end{figure}
+
+The one- and two-dimensional marginal distributions for the full posterior distribution are displayed in Figure~\ref{fig:GRN_posterior}.
+The posterior does not contain many interesting correlation structures, however several dimensions have leptokurtic tails which are difficult for the standard PAIS algorithm to sample from.
+The marginal densities also vary over very different scales, which might require us to use a variant of the LPAIS algorithm.
+
+\subsubsection{Implementation}
+
+%% proposal methods
+%For the posterior distribution in \eqref{eqn:chem_QEA_posterior}, we consider several proposal methods.
+First we implement both the PAIS and MH algorithms with a Gaussian proposal distribution.
+In the case of the PAIS algorithm, this is a Gaussian mixture proposal distribution.
+The proposal distribution uses a covariance matrix which has been constructed using the sample covariance of a sample produced using a standard RWMH algorithm.
+We also compare the PAIS and MH algorithms when using a transport map proposal distribution.
+This proposal method was discussed in detail in Chapter~\ref{sec:TransportPAIS}.
+%
+%
+%Figure~\ref{fig:chem_coupling} shows how we define a bijective map, $T'$, between parameter space $\mathcal{X}$ and a reference space $R$.
+In practice we cannot ensure that the approximate map $\tilde{T}'$ is uniquely invertible over the whole of $R$ and so $\tilde{T}'$ is not truly bijective.
+This leads to problems for our strictly positive state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in $R$ do not necessarily map back on to $\mathcal{X}$.
+This motivates the use of a third intermediate space.
+When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
+MH-RW (RWMH) and MH-TRW, as well as PAIS-RW and PAIS-TRW.
+%
+%We also consider how these algorithms perform when they are applied to the log of the reaction rates.
+We choose to use this log transformation since it converts our strictly positive parameter space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset \mathbb{R}^d$.
+This allows us to define $T$ between two subsets of $\mathbb{R}^d$, which means that even if $\tilde{T}$ is not uniquely invertible in some region of $R$, all possibilities are valid proposals.
+As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
+MH-logRW and MH-logTRW, with, PAIS-logRW and PAIS-logTRW.
+We apply the eight RW and logRW proposal distributions, with and without transport map acceleration, which were discussed in Section~\ref{sec:chem_multiscale} to this posterior.
+Again the intermediate log space is required so that our transport map is a function $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$, rather than a map from $\mathbb{R}_+^d$ to $\mathbb{R}^d$.
+
+%% choices made in transport map
+%For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=4$ as the basis functions, i.e.
+%\[
+%	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
+%\]
+%and
+%\[
+%	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
+%\]
+%Since $\mathcal{X}$ is four-dimensional, this yeilds a total of 125 map coefficients across the four index sets.
+This number can be reduced by using smaller index sets as discussed in~\cite{parno2014transport}.
+The transport map setup has not changed from the previous example.
+However, due to the scaling of the weight we have had to be more careful about which samples we use to build our map.
+We require that the objective function $C(T)$, from Equation~\eqref{eqn:TPAIS_objective} is convex, which requires that the second derivative, given in Equation~\eqref{eqn:TPAIS_hessian}, is positive definite.
+This positive definite property depends on all weights in our sample being strictly positive, which is not always possible on a computer.
+For this reason we filter out any samples from the optimisation problem where the weight is a numeric zero.
+This does not affect the validity of the map since a weight zero would not contribute to the expectation, and we do not require (and could not enforce) an exact map.
+
+In all eight algorithms we adapt the proposal variances during the burn-in phase of the algorithms to avoid behaviour like that seen in Example~\ref{sec:ex_W1} where we needed a huge ensemble size to approximate the posterior with our mixture of isotropic Gaussian kernels.
+
+%% choice of resampler?
+%As with the mixture model in Section~\ref{sec:app_mixture}, we will use the AMR resampler with an ensemble size of $M=500$.
+This increase in ensemble size from Chapter~\ref{sec:PAIS} compensates for the increase in parameter dimension.
+We again use the AMR resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size again compensates for the increase in dimension (from four to eight parameters.)
+
+%% analysis method
+%To measure the convergence of the sampling methods in this section, we will compare the convergence of the mean of each parameter.
+We approximate $\mathbb{E}(\mathbf{k})$ using 2.4 billion samples from the MH-RW algorithm.
+As in the previous chemical reaction example, we measure convergence of the algorithms only through the convergence of the mean, for computational ease.
+We use the sample Mahalanobis distance, with `true' covariance and `true' mean built from a sample of size 2.4 billion using the MH-RW algorithm.
+
+\subsubsection{Convergence results}
+
+The scaling parameter selection is done by optimising the acceptance rate for the MH algorithms, and optimising the effective sample size for the PAIS algorithms.
+The optimal parameters are given in Table~\ref{tab:grn_scaling_parameters}.
+
+\begin{table}[!h]
+	\centering
+	\begin{tabular}{lrrrrrrrr}
+	\toprule
+		 & \multicolumn{4}{c}{MH} & \multicolumn{4}{c}{PAIS} \\ \cmidrule(lr){2-5} \cmidrule(lr){6-9}
+		& RW & logRW & TRW & log-TRW & RW & logRW & TRW & log-TRW \\ \midrule
+		$\beta_{\%}^*$	 	& 1.0e-1 & 3.9e-1 & 1.0e-0 &6.0e-1 & - & - & - & - \\
+		$\beta_{\text{ESS}}^*$	& - 	        & -           & -            & -            & 1.0e-0 & 1.3e-0 & 8.0e-1 & 5.0e-1 \\
+		ESS/$M$		 		& - 	        & -           & -            & -            & 3.5e-4 & 4.7e-2 & 4.8e-2 & 1.6e-1  \\
+	\bottomrule
+	\end{tabular}
+	\caption{Scaling parameters for the sampling algorithms applied to the GRN example in Section~\ref{sec:grn}.}
+	\label{tab:grn_scaling_parameters}
+\end{table}
+
+From the effective sample sizes shown in Table~\ref{tab:grn_scaling_parameters} we can see the improvement to the efficiency of the algorithm both by proposing on the log space, and by transforming the parameter space into something closer to Gaussian.
+
+Due to the numerical cost of calculating the full relative $L^2$ errors for this posterior, we quantify the convergence speeds of these algorithms using the sample Mahalanobis distance between the sample mean and the `true' mean.
+This convergence analysis is shown in Figure~\ref{fig:grn_L2}.
+
+\begin{figure}[!htb]
+\centering
+\subfigure[Sampling algorithms on $\mathcal{X}$.]{\includegraphics[width=0.495\textwidth]{"images/GRN_convergence"}}
+\subfigure[Sampling algorithms on $\mathcal{Y}$.]{\includegraphics[width=0.495\textwidth]{"images/GRN_convergence_log"}}
+\caption{Convergence of the GRN example described in Section~\ref{sec:grn}.}
+\label{fig:grn_L2}
+\end{figure}
+
+The first thing to note is that the first PAIS measurement happens after 2500 samples, whereas the first MH measurement occurs after 50 samples.
+This is due to the much larger ensemble size required to sample in higher dimensions.
+We believe that an ensemble size of 2500 is over-cautious in this example and we could have used a smaller ensemble size.
+We also think that the required ensemble size to sample from these posteriors is reduced by the use of the transport map.
+This is due to the way the tail behaviour is improved by the transformation.
+
+The second obvious feature of these convergence plots is that the PAIS algorithms outperform the MH algorithms by a large margin - roughly a reduction of two orders of magnitude in the relative error over the same number of samples.
+A less positive result is that for this example the PAIS-TRW algorithm has found a map which has prevented further convergence of the mean.
+This is likely because one dimension of the transport map is sending proposals onto the negative real line and so all samples are being assigned a zero weight.
+This is behaviour which was anticipated, and it is always suggested that the transport map is constructed as a map $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$ rather than $T\colon\mathcal{X}\rightarrow\mathbb{R}^d$ where $\mathcal{X}\subset\mathbb{R}^d$.
+The large increase in the effective sample size observed between the PAIS-logRW and PAIS-logTRW algorithms is converted into an estimate which is twice as accurate after 1 million samples.
+
+A similar pattern is seen in the MH algorithms, where the MH-TRW performs the worst, and the MH-logTRW algorithm performs the best, although only marginally.
+
+\section{Conclusions}
+
+In this chapter we have demonstrated the process of applying the PAIS algorithm to realistic modelling situations.
+We saw how the standard PAIS algorithm can perform well for multimodal distributions in Section~\ref{sec:app_mixture}, in this case we only needed a RW proposal distribution, but it is possible to use the LPAIS algorithm or transport map based proposals if the tails of the posterior are not well behaved.
+
+We have also seen how the transport map converts posteriors which are badly scaled, have complicated correlation structures and possibly have non-Gaussian tails, into posteriors which we can sample effectively.
+This was demonstrated for two chemical reaction networks in Sections~\ref{sec:chem_multiscale} and \ref{sec:grn}
+
+These algorithms, when compared with the standard MC toolbox, have consistently performed the best, in the last GRN example realising estimates with more than two orders of magnitude better accuracy.
 

### 2019-11-01 Resubmission/TPAIS_Sept_17.tex
Resubmission folder and response.tex

--- 2019-11-01 Resubmission/TPAIS.tex
@@
 
+\section{Deterministic Resampling Methods}\label{sec:intro_optim}
+
+Optimal transportation arises in many industries. It deals with the efficient allocation
+of a particular resource to someone who wishes to use this resource. For example a fresh food
+supplier may have many depots around the country, each with a certain number of vans capable of
+carrying a fixed amount of product a particular distance in a day. The supplier may need to supply
+many supermarkets up and down the country with their produce. The supplier naturally wishes to
+perform this task quickly, and using as little fuel as possible. To achieve this goal the supplier
+can choose which depot or depots deliver to which supermarket, and whether a van stops at zero, one
+or more supermarkets on its route. The efficient solution of this problem can make a huge difference
+to the profits of the supplier.
+
+The supplier knows the initial distribution, $\mu$, where we have all of the fresh produce,
+and they also know where we want it to go, the target distribution, $\eta$. We need to find a map which converts
+the initial distribution, $\mu$, into the target, $\eta$. These
+maps are known as transport maps. We can also contrain the problem to one where we must use as little fuel as possible, the map satisfying this condition is known as the optimal transport map.
+
+\begin{dfn}[Transport Map~\cite{reich2013ensemble}]\label{def:transport_map}
+Given a random variable $X \sim \mu$, $X \in \mathcal{X}$, and a second probability measure $\nu$, a
+measurable function $T\colon \mathcal{X}\rightarrow\mathcal{X}$ is a transport map if the induced
+random variable $Y = T(X)$ satisfies
+\[
+	\int_\mathcal{X} \! f(y)\nu(\text{d}y) = \int_\mathcal{X} \! f(T(x))\mu(\text{d}x),
+\]
+for all functions $f\colon \mathcal{X} \rightarrow \mathbb{R}$.
+\end{dfn}
+
+Mathematically, the formulation of this assignment problem is quite abstract. The original
+formulation by Monge in 1781~\cite{monge1781memoire}, is to find the transport map, $T(\cdot)$,
+which minimises the integral
+\[
+	\int_\mathcal{X} \! c(s, T(s)) \, \text{d}\mu(s),
+\]
+subject to the constraint that $T(\mu) = \nu$. We are able to choose what we consider to be
+expensive by choosing the cost function $c(x,y)$ which is non-negative for all states
+$x,y\in\mathcal{X}$.
+
+A more widely used formulation of the Monge problem was presented by Kantorovitch in
+1942~\cite{kantorovitch1958translocation}. This formulation considers a space of joint probability
+distributions known as the space of transference plans, $\Pi(\mu, \nu)$.
+
+\begin{dfn}[Coupling, Transference Plan~\cite{reich2013ensemble}]\label{def:coupling}
+Given two distributions $\mu$ and $\nu$ on a space $\mathcal{X}$, a coupling of these distributions
+is a pair $Z = (X,Y)$ such that $X\sim\mu$ and $Y\sim\nu$. The joint measure $\eta$, where
+$Z\sim\eta$, on $\mathcal{Z}=\mathcal{X}\times\mathcal{X}$ is known as the transference plan for
+this coupling.
+\end{dfn}
+
+The optimisation problem then becomes
+\[
+	\int_{\mathcal{X}\times\mathcal{X}} \! c(x, y) \, \text{d}\gamma(x, y),
+\]
+subject to $\gamma(x,y) \in \Pi(\mu, \nu)$. Kantorovitch also developed the field of
+linear programming to help deal with solutions to this
+problem~\cite{kantorovich1960mathematical,kantorovitch1958translocation}.
+
+When the cost function is $c(x,y) = \|x - y\|^2$, if $\mu$ is absolutely continuous with respect to
+Lebesgue measure, and both $\mu$ and $\nu$ have finite second order moments, then it is guaranteed
+that there is a unique optimal transference plan, between $\mu$ and $\nu$~\cite{knott1984optimal}.
+
+%%%%%%%%%%%%
+
+\subsection{Coupling Probability Distributions}\label{sec:coupling}
+
+In this section we return to the idea of coupling probability distributions, and motivate their use
+in resampling algorithms. We defined a coupling in Definition~\ref{def:coupling} as a pair of random
+variables $Z = (X,Y)$, the joint probability distribution for this pair being a quantity of
+interest. In this section we assume that the marginal distributions of both $X$ and $Y$ are known,
+later on we relax this so that one of these distributions is not known but can be approximated by a
+sample. All distributions used in the later chapters will be absolutely continuous on $\mathcal{X} =
+\mathbb{R}^n$, so we have $\mu(\text{d}x) = \pi_0(x)\text{d}x$ and $\nu(\text{d}x) =
+\pi(x)\text{d}x$. Having absolutely continuous marginals does not mean that the coupling itself is
+absolutely continuous.
+
+\begin{figure}
+\centering
+\subfigure[$\rho = 0.1$]{\includegraphics[width=0.45\textwidth]{images/coupling_indep}}
+\subfigure[$\rho = 0.95$]{\includegraphics[width=0.45\textwidth]{images/coupling_correlated}}
+\caption{Couplings with different correlations.}
+\label{fig:cts_couplings}
+\end{figure}
+
+We now look at couplings between continuous random variables. As an example we consider coupling two
+univariate Gaussian distributions, $X_1 \sim \mathcal{N}(\mu_1, \sigma^2_1)$ and $X_2 \sim
+\mathcal{N}(\mu_2, \sigma^2_2)$. We know that the transference plan $\gamma(x_1, x_2)$ will be the
+bivariate Gaussian distribution,
+\[
+	Z = (X_1, X_2), \quad Z \sim \mathcal{N}\left(\begin{pmatrix}\mu_1\\\mu_2\end{pmatrix},
+		\begin{pmatrix} \sigma^2_1 & \rho\sigma_1\sigma_2 \\ \rho\sigma_1\sigma_2 & \sigma^2_2
+		\end{pmatrix}\right).
+\]
+The form of this bivariate Gaussian is restricted by the marginal distributions to have this
+covariance matrix, where we can choose the correlation $-1 \leq \rho \leq 1$. This unknown
+correlation means that there are infinitely many couplings between $X_1$ and $X_2$, some examples
+are given in Figure~\ref{fig:cts_couplings}, which leads us back to the idea of optimising the
+transport map to maximise some criterion. We might want to maximise the correlation between the two
+random variables. If we consider the conditional distribution $X_1|X_2 \sim \mathcal{N}(x_1; \mu,
+\sigma^2)$, where
+\[
+	\mu = \mu_1 - \frac{\rho\sigma_1\sigma_2}{\sigma^2_2}(\mu_2 - x_2) \quad \text{and} \quad \sigma^2 =
+		\sigma^2_1 - \frac{(\rho\sigma_1\sigma_2)^2}{\sigma_2^2} = (1-\rho^2)\sigma_1^2,
+\]
+we notice that as $\rho\rightarrow 1$ the variance $\sigma^2 \rightarrow 0$, and we obtain a
+deterministic coupling with transport map $T\colon x_1 \mapsto \mu_1 -
+\frac{\sigma_1}{\sigma^2}(\mu_2 - x_2)$. If we set $\mathcal{N}(\mu_2, \sigma^2_2) = \mathcal{N}(0,
+1)$ we see that this coupling matches our standard map for transforming standard Gaussian random
+variables into general Gaussian random variables.
+
+Here we consider an optimal coupling to be one which maximises the correlation between $X_1$ and
+$X_2$. Maximising the correlation is equivalent to minimising the expectation of $(X_2-X_1)^2$,
+\[
+	\mathbb{E}[(X_2-X_1)^2] = \mathbb{E}[X_1^2]+\mathbb{E}[X_2^2] - 2\mathbb{E}[X_1]\mathbb{E}[X_2]-
+		2\text{cov}(X_1, X_2).
+\]
+The coupling which maximises the correlation is therefore the solution to the Monge-Kantorovich
+problem with cost function
+\[
+	c(X_1,X_2) = \|X_1-X_2\|^2.
+\]
+This problem always has a solution since for any marginals densities $\pi_X(x)$ and $\pi_Y(y)$, the
+trivial coupling $\pi_Z(x,y) = \pi_X(x)\pi_Y(y)$ exists.
+
+We now look at couplings between two discrete random variables. Unlike in the continuous case where
+the coupling had support over all of $\mathcal{X}\times\mathcal{X}$, for couplings between discrete
+random variables this is rarely the case.
+
+Consider two discrete R.V.s
+\[
+	Y_1 \sim \mu_{Y_1} \quad \text{and} \quad Y_2 \sim \mu_{Y_2},
+\]
+defined over a parameter space $\mathcal{Y} = \left\{a_1, \dots, a_M\right\}$, $a_i \in \mathbb{R}$.
+Assign the mass functions
+\[
+	\mathbb{P}(Y_1 = a_i) = \frac{1}{M} \quad \text{and} \quad \mathbb{P}(Y_2 = a_i) = w_i,
+\]
+where $w_i \geq 0$ and $\sum_i \! w_i = 1$. All possible couplings $Z = (Y_1, Y_2)$ can be described
+by the matrix $T \in \mathbb{R}^{M\times M}_+$, such that
+\begin{equation}\label{eq:coupling_conditions}
+\mathbb{P}(Y_2) = MT\mathbb{P}(Y_1), \quad \text{subject to} \quad \sum\limits_{i=1}^M \! t_{ij} =
+\frac{1}{M}, \quad \sum\limits_{j=1}^M \! t_{ij} = w_i,
+\end{equation}
+where $[T]_{ij} = t_{ij}$.
+
+We present two special cases, first the map with zero correlation, or trivial map,
+\[
+T^t = \frac{1}{M}\begin{bmatrix} w_1 & \dots & w_1 \\ \vdots & & \vdots \\ w_M & \dots & w_M
+\end{bmatrix}
+\]
+It is clear that, after renormalisation, along any row we obtain a uniform distribution,
+$\mathbb{P}(Y_1)$, and along any column we obtain the distribution $\sum_{i=1}^M
+\delta_{a_i}(\cdot)w_i$, $\mathbb{P}(Y_2)$. Hence our conditionals are equal to our marginals and
+there is no correlation.
+
+The second case is more interesting. The coupling which maximises the correlation between the two
+marginals, $T^*$, is the solution to the optimisation problem
+\[
+	T^*  = \arg\min_T \! \sum\limits_{i,j=1}^M \! t_{ij}(a_i - a_j)^2,
+\]
+subject to the row and column constraints on $T$ in Equation~\eqref{eq:coupling_conditions}~\cite{reich2015probabilistic}.
+
+We now look at a simple example of a discrete coupling. Given a state space $\mathcal{Y} = \{a_1,
+\dots, a_M\}$ where $M=5$, and two probability distributions defined over $\mathcal{Y}$,
+\[
+\mathbb{P}(Y_1 = a_i) = \frac{1}{M} \quad \text{and} \quad \mathbb{P}(Y_2 = \mathbf{a}) = (0.1,
+0.25, 0.05, 0.3, 0.3)^\top.
+\]
+If we assume that the states $a_i$ are sorted so that $a_i < a_{i+1}$ for all $i=1,\dots,M-1$, then
+the optimal coupling is
+\[
+	T^* = \begin{bmatrix} 0.1 & 0      & 0    & 0    & 0 \\
+					       0.1 & 0.15 & 0    & 0    & 0 \\
+					       0    & 0.05 & 0    & 0    & 0 \\
+					       0    & 0      & 0.2 & 0.1 & 0 \\
+					       0    & 0      & 0    & 0.1 & 0.2
+		\end{bmatrix}.
+\]
+For an ordered state space we can construct this coupling following this staircase pattern by
+starting in the top left and filling first the row then the column heading towards the bottom right
+corner. We notice that this staircase structure corresponds to the shape seen in the highly
+correlated coupling in Figure~\ref{fig:cts_couplings} (b). The $2M-1$ constraints mean that we have
+at most $2M-1$ non-zero entries, so the support of the joint distribution is much smaller than the
+full space.
+
+The product $MT^*$ results in a column stochastic matrix $P$. If we want to draw an evenly weighted
+sample, $\{s_i\}_{i=1}^5$, which obeys the probability law $\mathbb{P}(Y_2)$, we can use the columns
+of $P$ as probability distributions,
+\[
+	s_1 \sim \frac{1}{2}\delta_{a_1}(\cdot)+\frac{1}{2}\delta_{a_2}(\cdot), \quad s_2 \sim
+		\frac{3}{4}\delta_{a_2}(\cdot)+\frac{1}{4}\delta_{a_3}(\cdot), \quad s_4 \sim
+		\frac{1}{2}\delta_{a_4}(\cdot)+\frac{1}{2}\delta_{a_5}(\cdot),
+\]
+and deterministically $s_3 = a_4$, $s_5 = a_5$. What we have done here is used an optimal coupling
+between weighted and unweighted random variables to resample from the weighted distribution.
+
+%%%%%%%%%%%%
+
+\subsection{Resampling with Filters}
+\label{sec:OT}
+
+Resampling is a method which is distinct from most of the Statistical literature in that rather than
+sampling from a known parametric distribution which is usually the aim, we instead attempt to sample
+from empirical distributions. These distributions often arise from a sample, $\{(w_i,
+\theta_i)\}_{i=1}^n$, in the form,
+\begin{equation}\label{eq:empirical_dist}
+\pi(\cdot) = \sum\limits_{i=1}^n \! w_i\delta_{\theta_i}(\cdot), \quad \text{where}
+\quad\sum\limits_{i=1}^n \! w_i = 1.
+\end{equation}
+The weights, $w_i$ may be equal to $1/n$ or they may all be different.
+\begin{dfn}[Resampling]\label{def:resampling}
+Given a discrete empirical distribution of the type in Equation~\eqref{eq:empirical_dist}. Resampling provides
+us a new discrete distribution of the form
+\[
+	\tilde{\pi}(\cdot) = \frac{1}{n}\sum\limits_{i=1}^n \! n_i\delta_{\theta_i}(\cdot), \quad
+		\text{where} \quad n_i \in \{0,\dots,n\}, \quad \sum\limits_{i=1}^n \! n_i = n.
+\]
+The density $\tilde{\pi}$ is in some way close to the original density $\pi$.
+\end{dfn}
+In Definition~\ref{def:resampling} the condition on the sum of the $n_i$ restricts $\tilde{\pi}$ to
+be composed of the same number of samples as $\pi$. This is not necessary although as mentioned in
+Section~\ref{sec:gmh} one can run into trouble when this is not the case. These problems arise
+because the empirical density $\pi$ is usually approximating some continuous distribution, $f$, and
+over reliance on one approximation will cause the tails of the $f$ to be underestimated. If a sample
+is required, one uses $n_i$ lots of $\theta_i$, otherwise statistics can be computed directly from
+$\tilde{\pi}$.
+
+%%%%%%
+
+\subsubsection{Bootstrap Resampling}
+
+One of the first resamplers capable of accomplishing the feat in Definition~\ref{def:resampling} was
+the bootstrap resampler (BSR)~\cite{efron1992bootstrap}.
+
+\begin{table}[htpb]
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+Obtain sample, $\{x_i\}_{i=1}^n$ from some distribution.\;
+Construct
+\[
+	\hat{F} = \sum\limits_{i=1}^n \! n_i\delta_{x_i}(\cdot), \quad n_i = \sum\limits_{k=1}^n \!
+		\text{I}(x_i = x_k).
+\]
+
+Sample with replacement $X^* = [x^*_1,\dots,x_n^*]^\top \sim \hat{F}$. i.e. $X^* \sim
+\text{Multi}(1, \frac{1}{n}[n_1,\dots,n_n]^\top)$.\;
+\caption{The Bootstrap Resampler~\cite{efron1992bootstrap}.\label{alg:bs_resampler}}
+\end{algorithm}
+\end{table}
+
+Algorithm~\ref{alg:bs_resampler} presents the BSR when we have an evenly weighted sample, this can
+just as easily be extended to a weighted sample, $\{(w_i, x_i)\}_{i=1}^n$. In this case the
+probability vector in the Multinomial distribution becomes $\tilde{\mathbf{w}} = [\tilde{w}_1,
+\dots, \tilde{w}_n]^\top$ where $\tilde{w}_i = \sum_{k=1}^n \! w_i\text{I}(x_i = x_k)$.
+
+%%%%%%
+
+\subsubsection{The Bootstrap Particle Filter}
+
+The data assimilation community is concerned with predicting the state of time dependent systems
+given noisy observations. This type of problem is known as the filtering problem. Usually there is a
+model for the system although it may be expensive to evaluate. We assume that we have a prediction,
+$x(t)=[x_1(t),\dots,x_N(t)]^\top$, from the model calculated using observations, $D_i(t-1)$. We also
+observe some data at time $t$, $D_i(t)$. An application of Bayes' formula allows us to combine these
+two pieces of information to refine the prediction.
+\begin{dfn}[Particle Filter]
+A particle filter is an algorithm which solves the filtering problem.
+\end{dfn}
+
+\begin{table}[htpb]
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+Define a system model,
+\[
+	x_i(t) = f_{t-1}(x_i(t-1), \xi_i(t-1)), \quad \text{where} \quad f_t\colon
+		\mathbb{R}^n\times\mathbb{R}^m\rightarrow\mathbb{R}^n, \quad \xi_i(t-1) \sim \mathcal{N}(0,
+\sigma^2).
+\]
+
+\nonl \textbf{Predict}\;
+Sample $\xi_i(t-1) \sim \mathcal{N}(0, \sigma^2)$.\;
+Calculate $x_i^*(t) = f_{t-1}(x_i(t-1), \xi_i(t-1))$.\;
+\nonl\;
+\nonl \textbf{Update}\;
+For each particle calculate
+\[
+	w_i = \frac{\ell(D_i(t)| x_i^*(t))}{\sum_{k=1}^N \! \ell(D_k(t)| x_k^*(t))}.
+\]
+
+Resample $x(t) \sim \text{Multi}(1, w/\|w\|_1)$, where $w = (w_1,\dots,w_N)^\top$.
+\caption{The Bootstrap Particle Filter~\cite{gordon1993novel}.\label{alg:bs_filter}}
+\end{algorithm}
+\end{table}
+The bootstrap filter is an early particle filter, consisting of two steps. The first is to propogate
+the prior's predictions through the system model. We then assign a weight to each particle based on
+how likely the data we have just received is given the prior prediction. Finally we use the
+bootstrap resampler to convert these weighted particles into the estimate of the current state. This
+current state can become the prior information for the next iteration.
+
+%%%%%%
+
+\subsubsection{Linear Ensemble Transform Filters}
+
+The bootstrap filter is a random transformation of the empirical density, $\pi$, and the transformed
+density, $\tilde{\pi}$, has the same support as $\pi$, with a number of copies of each
+particle. It is often undesirable for us to have copies since this means that we do not cover as
+much of the sample space as we could. This leads to wasted computation, poorer quality samples and a
+worse approximation of the densities of interest. One method of reducing the number of repeats in a
+sample is to allow $\tilde{\pi}$ to be built from linear combinations of the original sample. This
+means that the support of the density $\tilde{\pi}$ is not restricted to being the same as $\pi$ and
+reduces the likelihood of having two particles in the same position. Particle filters which use
+linear combinations to produce the transformed density are known as Linear Ensemble Transform
+Filters (LETFs).
+
+\begin{dfn}[Linear Ensemble Transform Filter]\label{def:LETF}
+A particle filter where the original sample, $\{\theta_i\}_{i=1}^M$, is transformed into a new
+sample, $\{\tilde{\theta}_i\}_{i=1}^M$, according to the relation
+\[
+	\tilde{\theta}_j = \sum\limits_{i=1}^M \! w_{ij}\theta_i, \quad w_{ij} \geq 0, \ \sum\limits_{i=1}^M
+		w_{ij} = 1,
+\]
+is known as a linear ensemble transform filter.
+\end{dfn}
+
+%%%%%%
+
+\subsubsection{The Ensemble Transform Particle Filter}
+
+The ensemble transform particle filter (ETPF) proposed by Reich in~\cite{reich2013nonparametric} is
+an ensemble transform method which combines the optimal coupling ideas from
+Section~\ref{sec:coupling} with the LETF framework to produce a particle filter which is consistent
+in the limit of the ensemble size $M\rightarrow\infty$. The transform takes a sample of weighted
+particles $\{(w_i,y_i)\}_{i=1}^M$ with law$(\mu_Y) = \pi_{Y}$ and converts it into a sample of
+evenly weighted particles $\{x_i\}_{i=1}^M$ with law$(\mu_{X}) = \pi_{X}$, by means of defining a
+coupling $T^*$ between the distributions $Y$ and $X$. The couplings used here are optimal in the
+sense discussed in Section~\ref{sec:coupling}. The coupling $T^*$ is the solution to a linear
+programming problem in $M^2$ variables with $2M-1$ constraints as defined in
+\cite{reich2013nonparametric}. Choosing to maximise the correlation means that a statistic $s(X)
+\rightarrow s(Y)$ as $M\rightarrow\infty$, and $s(X)$ converges faster than any other random
+variable induced by a linear coupling. The density $\pi_X$ is the optimal density for approximating
+$\pi_Y$ given that the mixture components have equal variance and equal weight.
+
+\begin{table}[htpb]
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+
+\nonl \textbf{Predict}\;
+Sample $\xi_i(t-1) \sim \mathcal{N}(0, \sigma^2)$.\;
+Calculate $x_i^*(t) = f_{t-1}(x_i(t-1), \xi_i(t-1))$.\;
+\nonl \textbf{Update}\;
+For each particle calculate
+\[
+	w_i = \frac{\ell(D_i(t)| x_i^*(t))}{\sum_{k=1}^N \! \ell(D_k(t)| x_k^*(t))}.
+\]
+
+Find the optimal coupling,
+\[
+	T^* = \arg\min\limits_{T} \sum\limits_{i,j=1}^M \! t_{ij}\|x_i^*(t) - x_j^*(t)\|^2_2.
+\]
+
+Set $x(t) = M(T^*)^\top x^*(t)$.\;
+
+\caption{The Ensemble Transform Particle Filter~\cite{reich2013nonparametric}.\label{alg:ETPF}}
+\end{algorithm}
+\end{table}
+
+The ETPF is outlined in Algorithm~\ref{alg:ETPF}. It follows the same structure as the bootstrap
+filter in Algorithm~\ref{alg:bs_filter}, but in this case the resampling step is performed by taking
+a linear combination of the prior states rather than sampling from a multinomial distribution. The
+conditions on the $t_{ij}$ are given in Equation~\eqref{eq:coupling_conditions}.
+
+%%%%%%
+
+\subsubsection{Approximate Multinomial Resampling}
+
+The ETPF described in the previous section is optimal in terms of maximising the correlation between
+the two marginal distributions. However the linear programming problem which must be solved for
+every new empirical distribution is very costly, especially as the ensemble size grows. For the
+purposes we require, discussed in Chapter~\ref{sec:PAIS}, we do not need as much accuracy as is
+provided by the ETPF. For this reason we introduce a new algorithm, the approximate multinomial
+resampler (AMR)~\cite{cotter2015parallel}, which shares many properties with the ETPF, but relaxes the constraint on
+finding the optimal coupling. Instead we find a coupling via a greedy method in which the
+correlation is high near particles with large weights, but is poor for particles with a low weight.
+
+To perform the resampling, we split the state space into $M$, not necessarily disjoint subspaces,
+$R_i\neq\emptyset$, each subspace containing at least one of the original samples $\{x_i\}$. We
+define $M$ new probability distributions, $p^i$, such that $\sum_{i=1}^M \! p^i = w$, where $w$ is
+the vector of the original probabilities corresponding to $x$. These probability distributions each
+satisfy the conditions
+\[
+	\tilde{p}_k^i \leq w_k, \quad \|\tilde{p}^i\|_1 = \frac{1}{M},\quad \text{and}\quad p^i =
+		M\tilde{p}^i.
+\]
+At this point, we could still be talking about the ETPF. The distinction comes in the way that we
+choose the $R_i$ and $\tilde{p}^i_k$. We choose to first pick the particle, $x_J$, with the highest
+weight and then expand a ball around it until the ball contains $1/M$ of the total probability mass.
+The particle furthest from $x_J$ inside the ball may have its weight reduced by the amount required
+for $\|\tilde{p}^i\|_1=1$. All other particles in the ball will have $\tilde{p}^i_k = w_k$. Once
+this is done, particles which have had all their probability mass assigned to a distribution $p^i$
+are removed, and we choose the particle with the next highest weight.
+
+This method ensures that areas with a high weight have a particle chosen in that area again. The
+greedy nature of the algorithm means that the correlation is sub optimal. In Algorithm~\ref{alg:AMR}
+we present the pseudo-code form of this algorithm.
+
+\begin{table}[htpb]
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+
+Obtain a weighted sample $\{(w_i, y_i)\}_{i=1}^M$.\;
+\For{$i=1,\dots,M$}{
+	$J = \arg\max_j \! w_j$.\;
+	$p_{J,i} = 1/M \wedge w_J$.\;
+	$w_J = w_J - p_{J,i}$.\;
+	\While{$\sum_j p_{j,i} < 1/M$}{
+		$K = \arg\min_{k\in\{k|w_k>0\}} \! \|y_J - y_k\|$.\;
+		$p_{K,i} = (1/M-\sum_j \! p_{j,i}) \wedge w_K$.\;
+		$w_K = W_K - p_{K,i}$.\;
+	}
+}
+$x = MP^\top y$.\label{algline:amr_mean}\;
+\caption{The Approximate Multinomial Resampler~\cite{cotter2015parallel}.\label{alg:AMR}}
+\end{algorithm}
+\end{table}
+
+In Line~\ref{algline:amr_mean} of Algorithm~\ref{alg:AMR} we see that the output of the algorithm,
+an evenly weighted sample, takes the form required by Definition~\ref{def:LETF} to be an LETF. We
+note that the AMR algorithm, like the ETPF algorithm preserves the mean of the sample,
+\[
+	\hat{\mu}_X = \frac{1}{M}\sum_i \! x_i = \frac{1}{M}\sum_i \sum_k \! p_{k,i}y_k = \frac{1}{M}\sum_k
+		w_ky_k = \hat{\mu}_Y.
+\]
+Here we use that the columns of $P$ are the probability distributions $p^i$, and the sum of these
+distributions is equal to the weights, $w$, hence
+\[
+	\sum_i \! p_{k,i}y_k = (p^1_k + \dots + p^M_k)y_k = w_ky_k.
+\]
+
+%%%%%%
+
+\subsubsection{Numerical Comparison of the Filter Resamplers}
+
+We have designed the AMR with the aims of being numerically cheaper than the ETPF, and more accurate
+than the BSR. Therefore we now present numerical examples which demonstrate that we have achieved
+this. To compare the algorithms, we write a simple MC algorithm which compares the ability of each
+algorithm to preserve the statistics of a weighted sample in an evenly weighted sample.
+
+\begin{table}[htpb]
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\BlankLine
+\For{$k=1,2,\ldots,\frac{N}{M}$}{
+	Sample $y_j \sim Q$, for $j = 1, \ldots, M$.\;
+	Weight $y_j$ under $F$,
+	\[
+		w_j = F(y_j) / Q(y_j).
+	\]
+
+	Resample 
+	\[
+		\hat{F}_k = \sum_j w_j\delta_{y_j}(\cdot) \rightarrow \sum_j n_j\delta_{x_j}(\cdot) = \hat{G}_k \approx \hat{F}_k.
+	\]
+
+	Calculate error in the statistic
+	\[
+		s_k = \frac{|s(\{x_j\}) - s(\{(w_j, y_j)\})|}{|s(\{x_j\})|}.
+	\]
+}
+Return mean($\{s_k\}$).\;
+\caption{Comparing Resamplers.\label{alg:ETMC}}
+\end{algorithm}
+\end{table}
+
+In Algorithm~\ref{alg:ETMC} we create random empirical weighted distributions, $\hat{F}$, which are an
+approximation of a chosen distribution $F$. We then resample from $\hat{F}$ to create a new
+distribution $\hat{G}$. The new empirical distribution, $\hat{G}$, follows roughly the same
+distribution but some error is introduced since we require an integer number,
+$n_j$, of each particle. We want to calculate how big the error is between the distributions of $\hat{F}$ and
+$\hat{G}$. To do this we calculate some chosen statistics of both $\hat{F}$ and $\hat{G}$ and
+calculate the relative difference between these statistics. This is done for each resampler with a fixed total sample size of $N$,
+and we vary $M$ to see how the relative error varies with the ensemble sizes.
+
+\begin{figure}[htb]
+\centering
+\subfigure[Relative error in $\mathbb{E}(X)$]{\includegraphics[width=0.45\textwidth]{"images/AMR_EX"}}
+\subfigure[Relative error in
+$\mathbb{E}(X^2)$]{\includegraphics[width=0.45\textwidth]{"images/AMR_EX2"}}\\
+\subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"images/AMR_EX3"}}
+\subfigure[Cost of resamplers per iteration]{\includegraphics[width=0.45\textwidth]{"images/AMR_speed"}}\\
+\caption{Relative errors representing the error between $\hat{F}$ and $\hat{G}$ as described in Algorithm~\ref{alg:ETMC}, where $\hat{G}$ is constructed using the Bootstrap, ETPF and AMR algorithms.}
+\label{fig:AMR}
+\end{figure}
+
+To test the resamplers we chose a proposal distribution, $Q = \mathcal{N}(1,2)$, and a target
+distribution $F = \mathcal{N}(2,3)$. Figure \ref{fig:AMR} (a)-(c) shows how the relative errors in
+the first three raw moments varies with ensemble size $M$. As expected, the AMR lies somewhere
+between the high accuracy of the ETPF and the less accurate BSR. Note that only the error for the
+BSR is presented for the first moment since both the ETPF and the AMR are mean preserving. Figure
+\ref{fig:AMR} (d) shows how the computational cost, measured in seconds, scales with the ensemble
+size. These results demonstrate that the AMR behaves how we wish, and importantly ensures that
+exactly one sample of the output will lie in each region with higher than average density.
+
+%%%%%%
+
+\subsubsection{Scaling of Resampling Algorithms with Dimension}
+Ideally the dimension of particles in a filter will not affect the convergence of the method.
+Unfortunately, the approximation of continuous density functions is computationally infeasible in
+high dimensions with the ensemble size required to achieve convergence growing exponentially in the
+dimension\cite{silverman1986density,snyder2008obstacles}.
+
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{images/ET_dim_scaling}
+\caption{Demonstration of the ETMC algorithm rapidly failing as the dimension increases in problem
+$G_1$ defined in Chapter~\ref{sec:PAIS}. The relative error shown is that between $\mathbb{E}_{\hat{F}}[X]$ and $\mathbb{E}_{\hat{G}}[X]$.}
+\label{fig:ET_dim_scaling}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:ET_dim_scaling} shows how quickly the ETMC algorithm loses its ability to accurately
+represent the mean of the target distribution as the dimension increases. Clearly for a simple
+Gaussian problem, we can not resample in much more than 10 dimensions and hope to recover a
+representative sample without massively increasing the ensemble size. This phenomenon is known as
+the curse of dimensionality. The AMR and BSR suffer in a similar manner as the dimension increases.
+
+%%%%%%%%%%%%
+
+\subsection{Couplings between Arbitrary Probability Measures}\label{sec:intro_TM}
+
+In the previous sections we have looked at couplings which convert a weighted sample into an
+unweighted sample while preserving the approximate empirical distribution. In this section we look at
+how we can couple two arbitrary measures. In the following discussion we concentrate on designing a
+coupling which maps from an arbitrary probability distribution onto a standard Gaussian
+distribution. The main difference here is that we would like to construct a map between the
+continuous distribution $F$ and a \emph{reference measure}, $\mathcal{N}(0, \text{I})$, rather than
+between the approximation of $F$, $\hat{F}$, and the approximation of $\hat{F}$, $\hat{G}$.
+
+Obviously constructing an exact map for arbitrary $F$ is infeasible, so we must again use an
+empirical distribution $\hat{F}$ where $\hat{F}\rightarrow F$ in the limit of infinite samples,
+however we can ensure that we are coupling $\hat{F}$ to the continuous distribution $\mathcal{N}(0,
+1)$. This convergence must be achieved by continual sampling from $F$, which for an arbitrary
+distribution can be expensive. To speed up this process~\cite{parno2014transport} proposes an MCMC
+algorithm which sequentially updates the distribution $\hat{F}$ and recalculates the map so that
+after each update, sampling from $F$ becomes cheaper. Calculating the transport map between
+$\hat{F}$ and $\mathcal{N}(0, \text{I})$ allows us to sample more efficiently because it reduces the
+sampling problem to sampling from a standard Gaussian distribution and then mapping these Gaussian
+random variables onto an arbitrary probability measure. Which for the class of maps considered is
+very cheap. These couplings are discussed further in Chapter~\ref{sec:map}.
+
 \section{Construction of transport maps in importance sampling} \label{sec:map}
@@
 PAIS is particularly useful if the posterior is multimodal, or inconveniently shaped such as the
-Rosenbrock density discussed in Chapter~\ref{sec:LPAIS}. However, we have had trouble when
+Rosenbrock density discussed in Chapter~\ref{chapter:HMC}. However, we have had trouble when
 attempting to increase the dimension of the parameter space. This was seen when we attempted to
@@
 
-In this Chapter we look at a method which will allow us to improve efficiency in our proposal distributions. The ideal sampling
-situation in MCMC is to explore an $d$-dimensional Gaussian distribution (or other standard distribution) using a near proportional
+In this Chapter we look at a method which will allow us to improve efficiency in low dimensions, and
+will also allow us to sample more efficiently in higher dimensions. The ideal sampling
+situation in MCMC is to explore an $n$-dimensional Gaussian distribution (or other standard distribution) using a near proportional
 proposal distribution.
@@
 %functions used in the expansion, and increasing the number of samples used in the optimisation
-%problem. In practice it appears that we need not aim to achieve a high accuracy solution. Excellent
+%problem. In practise it appears that we need not aim to achieve a high accuracy solution. Excellent
 %results have been observed using monomial basis functions of total order 3, and we need a sample of
@@
 
-\section{The Transport Map}
+\subsection{The Transport Map}
 
@@
 \begin{dfn}[Exact Transport Map $T$]
-	A transport map $T$ is exact if the {\it pullback} of the standard Gaussian density, $\phi(\cdot)$,
+	A transport map $T$ is exact if the {\it pullback} of the standard Gaussian density,
 	\begin{equation}\label{eq:pullback}
@@
 	\end{equation}
-	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
-	\[
-		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
-	\]
+	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$.
 \end{dfn}
 
-In the case that we have the exact map, $T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible maps, we are able to draw a sample from $\pi_r = \phi$, the density of the standardised Gaussian distribution, and map these samples back onto target space using $T^{-1}$.
+In the rare case that we have the exact map, $T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible maps, we are able to draw a sample from $\pi_r = \phi$, the density of the standardised Gaussian distribution, and map these samples back onto target space using $T^{-1}$.
 These proposed samples are distributed according to the target distribution.
-
-\begin{dfn}[Target and Reference Space]
-	The transport map pushes a particle from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target measure $\mu_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference measure $\mu_r$.
-\end{dfn}
 
@@
 \[
-	\tilde{T}\mu_{\theta} \approx \mu_r.
-\]
-Relaxing the equality constraint to finding the approximate map $\tilde{T} \in \mathcal{T}^{\ltri}$ which optimises $C(T)$, allows us to formulate an efficient sampling method, in terms of the linear algebra required, and experimentation has shown that the pushforward of the target is close enough to Gaussian to achieve efficient sampling.
-
-\subsection{The optimisation problem}
+	\mu_{\theta} \approx \tilde{T}\mu_r.
+\]
+Relaxing the equality constraint to; finding the approximate map $\tilde{T} \in \mathcal{T}^{\ltri}$ which minimises the distance, allows us to formulate an efficient sampling method in terms of the linear algebra required, and experimentation has shown that the departure of the reference space from Gaussian is not too severe.
+
+\subsubsection{The optimisation problem}
 
@@
 \]
-This divergence results in some nice properties which we will explore in the following derivation.
-The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to it's relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
-\[
-	D_{KL}(p\|q) \geq D_H^2(p,q),
-\]
-which is a true metric between probability distributions $p$ and $q$.
-Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
+Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map, the divergence becomes
 \[
@@
 \]
-We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
-This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown.
-However this is exactly the scenario for which we would turn to MCMC methods for a solution.
+We note the posterior density is constant in $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
+This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown, however this is exactly the scenario for which we would turn to MCMC methods for a solution.
 
@@
 \]
-which has a unique solution since the cost function is convex.
+which has a unique solution.
 
@@
 $\beta=1$ is sufficient for most problems. The form of the penalisation term promotes maps which are
-closer to the identity.
-
-\subsection{The structure of the map}
+close to the identity.
+
+\subsubsection{The structure of the map}
 
@@
 of the map in more detail.
-The lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density, as well as the inverse of the map, $\tilde{T}^{-1}$.
+The lower lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density as well as the inverse of the map, $\tilde{T}^{-1}$.
 The Jacobian of $\tilde{T}$ is a lower triangular matrix,
 \[
-	J_T(\theta) = \begin{bmatrix}
+	DT(\theta) = \begin{bmatrix}
 		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & \partial_{\theta_d} \tilde{T}_1(\theta)\\
@@
 Inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map.
-The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, so we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$.
+The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$.
 This inversion tells us the value of $\theta_1$, which means the next component is again a univariate polynomial, $T_2(\theta_2; \theta_1)=r_2$.
-We can then perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.
+This means that we can perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.
 
@@
 Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$.
-These index sets are discussed
+These index sets are discussed further
 in~\cite{parno2014transport}.
@@
 
-\subsection{Implementation of the optimisation problem}\label{sec:transport_implementation}
+\subsubsection{Implementation of the optimisation problem}\label{sec:transport_implementation}
 
@@
 positive and so this functional is always finite.
-In practice it is infeasible to enforce this condition across the whole parameter space.
+In practise it is infeasible to enforce this condition across the whole parameter space.
 We instead enforce this condition by ensuring that the derivatives are positive at each sample point.
@@
 	&\text{subject to} \quad \frac{\partial T_i}{\partial\theta_i}(\theta^{(k)};
-		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,d.
+		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,n.
 		\notag
@@
 
-\subsection{Implementation of the optimisation problem in PAIS}
+\subsubsection{Implementation of the optimisation problem in PAIS}
 
@@
 We introduce a diagonal matrix $W = \text{diag}(w)$, where $w = (w_1,\dots,w_K)^\top$ into the log of the reference density, and replace the ones vector, $c$, with $w$.
-Analogously to the unweighted case, we include a $1/K$ to the regularisation term so that it has less influence as we obtain more samples.
-If we allowed this regularisation term to exert influence at all updates we would converge to a suboptimal map.
-We should converge to the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation.
+We also include a $1/K$ to the regularisation term to allow it to be drowned out.
+We should obtain the same optimal map with a weighted sample as we do with an unweighted sample since we are trying to approximate the same expectation.
 The weights are strictly
@@
 
-\section[Transport map MCMC]{Using the transport map to propose states in MCMC algorithms}\label{sec:transport_algs}
-
-Given samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
-target measure to a reference measure. We now consider how to implement an MCMC algorithm which uses
-these maps to propose new states. These algorithms fit into the adaptive framework described in
-Section~\ref{sec:adapt} since we use the full history of the chain when updating the map, and this update changes the shape of our proposal
-distribution. Convergence of this adaptation is shown in~\cite{parno2014transport}.
-
-\subsection{Transport map MH algorithms}
-
-As introduced in~\cite{parno2014transport}, a MH algorithm using the transport map to propose new
-states can be designed, as given in Algorithm~\ref{alg:TransportMH}. These algorithms
-map a sample from the target onto the reference space, a new state is then efficiently
-proposed using a Gaussian kernel and is mapped back onto target space. This can
-result in rapid mixing, where step size and direction are now a function of the current particles location in state space, similar to the ideas of Riemann manifold Hamiltonian Monte Carlo~\cite{girolami2011riemann}.
-
-\begin{table}
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\BlankLine
-Initialise state $\theta^{(1)} = \theta_0$.\;
-Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
-\For{$k \leftarrow 1, \dots, L-1$}{
-	Compute $r = \tilde{T}(\theta^{(k)}; \bar{\gamma}^{(k)})$.\;
-	Sample $r' \sim q_r(\cdot; r)$.\;
-	Invert $\theta' = \tilde{T}^{-1}(r'; \bar{\gamma}^{(k)})$.\;
-	Calculate:
-	\[
-		\alpha = 1 \wedge
-			\frac{\pi(\theta')}{\pi(\theta^{(k)})}\frac{q_r(r|r')|J_{\tilde{T}}(\theta^{(k)};
-			\bar{\gamma}^{(k)})|}{q_r(r'|r)|J_{\tilde{T}}(\theta';\bar{\gamma}^{(k)})|}.
-	\]
-
-	Sample $u \sim U[0,1]$.\;
-	Set $\theta^{(k+1)}$ to $\theta'$ with probability $\alpha$, otherwise $\theta^{(k+1)} =
-		\theta^{(k)}$.
-
-	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
-		\For{$i \leftarrow 1, \dots, n$}{
-			Solve \eqref{eq:final_map_optim} with $\{\theta^{(1)}, \dots, \theta^{(k+1)}\}$ and
-				update $\gamma_i^{(k+1)}$.\;
-		}
-	}{
-		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
-	}
-}
-\caption{MH algorithm with adaptive transport map~\cite{parno2014transport}\label{alg:TransportMH}}
-\end{algorithm}
-\end{table}
-
-Initially we begin with the identity map, $\iota$, while we build enough samples to produce an approximate
-map. For simplicity, we update the map at uniform intervals of $K_U$ iterations, given an initial
-burn-in phase. In order to satisfy the conditions for adaptive algorithms to be ergodic, discussed
-in Section~\ref{sec:adapt}, this adaptive phase must come to an end. It is sensible to stop
-adaptation once the map has sufficiently converged, e.g. when updates do not change the coefficients, $\gamma$,
-above some tolerance, or as suggested by~\cite{parno2014transport} when the variance of $D_\text{KL}(\pi\|\tilde{\pi})$ is sufficiently close to zero.
-
-Once the map is sufficiently converged, we can decide whether we would like to continue using a local proposal distribution, or whether to use a distribution which is independent of the current state.
-An independent proposal is more efficient
-if the reference space is close to Gaussian; however, if it is not then certain regions of the target
-space could be less likely to be visited than in a standard MH algorithm. A compromise can be found by randomly selecting which kernel to sample from, i.e. the proposal distribution
-\[
-	q(x, \cdot) = p\mathcal{N}(x, \Sigma)+(1-p)\mathcal{N}(0,\text{I}), \quad p \in [0, 1].
-\]
-
-\subsection{Transport map PAIS algorithms}
+\section{T-PAIS}\label{sec:TPAIS}
+%Plug into PAIS
 
@@
 
-\section[Convergence of transport MCMC]{Convergence of the transport proposal based MCMC algorithms}
-
-In this section we study the convergence of the transport based proposal distributions which we have described in Section~\ref{sec:transport_algs}.
-We follow a similar strategy to that in Chapter~\ref{sec:PAIS}.
-We begin by finding optimal values for the scaling parameters by performing a series of simulations with differing scaling parameters.
-We then produce 1 million samples from the target distribution with the optimal scaling parameter and an ensemble size of $M=150$.
-We take 32 repeats of this optimal simulation and present the geometric average of the convergence rates.
-This process is then repeated for each of the algorithms we have discussed, Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2}.
-
-In this section we return to the Rosenbrock banana-shaped density which in Chapter~\ref{sec:LPAIS} we labelled $R_1$. This target density is
-\begin{equation}\label{eqn:R2_repeat}
-	\pi(\theta) = \frac{\sqrt{10}}{\pi}\exp\left\{ -(1 - \theta_1)^2 - 10(\theta_2 - \theta_1^2)^2 \right\}.
-\end{equation}
-A contour plot of the target density is given in Figure~\ref{fig:R2_posterior}.
-\begin{figure}
-\centering
-\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"images/R2_marginal_1"}}
-\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"images/R2_marginal_2"}}
-\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/R2_posterior}}
-\caption{Visualisation of the density of example $R_1$, as given in Equation~\eqref{eqn:R2_repeat}.}
-\label{fig:R2_posterior}
-\end{figure}
-
-\subsection{Implementation details}
-
-We now demonstrate some properties of the transport maps we will be using in our MCMC algorithms.
-We draw 1 million samples from the density in \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:transport_implementation} to build a transport map.
-We use this map to push forward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian.
-We then pull the sample back on to target space using the inverse map to check that our map is invertible and well behaved.
-
-For this example, we use an index set of total order 3 with monomial basis functions. This results in a map of the form
-\[
-	T(\theta_1, \theta_2) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \end{bmatrix},
-\]
-where
-\begin{align*}
-		T_1(\theta_1) &= \gamma_{1,1} + \gamma_{1,2}\theta_1 + \gamma_{1,3}\theta_1^2 + \gamma_{1,4}\theta_1^3, \\
-		T_2(\theta_1, \theta_2) &= \gamma_{2,1} + \gamma_{2,2}\theta_1 + \gamma_{2,3}\theta_1^2 + \gamma_{2,4}\theta_1^3
-					+ \gamma_{2,5}\theta_2 + \gamma_{2,6}\theta_1\theta_2 \\
-				 & \qquad \quad + \gamma_{2,7}\theta_1^2\theta_2 + \gamma_{2,8}\theta_2^2 + \gamma_{2,9}\theta_1\theta_2^2 +
-					 \gamma_{2,10}\theta_2^3.
-\end{align*}
-Clearly even with only basis functions of total order 3, we have a large number of unknowns in our optimisation problem, $\bar{\gamma} \in \mathbb{R}^{14}$.
-If we were to increase the dimension of $\theta$ further we would need to reduce the number of terms we include in the expansion by, for example, removing all the cross terms.
-This reduces the quality of our map but since we only require an approximate map we can afford to reduce the accuracy.
-
-\begin{figure}[htpb]
-\centering
-\subfigure[Original sample $\theta$ from RWMH algorithm.]{\includegraphics[width=0.31\textwidth]{"images/R2_transport_orig"}}\quad
-\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/R2_transport_ref"}}\quad
-\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/R2_transport_pullback}}
-\caption{Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
-\label{fig:R2_transport}
-\end{figure}
-
-Figure~\ref{fig:R2_transport} shows the efficiency of the transport map.
-Even though we have truncated the infinite expansion in the monomial basis down to 4 and 10 terms in respective dimensions, the push forward of the sample is still a unimodal distribution centred at the origin with standard deviation 1.
-As you move out into the tails of the reference density more non-Gaussian features form, but these are not too much of a problem when using a suitable proposal distribution.
-The pullback from reference space, in Figure~\ref{fig:R2_transport}, is an exact match of the original sample since we have not perturbed the sample in reference space.
-This inversion is well defined in the sampling region, although not necessarily outside.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{images/zero_jacobian}
-\caption{An example transport map which does not allow exploration of the full target distribution, and has a non-invertible point.}
-\label{fig:TM_zero_jacobian}
-\end{figure}
-
-If we consider a map built from polynomials of total order $p=2$, it is clear that the map $\tilde{T}$ does not map onto $\mathbb{R}^d$, but on to a subset of it.
-Figure~\ref{fig:TM_zero_jacobian} illustrates this point.
-We see that in this dimension, the $\tilde{T}^{-1}$ will not map any proposal $r > 0.8$ to a point in the target space.
-The point at which no inverse exists for a point $r_1$ can be made arbitrarily far out in the tails of the posterior distribution, but will be finite.
-The point at which the Jacobian is zero is also a non-invertible point since the logarithm in the pullback becomes infinite.
-This means that (1) we will never sample from the tail of the target distribution which lies beyond this turning point, and (2) all points in reference space $r < 0.8$ have the possibility of being mapped to two distinct points in target space.
-We can choose to truncate our parameter space at this turning point which allows our map to be bijective, however we will not be able to sample from the true posterior.
-For this reason it is best to chose an index set where the maximum order is an odd number.
-
-\subsection[Numerical results]{Numerical results for convergence of transport map based algorithms}
-
-We first find the optimal scaling parameters for the individual algorithms.
-This is done as before by optimising the effective sample size in the PAIS algorithm, and by tuning the relative $L^2$ error in the MH algorithm.
-There is currently no guidance on the best way of tuning the MH algorithm with transport map proposals although one might expect results similar to the standard MH results.
-As in the PAIS algorithm, the effective sample size might be the best option.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{lrrr}
-\toprule
-	Statistic \quad / \quad Algorithm & \ref{alg:TransportMH} & \ref{alg:TransportPAIS1} & \ref{alg:TransportPAIS2}  \\ \cmidrule(lr){1-4}
-	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
-	$\delta_{\text{ESS}}$				 & - & 1.0e-1 & 5.2e-1 \\ \cmidrule(lr){1-4}
-	Acc. rate							 & 0.23 & - & - \\
-	ESS ratio							 & - & 0.62 & 0.71 \\
-\bottomrule
-\end{tabular}
-\caption{Optimal scaling parameters for the transport map based algorithms applied to $R_1$.}
-\label{tab:R2_opt_scaling}
-\end{table}
-
-The optimal scaling parameters are given in Table~\ref{tab:R2_opt_scaling}.
-Here we see that the effective sample size is much lower than we see in the one-dimensional examples with the PAIS algorithms.
-However, in $R_1$ we are dealing with a much more complicated correlation structure, as well as a very slowly decaying tail in $\theta_2$.
-We have seen in Section~\ref{sec:LPAIS_R1} that the standard PAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.
-
-\begin{figure}[!h]
-\centering
-\subfigure[Comparison of the two Transport PAIS options.]{\includegraphics[width=0.45\textwidth]{images/R1_L2}}
-\subfigure[Comparison of Transport PAIS option (2) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/R1_L2_all}}
-\caption{Convergence of Algorithms~\ref{alg:TransportMH}, \ref{alg:TransportPAIS1}, \ref{alg:TransportPAIS2} for $R_1$. Ensemble size $M=150$, resampling performed using the AMR algorithm.}
-\label{fig:R2_l2_convergence}
-\end{figure}
-
-The convergence of the three algorithms is displayed in Figure~\ref{fig:R2_l2_convergence}.
-Figure (a) shows that the two variations of the transport based PAIS algorithms converge with similar rates.
-The second version, which performs the resampling stage in reference space rather than target space, has a slightly higher ESS, and is more stable than option (1).
-This version also has a property that we can exploit in Section~\ref{sec:TPAIS_higher_dim}.
-
-\section{Sampling in higher dimensions}\label{sec:TPAIS_higher_dim}
+\subsection{Sampling in higher dimensions}\label{sec:TPAIS_higher_dim}
 
@@
 
-\begin{table}[!htpb]
+\begin{table}
 \begin{algorithm}[H]
@@
 \]
-the likelihood of a reaction, $R_i$, occurring is determined by the \emph{propensity function}
-\[
-	\alpha_i(\mathbf{X}(t); \mathbf{k}) = \lim\limits_{\Delta t\rightarrow 0} \! \frac{\mathbb{P}(\text{Reaction $i$ occurs in $[t, t+\Delta t)$})}{\Delta t}
-\]
-which depends on the current state of the system and the reaction rates $\mathbf{k}$.
+the likelihood of a reaction, $R_i$, occurring is determined by the \emph{propensity function} $\alpha_i(\mathbf{X}(t); \mathbf{k})$ and depends on the current state of the system and the reaction rates $\mathbf{k}$.
 \end{dfn}
@@
 There is a propensity function associated with each of the $d$ possible reactions.
-The form of these functions is dependent on the type of reaction which occurs.
+The form of these functions is dependent on the the type of reaction which occurs.
 This notation can become cumbersome and so the dependence on the current state and reaction rates may be omitted.
@@
 
-The Gillespie algorithm is a stochastically exact algorithm which simulates a Markov jump process about the chemical population state space.
-The algorithm assumes that reactions occur in intervals which follow an exponential distribution.
+The Gillespie algorithm simulates a Markov jump process about the chemical population state space.
+The algorithm assumes that reactions occur at times which follow an exponential distribution.
 The rate associated with this exponential distribution is linked to the total system propensity $\alpha_0$.
 Which reaction occurs at each of these times depends on the distribution $\mathbf{p} = [p_1, \dots, p_d]^\top$.
+This assumption is the same as that in the formulation of the chemical master equation.
 
@@
 		Set $t_{j+1} \leftarrow t_j + \Delta t$.\;
-		Update populations $\mathbf{X}(t_{j+1}) = \mathbf{X}(t_j) + \nu_{r(t_{j+1})}$.\;
+		Update populations $\mathbf{X}(t_{j+1}) = f_{r(t_{j+1})}(\mathbf{X}(t_j))$.\;
 		$j \leftarrow j+1$.\;
@@
 The second, $\Delta t$, is from the exponential distribution with rate $\alpha_0$, and tells us how long until this reaction occurs.
-Finally we update the system's state, $\mathbf{X}(t_{j+1})$, according to which reaction has occurred, using the reaction state change vector, $\nu_i$.
-
-\begin{dfn}[State change vectors and the stoichiometric matrix]
-	The {\it state change vector} of a reaction $R_i$ is the vector, $\nu_i$, where the $j$th entry corresponds to the change made to the $j$th species by the reaction $R_i$.
-e.g.
-In a system with two chemicals $\mathbf{X} = [X_1, X_2]^\top$, and a reaction $R_1\colon X_1 \rightarrow X_2$, the state change vector of $R_1$ is $v_1 = [-1, 1]^\top$.
-The columns of the {\it stoichiometric matrix} are formed by the state change vectors for each reaction.
-\end{dfn}
+Finally we update the system's state, $\mathbf{X}(t_{j+1})$, taking into account which reaction has occurred, using the transition function $f_{r(t_j)}$.
+These steps result in a stochastically exact trajectory from which we can determine the probability of being in any particular state.
 
@@
 
-In many chemical systems, we can find variables which change on more than one time scale.
-These variables might be a single species, or a linear combination of different species.
-It is often possible to isolate which variables are changing more frequently (the \textit{fast variables}) and those which are changing less frequently (the \textit{slow variables}).
-In many cases we are interested in the dynamics of the slow variables, and the full simulation of the fast variables can be considered a waste of computational resources.
-In real world experiments it can also be difficult to measure the quantities of the fast variables given the relatively short times the system stays in a single state.
+In real world chemical systems, reactions will occur on more than one time scale.
+It is often possible to isolate which reactions are occurring more frequently (the \textit{fast reactions}) and those which are occurring less frequently (the \textit{slow reactions}).
+In many cases we are interested in the dynamics of the slow reactions, and the full simulation of the fast reactions can be considered a waste of computational resources.
 In this thesis we consider two separate simplifying approaches for these multiscale systems.
@@
 \begin{dfn}[Quasi-equilibrium assumption]
-	The quasi-equilibrium assumption (QEA) is the assumption that the fast variables converge in distribution on a timescale which is negligible with respect to the rate of change of the slow variables.
+	The quasi-equilibrium assumption (QEA) is the assumption that the fast reactions converge in distribution on a timescale which is negligible with respect to the rate of occurrence of the slow reactions.
 \end{dfn}
 
-This assumption allows us to approximate the dynamics of the slowly changing quantities in the system by assuming that the fast quantities are in equilibrium with respect to the fast variables in isolation, i.e.
-the slow rates $\rightarrow 0$.
+This assumption allows us to approximate the dynamics of the slowly changing quantities in the system by assuming that the fast quantities are in equilibrium with respect to the fast reactions in isolation.
 We also consider the constrained multiscale approach (CMA) described in \cite{cotter2011constrained,cotter2016constrained}.
-The CMA is similar to the QEA except that it takes into account changes in the slow variables which also change the fast variables.
-We briefly define the CMA as is given in \cite{cotter2016constrained}.
-
-\begin{dfn}[Constrained Projector]\label{dfn:constrained_projector}
-	Given a partition of the state space with $N$ variables into $N_s$ slow variables $\mathbf{S} = [S_1,\dots, S_{N_s}]$ and $N_f$ fast variables $\mathbf{F} = [F_1,\dots, F_{N_f}]$.
-A constrained projector $\mathcal{P}_\mathbf{S}\colon\mathbb{N}^N_0\rightarrow\mathbb{N}^N_0$ maps the values of the slow variables to a given value $\mathbf{S}$, while preserving the values of the fast variables.
-\[
-	\mathcal{P}_\mathbf{S}([\hat{\mathbf{S}}, \hat{\mathbf{F}}]) = [\mathbf{S}, \hat{\mathbf{F}}], \quad \forall \, [\mathbf{S},\mathbf{F}] \in \mathbb{N}_0^N.
-\]
-\end{dfn}
-
-\begin{dfn}[Constrained Stoichiometric Projector]
-	Given the partition in Definition~\ref{dfn:constrained_projector}, the constrained stoichiometric projector $\mathcal{P}\colon\mathbb{N}_0^{N}\rightarrow\mathbb{N}_0^{N}$ maps all non-zero values of the slow variables to zero.
-\[
-	\mathcal{P}([\mathbf{S}, \mathbf{F}]) = [\mathbf{0}, \mathbf{F}], \quad \forall \, [\mathbf{S}, \mathbf{F}] \in \mathbb{N}_0^{N}.
-\]
-\end{dfn}
-
-\begin{dfn}[Constrained subsystem]
-	The constrained subsystem is defined for a chemical reaction system, such as those we have been discussing so far, and is the system which arises from partitioning this chemical reaction system as in Definition~\ref{dfn:constrained_projector} and applying the constrained projector, $\mathcal{P}_\mathbf{S}$ to the state vector $\mathbf{X} = [\mathbf{S}, \mathbf{F}]$ after every new reaction, or equivalently applying the constrained stoichiometric projector, $\mathcal{P}$, to the change in state resulting from each reaction.
-When a stoichiometric vector becomes null by application of $\mathcal{P}$, that reaction can be removed from the system.
-If a reaction with one of the new stoichiometries would result in a species with a negative population, the propensity of this reaction is set to zero.
-\end{dfn}
-
 The CMA accounts for the differences in the invariant distribution of the fast species which are due to the occurrence of slow reactions.
 When the fast and slow subsystems operate on very different timescales, the CMA and QEA produce near identical results, however when the systems are not so clearly separated the CMA provides superior accuracy.
+Other simulation methods such as the $\tau$-leap method introduce significant bias into the approximations.
 
@@
 The data, $D$, is a complete trajectory of a molecular system which includes the times of each reaction as well as which reaction occurs.
-In practice it might not be obvious which reaction has occurred just by looking at the changes in chemical populations, however we ignore this problem for now.
+In practise it might not be obvious which reaction has occurred just by looking at the changes in chemical populations, however we ignore this problem for now.
 
@@
 
-From this formulation, we see that the random variables $(\Delta t_j, r(t_j))$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are Markovian.
-This conditional independence means that we can group events together by what state the system was in when the event happened.
+From this formulation, we see that the random variables $(\Delta t_j, r(t_j))$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are independent of each other.
+This independence means that we can group events together by what state the system was in when the event happened.
 We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
@@
 \[
-	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
+	\ell(\mathbf{R},~ \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{r}(\mathbf{Y}_i); \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
 \]
@@
 
-\section{Mono-molecular examples}
+\subsection{Mono-molecular examples}
 
@@
 
-\subsection{A Multiscale Chemical System}\label{sec:chem_multiscale}
+\subsubsection{A Multiscale Chemical System}\label{sec:chem_multiscale}
 
@@
 parameter regime such that the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
-more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
+much much more frequently than the other reactions, $R_1\colon \emptyset \rightarrow S_1$, and $R_4\colon S_2 \rightarrow \emptyset$. Notice that both
 chemical species are involved in fast reactions. However, the quantity
-$X = X_1 + X_2$, corresponding to an artificial species $S$, is conserved by both of the fast reactions,
+$S = X_1 + X_2$ is conserved by both of the fast reactions,
 and as such, this is the slowly changing quantity in this system.
-When recording our data, we are only able to take readings of the reactions which change the slow variable $S$. 
 The effective dynamics of $S$ can be represented as follows
@@
 application of the two approaches mentioned in Section~\ref{sec:SSAs}.
-Under the QEA, the value of $\mathbb{E}\left[k_4X_2|S=s\right]$ is found by computing the invariant distribution of the fast subsystem of reactions~\cite{jahnke2007solving},
+Under the QEA, the value of $\mathbb{E}\left[k_4X_2|S=s\right]$ is approximated by finding the steady state of the ODE representing the fast subsystem of
+reactions:
 \[
@@
 \]
-which tells us that $X_2$ is distributed according to Bin$(S, k_3(k_3+k_4)^{-1})$ and so
+we arrive at
 \[
@@
 We can use the posterior distribution derived in Section~\ref{sec:chem_stoch_posterior} to specify the posterior distribution for each of the three scenarios we are interested in.
-We would like to find out how the posterior distributions we obtain from the two models, the QEA and CMA, differ to the posterior for a complete data set.
+We are interested in the posterior distribution when we have a complete data set telling informing us of every reaction which occurs as well as perfect readings of the numbers of reactants.
+We are also interested in the two simplifying assumptions we saw in the previous section.
 
@@
 % choices made in transport map
-For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=3$ as the basis functions, i.e.
+For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=4$ as the basis functions, i.e.
 \[
@@
 \]
+Since $\mathcal{X}$ is four-dimensional, this yields a total of 125 map coefficients across the four index sets.
+This number can be reduced by using smaller index sets as discussed in~\cite{parno2014transport}.
 
@@
 Another option is to increase the total order of our index set.
-For these numerics we have chosen $p=3$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
+For these numerics we have chosen $p=4$, but we know that we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
 
@@
 Since the two posteriors have been approximated using an MCMC sample, there is a significant amount of noise, particularly in the tails of the distributions We can see that the differences in the marginals for $k_1, k_2,$ and $k_3$ are relatively small and the differences are largely positive.
-These positive differences tell us that the constrained approach claims to be more informative about the values of these first three parameters.
+This means that the constrained approach is slightly more informative about the values of these first three parameters.
 However the marginal for $k_4$ varies by a significant amount and is more negative.
-This means that the mean estimates for $k_4$ differ by a large amount, the posterior for the QEA is most peaked.
+This means that the mean estimates for $k_4$ vary by a large amount, and we should be more confident in the solution given by the QEA.
 
@@
 Similarly, the constrained approach tells us that we are able to observe the quantity $\hat{k}_4^{\text{CMA}} = k_2k_4/(k_2+k_3+k_4)$.
-To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most accurate and informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model, with all reactions observed.
+To validate our inferences on $k_2$, $k_3$ and $k_4$ we would like to discover which model is most informative about these parameters, and which model gives us results which are most similar to what we can obtain from the full model.
 
@@
 Again the constrained model is not substantially more attractive than the QEA model when we consider the Bayes factor $B_{1,2} = 1.96 < 3.2$.
-The Bayes factor $B_{0,2} = 4.1 > 3.2$ does tell us that we should significantly prefer the full model to the QEA approximation of the posterior.
+The Bayes factor $B_{0,2} = 4.1 > 3.2$ does tell us that we should significantly prefer the full model to the QEA model.
 These Bayes factors present a weak argument that that the constrained model provides us with a better description of the data than the QEA model.
-We can also present this information graphically, which might provide us with a greater justification for preferring the constrained approximation of the posterior.
+We can also present this information graphically, which might provide us with a greater justification for preferring the constrained model.
 
@@
 \subsection{Gene Regulatory Networks}\label{sec:grn}
-In this example, we look at a model of a gene regulatory network (GRN) mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering} used by a particular cell to regulate the amount of a protein, $P$, present.
-In this model, we again have reactions occurring at particular rates, $k_i$, between four different species.
+In this example, we look at a model of a gene regulatory network (GRN) mechanism used by a particular cell to regulate the amount of a protein, $P$, present.
+In this model, we again have reactions occurring at particular rates, $k$, between four different species.
 Here there is a special species, $G$, representing the gene responsible for building the protein, of which there is only one at a time, and this species can either be turned on or off.
@@
 
-In this example protein is created by a molecule of mRNA at a rate $k_6$.
+In this example, similar to those found in ***citations***, protein is created by a molecules of mRNA at a rate $k_6$.
 The amount of protein in the cell is then governed by the rate of decay of $P$, and the number of mRNA molecules in the cell creating $P$.
-In turn the number of mRNA molecules is controlled by its own decay rate and whether or not the gene $G$, which produces the mRNA, is active.
+In turn the number of mRNA molecules is controlled by its own decay rate and whether or not the gene $G$, which produces the mRNA is active.
 The gene can be turned on and off depending on how many dimers are present in the system, which is related to how many protein molecules there are.
 This completes a feedback cycle whereby if the protein population gets too high then the gene will turn off and the rate of new protein formation will decrease, causing the level to drop and the gene to turn back on.
-Figure~\ref{fig:GRN_cycle} shows what happens to the molecules in the system as the gene turns on and off.
 
@@
 	\caption{Diagram of the effects of molecule populations in a gene regulatory network.}
-	\label{fig:GRN_cycle}
+	\label{fig:grn_cycle}
 \end{figure}
 
-%As in Section~\ref{sec:chem_multiscale} we use the constrained multiscale approach to model the system.
-Due to the small scale of the system and the speed of the reactions, we are only able to measure the total amount of protein, $T$, and mRNA, $M$, while the number of dimers and whether or not the gene is switched on must be considered to be missing data.
-For this scenario, the likelihood function is intractable and so we must use a multiscale approximation of the system.
-Here we choose the constrained multiscale approximation due to its success in the previous example.
-
+As in Section~\ref{sec:chem_multiscale} we use the constrained multiscale approach to model the system.
 The formulation here is slightly more complex.
-We define a fast subsystem which acts on the fast variables $\mathcal{F} = [P, D, G]$ and incorporates reactions $R_1$-$R_4$.
-For the true parameter values we have chosen, 90\% of reactions which occur in a particular time frame are one of these four reactions.
+We define a fast subsystem which acts on the fast variables $\mathcal{F} = [P, D, G]$ and incorporates reactions $R_1$-$R_4$, 90\% of reactions which occur in a particular time frame are one of these four reactions.
 \[
@@
 \begin{align}
-	\hat{\alpha}_5 = k_5\mathbb{E}[G|T=t],\quad&\quad \alpha_6 = k_6M, \notag\\
-	\hat{\alpha}_7 = k_7\mathbb{E}[P|T=t], \quad&\quad \alpha_8 = k_8M.\label{eqn:grn_propensities}
+	\hat{\alpha}_5 = k_5\mathbb{E}[G|T],\quad&\quad \alpha_6 = k_6M, \notag\\
+	\hat{\alpha}_7 = k_7\mathbb{E}[P|T], \quad&\quad \alpha_8 = k_8M.\label{eqn:grn_propensities}
 \end{align}
-
-To calculate these expectations, we note that the fast subsystem is a deficiency zero system~\cite{anderson2010product}, and is also reversible. i.e. The system has four complexes, 
+To calculate these expectations, we note that the fast subsystem is a deficiency zero system ***citations***, and is also reversible. i.e. The system has four complexes, 
 \[
@@
 \]
-hence the deficiency is $d = 4 - 2 - 2 = 0$. This means that, by \cite{anderson2010product}, the conditional probabilities have the closed form equation,
-\[
-	\mathbb{P}[\mathcal{F}|T=t] \propto \text{Poisson}(P_t;\bar{p})\text{Poisson}(D_t;\bar{d})\text{Poisson}(G_t;\bar{g})\text{Poisson}(1-G_t;1-\bar{g}),
-\]
-from which we can calculate the expected values of $P_t$, $D_t$, and $G_t$.
-The fast variables $P_t$, $D_t$ and $G_t$ are constrained to the line defined by $T=P+2D+2(1-G)=t$.
+hence the deficiency is $d = 4 - 2 - 2 = 0$. This means that the conditional probabilities have the closed form equation,
+\[
+	\mathbb{P}[\mathcal{F}|T] \propto \text{Poisson}(P_T;\bar{p})\text{Poisson}(D_T;\bar{d})\text{Poisson}(G_T;\bar{g})\text{Poisson}(1-G_T;1-\bar{g}),
+\]
+from which we can calculate the expected values of $P$, $D$, and $G$.
+The fast variables $P_T$, $D_T$ and $G_T$ are constrained to the line defined by $T=P+2D+2(1-G)=t$.
 The concentrations $\bar{p}$, $\bar{d}$ and $\bar{g}$, must be calculated as the solution to the nonlinear system produced by examining the steady state behaviour of the fast subsystem,
@@
 \end{align}
-subject to $p \in [0, T]$, $g \in [0, 1]$, and $0 \leq d \leq \text{floor}(T/2)$.
+subject to $p \in [0, T]$.
 
@@
 A less positive result is that for this example the PAIS-TRW algorithm has found a map which has prevented further convergence of the mean.
-This is likely because one dimension of the transport map is sending proposals onto the negative real line and so all samples are being assigned a zero weight.
+This is likely because the one dimension of the transport map is sending proposals onto the negative real line and so all samples are being assigned a zero weight.
 This is behaviour which was anticipated, and it is always suggested that the transport map is constructed as a map $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$ rather than $T\colon\mathcal{X}\rightarrow\mathbb{R}^d$ where $\mathcal{X}\subset\mathbb{R}^d$.
-The large increase in the effective sample size observed between the PAIS-logRW and PAIS-logTRW algorithms is converted into an estimate which is twice as accurate after 1 million samples.
+Despite the large increase in the effective sample size observed between the PAIS-logRW and PAIS-logTRW algorithms, this improvement is  converted into an estimate which is twice as accurate after 1 million samples.
 
 A similar pattern is seen in the MH algorithms, where the MH-TRW performs the worst, and the MH-logTRW algorithm performs the best, although only marginally.
-
-\section{Conclusions}
-
-In this chapter we have demonstrated the process of applying the PAIS algorithm to realistic modelling situations.
-We saw how the standard PAIS algorithm can perform well for multimodal distributions in Section~\ref{sec:app_mixture}, in this case we only needed a RW proposal distribution, but it is possible to use the LPAIS algorithm or transport map based proposals if the tails of the posterior are not well behaved.
-
-We have also seen how the transport map converts posteriors which are badly scaled, have complicated correlation structures and possibly have non-Gaussian tails, into posteriors which we can sample effectively.
-This was demonstrated for two chemical reaction networks in Sections~\ref{sec:chem_multiscale} and \ref{sec:grn}
-
-These algorithms, when compared with the standard MC toolbox, have consistently performed the best, in the last GRN example realising estimates with more than two orders of magnitude better accuracy.
 

### 2019-12-04 Resubmission/response.tex
Started resubmission

@@
 
-We would like to thank the referees and editors for their time in considering our manuscript for publication in SIAM JUQ.
+We believe that the manuscript is now tighter and improved, and we would like to thank the referees and editors for their time in considering our manuscript for publication in SIAM JUQ.
 
@@
 With the length of this paper in mind, I would suggest shortening Sections 6 and 7 in order to focus more on numerical comparisons of the new algorithm and less on stochastic reaction networks.
-Details of the multiscale material could be moved into supplementary material or even into another manuscript.}{}
+Details of the multiscale material could be moved into supplementary material or even into another manuscript.}{We agree that the paper is lengthy, and welcome suggestions as to how to shorten it to make it more accessible.
+The method was designed with the application of inference of multiscale stochastic networks in mind, and the main examples in the manuscript concern this topic.
+However we agree that this is not the main focus of the paper, and so we have taken the referees suggestion onboard with the aim of reducing the amount of reaction network material.
+We are however keen to keep the results as shown in Figure 7.6 in the paper, since this was only possible with the aid of the methodology presented, but yet we do not believe it warrants a manuscript of its own.
+We have also rebalanced the introduction to make the main focus the MCMC methodology.
+Any required references for the material regarding multiscale methods for stochastic reaction networks are given Section 6.
+
+The biggest change we have made following this comment is the heavy redaction of Section 6.
+This keeps the focus on the numerical methodology.
+The more detailed version of Section 6 is now included as supplementary material for interested readers.}
+
+****DO SUPPLEMENTARY MATERIAL*****
+
+%I have reduced the introduction and it also flows better now.
 
@@
 With a sufficiently large number of samples, decreasing $\beta$ should result in a more accurate transport map, so why does a smaller value of $\beta$ not yield a more efficient algorithm?
-With $\beta=1$, is the map somehow able to capture enough of the posterior structure to enable effective importance sampling?}{}
+With $\beta=1$, is the map somehow able to capture enough of the posterior structure to enable effective importance sampling?}{The regularisation term is present to prevent overfitting of the transport map to what is likely to be an incomplete MCMC sample.
+The stage at which the map is being constructed coincides with the burn-in of the ETAIS algorithm, and as such overfitting to incomplete data could produce a map which hinders further exploration of the state space.
+The value used has been shown to work well, as demonstrated in the numerical experiments presented.
+A discussion along these lines has been added to after equation (3.2).}
+
+%DONE
 
@@
 Is there a reason for sticking with the random walk proposals?
-I think adding a comparison of proposal types, even if only on the Rosenbrock toy problem, would help readers understand the importance of each part in the transport map ETAIS algorithm: the proposal, map, resampler, and ensemble size.}
+I think adding a comparison of proposal types, even if only on the Rosenbrock toy problem, would help readers understand the importance of each part in the transport map ETAIS algorithm: the proposal, map, resampler, and ensemble size.}{The referee is entirely correct that in the situation that the transport map is able to capture all of the posterior features, that independent sampling from the reference Gaussian will produce near iid samples from the density.
+However, as demonstrated in the examples in the paper, for example Figure 7.4, the pushforward distribution for complex examples is often quite far from the Gaussian reference distribution, which would lead to a poorly performing algorithm.
+The whole basis of combining the ETAIS approaach with the transport maps is that for complex examples, the push forward density itself has complex structure which benefits from adaptive importance sampling.
+We have added a substantial discussion on this topic to just before Section 7.1.1, and also to the conclusions and abstract, since this is a key justification for the approach *****DO THIS*******.
+
+  The referee suggests using the Rosenbrock example to demonstrate this, but as can be seen in Figure ****, the transport map does a very good job for this example, and so experiments of the type suggested would not be sufficient to demonstrate this point.
+Conducting these experiments on the stochastic reaction networks example would be a substantial body of work, for which unfortunately we no longer have the resources to undertake.
+*****check and rewrite this******}
+
+%First part done, need to think more carefully about the second part. We REALLY don't want to do more numerics...
 
@@
 How is the ensemble size chosen?
-The authors mention that a smaller ensemble can be used when the transport map is accurate, but are there any guidelines on how to choose the ensemble size given the map?}{}
+The authors mention that a smaller ensemble can be used when the transport map is accurate, but are there any guidelines on how to choose the ensemble size given the map?}{It is relatively easy to assess when the ensemble size is too small, since it leads to poor importance samples and larger variances in the weights.
+If the transport map is doing a good, then the push forward distribution will be close to Gaussian, and then rules of thumb regarding the number of samples required to characterise an $n$-dimensional Gaussian can be used to suggest ensemble sizes.
+If variances of weights are very high, then it is likely that ensemble sizes will need to be increased.
+In general, the bigger the ensemble, the better the proposal, and the lower the variance of the resulting importance weights.
+However this has to be balanced with the increased costs of resampling, which grow polynomially in the ensemble size, depending on the method chosen.
+We have added a discussion on this to the end of Section 4.}
+
+%DONE
 
@@
 When reading the manuscript, I felt like I walked off the end of the manuscript.
-I think a sentence or two that ties things together would provide a better closing statement.}
+I think a sentence or two that ties things together would provide a better closing statement.}{We have rewritten this Section and ******.}
 
-\comment{I would prefer numbering all equations, but leave that to the author's discretion.}{}
+%REDO THE CONCLUSIONS AFTER EVERYTHING ELSE HAS BEEN REWRITTEN
 
-\comment{Subscript in Algorithm 1 on line 3. Should the proposal be centered at $\theta_j$ or $\theta_i$}{}
+\comment{I would prefer numbering all equations, but leave that to the author's discretion.}{We had no strong opinion on this so are happy to go with this suggestion.}
 
-\comment{Second equation on page 4. Should the sample be labeled with a $\hat{ }$?}{}
+%DONE
 
-\comment{When discussing monotonicity of the map, it might be worth pointing out the monotone parameterizations introduced in "An Introduction to Sampling via Measure Transport" by Marzouk et al.}{}
+\comment{Subscript in Algorithm 1 on line 3. Should the proposal be centered at $\theta_j$ or $\theta_i$}{Corrected, thanks.}
+
+%DONE
+
+\comment{Second equation on page 4. Should the sample be labeled with a $\hat{ }$?}{Corrected, thanks.}
+
+%DONE
+
+\comment{When discussing monotonicity of the map, it might be worth pointing out the monotone parameterizations introduced in "An Introduction to Sampling via Measure Transport" by Marzouk et al.}{Thanks for this suggestion, we have added a couple of sentences on this to ******.}
+
+%STILL TO DO
 
 \comment{In Section 4, the notation for proposed points on the reference space (marked with an apostrophe) and proposed points on the target space (marked with a $\hat{ }$) are different.
-It might be easier to follow if these were consistent, but that is up to the authors.}{}
+It might be easier to follow if these were consistent, but that is up to the authors.}{We have made a thorough review of the notation in this Section *******}
+%STILL TO DO
 
-\comment{Equation (6.2) $Y_j$ is used in the text but $X$ is used in the equation. Is that a typo?}{}
+\comment{Equation (6.2) $Y_j$ is used in the text but $X$ is used in the equation. Is that a typo?}{Correct, thanks.}
 
-\comment{Sentence before equation (6.3). "variable" should be plural "variables"}{}
+%DONE
+
+\comment{Sentence before equation (6.3). "variable" should be plural "variables"}{Corrected, thanks.}
+
+%DONE
 
@@
 On that note, the introduction takes a rather long time to get to the point of combining transport-ETIAS with partially observed stochastic reaction networks.
-Perhaps the necessary link here is “sloppiness,” mentioned on page 3?}{}
+Perhaps the necessary link here is “sloppiness,” mentioned on page 3?}{We agree with this comment.
+In particular, we have entirely removed the discussion on stochastic reaction networks from the introduction.
+This change moves the focus more firmly to that of the proposed sampling methods.
+As suggested by the referee we keep the discussion on sloppiness and unidentifiability, and briefly discuss multiscale systems as an example where this sloppiness often presents itself, leading to posterior densities which are highly concentrated near to lower-dimensional manifolds.}
 
-\comment{Also, it should be made more clear what exactly is new within the stochastic reaction networks portion of the paper}{}
+% 1/2 DONE, still to do - move some of the olde introduction into the stochastic networks section
+
+%Done now I think?
+
+\comment{Also, it should be made more clear what exactly is new within the stochastic reaction networks portion of the paper}{Thanks to the referee for this comment.
+The main novel result in this portion of the paper is that summarised by the plots in Fig *****.
+The multiscale methods themselves have been presented in other work which was already heavily cited in this paper.
+I am not aware of anybody having formulated inverse problems using these methods before, but the treatment is natural.
+But the ability to sample from the complex posteriors arising from these problems, and therefore the ability to compare the posteriors arising from different multiscale approximations is new.
+Following this comment and comments from referee 1, we have heavily reduced the volume of material discussing the multiscale approximations in Section 6, which now appears in supplementary material.}
+
+%DONE?
 
@@
 And relative to what?
-Similarly, can you define $\delta_{ESS}$ as a function of the ESS?}
+Similarly, can you define $\delta_{ESS}$ as a function of the ESS?}{Thanks for this comment, we have added more thorough descriptions of these quantities to *******.}
 
@@
 1
-which case I think plotting all the runs in the same figure would be more illustrative). These are Monte Carlo schemes, so I expect the error in any given running estimate to be noisy.}{}
+which case I think plotting all the runs in the same figure would be more illustrative).
+These are Monte Carlo schemes, so I expect the error in any given running estimate to be noisy.}{These plots are produced by averaging over ***** repeats.
+The plots show comparison of average convergence for several different algorithms.
+We tried creating the plots that the referee has suggested but the sheer number of lines led to plots which were in our opinion far less informative.
+We have given more thorough details of how the plots were created in ******.}
 
@@
 Similarly, what is meant by \# of samples in the ETIAS error plots?
-Using “samples” without context is confusing, as there are many intermediate populations of samples at hand.}{}
+Using “samples” without context is confusing, as there are many intermediate populations of samples at hand.}{For the MCMC methods we mean the number of iterations, and for ETAIS, we mean the number of iterations multiplied by the ensemble size.
+We have added a description along these lines to ******.}
 
-\comment{To complement these plots, I think that a very useful and widely used diagnostic, applicable both to MCMC and ETIAS and their transport-accelerated versions, would be ESS per posterior density evaluation, or simply ESS per unit of computational time.}{}
+\comment{To complement these plots, I think that a very useful and widely used diagnostic, applicable both to MCMC and ETIAS and their transport-accelerated versions, would be ESS per posterior density evaluation, or simply ESS per unit of computational time.}{The ESS is a very commonly used and useful diagnostic, but we are of the opinion that it is flawed and can lead to misleading results, in particular in the scenarios we are considering where the posterior density is highly concentrated around a low dimensional manifold.
+The ESS for an MCMC sample which has sampled efficiently locally but failed to explore the whole of the state space, can give a deceptively high value.
+To avoid this, we have used a much more challenging metric, the relative $L^2$ error in the density, approximated on a uniform mesh.
+This is challenging both in terms of computing it, but it is also a challenging metric for the Monte Carlo methods, since it requires global convergence of the chains.
+We understand the reasons that the referee has asked for this metric to be considered, not least because it's use is ubiquitous but we believe that the one we have used is actually a more rigorous measure.}
 
@@
 If by the true map one means the KR rearrangement, then it is invertible.
-Maybe I am confused about what is meant by “true map?”}{}
+Maybe I am confused about what is meant by “true map?”}{Thanks to the referee for highlighting our confusion about this, we have ammended the discussions in ****** and ****** to reflect this.}
 
@@
 More broadly, this reflects the close link between independence Metropolis algorithms and importance sampling.
-Some comments (please agree or disagree) on this front would be illuminating.}{}
+Some comments (please agree or disagree) on this front would be illuminating.}{The first referee made a similar comment on this on Page ***** of this document, which we hope addresses this comment also.
+We also refere the referee to the discussion we have added on this in the manuscript at ********.
+The problems we are attempting to sample from in this work are highly challenging, with the density highly concentrated on a low dimensional curved manifold, and in the majority of cases the low dimensional approximations of the transport map are not sufficient to make independent sampling efficient.}
 
@@
 This relates to my broader comment above.
-It would be good to have a clearer picture of where exactly this paper’s key contributions lie, and even of their relative weight.}{}
+It would be good to have a clearer picture of where exactly this paper’s key contributions lie, and even of their relative weight.}{We have not seen the derivation of the likelihood anywhere, but it is a natural treatment of the likelihood of a continuous time Markov chain, and so following comments from both referees we have removed this from the paper.
+It now appears in the supplementary material for interested readers.}
 
-\comment{Here are several suggestions for content that could be removed entirely or shortened, along with a few areas that could be expanded:
-  \begin{itemize}
-    \item The use of log-transformations to map strictly positive variables to the real line is a standard trick in MCMC, and has also been employed in previous work on transport.
+\comment{Here are several suggestions for content that could be removed entirely or shortened, along with a few areas that could be expanded:}{Thanks for these suggestions, we will address each in turn below.}
+
+\comment{Suggestion for removal: The use of log-transformations to map strictly positive variables to the real line is a standard trick in MCMC, and has also been employed in previous work on transport.
 One could simply mention this and move forward.
 Thus I don’t see the need of including any results without the log preconditioner in problems where it is warranted.
-Log- preconditioning is well recognized as a good thing to do.
-\item A lot of Section 3 could be abridged by appropriate referencing of [46].
-\item Section 4 is quite terse, but the“second option” (Algorithm 3)—both proposing and resampling on the reference space—seems to be the cleaner algorithm and, if I understand the authors, the more effective one.
+Log- preconditioning is well recognized as a good thing to do.}{Thanks to the referee for these very helpful suggestions.
+We agree that the log transformation is pretty standard, and so we have vastly reduced the discussion and numerical experiments presented on this topic.}
+
+%Have I?
+
+\comment{Suggestion for removal: A lot of Section 3 could be abridged by appropriate referencing of [46].}{
+We have removed as much as possible, but we wanted the paper to remain as standalone as possible to make it more accessible and easier to read.
+We also need to highlight the differences in which we construct the map, the material in section 6.****.
+However we have removed some of the details from sections *** and ****.}
+
+\comment{Suggestion for removal: Section 4 is quite terse, but the“second option” (Algorithm 3)—both proposing and resampling on the reference space—seems to be the cleaner algorithm and, if I understand the authors, the more effective one.
 So perhaps the first option can be mentioned in passing, and all the results can focus on the second?
-Is there any good reason to use Algorithm 2 instead of Algorithm 3?
-\item The description of the ETIAS algorithm in Section 2 is very nice, but here an illustrative figure might help. \end{itemize}}{}
+Is there any good reason to use Algorithm 2 instead of Algorithm 3?}{
+We now only briefly describe algorithm 2, and no longer give pseudocode for this option. We have kept the numerics in which the results are compared and a brief discussion.}
 
-\comment{End of Section 2, “learn local covariances across the whole of the domain”: I believe I under- stand what is meant, but this awkwardly phrased}{}
+\comment{Suggestion for removal: The description of the ETIAS algorithm in Section 2 is very nice, but here an illustrative figure might help.}{****Illustrative figure? Like what?*****}
+
+\comment{End of Section 2, “learn local covariances across the whole of the domain”: I believe I under- stand what is meant, but this awkwardly phrased}{Thanks, we have rerwritten this *****.}
 

### 2019-12-04 Resubmission/Supplementary.tex
Started resubmission

--- 2019-01-15 TETAIS.tex
@@
-\documentclass[final]{siamltex}
-%test change
-\usepackage{cite}
-\usepackage{graphicx,bbm,pstricks,soul}
-\usepackage{pifont}
-\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
-\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
-\usepackage{tikz,hhline}
-\usepackage{tabularx}
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\setlength{\parindent}{0in}
-\usepackage{amsmath,amsfonts,amsbsy,amssymb}
-\newcommand{\RARR}[3]{#1
-  \;\displaystyle\mathop{\displaystyle\longrightarrow}^{#3}\; #2}
-\newcommand{\RARRlong}[3]{#1
-  \;\displaystyle\mathop{-\!\!\!-\!\!\!-\!\!\!-\!\!\!-\!\!\!\!\displaystyle
-  \longrightarrow}^{#3}\; #2}
-\newcommand{\LARR}[3]{#1
-  \;\displaystyle\mathop{\displaystyle\longleftarrow}^{#3}\; #2}
-\newcommand{\LRARR}[4]{{\mbox{ \raise 0.4 mm \hbox{$#1$}}} \;
-  \mathop{\stackrel{\displaystyle\longrightarrow}\longleftarrow}^{#3}_{#4}
-  \; {\mbox{\raise 0.4 mm\hbox{$#2$}}}}
-\newcommand{\bX}{{\bf X}}
-\newcommand{\vecx}{{\mathbf x}}
-\newcommand{\vecy}{{\mathbf y}}
-\newcommand{\vecz}{{\mathbf z}}
-\newcommand{\vecq}{{\mathbf q}}
-\newcommand{\bs}{{\mathbf s}}
-\newcommand{\vecr}{{\mathbf r}}
-\newcommand{\vecX}{{\mathbf X}}
-\newcommand{\vecv}{{\mathbf v}}
-\newcommand{\tick}{\ding{52}}
-\newcommand{\cross}{\ding{54}}
-\newcommand{\vecn}{{\mathbf n}}
-\newcommand{\vecp}{{\mathbf p}}
-\newcommand{\cT}{{\mathcal T}}
-\newcommand{\dt}{{\mbox{d}t}}
-\newcommand{\dx}{{\mbox{d} \vecx}}
-\newcommand{\boldnu}{{\boldsymbol \nu}}
-\newcommand{\er}{{\mathbb R}}
-\renewcommand{\div}{{\rm div}}
-\newcommand{\bnu}{{\bf \nu}}
-\newcommand{\divergence}{\mathop{\mbox{div}}}
-\renewcommand{\P}{\mathbb{P}}
-\newcommand{\FP}{P_{\rm{FP}}}
-\newcommand{\ME}{P_{\rm{ME}}}
-\newcommand{\MEs}{P_{\rm{ME}_{S}}}
-\newcommand{\D}{\mathcal{D}}
-\newcommand{\G}{\mathcal{G}}
-\newcommand{\N}{\mathcal{N}}
-\newcommand{\X}{{\mathbf X}}
-\newcommand{\Y}{{\mathbf Y}}
-\newcommand{\W}{{\mathbf W}}
-\newcommand{\data}{D}
-\newcommand{\neff}{n_{\text{eff}}}
-\newcommand{\E}{{\mathbb E}}
-\renewcommand{\b}[1]{{\bf #1}}
-\DeclareMathOperator*{\argmin}{arg\,min}
-\DeclareMathOperator*{\argmax}{arg\,max}
-
-\newcommand{\picturesAB}[6]{
-\centerline{
-\hskip #4
-\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
-\hskip #5
-\epsfig{file=#1,height=#3}
-\hskip #6
-\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
-\hskip #5
-\epsfig{file=#2,height=#3}
-}}
-\newcommand{\picturesCD}[6]{
-\centerline{
-\hskip #4
-\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
-\hskip #5
-\epsfig{file=#1,height=#3}
-\hskip #6
-\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
-\hskip #5
-\epsfig{file=#2,height=#3}
-}}
-
-\makeatletter  
-\newcommand{\xleftrightarrows}[2][]{\mathrel{%  
- \raise.40ex\hbox{$  
-       \ext@arrow 3095\leftarrowfill@{\phantom{#1}}{#2}$}%  
- \setbox0=\hbox{$\ext@arrow 0359\rightarrowfill@{#1}{\phantom{#2}}$}%  
- \kern-\wd0 \lower.4ex\box0}}
-
-\newcommand{\xrightleftarrows}[2][]{\mathrel{%  
- \raise.40ex\hbox{$\ext@arrow 3095\rightarrowfill@{\phantom{#1}}{#2}$}%  
- \setbox0=\hbox{$\ext@arrow 0359\leftarrowfill@{#1}{\phantom{#2}}$}%  
- \kern-\wd0 \lower.4ex\box0}}
-
-\def\leftrightarrowfill@{%
- \arrowfill@\leftarrow\relbar\rightarrow%
- }
-\newcommand*{\centerfloat}{%
-  \parindent \z@
-  \leftskip \z@ \@plus 1fil \@minus \textwidth
-  \rightskip\leftskip
-  \parfillskip \z@skip}
-\makeatother
-\makeatother
-
-\newcommand\irregularcircle[2]{% radius, irregularity
-  \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
-  +(0:\len pt)
-  \foreach \a in {10,20,...,350}{
-    \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
-    -- +(\a:\len pt)
-  } -- cycle
-}
-
-\newcommand{\into}{\operatornamewithlimits{\longrightarrow}}
-\newtheorem{dfn}{Definition}[section]
-
-\newcommand{\ltri}{%
-\,\resizebox{!}{0.25\baselineskip}{%
-\begin{tikzpicture}%
-\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
-\end{tikzpicture}%  
-}\xspace%
-}%
-
-\newcommand{\smallltri}{%
-\,\resizebox{!}{0.15\baselineskip}{%
-\begin{tikzpicture}%
-\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
-\end{tikzpicture}%  
-}\xspace%
-}%
-
-\author{Simon L. Cotter\thanks{School of
-    Mathematics, University of Manchester, Manchester, UK. e:
-    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
-    grant award EP/L023393/1. SLC would like to thank the Isaac Newton
-    Institute for Mathematical Sciences for support and hospitality
-    during the programme ``Uncertainty quantification for complex systems: theory and methodologies'' when work on this paper was undertaken. This work was supported by:
-EPSRC grant number EP/K032208/1.} \and Ioannis
-G. Kevrekidis\thanks{Chemical and Biomolecular Engineering \& Applied
-  Mathematics and Statistics, Johns Hopkins University, Baltimore, MD
-  21218. The work of IGK was partially supported by the US National
-  Science Foundation, by the US AFOSR, and by DARPA.} \and Paul
-  Russell\thanks{School of
-    Mathematics, University of Manchester, Manchester, UK.}}
-
-\title{Transport map accelerated adaptive importance sampling, and application to inverse problems arising from
-  multiscale stochastic reaction networks}
-\begin{document}
-\maketitle
-\begin{abstract}
-In many applications, Bayesian inverse problems can give rise to
-probability distributions which contain complexities due to the
-Hessian varying greatly across parameter space. This complexity often manifests itself as lower
-dimensional manifolds on which the likelihood function is
-invariant, or varies very little. This can be due to trying to infer
-unobservable parameters, or due to sloppiness in the model which is
-being used to describe the data. In such a situation, standard
-sampling methods for characterising the posterior distribution, which
-do not incorporate information about this structure, will be highly
-inefficient.
-
-In this paper, we seek to develop an approach to
-tackle this problem when using adaptive importance sampling methods,
-by using optimal transport maps to simplify
-posterior distributions which are concentrated on lower dimensional
-manifolds. This approach is applicable to a whole range of
-problems for which Monte Carlo Markoc chain (MCMC) methods mix slowly.
-
-We demonstrate the approach by considering inverse problems arising
-from partially observed stochastic reaction networks. In particular,
-we consider systems which exhibit multiscale behaviour, but for which
-only the slow variables in the system are observable. We demonstrate
-that certain multiscale approximations lead to more consistent
-approximations of the posterior than others. The use of optimal
-transport maps stabilises the ensemble transform adaptive importance
-sampling (ETAIS) method, and allows for efficient sampling with
-smaller ensemble sizes.
-\end{abstract}
-
-\section{Introduction}
-The importance of Markov chain Monte Carlo (MCMC) methods is becoming
-increasingly apparent, in a world replete with datasets which need to be
-combined with complex dynamical systems in order for us to be able
-make progress in a range of scientific fields. Different MCMC methods
-have been designed with different challenges in mind, for example high dimensional
-targets\cite{cotter2013mcmc}, or large data
-sets\cite{bierkens2016zig}. New types of stochastic processes
-have been considered, such as piecewise deterministic Markov processes
-\cite{bouchard2018bouncy}.
-Other methods are able to make very large moves informed by the
-data\cite{teh2016consistency,duane1987hybrid}, and others are able to exploit
-the geometry of the target\cite{girolami2011riemann}. Population Monte Carlo
-(pMC) methods use efficient importance sampling proposals informed by an
-ensemble of particles which have already learned about the structure
-of the target\cite{cappe2012population,cappe2008adaptive,douc2007convergence,douc2007minimum, martino2017layered, cornuet2012adaptive, martino2015adaptive}.
-These methods are particularly useful in the context of
-low dimensional target distributions which have complex\footnote{We
-  use ``complex'' with the meaning ``complicated'' rather than
-  containing imaginary numbers.} structure, for
-which traditional Metropolis-Hastings methods tend to struggle, for
-example when the target is multimodal or if the density is
-concentrated on a lower-dimensional manifold. Moreover, they are an
-excellent candidate for parallelisation, since the task of computing
-the likelihood for each particle in the ensemble can be distributed
-across a set of processors. This is of utmost importance in an age
-where increases in computational power are most likely to come from
-large-scale parallelisation as opposed to the large speed-ups in
-individual processors that we saw throughout the 20th century.
-
-In \cite{cotter2015parallel}, we explored ideas from adaptive MCMC
-within adaptive importance samplers, in order to construct a method which could automatically adapt to optimise
-efficiency. Alongside this, we explored the use of state-of-the-art
-resamplers\cite{reich2013nonparametric}, and the multinomial transform
-(MT), a greedy approximation of the ensemble transform method, in order to improve the quality of
-the importance sampler proposal distribution.
-
-Other approaches to challenging posterior structure have been
-proposed, including the work of Marzouk and Parno\cite{parno2014transport}, who
-demonstrated how invertible transport maps can be constructed which
-map the target distribution close to an easily-explored reference
-density, such as a $d$-dimensional Gaussian. Since the map is
-invertible, Gaussian proposal densities on the reference space can be
-mapped to highly informed moves on the parameter space. A related idea
-is the use of the anisotropic diffusion maps\cite{singer2008non}, in
-which a map which describes the manifold is constructed using
-principle component analysis. These ideas have also been applied to
-stochastic reaction
-networks\cite{singer2009detecting,dsilva2016data}. In
-\cite{chiavazzo2017intrinsic}, these maps were used to propagate
-data-mining into unexplored regions which were predicted to be close
-to the lower-dimensional manifold. The idea is similar to that of the
-Lamperti or Girsanov transformations for SDEs\cite{flandrin2003stationarity,gardiner2009stochastic},  the aim being to map complex
-processes onto a standard one which is well understood, through a
-change of measure.
-
-In this paper, we particularly concern ourselves with the subject of
-inverse problems in stochastic reaction networks. This type of model
-is used in situations where ODE or diffusion\cite{gillespie2000chemical}
-approximations of chemical reaction kinetics are not viable, usually
-because of the low concentration of one or more chemical
-species\cite{gillespie1977exact}. Biochemical networks typically fall into
-this category, because of the small volume of the reactors involved (a
-single cell).
-
-One challenge with networks arising from these applications is that
-they often have some reactions which are occurring many times on a
-timescale in which other reactions in the system are unlikely to
-occur. This multiscale behaviour throws up two main challenges, both
-in observability in an experimental setting, but also in the context
-of modelling, where multiscale systems can make the simulation of the
-continuous time Markov chain computationally intractable.
-
-A range of numerical methods have been suggested to tackle this second
-problem, including tau-leaping\cite{rathinam2003stiffness,cao2006efficient}, diffusion approximations\cite{erban2006gene,cotter2011constrained,cotter2016error}, quasi-steady state
-approximations\cite{weinan2007nested,cao2005slow}, and other averaging
-arguments\cite{cotter2016constrained}. Some of these approaches are more
-computationally intensive than others, but this can be mitigated with the
-use of analytical results about the invariant densities of certain
-types of fast subsystems\cite{jahnke2007solving,anderson2010product,anderson2016product}.
-
-We would not expect the fast processes in such a system to be
-experimentally observable. In such a situation where only the slow
-variables in a system are identifiable, in an ideal world, we would be
-be able to integrate over the space of all possible trajectories of
-the fast processes. Since this is an insurmountable computational
-task, we might instead wish to approximate the likelihoods arising
-from such a set of observations by the use of an appropriate
-multiscale methodology.
-
-The partial observability of these systems lead to only
-partial-observability of the unknown parameters in the system, for
-example the reaction rates of the fast reactions. The resulting
-multiscale approximations, and the resulting approximations of the
-likelihoods arising from the inverse problem for such a network, 
-often exhibit 
-insensitivity to changes in certain directions of
-parameter space. This causes problems when aiming to sample from such
-a target, since the density is concentrated on a manifold whose
-tangential vectors point in directions of insensitivity. This
-\emph{sloppiness}\cite{gutenkunst2007universally,apgar2010sloppy} of the likelihood is a challenging problem
-that appears in many applications in science and engineering\cite{constantine2014active}.
-
-In this paper, we aim to incorporate transport maps into an ensemble
-importance sampling setting, in order to develop stable algorithms which are
-capable of sampling from probability distributions with complex
-structures. These maps are constructed using an importance sample of
-the posterior distribution, and are used to simplify the posterior
-distribution, mapping them as close as possible to a reference
-Gaussian measure. Through this map, straightforward Gaussian proposal
-densities are transformed into complex densities, which match as well
-as possible the target density. Less well-informed proposal kernels means that complex
-structures in the density can only be well represented with a large
-increase in the ensemble size, increasing computational cost. If
-computational resources do not allow for such an increase, this can
-lead to stability issues in the algorithm, due to the proposal
-distribution in the importance sampler not representing well the
-target density, leading to high variance in the weights, causing slow
-convergence. Better informed proposal densities
-leads to more stable and faster converging importance sampling
-schemes, and we will demonstrate this through examples
-arising from Bayesian inverse problems on multiscale stochastic
-reaction networks.
-
-%Topics to cover: parallel MCMC, pMC, PAIS
-%Transport maps, transport map MCMC
-%Stochastic reaction networks
-%Multiscale approximations, QSSA/QEA, CMA
-
-In Section \ref{sec:ETAIS}, we will briefly reintroduce the ensemble
-transport adaptive importance sampling (ETAIS) method. In Section \ref{sec:map} we show how an appropriate transport map can
-be constructed from importance samples which maps the posterior close
-to a reference Gaussian measure. In Section \ref{sec:TETAIS} we show
-how such a map can be incorporated into a sophisticated parallel MCMC
-infrastructure in order to accelerate mixing, and improve stability. In Section
-\ref{sec:conv} we seek to show the advantages of this approach through
-the analysis of a challenging test problem. In Section
-\ref{sec:multi} we consider how likelihoods can be approximated using
- multiscale methodologies in order to carry out inference for
-multiscale and/or partially observed stochastic reaction networks. In
-Section \ref{sec:num} we present some numerical examples, which serve
-to demonstrate the increased efficiency of the described sampling
-methodologies, as well as investigating the posterior approximations
-discussed in the previous section. We conclude with a discussion in
-Section \ref{sec:conc}.
-
-\section{Ensemble Transport Adaptive Importance Sampling}\label{sec:ETAIS}
-Ensemble transport adaptive importance sampling
-(ETAIS)\cite{cotter2015parallel} is an adaptive importance sampling framework
-which uses an ensemble of particles and state-of-the-art resampling
-methods to construct a mixture proposal distribution which closely
-matches the target distribution.
-
-In importance sampling, we aim to sample from a target $\pi$, by
-sampling from a chosen proposal distribution $\chi$. Each sample $\theta^{(k)} \sim \chi$ is then weighted by
-$w_k = \frac{\pi(\theta^{(k)})}{\chi(\theta^{(k)})}$, to take account of the bias of
-sampling from a different distribution to $\pi$. Monte Carlo estimates
-using a sample of size $N$
-of a function $f$ with respect to $\pi$ can then be made through the
-formula
-\[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
-  w_kf(\theta^{(k)}),\]
-where $\bar{w} = \sum_{k=1}^N w_k$ is the total weight.
-This method works well when $\pi$ and $\chi$ are close, but can be
-excruciatingly slow when they are not. The idea behind adaptive
-importance sampling methods is
-to construct a good proposal distribution, either from the entire
-history of the algorithm up to the current point, or to use the
-current state of a whole ensemble of $M$ particles in the system.
-
-In ETAIS, the proposal distribution $\chi$ is chosen to be the equally
-weighted mixture of set of proposal kernels. If $\theta^{(k)} = [\theta_1^{(k)},
-\theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
-ensemble, and we wish to use a mixture of the proposal density $q(\cdot ;
-\cdot, \beta)$ then 
-\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)},
-\beta).\]
-Usually the kernel $q(\cdot ;
-x,\beta)$ is centred at, or has mean equal to $x$, and whose variance
-scales with $\beta$. Good values for the
-algorithmic parameter(s) $\beta$ can be
-found by optimising for the effective sample size of the importance
-sample that is produced (see \cite{cotter2015parallel, russ2017parallel} for more details).
-
-If the ensemble is large enough, and the chain has entered
-probabilistic stationarity, then the current state of the ensemble is
-a good rough discrete approximation of the target density, and in turn
-$\chi^{(k)}$ is close enough to $\pi$ to produce an efficient importance
-sample $\{(\hat{\theta}^{(k)},w^{(k)})\}$,  where $\theta^{(k)} =
-[\theta_1^{(k)}, \ldots, \theta_M^{(k)}]^\top$ and $w^{(k)} =
-[w_1^{(k)}, \ldots, w_M^{(k)}]^\top$. It can be advantageous to use stratified sampling of the
-mixture in order to
-ensure that the sample made is as representative as possible of the
-target density, i.e.
-\[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
-for each $i = 1,2,\ldots,M$.  We now have a weighted sample, and it
-would be inadvisable to use an equally weighted mixture of proposal
-distributions from each of these points. Therefore, before starting
-the next iteration, the importance sample
-$\{(\theta^{(k)},w^{(k)})\}$ is resampled to produce an equally
-weighted sample, ready for the next iteration of the algorithm. In
-ETAIS, a state-of-the-art resampler is used, which uses optimal
-transport methods to find the equally weighted discrete sample which
-best represents the statistics of the importance
-sample\cite{reich2013nonparametric}. For larger ensemble sizes $M$ this can become
-expensive, in which case a greedy approximation of this algorithm, the
-Multinomial Transformation (MT) can be
-implemented\cite{cotter2015parallel}. The output of the resampler is then denoted
-$\theta^{(k+1)}$ and the algorithm is ready for the next
-iteration. The ETAIS is summarised in Algorithm \ref{alg:ETAIS}, where
-the importance samples $\{(\hat{\theta}^{(k)},w^{(k)})\}$ are stored as
-the output.
-
-\begin{table}[!h]
-\centering
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\BlankLine
-	Initialise $\theta^{(1)} \sim \pi_0$.\;
-	\For{$k=1,\ldots,N$}{
-		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_j^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:ETAIS_propose}\;
-		Calculate $w^{(k)} = (w_1^{(k)}, \ldots, w_M^{(k)})^\top$, where
-			\[
-				w^{(k)}_i = \frac{\pi(\hat{\theta}^{(k)}_i)}{\chi^{(k)}(\hat{\theta}^{(k)}_i; \theta^{(k)}, \beta)}.
-			\]
-		\label{algline:ETAIS_weights}
-
-		Resample $\theta^{(k+1)} \leftarrow \| w^{(k)}\|_1^{-1}\sum_{j=1}^M w_j^{(k)}\delta_{\hat{\theta}_j^{(k)}}(\cdot)$.\label{algline:ETAIS_resample}\;
-	}
-	Output $\{(\hat{\theta}^{(n)},w^{(n)})\}_{n=1}^N$.
-\caption{The ETAIS Algorithm.\label{alg:ETAIS}}
-\end{algorithm}
-\end{table}
-
-One problem with the ETAIS and other similar methods can become apparent if
-the target density has highly curved manifolds on which the density is
-concentrated. In this
-case, unless the proposal densities $q$ are informed by this local
-structure, the mixture distribution proposal may not well approximate
-$\pi$ without a very large ensemble size $M$, which can become
-prohibitively expensive. Some methods have been
-proposed\cite{douc2007minimum}
-which use samples local to each particle to inform local covariance
-structure.
-
-In this paper, we investigate the use of transport maps to learn local
-covariances across the whole of the domain, in order to stabilise
-adaptive importance sampling methods, and make these methods more applicable to a wider
-range of more challenging inference problems.
-
-\section{Construction of transport maps in importance sampling} \label{sec:map}
-%Follow the other papers, show adaptation(s)
-In this Section, we describe the construction of transport maps which
-allow for the simplification of complex posterior distributions in
-order to allow for improved sampling, in particular for methods based
-on importance sampling.
-In~\cite{el2012bayesian} the transport map was introduced to provide a transformation from the prior
-distribution to the posterior distribution, the idea being that one could draw a moderately sized
-sample from the prior distribution and use this sample to approximate a map onto the target space.
-Once this map was known to the desired accuracy, a larger sample from the prior could be used to
-investigate the posterior distribution. This
-methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
-algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
-distribution, the map transforms a sample from the posterior onto a reference space.
-The reference density is chosen to allow efficient proposals using a simple proposal
-distribution such as a Gaussian centred at the previous state. Proposed states can then be mapped back into a sample from the posterior by applying the inverse of the transport map.
-
-Proposing new states in this way allows us to make large, informed
-moves, even when sampling from probability distributions with complex structure.
-If a very good such map can be found, it is also possible that the
-push forward of the posterior density through the map is close enough
-to the reference Gaussian that we can efficiently
-propose moves using a proposal distribution which is independent of
-the current state, for example sampling from the reference Gaussian
-itself, or a slightly more diffuse distribution.
-
-In this Section we outline the methodology in
-\cite{parno2014transport} for approximately coupling the target
-measure with density
-$\pi_{\theta}$, with the reference distribution with density $\pi_r$, and
-show how the map can be constructed using a weighted sample
-and hence how we can incorporate the map into importance sampling schemes.
-
-\begin{dfn}[Transport Map $T$]
-	A transport map $T$ is a smooth function $T\colon
-        \mathcal{X}\rightarrow\mathbb{R}^d$ such that the {\it
-          pullback} of the reference measure with density $\phi(\cdot)$,
-	\begin{equation}\label{eq:pullback}
-		\tilde{\pi}(\theta) = \phi(T(\theta))|J_T(\theta)|,
-	\end{equation}
-	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
-	\[
-		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
-	\]
-\end{dfn}
-
-% In the case that we have an invertible map,
-% $T\in\mathcal{T}$ where $\mathcal{T}$ is the space of all invertible
-% maps, we are able to draw a sample from $\pi_r = \phi$, the density of
-% the reference distribution, which could be picked to be the
-% standardised Gaussian distribution for example, and map these samples back
-% onto target space using $T^{-1}$. These proposed samples are then distributed according to the target distribution.
-
-\begin{dfn}[Target and Reference Space]
-	The transport map pushes a sample from a {\it target space} $\mathcal{X}$, that is a subset of $\mathbb{R}^d$ equipped with a target density $\pi_{\theta}$, onto a {\it reference space}, $R$, again a subset of $\mathbb{R}^d$ equipped with the reference density $\pi_r$.
-\end{dfn}
-
-Armed with such a map, independent samples can be made of the target
-measure, using the pullback of the reference density $\phi$ through $T^{-1}$.
-Clearly the pullback only exists when $T$ is monotonic, i.e. has a positive definite Jacobian, and has continuous first derivatives.
-Not all maps satisfy these conditions, so we define a smaller space of
-maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all
-invertible maps. This space does not necessarily contain an exact
-coupling between target and reference space, and so we are motivated to formulate an optimisation problem to find the map $\tilde{T}
-\in \mathcal{T}^\uparrow$ which most closely maps the target density
-to the reference density.
-
-% We attempt to find the deterministic coupling of two continuous
-% probability distributions, $(\mu_\theta, \hat{\mu_r})$, such that
-% $\hat{\mu_r} = T\mu_\theta$ where the distance between $\mu_r$ (the
-% desired reference measure) and $\hat{\mu_r}$ (the achieved reference
-% measure) is minimised, for $T \in \mathcal{T}^\uparrow$. As in
-% \cite{parno2014transport} we aim to minimise the Kullback-Liebler (KL)
-% divergence between the density of $\mu_{\theta}$ and the pullback of
-% the density of $\mu_{r}$, i.e. the distance between $\pi(\theta)$ and
-% $\tilde{\pi}(\theta)$. For two absolutely continuous measures with
-% densities $\pi_1$ and $\pi_2$ respectively, the KL
-% divergence is given by
-% \[D_\text{KL}(\pi_1\|\pi_2) = 
-% 		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\]
-% The KL divergence is not itself a norm, since
-% it is not symmetric, i.e. $D_\text{KL}(\pi_1\|\pi_2) \neq
-% D_\text{KL}(\pi_2\|\pi_1)$ in general. However, it is still a useful
-% measure of the similarity of two probability distributions, not least since the
-% square root of the KL-divergence is an upperbound to the Hellinger
-% distance metric.
-
-As in previous work in \cite{parno2014transport}, we can ensure
-invertibility if we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
-\[
-	\tilde{T}(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
-		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
-\]
-where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. % We assume that the target and reference probability densities are absolutely continuous on
-% $\mathbb{R}^d$. Under this formulation, with appropriate
-% regularisation (which will be detailed later), we are guaranteed a unique invertible map $\tilde{T}$ with the property that
-% \[
-% 	\tilde{T}\mu_{\theta} \approx \mu_r.
-% \]
-% % Relaxing the equality constraint to finding the approximate map
-% % $\tilde{T} \in \mathcal{T}^{\ltri}$ which minimises $C(T)$, gives us a
-% % practical route to finding a good candidate map.
-
-\subsection{The optimisation problem}
-Our aim is now to find the lower triangular map $\tilde{T} \in
-\mathcal{T}^{\ltri}$ such that the difference between target
-density and the pullback of the reference density is minimised. As in
-\cite{parno2014transport}, we choose the cost function
-to be the Kullback-Leibler (KL) divergence between the posterior density and the pullback density,
-\[
-	D_\text{KL}(\pi\|\tilde{\pi}) =
-		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
-\]
-This divergence results in some nice properties which we will explore in the following derivation.
-The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to its relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
-\[
-	D_{KL}(p\|q) \geq D_H^2(p,q),
-\]
-which is a true distance metric on probability distributions.
-Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
-\[
-	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
-		\log\left|J_{\tilde{T}}(\theta)\right|\right].
-\]
-We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
-This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown.
-However this is exactly the scenario for which we would turn to MCMC methods for a solution.
-
-To find the best coupling, $\tilde{T} \in \mathcal{T}^{\ltri}$, we solve the optimisation problem,
-\[
-	\tilde{T} = \arg\min_{T \in \mathcal{T}^{\smallltri}} \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
-		\log\left|J_T(\theta)\right|\right]
-\]
-which has a unique solution since the cost function is convex.
-We also include a regularisation term, which is required for reasons which will become clear later.
-The optimisation problem now takes the form
-\begin{equation}\label{eq:gen_map_optim}
-	\tilde{T} = \arg\min_{T\in\mathcal{T}^{\smallltri}} \left[
-		 \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
-		\log\left|J_T(\theta)\right|\right] + \beta\mathbb{E}(T(\theta)- \theta)^2 \right].
-\end{equation}
-The parameter $\beta>0$ does not need to be tuned, as experimentation has shown that the choice
-$\beta=1$ is sufficient for most problems. This expectation can be
-approximated by using an MCMC approximation. The form of the penalisation term promotes maps which are
-closer to the identity, and so prevents overfitting when the quality
-or size of the current sample from the posterior is not sufficient.
-
-\subsection{The structure of the map}
-
-Before we continue with the derivation of the optimisation problem, we consider the structure
-of the map in more detail.
-The lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density, as well as the inverse of the map, $\tilde{T}^{-1}$.
-The Jacobian of $\tilde{T}$ is a lower triangular matrix,
-\[
-	J_T(\theta) = \begin{bmatrix}
-		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & \partial_{\theta_d} \tilde{T}_1(\theta)\\
-		\vdots & \ddots & \vdots \\
-		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
-	\end{bmatrix} = \begin{bmatrix}
-		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & 0\\
-		\vdots & \ddots & \vdots \\
-		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
-	\end{bmatrix}
-\]
-since $\partial_{\theta_n} \tilde{T}_k(\theta) = 0$ for all $n > k$.
-This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
-\begin{equation}\label{eqn:separable_jacobian}
-	\log\left|J_{\tilde{T}}(\theta)\right| = \sum\limits_{i=1}^d \! \log \partial_{\theta_i} \tilde{T}_i(\theta),
-\end{equation}
-where we note that this term is separable in terms of the dimension $i$.
-
-Inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map.
-The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, so we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$.
-This inversion tells us the value of $\theta_1$, which means the next component is again a univariate polynomial, $T_2(\theta_2; \theta_1)=r_2$.
-We can then perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.
-
-We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions.
-Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not necessarily orthogonal with respect to the target distribution.
-Each component of the map is defined as a multivariate polynomial expansion,
-\begin{equation}\label{eq:map_defn}
-	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
-\gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
-\end{equation}
-The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
-$\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in
-\mathcal{J}_i \subset \mathbb{N}_0^i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
-\[
-	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
-\]
-and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$.
-Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.
-
-The cardinalities of the multi-index sets, $M_i = \text{card}(\mathcal{J}_i)$, give the number of unknowns in our
-optimisation problem, and so we would like to keep this number as small as possible. One option is
-to use polynomials of total order $p$,
-\[
-	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
-\]
-which is optimal in terms of the amount of information captured by the map about the target.
-The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = \binom{i+p}{p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$.
-Smaller optimisation problems can be produced by constructing subsets of $\mathcal{J}_i^\text{TO}$.
-These index sets are discussed
-in~\cite{parno2014transport}.
-Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}.
-Here we stick with polynomials of total order $p$ since we work with low dimensional problems with the ETAIS algorithm.
-
-\subsection{Implementation of the optimisation problem}\label{sec:transport_implementation}
-
-We now discuss how we can evaluate the cost function in
-Equation~\eqref{eq:gen_map_optim}. In \cite{parno2014transport}, this
-expectation is approximated using an MCMC estimator, such that
-\begin{align}
-	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
-			\beta\mathbb{E}(T(\theta)-\theta)^2 \notag \\
-		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
-			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. 
-\end{align}
-Here we diverge from previous work, as we aim to build a map from
-samples from an importance sampling scheme. Such samples no longer
-carry equal weight, and as such the Monte Carlo estimator becomes
-\begin{equation}\label{eqn:TM_full_cost}
-	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
-		w_k \left[-\log\pi_r(T_i(\theta^{(k)})) -
-			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
-\end{equation}
-where $w_k$ are the weights associated with each sample $\theta^{(k)}$, and $\bar{w}$ is the sum of
-all these weights.
-Optimisation of this cost function results in a map from $\pi$ to some reference density $\pi_r$.
-By choosing the reference density to be a Gaussian density, we can simplify this expression greatly.
-Substitution of the Gaussian density into Equation~\eqref{eqn:TM_full_cost} leads to
-\begin{equation}\label{eqn:TETAIS_objective}
-	C(T) = \frac{1}{\bar{w}}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K
-		w_k\left[\frac{1}{2}T_i^2(\theta^{(k)}) - \log\frac{\partial
-		T_i}{\partial\theta_i}(\theta^{(k)}) + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right],
-\end{equation}
-
-Note that since we assume that the map is monotonic, the derivatives of each component are
-positive and so this functional is always finite.
-In practice it is infeasible to enforce this condition across the whole parameter space.
-We instead enforce this condition by ensuring that the derivatives are positive at each sample point.
-This means that when we sample away from these support points while in reference space, it is possible to enter a region of space where the map is not monotonic.
-
-We now return to the structure of the map components given in Equation~\eqref{eq:map_defn}. Since the basis functions are
-fixed, the optimisation problem in \eqref{eq:gen_map_optim} is really over the map components $\bar{\gamma} = (\gamma_1, \dots,
-\gamma_d)$ where $\gamma_i \in \mathbb{R}^{M_i}$.
-Note that $C(T)$ is the sum of $d$ expectations, and these expectations each only concern one dimension.
-Therefore we can rewrite \eqref{eq:gen_map_optim} as $d$ separable optimisation problems.
-\begin{align}\label{eq:gamma_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{\bar{w}}\sum\limits_{k=1}^K
-		w_k \left[\frac{1}{2}T_i^2(\theta^{(k)}; \gamma_i) - \log\frac{\partial T_i}{\partial\theta_i}(\theta^{(k)}; \gamma_i) + \beta(T_i(\theta^{(k)};
-		\gamma_i)-\theta^{(k)})^2\right], \\
-	&\text{subject to} \quad \frac{\partial T_i}{\partial\theta_i}(\theta^{(k)};
-		\gamma_i) > 0 \ \text{for all}\ k=1,\dots,K,\ i=1,\dots,d.
-		\notag
-\end{align}
-The sum in Equation~\eqref{eq:map_defn} is an inner
-product between the vector of map coefficients, and the evaluations of the basis function at a
-particular $\theta^{(k)}$. If we organise our basis evaluations into two matrices,
-\[
-	(F_i)_{k,\mathbf{j}} = \psi_\mathbf{j}(\theta^{(k)}), \quad \text{and} \quad (G_i)_{k,\mathbf{j}} =
-\frac{\partial\psi_\mathbf{j}}{\partial\theta_i}(\theta^{(k)}),
-\]
-for all $\mathbf{j}
-\in \mathcal{J}_i^\text{TO}$, and $k = 1,\dots,K$, then we have that
-\[
-	T_i(\theta^{(k)}) = (F_i)_{k\cdot}\gamma_i \quad \text{and} \quad \frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}; \gamma_i) = (G_i)_{k\cdot}\gamma_i,
-\]
-so \eqref{eq:gamma_map_optim} becomes
-\begin{align}\label{eq:blas_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}}
-          \frac{1}{2}(F_i\gamma_i)^\top W (F_i\gamma_i) -
-		{\bf w}^\top\log(G_i\gamma_i) + \frac{\beta}{\bar{w}}\sum\limits_{k=1}^K \!
-		w_k(F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)}), \\
-	&\text{subject to} \quad G_i\gamma_i > 0. \notag
-\end{align}
-In this expression, the vector ${\bf w} = [w_1, w_2, \ldots,
-w_K]^\top$ is the vector of the weights, $W$ is the diagonal matrix $W
-= \text{diag}(w)$ and $\log(G_i\gamma_i)$ is to be
-evaluated element-wise. As more importance samples are made, new rows can be appended to the
-$F_i$ and $G_i$ matrices, and $F_i^\top W F_i$ can be efficiently updated via the addition of rank-1 matrices.
-
-The regularisation term in Equation~\eqref{eq:blas_map_optim} can be approximated using Parseval's identity,
-\[
-	\frac{1}{\bar{w}}\sum\limits_{k=1}^K \! w_k
-        (F_i\gamma_i-\theta^{(k)})^\top(F_i\gamma_i-\theta^{(k)})
-        \xrightarrow[K \to \infty]{}
-		\int_{\mathbb{R}^n} |T(\theta)-\theta|^2 \text{d}\pi_\theta =
-		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
-\]
-where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
-the polynomial family $\mathcal{P}(\theta)$ is chosen to be orthonormal with respect to $\pi_\theta$; however this
-approximation prevents the map from collapsing onto a Dirac when the expectation is badly approximated by a small number of samples.% If we do not normalise the MC estimator by $K$, we can allow this regularisation term to be dominated by the rest of the cost function as $K$ increases.
-
-These simplifications result in the efficiently implementable, regularised optimisation problem for
-computing the map coefficients, 
-\begin{align}\label{eq:weighted_map_optim}
-	&\arg\min_{\gamma_i\in\mathbb{R}^{M_i}} \frac{1}{2\bar{w}}\gamma_i^\top F_i^\top WF_i\gamma_i -
-		\frac{w^\top}{\bar{w}}\log(G_i\gamma_i) + \beta\|\gamma_i-\iota\|^2, \\
-	&\text{subject to} \quad G_i\gamma_i > 0. \notag
-\end{align}
-%where additionally we have introduced a factor of $\frac{1}{K}$ to the
-%regularisation term to prevent over-smoothing as the number of samples increases.
-This optimisation problem can be efficiently solved using Newton iterations. It is suggested
-in~\cite{parno2014transport} that this method usually converges in around 10-15 iterations, and we
-have seen no evidence that this is not a reasonable estimate. When calculating the map several times
-during a Monte Carlo run, using previous guesses of the optimal map to seed the Newton algorithm
-results in much faster convergence, usually taking only a couple of iterations to satisfy the stopping
-criteria.
-
-The Hessian takes the
-form
-\begin{equation}\label{eqn:TETAIS_hessian}
-	HC_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i + G_i^\top
-		W\text{diag}([G_i\gamma_i]^{-2})G_i\right] + \beta I,
-\end{equation}
-where $[G_i\gamma_i]^{-2}$ is to be taken element-wise, and $I$ is the $M_i\times M_i$
-identity matrix. The first derivative of $C_i(T)$ is
-\[
-	\nabla C_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i\gamma_i - G_i^\top
-		W[G_i\gamma_i]^{-1}\right] + \beta(\gamma_i - \iota),
-\]
-again $[G_i\gamma_i]^{-1}$ is taken element-wise.
-
-\section[Transport map MCMC]{Transport map usage in ETAIS
-  and other importance sampling algorithms}\label{sec:TETAIS}
-
-Given importance samples from the target distribution, we have demonstrated how to construct an approximate transport map from the
-target measure to a reference measure. We now consider how to
-implement an ensemble importance sampling algorithm which uses
-these maps to propose new states. In \cite{parno2014transport} it was
-shown how approximate transport maps can be used to accelerate
-Metropolis-Hastings methods, with the map being periodically updated
-with the samples produced from the target measure. Convergence of this
-adaptation is shown in~\cite{parno2014transport}. In this Section, we
-will show how similarly, these maps can be used to construct highly
-efficient importance sampling schemes.
-
-In particular, we will show how we can use the transport map derived in Equation~\eqref{eq:weighted_map_optim} to
-design a proposal scheme for the ETAIS algorithm. In this case we have a choice in how to proceed; we
-propose new samples on reference space and resample on target space, or we both propose and resample on reference space, mapping onto target space to output the samples.
-The first option allows us to reuse much of the framework
-from the standard ETAIS algorithm and in the numerics later we see that this performs better than
-both the Transport MH algorithm, and the standard ETAIS algorithm. The second option requires some
-restructuring, but can result in improved performance from the resampler.
-
-\begin{table}
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\BlankLine
-Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
-Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
-\For{$k \leftarrow 1, \dots, L-1$}{
-	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
-	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
-	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
-	Calculate:
-	\[
-		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
-	\]
-
-	Resample $\theta^{(k+1)} \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{\hat{\theta}^{(k)}_j}(\cdot)$.\;
-
-	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
-		\For{$i \leftarrow 1, \dots, n$}{
-			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
-				and update $\gamma_i^{(k+1)}$.\;
-		}
-	}{
-		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
-	}
-}
-\caption{ETAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportETAIS1}}
-\end{algorithm}
-\end{table}
-
-The first option is given in Algorithm~\ref{alg:TransportETAIS1}. We
-denote the ensembles of states in target space $\theta^{(k)} =
-\{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the
-reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble
-size. Similarly, the proposal states are denoted $r' =
-\{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) =
-\{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M,
-\hat{\theta}^{(k)}_M)\}$, where these pairs are the states which
-together form our sample from the target distribution. As in the standard version of the ETAIS algorithm we use the deterministic mixture weights.
-
-\begin{table}
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\BlankLine
-Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
-Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
-\For{$k \leftarrow 1, \dots, N-1$}{
-	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
-	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
-	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
-	Calculate:
-	\[
-		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
-	\]
-
-	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\label{algline:TETAIS_resample}\;
-	Invert $\theta^{(k+1)}_i = \tilde{T}^{-1}(r^*_i)$.\;
-	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
-		\For{$i \leftarrow 1, \dots, n$}{
-			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
-				and update $\gamma_i^{(k+1)}$.\;
-		}
-	}{
-		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
-	}
-}
-\caption{ETAIS algorithm with adaptive transport map. Option 2.\label{alg:TransportETAIS2}}
-\end{algorithm}
-\end{table}
-
-The second option, Algorithm~\ref{alg:TransportETAIS2}, is similar to
-the first except on Line~\ref{algline:TETAIS_resample} where rather
-than resampling in target space we resample in reference
-space. Assuming a good transport map, the push forward of the target
-measure onto 
-reference space has no strong correlations, and the
-Gaussian marginals are easy to approximate with fewer ensemble
-members. This means that the resampling step will be more efficient in
-moderately higher dimensions, which we discuss in Section~\ref{sec:TETAIS_higher_dim}.
-
-\section[Convergence of transport MCMC]{Convergence of the transport proposal based MCMC algorithms}\label{sec:conv}
-
-In this Section we study the convergence of the transport based
-proposal distributions which we have described in
-Section~\ref{sec:TETAIS}.
-We take as a test problem the ubiquitous Rosenbrock banana-shaped
-density. This target density is given by
-\begin{equation}\label{eqn:R2_repeat}
-	\pi(\theta) = \frac{\sqrt{10}}{\pi}\exp\left\{ -(1 - \theta_1)^2 - 10(\theta_2 - \theta_1^2)^2 \right\}.
-\end{equation}
-A contour plot of the target density is given in Figure~\ref{fig:R2_posterior}.
-\begin{figure}
-\centering
-\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_1"}}
-\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_2"}}
-\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_posterior}}
-\caption{Visualisation of the Rosenbrock density as given in Equation~\eqref{eqn:R2_repeat}.}
-\label{fig:R2_posterior}
-\end{figure}
-This problem is challenging to sample from since it has a highly peaked and curved ridge, and is often used
-as a test problem in optimisation and MCMC communities.
-
-\subsection{Implementation details}
-
-Before looking at the performance of the algorithms, we demonstrate some properties of the transport maps we will be using.
-We draw 1 million samples from the density \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:map} to build a transport map.
-We use this map to push forward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian.
-We then pull the sample back on to target space using the inverse map to check that our map is invertible and well behaved.
-
-For this example, we use an index set of total order 3 with monomial
-basis functions. It is important that total order is an odd number,
-since otherwise the map will not be surjective. This results in a map of the form
-\[
-	T(\theta_1, \theta_2) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \end{bmatrix},
-\]
-where
-\begin{align*}
-		T_1(\theta_1) &= \gamma_{1,1} + \gamma_{1,2}\theta_1 + \gamma_{1,3}\theta_1^2 + \gamma_{1,4}\theta_1^3, \\
-		T_2(\theta_1, \theta_2) &= \gamma_{2,1} + \gamma_{2,2}\theta_1 + \gamma_{2,3}\theta_1^2 + \gamma_{2,4}\theta_1^3
-					+ \gamma_{2,5}\theta_2 + \gamma_{2,6}\theta_1\theta_2 \\
-				 & \qquad \quad + \gamma_{2,7}\theta_1^2\theta_2 + \gamma_{2,8}\theta_2^2 + \gamma_{2,9}\theta_1\theta_2^2 +
-					 \gamma_{2,10}\theta_2^3.
-\end{align*}
-Clearly even with only basis functions of total order 3, we have a large number of unknowns in our optimisation problem, $\bar{\gamma} \in \mathbb{R}^{14}$.
-If we were to increase the dimension of $\theta$ further we would need to reduce the number of terms we include in the expansion by, for example, removing all the ``cross'' terms.
-This reduces the quality of our map but since we only require an approximate map we can afford to reduce the accuracy.
-
-\begin{figure}[htpb]
-\centering
-\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
-\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
-\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
-\caption{The effect of the approximate transport map $\tilde{T}$ on a
-  sample from the Rosenbrock target density as described in Equation~\eqref{eqn:R2_repeat}.}
-\label{fig:R2_transport}
-\end{figure}
-
-Figure~\ref{fig:R2_transport} shows the output of the approximate
-transport map. Even though we have truncated the infinite expansion in
-the monomial basis down to 4 and 10 terms in respective dimensions,
-the push forward of the sample is still a unimodal distribution
-centred at the origin with standard deviation 1. As one moves out into
-the tails of the reference density more non-Gaussian features are
-clearly visible. However, overall, the push forward of the target
-density does not look a challenging one to sample from, with even
-relatively simple MCMC methods such as Metropolis-Hasting random walk (MH-RW). The pullback from
-reference space, in Figure~\ref{fig:R2_transport}, is an exact match
-of the original sample since we have not perturbed the sample in
-reference space. This inversion is well defined in the sampling
-region, although not necessarily outside\cite{parno2014transport}.
-
-\subsection[Numerical results]{Numerical results for convergence of
-  transport map based algorithms on the Rosenbrock density}
-
-We first find the optimal scaling parameters for the individual
-algorithms. This is done, as in \cite{cotter2015parallel}, by optimising for the effective sample size in
-the ETAIS algorithm, and by tuning the relative $L^2$ error in the MH
-algorithm. There is currently no
-guidance on the best way of tuning the MH algorithm with transport map
-proposals although one might expect results similar to the standard MH
-results, especially if adaptation of the map is stopped after a given
-point. As in the ETAIS algorithm, optimising for the effective sample size might be the best option.
-
-\begin{table}[!ht]
-\centering
-\begin{tabular}{lrrr}
-\toprule
-	Statistic \quad / \quad Algorithm & Transport M-H &
-                                                            Alg. \ref{alg:TransportETAIS1} & Alg. \ref{alg:TransportETAIS2}  \\ \cmidrule(lr){1-4}
-	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
-	$\delta_{\text{ESS}}$				 & - & 1.0e-1 & 5.2e-1 \\ \cmidrule(lr){1-4}
-	Acc. rate							 & 0.23 & - & - \\
-	ESS ratio							 & - & 0.62 & 0.71 \\
-\bottomrule
-\end{tabular}
-\caption{Optimal scaling parameters for the transport map based
-  algorithms applied to $R_1$, optimising for the $L^2$ error and
-  the effective sample size (ESS) for ETAIS algorithms, and for average
-  acceptance rate for MH algorithms.}
-\label{tab:R2_opt_scaling}
-\end{table}
-
-The optimal scaling parameters are given in
-Table~\ref{tab:R2_opt_scaling}. Here we see that the effective sample
-size is much lower than we see in the one-dimensional examples with
-the ETAIS algorithms. However, in the Rosenbrock density
-\eqref{eqn:R2_repeat} we are dealing with a much more complicated
-curved structure, as well as a very slowly decaying tail in
-$\theta_2$.
-From our experiments, we have observed that the standard ETAIS-RW required an ensemble size of $M=500$ to overcome the problems in this density, however the transport map transforms the tails to be more like those of a Gaussian which can be approximated well by a smaller ensemble size of $M=150$.
-
-\begin{figure}[!ht]
-\centering
-\subfigure[Comparison of the two Transport ETAIS
-options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
-\subfigure[Comparison of Transport ETAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
-\caption{Convergence of algorithms; Transport M-H,
-  algorithm \ref{alg:TransportETAIS1}, algorithm \ref{alg:TransportETAIS2} for
-  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the MT algorithm.}
-\label{fig:R2_l2_convergence}
-\end{figure}
-
-The convergence of the three algorithms is displayed in
-Figure~\ref{fig:R2_l2_convergence}. Figure (a) shows that the two
-variations of the transport based ETAIS algorithms converge with
-similar rates. The second version, which performs the resampling stage
-in reference space rather than target space, has a slightly higher
-ESS, and is more stable than option (1). This version also has a
-property that we can exploit in Section~\ref{sec:TETAIS_higher_dim}.
-
-Figure~\ref{fig:R2_l2_convergence} (b) compares the performance of
-RWMH, transport map-accelerated RWMH, ETAIS-RW and transport
-map-accelerated ETAIS-RW. The difference in performance between RWMH
-and its accelerated version is relatively negligible, in part because
-this version of the Rosenbrock is not overtly pathological. The story
-is different for the ETAIS algorithms. The slowly decaying tails of
-the Rosenbrock cause a problem for ETAIS since these are not
-well-represented in the proposal distributions, leading to high weight
-variance and poor convergence. A much larger
-ensemble would be required to ensure a stable ETAIS implementation for this
-problem, which leads to high overhead costs for the resampling step
-and evaluation of the importance weights. The transport map-accelerated
-ETAIS, on the other hand, is stabilised by the map, which transforms the proposal
-distributions in order to adequately approximate the slowly decaying
-tails of the target. This stabilisation allows us to benefit from the
-improved convergence of the ETAIS methodology, with a smaller ensemble
-size.
-
-%\section{Transport map-accelerated Parallel Adaptive Importance
-  %Sampling (TETAIS)}\label{sec:TETAIS}
-
-\section{Multiscale Methods for Stochastic Chemical Reaction
-  Networks}\label{sec:multi}
-In this Section, we discuss some recent advances in multiscale methods
-for stochastic reaction networks. Inverse problems arising in this
-area often lead to complex posterior
-distributions, which traditional MCMC methods can struggle to sample
-from. We will then go on to solve some inverse problems related to
-this in Section \ref{sec:num}, using the transport map versions of the
-ETAIS algorithm, as described in Section \ref{sec:TETAIS}.
-
-We consider chemical reaction networks of $N_s$ chemical species $\{S_j\}_{j=1}^{N_s}$,
-with population numbers given by $X(t) = [X_1(t), X_2(t), \ldots, X_{N_s}(t)]^\top \in
-\mathbb{N}_0^{N_s}$ reacting in a small reactor, through $N_r$ different reaction
-channels. When population numbers of one or more of the chemical
-species in the reactor is small, as is the case with chemical
-reactions occurring within living cells, the sporadic and discrete way
-in which reactions occur can not be well modelled by deterministic
-continuous models, such as ODEs. In such a situation, we may wish to
-model the dynamics through a discrete stochastic model, such as a
-continuous time Markov chain.
-
-For each reaction $R_j$, for $j = 1,2,\ldots N_r$, there is a
-propensity or hazard function $\alpha_j(X(t))$ which indicates how
-likely that reaction is to fire, defined by
-\[\alpha_j(X(t)) = \lim_{dt \to 0} \mathbb{P}(\text{Reaction $R_j$ in
-    the time interval  } \quad s \in [t, t+ dt] ).\] If a system
-satisfies what is called \emph{mass action kinetics}, then the form of
-the function $\alpha_j$ is determined, up to a rate constant, by the
-reactants involved in that reaction:
-\begin{equation}\label{eq:MAK}
-\alpha_j(X) = k_j \prod_{m=1}^{N_s} \prod_{n=0}^{\nu_{j,m} -1} (X_m - n),
-\end{equation}
-where $\nu_{j,m}$ is the $m$th component of the stoichiometric vector
-$\nu_j$, $X_m$ is the $m$th component of the state vector $X$, and the
-$k_j$ are rate constants.
-
-Following each reaction $R_j$ there is an instantaneous change in the
-current state, as the reactants of the reaction are consumed, and the
-products produced. This is modelled by a change in the state vector
-$X(t) = X(t) + \nu_j$ where $\nu_j$ is that stoichiometric vector for
-reaction $R_j$.
-
-The model can be represented as the following expression involving
-$N_r$ different unit rate Poisson
-processes\cite{anderson2011continuous} $Y_j$, given by:
-\begin{equation}\label{eq:RTC}
-X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j \int_0^t \alpha_j(X(s)) ds.
-\end{equation}
-
-The master equation for this model is only solvable in certain
-situations, for example monomolecular networks\cite{jahnke2007solving}, or for
-steady-state solutions for weakly reversible deficiency
-zero networks\cite{anderson2010product,anderson2016product}. Trajectories for this system can be
-sampled exactly, for instance using the Gillespie SSA\cite{gillespie1977exact}, or
-its variants\cite{gillespie2007stochastic,cao2004efficient,anderson2007modified}. However, if the system is stiff, i.e. there
-are some reactions which are firing many times on a timescale for
-which others are unlikely to fire at all, then trajectories can become
-prohibitively expensive to simulate, since these methods simulate
-every single reaction event with the same cost. In such a system, one
-might employ multiscale methods in order to approximate trajectories
-at a lower cost than the exact algorithms.
-
-One common approach is to induce the quasi-equillibrium approximation
-QEA. This approximation makes the assumption that fast reactions
-enter quasi-equilibrium on a timescale which is negligible with
-respect to the timescale on which the slow dynamics in the system are
-evolving. This amounts to taking the asymptotic limit that the
-propensities of the ``slow'' reaction channels are equal to zero. This
-allows us to sample approximate trajectories of the slow reactions
-without the need to compute many fast reaction events.
-
-This approach can work well when there is a large timescale gap
-between the fast and slow reactions in a system, but where the
-timescale gap is less pronounced, it can lead to significant
-errors\cite{cotter2016error}. Another approach is the
-constrained multiscale algorithm (CMA)\cite{cotter2011constrained,cotter2016constrained}, based in part on the
-equation-free approach to multiscale computations\cite{kevrekidis2003equation,erban2006gene}. This approach also assumes quasi-equilibrium in the
-system, but takes into account the effect that the slow reactions can
-have on the invariant distribution of the fast variables in the
-system. For a more detailed description of this method, please refer
-to the literature\cite{cotter2016constrained}.
-
-Multiscale methods are not only of use when forward evaluations of the
-dynamics are costly, but can also be of use where we attempt to solve
-an inverse problem where the fast variables are unobservable. One
-approach in this situation would be to integrate over all possible
-trajectories of the fast variables, but this would almost always be
-prohibitively expensive. Another approach would be to use an
-appropriate multiscale approximation, so that the effective dynamics
-of the slow variables can be approximate without the need to consider
-the rapid fluctuations of the fast variables in the system.
-
 \subsection{Likelihoods arising from stochastic reaction networks}
@@
 next reaction, which is given by an exponential random variable
-\[\tau_j \sim \exp \left ( \frac{1}{\alpha_0(X(t))} \right ),\]
+\begin{equation}\tau_j \sim \exp \left ( \frac{1}{\alpha_0(X(t))} \right ),\end{equation}
 where $\alpha_0(X(t)) =\sum_{i}^{N_r} \alpha_i(X(t))$ is the total
@@
 $i \in \{1,2,\ldots,N_r\}$ with probability
-\[\mathbb{P}(r_j = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\]
+\begin{equation}\mathbb{P}(r_j = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\end{equation}
 As such, in order to compute the likelihood of a particular trajectory
@@
 We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
-\[
+\begin{equation}
 	T(\mathbf{Y}) = \sum\limits_{j=1}^M \tau_j\; \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y}).
-\]
+\end{equation}
 This random variable, $T(\mathbf{Y})$, is a sum of exponentially
-distributed random variable, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
+distributed random variables, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
 \begin{equation}\label{eqn:chem_time_dist}
@@
 Similarly, we can define the reactions which occurred when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_{N_r}(\mathbf{Y})]^\top$ where
-\[
+\begin{equation}
 	r_i(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(r_j = i \textbf{ and }\mathbf{X}(t_{j-1}) = \mathbf{Y}).
-\]
+\end{equation}
 Here all the random variables $r_j$ follow the same multinomial distribution, and so
@@
 With these definitions we define two new structures
-\[
+\begin{equation}
 	\mathbf{T} = [T(\mathbf{Y}_1), \dots, T(\mathbf{Y}_K)]^\top, \quad \text{and} \quad \mathbf{R} = [\mathbf{r}(\mathbf{Y}_1), \dots, \mathbf{r}(\mathbf{Y}_K)]^\top,
-\]
+\end{equation}
 where $K = |\mathcal{S}|$, the number of states in $\mathcal{S}$, and each state $\mathbf{Y}_i \in \mathcal{S}$ has been enumerated. We use these structures to define shorter notation,
-\[
+\begin{equation}
 	\mathbf{T}_i = T(\mathbf{Y}_i), \quad \mathbf{R}_{ij} = r_j(\mathbf{Y}_i), \quad \text{and} \quad \mathbf{K}_i = K(\mathbf{Y}_i).
-\]
+\end{equation}
 
@@
 Given the distributions in Equations~\eqref{eqn:chem_time_dist} and~\eqref{eqn:chem_react_dist} for our data, the likelihood of observing the data $\mathbf{R}$ and $\mathbf{T}$ is
-\[
+\begin{equation}
 	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{R}_{i\cdot}; \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
-\]
+\end{equation}
 where again $\mathbf{Y}_i \in \mathcal{S}$ and the propensities
@@
 occurrence of slow reactions.
-\section{Numerical Examples}\label{sec:num}
-In this section we look at two examples of chemical systems to
-demonstrate the effectiveness of transport ETAIS algorithms for
-sampling from the
-challenging posterior distribution on the reaction parameters that
-arise.
-
-% The first example will only contain at most mono-molecular reactions,
-% i.e. reactions which convert one species into another species at a
-% particular rate. As shown in~\cite{jahnke2007solving} these systems
-% have a Poisson invariant distributions on the populations, given by
-% \begin{equation}\label{eqn:mono_stat_dist}
-% 	\mathbb{P}(X_1=x_1, \ldots, X_n=x_n) = \left(\prod_{i=1}^n \! \frac{\bar{c}_i^{x_i}}{x_i!}\right)\exp\left(-\sum_i^n \! \bar{c}_i\right),
-% \end{equation}
-% where $x_i \in \mathbb{N}$ and $\bar{c}_i$ are the complex balances or the concentration of chemical $X_i$ according to the solution of a steady state approximation of the system.
-We make use of this distribution to formulate posteriors for the chemical reaction rates using the constrained multiscale approach, as in~\cite{cotter2016constrained}.
-
-%%%%%
-
-\subsection{A Multiscale Chemical System}\label{sec:chem_multiscale}
-
-First we consider a simple multiscale example involving two chemical
-species $S_1$ and $S_2$,
-\begin{equation}\label{eqn:full_system}
-	\emptyset \xrightarrow{k_1} S_1 \xleftrightarrows[k_3]{k_2} S_2 \xrightarrow{k_4} \emptyset.
-\end{equation}
-Each arrow represents a reaction from a reactant to a product, with
-some rate constant $k_i$, and where mass action kinetics is
-assumed. The parameters $k_i$ are non-negative, and $\mathbf{k} =
-[k_1,\dots,k_4]^\top \in \mathbb{R}_+^4 = \mathcal{X}$. We denote the
-population count of
-species $S_i$ by $X_i \in \mathbb{N}_0$. 
-We consider this system with the following parameters:
-\begin{equation}\label{eq:params1}
-k_1V = 100, \qquad k_2 = k_3 = 10, \qquad k_4 = 1,
-\end{equation} 
-where $V$ is the volume of the reactor.
- In this 
-parameter regime the reactions $R_2\colon S_1\rightarrow S_2$ and $R_3\colon S_2\rightarrow S_1$ occur
-more frequently than the other reactions. Notice that both
-chemical species are involved in fast reactions. However, the quantity
-$S = X_1 + X_2$, is conserved by both of the fast reactions,
-and as such, this is the slowly changing quantity in this system.
-
-We observe this system over the time period $t \in [0,500]$ with
-initial condition $S_1(0) = S_2(0) = 0$, but we assume that we are only able to observe
-the slow variable $S(t) = S_1(t) + S_2(t)$. 
-The effective dynamics of $S$ can be approximated by a system of only
-two reactions:
-\begin{equation}\label{eqn:QSSA_system}
-	\emptyset \xrightarrow{\alpha_1(S)} S \xrightarrow{\bar{\alpha}_4(S)} \emptyset,
-\end{equation}
-where here the propensities are shown as opposed to the rates.
-
-The effective propensity $\bar{\alpha}_4(S)$ can be  approximated through
-application of a multiscale approximation, as detailed in Section
-\ref{sec:multi}, and in more detail in \cite{cotter2016constrained}. These effective
-propensities can often be computed without the need for expensive
-simulations, in particular when the fast subsystem is deficiency zero,
-which is often the case\cite{anderson2010product,anderson2016product}. For this very
-simple example, the dependance of these effective
-propensities on the original rate parameters can be explicitly
-understood. Under the QEA,
-the effective rate of degradation of the slow variable $S$ is given by
-\[
-	\bar{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3}.
-\]
-Similarly, the analysis of the constrained system as discussed in~\cite{cotter2016constrained} yields the effective propensity
-\begin{equation}\label{eqn:chem_CMA_rate}
-	\bar{\alpha}_4^{\text{CMA}}(s) = \mathbb{E}_{\text{CMA}}\left[k_4X_2|S=s\right] = \frac{k_2k_4s}{k_2+k_3+k_4}.
-\end{equation}
-
-Our observations are uninformative about the reaction rates $k_2,
-k_3$, and $k_4$, as for each multiscale approximation there
-is a manifold $\mathcal{M} \subset \mathcal{X}$ along which the effective propensity
-$\hat{\alpha}_4$ is invariant, leading to a highly ill-posed inverse
-problem. This type of problem is notoriously difficult to
-sample from using standard MH algorithms, as the algorithms quickly
-gravitate towards the manifold $\mathcal{M}$ but then
-exploration around $\mathcal{M}$ is slow.
-
-We aim to recover three different posterior distributions. In the
-first, which is of the form \eqref{eqn:chem_posterior}, we assume that we can fully and exactly observe the whole state
-vector for the whole of the observation time window $t \in [0,T]$. In
-the second and
-third, the same data are used, but projected down to one dimensional observations of the slow
-variable $S = X_1 + X_2$. In these last two examples, we approximate the
-posterior using \eqref{eq:Apos} with QEA and CMA approximations of the
-effective dynamics respectively.
-
-In all three cases, we find the posterior distribution for $\mathbf{k}
-\in \mathcal{X} = \mathbb{R}_+^4$. These four parameters are assigned Gamma prior distributions,
-\[
-	k_i \sim \text{Gamma}(\cdot; \alpha_i, \beta_i), \quad \text{for} \quad i = 1, \dots, 4.
-\]
-The hyper-parameters corresponding to each of these prior distributions are given in Table~\ref{tab:multiscale_priors}. These priors are the same for each of the three posterior distributions.
-
-\begin{table}
-\centering
-\begin{tabular}{|l|R{2em}R{2em}R{2em}R{2em}|}
-	\hline
-	Parameter$\backslash$Dimension $i$ & 1 & 2 & 3 & 4 \\ \hline
-	$\alpha_i$ & 150 & 5 & 5 & 3 \\
-	$\beta_i$ & $\frac{15}{9}$ & $\frac{5}{12}$ & $\frac{5}{12}$ & $1$ \\ \hline 
-\end{tabular}
-\caption{Hyper-parameters in the prior distributions for the multiscale problem described in Section~\ref{sec:chem_multiscale}.}
-\label{tab:multiscale_priors}
-\end{table}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
-\caption{CMA approximation \eqref{eq:Apos} of the posterior arising from observations of
-  the slow variable $S = X_1 + X_2$ for system
-  \eqref{eqn:full_system}  with true rates given by \eqref{eq:params1}.}
-\label{fig:chem_CMA_posterior}
-\end{figure}
-
-Figure~\ref{fig:chem_CMA_posterior} shows how this posterior looks
-when we use the CMA to model the effective degradation propensity
-$\bar{\alpha}_4$. Although this posterior does not appear to be
-pathologically difficult to sample using MCMC from these plots, it most certainly is. The posterior
-density is concentrated close to a curved manifold. Since the manifold
-is curved, the joint marginals do not look unreasonable, where in fact
-the density has very differing derivatives in different directions
-when evaluated close to this manifold.
-
-We consider several different algorithms for sampling from these
-distributions. First we implement both the ETAIS and MH algorithms with
-a Gaussian proposal distribution. In the case of the ETAIS algorithm,
-this is a Gaussian mixture proposal distribution, and in the MH
-algorithm this is equivalent to a vanilla random walk. The proposal
-distribution uses a covariance matrix which has been constructed using
-the sample covariance of a sample produced using a standard MH-RW
-algorithm. This is necessary since these algorithms otherwise fail to
-converge in a reasonable time for
-this problem. We also compare the ETAIS and MH algorithms when using a
-transport map proposal distributionm, as was discussed in detail in Chapter~\ref{sec:TETAIS}.
-
-\begin{figure}
-	\centering
-	\begin{tikzpicture}
-		\pgfmathsetseed{1}
-
-		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
-		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$R$};
-
-		\draw (A) \irregularcircle{1cm}{2.5mm};
-		\draw (B) \irregularcircle{1cm}{0.5mm};
-
-		\draw [->] (A) -- (B);
-
-		\node at (2.5,1) {$T\colon\mathcal{X}\rightarrow R$};
-
-		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
-		\node at (5,-1.5) {$r\in R\subset\mathbb{R}^d$};
-	\end{tikzpicture}
-
-	\begin{tikzpicture}
-		\pgfmathsetseed{1}
-
-		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
-		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
-		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};
-
-		\draw (A) \irregularcircle{1cm}{2.5mm};
-		\draw (B) \irregularcircle{1cm}{1.1mm};
-		\draw (C) \irregularcircle{1cm}{0.5mm};
-
-		\draw [->] (A) -- (B);
-		\draw [->] (B) -- (C);
-
-		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
-		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};
-
-		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
-		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
-		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
-	\end{tikzpicture}
-	\caption{Couplings between parameter space and reference
-          when using the transport map with (bottom) and without (top)
-        a log preconditioner.}
-	\label{fig:chem_coupling}
-\end{figure}
-
-Figure~\ref{fig:chem_coupling} (top) shows how we define a bijective map,
-$\hat{T}$, between parameter space $\mathcal{X}$ and a reference space
-$R$.
-When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
-MH-RW  and MH-TRW, as well as ETAIS-RW and ETAIS-TRW.
-In practice we cannot ensure that the approximate map $\tilde{T}$
-is uniquely invertible over the whole of $R$ and so $\tilde{T}$ is not
-truly bijective. This leads to problems for our strictly positive
-state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in
-$R$ do not necessarily map back onto $\mathcal{X}$.
-
-Therefore we also consider how these algorithms perform when the transport map
-is preconditioned by the logarithmic function applied to the parameters. We choose to use this log
-preconditioner since it converts our strictly positive parameter
-space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset
-\mathbb{R}^d$. This allows us to define $T$ between two subsets of
-$\mathbb{R}^d$, which means that all proposals made on the reference
-space are mapped to valid
-proposals on the true parameter space. It also reduces the work required of the map, which since
-we use only a finite dimensional approximation, can improve
-performance and stability.
-As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
-MH-logRW and MH-logTRW, with, ETAIS-logRW and ETAIS-logTRW.
-
-Figure~\ref{fig:chem_coupling} (bottom) displays the composition of maps
-when we include this $\log$ preconditioner leading to an intermediate space $\mathcal{Y}$.
-The inclusion of this additional map means that we must again alter our importance weight definition to reflect the pullback from $R$ through $\mathcal{Y}$.
-The weight is now
-\[
-	w_i(\theta) = \frac{\pi(\theta|\mathbf{R},\mathbf{T})}{\chi(\tilde{T}\circ\log(\theta)|\tilde{T}\circ\log(\theta^{(i-1)}))|J_{\tilde{T}\circ\log}(\theta)|},
-\]
-where $\theta$ is a proposed new state, $\theta^{(i-1)}$ is the ensemble of states from the previous iteration, and $J_{\tilde{T}\circ\log}(\theta)$ is the Jacobian of the composition of the two maps.
-This Jacobian is straightforward to calculate,
-\[
-	|J_{\tilde{T}\circ\log}(\theta')| = |J_{\tilde{T}}(\log(\theta'))||J_{\log}(\theta')|,
-\]
-where the first determinant is computed as in
-\eqref{eqn:separable_jacobian}, and the second is given by
-\[
-	|J_{\log}(\theta')| = \prod\limits_{i=1}^d \frac{1}{\theta'_i}.
-\]
-
-% choices made in transport map
-For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=3$ as the basis functions, i.e.
-\[
-	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
-\]
-and
-\[
-	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
-\]
-
-% choice of resampler?
-We will use the MT\cite{russ2017parallel} resampler with an ensemble size of
-$M=500$. This increase in ensemble size in comparison with the previous
-example in Section \ref{sec:conv} is to allow for the increase in parameter dimension.
-
-% analysis method
-To measure the convergence of the sampling methods in this section, we
-will compare the convergence of the mean of each parameter. We
-approximate $\mathbb{E}(\mathbf{k})$ using 2.4 billion samples from
-the MH-RW algorithm. We do this for each of the eight algorithms we
-have so far discussed. Convergence is shown only for the CMA
-approximation of the posterior \eqref{eq:Apos} with effective rate for
-the degradation of $S$ given by \eqref{eqn:chem_CMA_rate}, but we
-expect very similar results for the other multiscale approximation of
-posterior distributions
-discussed.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{rrrrrr}
-\toprule
-	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{ETAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
-	& & $\delta$ & & $\delta$ & ESS \\ \midrule
-	RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
-	TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
-	logRW & & 1.7e-1 & & 1.2e-0 & 6.0e-2 \\
-	logTRW & & 2.7e-2 & & 1.5e-1 & 3.5e-1 \\
-\bottomrule
-\end{tabular}
-\caption{Optimal scaling parameters for the MH and ETAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}.
-MH parameters optimised by acceptance rate, and ETAIS parameters optimised using effective sample size.}
-\label{tab:chem_multiscale_scaling}
-\end{table}
-
-The optimal scaling parameters for the proposal distributions are
-given in Table~\ref{tab:chem_multiscale_scaling}. These are
-precomputed by running the algorithms for different parameter
-values. MH-RW based algorithms are tuned to an acceptance rate close
-to the optimal $23.4\%$, and ETAIS algorithms are tuned to maximise the
-effective sample size (ESS). The optimal parameter values for the MH
-algorithms are not particularly informative about their performance,
-since they relate to algorithms operating on different
-transformations. The ESS is highest for the logTRW variant
-of the ETAIS algorithm, as we would hope. We note here that as
-outlined in \cite{cotter2015parallel}, it is possible to construct
-adaptive versions of ETAIS which automatically tune these algorithmic
-parameters, but we compare against the static optimal parameter values
-for all algorithms to ensure fairness.
-
-\begin{figure}[!htb]
-\centering
-\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
-\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
-\caption{Convergence of eight different MCMC algorithms for the CMA
-  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}.}
-\label{fig:chem_multiscale_L2}
-\end{figure}
-
-Convergence of the eight algorithms for this example are shown in
-Figure~\ref{fig:chem_multiscale_L2}. We first note the poor
-performance of the MH based algorithms, despite being preconditioned
-with a pre-computed sample covariance.
-Only the MH-TRW is at all competitive with the ETAIS algorithms.
-During the simulation interval, the MH-TRW algorithm has not settled down to the expected $\mathcal{O}(1/\sqrt{N})$ rate which means that the estimate is still biased by the long burn-in time.
-As we have seen in previous examples, the burn-in time for the ETAIS algorithm is negligible.
-
-The ETAIS variants with RW and TRW proposals perform similarly on both
-sample spaces. When sampling without a preconditioner, the transport map is not
-quite as efficient, largely due to the difficulties discussed in the
-previous section i.e. many proposals are made which do result in
-negative reaction rates. Sampling with transport maps with a $\log$
-preconditioner leads to more comparable Monte Carlo errors, with the
-logTRW being apparently slightly less stable. This proposal method
-becomes more stable as we increase either the ensemble size, or the
-number of iterations between updates of the transport map,
-$T$. Overall we see the smallest Monte Carlo errors for a given number
-of likelihood evaluations coming from the ETAIS-logTRW algorithm.
-
-\begin{figure}[!htb]
-\centering
-\subfigure[Push forward of the posterior sample through $\hat{T}$ with
-no log
-preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
-\\
-\subfigure[Push forward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
-\caption{Push forward of the target measure on reference space for the the CMA
-  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}, found using the TRW and logTRW proposal distributions.
-Components are linked by the relation $r_i = T_i(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
-\label{fig:chem_reference_spaces}
-\end{figure}
-
-We now look at the proposal distributions of the transport map accelerated algorithms.
-In Figure~\ref{fig:chem_reference_spaces}, we see the reference spaces found by; (a) mapping the posterior through the map $\hat{T}$, (b) by mapping the posterior through $T\circ\log$.
-For the most part, each of these marginal distributions can be recognised as a Gaussian.
-However, with the exception of $\mathbb{P}(r_2,r_3)$, we would not consider them to be close to a standardised $\mathcal{N}(0, \text{I})$ distribution.
-Before thinking that the transport map has not helped us to find a `nicer' space on which to propose new values we should consider that the dimensions are now (1) largely uncorrelated, and (2) the variances in each dimension are much more similar than they are in Figure~\ref{fig:chem_CMA_posterior}.
-
-In Figure~\ref{fig:chem_reference_spaces} (b) we see that
-$\text{var}(r_1)$ and $\text{var}(r_4)$ are much smaller than
-$\text{var}(r_2)$ and $\text{var}(r_3)$. To combat this we have a
-number of choices. We might wish to learn the covariance structure of
-the push forward of the target onto the reference space, and
-incorporate this information into the covariances of the mixture
-components in an adaptive scheme.
-Another option is to increase the total order of our index set.
-For these numerics we have chosen $p=3$, but we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
-
-\subsubsection{Comparison of the Constrained and QEA approaches}
-
-The convergence analysis has been performed for the constrained
-approximation of the posterior distribution. We now look at the
-differences in the approximations of the posterior distribution
-arising in this problem between the CMA and QEA. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
-\[
-	\hat{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{\alpha}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
-\]
-This difference in the denominator causes a shift in the manifold on
-which the posterior is concentrated, as can be seen in
-Figure~\ref{fig:chem_diff}. The figure shows the difference in
-posteriors densities
-\begin{equation}\label{eqn:chem_diff}\pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
-\end{equation}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.9\textwidth]{"images/Applications/Diff_CMA_QEA"}
-\caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
-\label{fig:chem_diff}
-\end{figure}
-
-Since the two posteriors have been approximated using an MCMC sample,
-there is a significant amount of noise, particularly in the tails of
-the distributions. We can see that the differences in the marginals
-for $k_1, k_2,$ and $k_3$ are relatively small, but the marginal for
-$k_4$ varies by a significant amount.
-
-In the QEA approximation, the effective rate of degradation of $S$ is
-given by $\hat{k}_4^{\text{QEA}} = k_2k_4/(k_2+k_3)$, and as such, our
-observations should be informative about this quantity. Using the CMA,
-this effective rate is approximated by $\hat{k}_4^{\text{CMA}} =
-k_2k_4/(k_2+k_3+k_4)$. To validate our inferences on $k_2$, $k_3$ and
-$k_4$ we would like to discover which model is most accurate and
-informative about these quantities, and which model gives us results
-which are most similar to that which can be obtained from the full
-model, with all reactions (including fast reactions) observed.
-
-A conventional way to compare two models under a Bayesian framework is
-to calculate the Bayes factors~\cite{chen2012monte}. The Bayes factor,
-$B_{1,2}$, between two models, $\mathcal{M}_1$ and $\mathcal{M}_2$,
-given data $D$, can be interpreted as a ratio of the normalisation constants of the posterior distributions given each model,
-\[
-	B_{1,2} = \frac{\mathbb{P}(D|\mathcal{M}_1)}{\mathbb{P}(D|\mathcal{M}_2)}, \quad \text{where} \quad \mathbb{P}(D|\mathcal{M}_k) = \int_\mathcal{X} \mathbb{P}(D|\theta_k,\mathcal{M}_k)\mathbb{P}(\theta_k|\mathcal{M}_k) \, \text{d}\theta_k.
-\]
-Under the ETAIS framework, it is straightforward to calculate these factors using the Monte Carlo estimator for the normalisation constants.
-From~\cite{robert2013monte} these normalisation constants take the form
-\[
-	Z_k \approx \hat{Z}_k\frac{1}{NM}\sum_{i=1}^N\sum_{j=1}^M w_j^{(i)}(\mathcal{M}_k),
-\]
-where $w_j^{(i)}(\mathcal{M}_k)$ is the weight under model $\mathcal{M}_k$ corresponding to the $j$-th ensemble member on the $i$-th iteration.
-Hence, $B_{1,2} \approx \hat{Z}_1/\hat{Z}_2$.
-We can compare more than two models in this way by selecting the model with the largest marginal distribution as the best model.
-
-We now label the constrained model as $\mathcal{M}_1$, the model
-arising from the QEA as $\mathcal{M}_2$, and the full data model as
-$\mathcal{M}_0$. Model $\mathcal{M}_1$ is dependent on the parameter
-$\theta_1 = (k_1, \hat{k}_4^{\text{CMA}})^\top$, and model
-$\mathcal{M}_2$ is dependent on the parameters $\theta_2 = (k_1,
-\hat{k}_4^{\text{QEA}})^\top$. The full model $\mathcal{M}_0$ is
-dependent on all four parameters $\theta_0 = (k_1, k_2, k_3, k_4)^\top$.
-The marginal densities for the data, evaluated at the observed data, given the models are displayed in Table~\ref{tab:chem_Bayes_marginals}.
-
-\begin{table}[!htb]
-\centering
-\begin{tabular}{cccc}
-	\toprule
-	Model description & $k$ & \quad $\mathbb{P}(D =
-                                  (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$
-  & $B_{0,k}$ \\ \cmidrule(lr){1-4}
-	Full & 0 & 6.8e-3 & 1 \\
-	CMA & 1 & 3.2e-3 & 2.09 \\
-	QEA & 2 & 1.7e-3 & 4.10 \\ \bottomrule
-\end{tabular}
-\caption{Marginal distributions and Bayes factors for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:chem_multiscale}.}
-\label{tab:chem_Bayes_marginals}
-\end{table}
-
-Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals}
-we see that we should of course prefer the model $\mathcal{M}_0$ in
-which we observe all reactions perfectly. However this model is not
-significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$,
-~\cite{kass1995bayes}). The Bayes factor $B_{0,2} = 4.1 > 3.2$ tells
-us that we should significantly prefer the full model to the QEA
-approximation of the posterior. These Bayes factors present a
-relatively weak
-argument that the constrained approximation of the posterior is
-superior to the QEA approximation.
-
-\begin{figure}[!htb]
-\centering
-\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/QEA_approx_marginal"}}
-\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_approx_marginal"}}
-\caption{Comparison of the approximate marginal densities for the
-  quantities $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$
-  across the three posterior densities $\pi_i = \mathbb{P}(\theta_i|D,\mathcal{M}_i)$ for $i = 0, 1, 2$.}
-\label{fig:chem_model_comp}
-\end{figure}
-
-Figure~\ref{fig:chem_model_comp} displays more compelling evidence in
-this direction. Here we show the marginal distributions for the
-quantities $\hat{k}_4^{\text{QEA}}$ and $\hat{k}_4^{\text{CMA}}$ with
-respect to each of the three posterior distributions, $\pi_i(\cdot|D,
-\mathcal{M}_i)$, $i = 0, 1, 2$.
-
-Firstly, in
-Figure~\ref{fig:chem_model_comp} (a), displaying the marginals for
-$\hat{k}_4^{\text{QEA}}$ we see that $\pi_0$, the density arising from
-the full model with all reactions observed, is peaked a distance away
-from $\pi_2$, QEA approximation. The CMA approximation $\pi_1$, in contrast, is
-peaked close to peak of $\pi_0$, but there is more uncertainty here
-since we do not observe the fast reactions.
-
-In contrast, in Figure~\ref{fig:chem_model_comp} (b), the marginal
-densities for $\hat{k}_4^{\text{CMA}}$ for $\pi_0$ and $\pi_1$ are
-barely distinguishable, even with the vastly reduced data available in
-the posterior $\pi_1$ with CMA approximation. The QEA approximation
-$\pi_2$ is again peaked far away from both of the other distributions.
-This appears to demonstrate that the data regarding the slow variable
-$S$ is only informative about $k_1 $ and $\hat{k}_4^{\text{CMA}}$, as
-predicted by the CMA multiscale approximation.
-
-\subsection{Gene Regulatory Network}\label{sec:grn}
-In this example, we look at a model of a gene regulatory network (GRN)
-mechanism~\cite{kaern2005stochasticity,guido2006bottom,becskei2000engineering}
-used by a particular cell to regulate the amount of a protein, $P$,
-present. Proteins can bind in pairs to form a dimer $D$. The dimer can
-bind to the gene $G$ to a ``switched off'' state $G'$, in which the
-production of mRNA molecules $M$ is inhibited. The mRNA encodes for
-production of the protein $P$, and both $P$ and $M$ can degrade. 
-The eight reactions are given in Equation~\eqref{eqn:GRN_reactions}.
-\begin{align}
-	P+P \xleftrightarrows[k_2]{k_1} D \quad & \quad G+D \xleftrightarrows[k_4]{k_3} G' \nonumber\\
-	G \xrightarrow{k_5} G+M \quad & \quad M\xrightarrow{k_6} M+P  \label{eqn:GRN_reactions} \\
-	P \xrightarrow{k_7} \emptyset \quad & \quad M \xrightarrow{k_8} \emptyset \nonumber
-\end{align}
-
-We create trajectories of this system with the following parameters:
-\begin{eqnarray*}
-k_1 = 0.04, \qquad
-k_2 = 5000, \qquad
-k_3 = 100, \qquad
-k_4 = 1, \\ 
-k_5 = 0.5, \qquad
-k_6 = 2, \qquad
-k_7 = 0.2, \qquad
-k_8 = 0.05,
-\end{eqnarray*}
-over the time period $t \in [0,500]$.
-
-T-cells can be manufactured which bind to
-proteins of a certain type, and cause them to phosphoresce, allowing
-for approximate observations of the concentration levels of that
-protein. However, the T-cells will not necessarily differentiate
-between monomers and dimers of the protein. We replicate such a
-scenario, in a zero-observational noise setting, by assuming that we
-can only observe $T=P + 2D + 2G'$, the total number of protein
-molecules in the system, alongside $M$, the concentration of mRNA. The
-number of dimers, and the status of the genetic switch, are assumed
-unobservable. In
-order to sample from the posterior distribution on the reaction
-parameters in the system given these observations, we can either integrate over all possible
-trajectories of the fast variables (which is computationally
-intractable), or as in Section \ref{sec:chem_multiscale} we can use a
-multiscale method to approximate the posterior.
-
-The effective dynamics of $T$ and $M$ are then given by:
-\begin{align}
-	\emptyset \xrightarrow{\hat{\alpha}_5} M \quad & \quad M\xrightarrow{\alpha_6} M+T  \label{eqn:GRN_eff} \\
-	T \xrightarrow{\hat{\alpha}_7} \emptyset \quad & \quad M \xrightarrow{\alpha_8} \emptyset \nonumber
-\end{align}
-Here, the propensities are displayed, as opposed to the rates, with
-approximations of the propensities $\hat{\alpha}_5$ and
-$\hat{\alpha}_7$ to be computed. We omit the derivation of these
-effective propensities, using the CMA methodology, for brevity, but
-interested readers can see more details in \cite{russ2017parallel}.
-
-\subsubsection{Target Distribution}
-As alluded to in the previous section, we aim to sample from the CMA
-approximation of the posterior distribution as defined in
-Equation~\eqref{eqn:chem_posterior}. This example is more involved
-than the previous one, since we have an eight dimensional parameter
-space corresponding to the eight reaction rates in the system
-\eqref{eqn:GRN_reactions}. However half of those reactions are
-unobservable, and their effect on the posterior is only felt through
-their effect on the effective propensities $\hat{\alpha}_5$ and $\hat{\alpha}_7$.
-
-The priors for this problem were chosen to be fairly uninformative with respect to the likelihood function for each reaction rate.
-A list of the hyper-parameters corresponding to each Gamma prior can be found in Table~\ref{tab:grn_priors}.
-
-\begin{table}[!h]
-\centering
-\begin{tabular}{|l|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|}
-	\hline
-	Dimension $i$ & 1 & 2& 3& 4 & 5 & 6 & 7 & 8 \\ \hline
-	$\alpha_i$ & 2 & 100 & 100 & 3 & 3 & 3 & 2 & 2\\ \hline
-	$\beta_i$ &50 & 0.02 & 1 & 1 & 0.6 & 1 & 50 & 50 \\ \hline
-\end{tabular}
-\caption{Hyper parameters for the Gamma priors on each of the reaction rates in the GRN example in Section~\ref{sec:grn}.}
-\label{tab:grn_priors}
-\end{table}
-
-\begin{figure}[htb]
-\centering
-\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/Applications/GRN_posterior"}}%
-\caption{Marginal densities of the CMA approximation of the posterior
-  distribution for the GRN example in Section~\ref{sec:grn}.}
-\label{fig:GRN_posterior}
-\end{figure}
-
-The one- and two-dimensional marginal distributions for the full
-posterior distribution are displayed in
-Figure~\ref{fig:GRN_posterior}. The posterior does not contain many
-interesting correlation structures, however several dimensions have
-leptokurtic tails which are difficult for the standard ETAIS algorithm
-to sample from. The marginal densities also vary over very different
-scales, which could lead to problems with the algorithms which do not
-benefit from the transport map.
-
-\subsubsection{Implementation}
-
-We apply the eight RW and logRW proposal distributions, with and
-without transport map acceleration, which were discussed in
-Section~\ref{sec:chem_multiscale} to this posterior. As with this
-example, the log conditioner is required so that our transport map is a function $T\colon\mathbb{R}^d\rightarrow\mathbb{R}^d$, rather than a map from $\mathbb{R}_+^d$ to $\mathbb{R}^d$.
-
-The transport map setup has not changed from the previous example.
-However, due to the scaling of the weight we have had to be more careful about which samples we use to build our map.
-We require that the objective function $C(T)$, from Equation~\eqref{eqn:TETAIS_objective} is convex, which requires that the second derivative, given in Equation~\eqref{eqn:TETAIS_hessian}, is positive definite.
-This positive definite property depends on all weights in our sample being strictly positive, which is not always possible on a computer.
-For this reason we filter out any samples from the optimisation problem where the weight is a numeric zero.
-This does not affect the validity of the map since a weight zero would not contribute to the expectation.
-
-In all eight algorithms we adapt the proposal variances during the
-burn-in phase of the algorithms to avoid the need for a huge ensemble size to approximate the posterior with a mixture of isotropic Gaussian kernels.
-We use the MT resampler, this time with an ensemble size of $M=2500$. The increase in ensemble size compensates for the increase in dimension from four to eight.
-
-As in the previous chemical reaction example, we measure convergence of the algorithms only through the convergence of the mean.
-We use the sample Mahalanobis distance, with `true' covariance and `true' mean built from a sample of size 2.4 billion using the MH-RW algorithm.
-
-\subsubsection{Convergence results}
-
-The scaling parameters used are selected by optimising the acceptance
-rate for the MH algorithms, and optimising the effective sample size
-for the ETAIS algorithms. The optimal parameters are given in
-Table~\ref{tab:grn_scaling_parameters}. Here, for Metropolis-Hastings
-variants, $\beta_{\%}^*$ is the variance of the proposal, optimised
-for an acceptance rate of $50\%$. For the ETAIS variants,
-$\beta_{\text{ESS}}^*$ is the variance of each of the proposal kernels
-around each particle, optimised numerically for the maximal effective
-sample size (ESS). ESS/$M$ denotes the ratio of ESS to the number of
-particles, with values closer to one denoting a more efficient algorithm.
-
-\begin{table}[!h]
-	\centering
-	\begin{tabular}{lrrrrrrrr}
-	\toprule
-		 & \multicolumn{4}{c}{MH} & \multicolumn{4}{c}{ETAIS} \\ \cmidrule(lr){2-5} \cmidrule(lr){6-9}
-		& RW & logRW & TRW & log-TRW & RW & logRW & TRW & log-TRW \\ \midrule
-		$\beta_{\%}^*$	 	& 1.0e-1 & 3.9e-1 & 1.0e-0 &6.0e-1 & - & - & - & - \\
-		$\beta_{\text{ESS}}^*$	& - 	        & -           & -            & -            & 1.0e-0 & 1.3e-0 & 8.0e-1 & 5.0e-1 \\
-		ESS/$M$		 		& - 	        & -           & -            & -            & 3.5e-4 & 4.7e-2 & 4.8e-2 & 1.6e-1  \\
-	\bottomrule
-	\end{tabular}
-	\caption{Scaling parameters for the sampling algorithms applied to the GRN example in Section~\ref{sec:grn}.}
-	\label{tab:grn_scaling_parameters}
-\end{table}
-
-From the effective sample sizes shown in
-Table~\ref{tab:grn_scaling_parameters} we can see the improvement to
-the efficiency of the algorithm both by transforming the parameter
-space into something closer to Gaussian, and additionally by utilising
-a log preconditioner.
-
-Due to the numerical cost of calculating the full relative $L^2$ errors for this posterior, we quantify the convergence speeds of these algorithms using the sample Mahalanobis distance between the sample mean and the `true' mean.
-This convergence analysis is shown in Figure~\ref{fig:grn_L2}.
-
-\begin{figure}[!htb]
-\centering
-\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
-\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
-\caption{Convergence of eight different MCMC algorithms for the CMA
-  approximation of the  posterior arising from the example in Section \ref{sec:grn}.}
-\label{fig:grn_L2}
-\end{figure}
-
-The first thing to note is that the first ETAIS measurement happens
-after 2500 samples, whereas the first MH measurement occurs after 50
-samples. This is due to the much larger ensemble size required to
-sample in higher dimensions. We believe that an ensemble size of 2500
-is over-cautious in this example and we could have used a smaller
-ensemble size. We also believe that the required ensemble size to sample
-from these posteriors is reduced by the use of the transport map.
-
-The second obvious feature of these convergence plots is that the ETAIS
-algorithms outperform the MH algorithms by a large margin - roughly a
-reduction of two orders of magnitude in the relative error over the
-same number of samples. A less positive result is that for this
-example the ETAIS-TRW algorithm has found a map which has prevented
-further convergence of the mean. This is likely because one dimension
-of the transport map is sending proposals onto the negative real line
-and so all samples are being assigned a zero weight. This is behaviour
-which was anticipated, and it is always suggested that appropriate
-preconditioners are used where the posterior is not absolutely
-continuous with respect to Lebesgue measure, as is the case when the
-posterior is only non-zero on the positive quadrant of $\mathbb{R}^d$
-as in these examples. The large increase in the effective sample size
-observed between the ETAIS-logRW and ETAIS-logTRW algorithms is
-converted into an estimate which is twice as accurate after 1 million
-samples.
-
-A similar pattern is seen in the MH algorithms, where the MH-TRW
-performs the worst, and the MH-logTRW algorithm performs the best,
-although only marginally.
-
-\section{Discussion}\label{sec:conc}
-In this paper, we have investigated the use of transport maps to
-accelerate and stabilise ensemble importance sampling schemes. In
-particular we looked at the application of these methods to problems
-arising in inverse problems for multiscale stochastic chemical
-networks, where the posterior distributions which arise can be highly
-concentrated along curved manifolds in parameter space.
-Below we highlight some areas for future work.
-
-\subsection{Sampling in higher
-  dimensions}\label{sec:TETAIS_higher_dim}
-
-One major issue with importance sampling-based algorithms is the curse of
-dimensionality. This curse is two-fold, since as the target
-distribution's dimension is increased, a larger ensemble size is often
-required in order to capture the addition complexity of the density
-that the higher dimension enables. Furthermore, the cost of the
-resampling step within the algorithm grows with the ensemble size. For
-example, the ETPF algorithm is $\mathcal{O}(n^2)$. However, the
-simplification that the transport map makes has some potential to go
-some way to tackling this problem.
-
-Here, we will briefly discuss how transport maps could aid
-with making moderately higher dimensional problems accessible to this
-family of methods. It is important to emphasise here that these
-methods are unlikely to be suitable in anything other than low
-dimensional problems, but the idea that we present here, which is the
-subject of future investigation, could make them applicable to more
-problems which have more than a handful of parameters.
-
-Algorithm~\ref{alg:TransportETAIS2} allows us to decorrelate the
-dimensions of our random parameter on reference space, where we then
-can resample and map the resulting ensemble back onto target
-space. Since, on reference space, the dimensions are approximately uncorrelated, we
-are able to resample in each dimension separately. Resampling in a
-single dimension allows for optimisations in resampling code, and also
-means that the resampler is not affected by the curse of
-dimensionality.
-
-In one
-dimension the ETPF algorithm can be implemented very efficiently. As
-described in~\cite{reich2013nonparametric}, the coupling matrix has
-all non-zero entries in a staircase pattern when the state space is
-ordered. We can exploit this knowledge to produce
-Algorithm~\ref{alg:ETPF_1d}, which is much faster than using the
-simplex algorithm to minimise the associated cost function, and faster
-than the MT algorithm\cite{cotter2015parallel}. Moreover, the task of
-running this resampler can then be parallelised over $d$
-processors. This could be an area for further study.
-
-\begin{table}[!htpb]
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\BlankLine
-Sort the states, $\{(w_i, x_i)\}_{i=1}^M$, into ascending order.\;
-Normalise the weights $p_i = w_i/\|w\|_1$.\;
-Set $y_i \leftarrow 0$ for all $i=1,\dots,M$.\;
-Set $c \leftarrow 0$\;
-\For{$i \leftarrow 1, \dots, M$}{
-	Set $t \leftarrow p_i$\;
-	\While{$j \leq M$ and $t > 0$}
-	{
-		Set $s \leftarrow \left(M^{-1}-c\right) \wedge t$\;
-		Increase $y_j$ by $M\times s\times x_i$.\;
-		Decrease $t$ by $s$.\;
-		Increase $c$ by $s$.\;
-		\If{$t>0$}
-		{
-			Increase $j$ by 1.\;
-			Set $c \leftarrow 0$.
-		}
-	}
-}
-Return $y$.\;
-\caption{ETPF algorithm in one dimension.\label{alg:ETPF_1d}}
-\end{algorithm}
-\end{table}
-
-\subsection{Invertibility of the transport map}
-In this paper, we have considered an approach similar to that
-presented in \cite{parno2014transport}, where we aim to find an
-invertible map which approximates the true transport map between the
-reference measure and the target measure, however this true map is
-invariably itself not invertible. However, by considering embedding
-the map in higher dimensions\cite{whitney_self-intersections_1994,nash1956imbedding,takens_detecting_1981}, these issues of non-invertibility may be
-avoided, and higher quality maps may be found which better aid
-exploration and sampling of the whole state space.
-
-% \subsection{The chicken and egg paradox}
-% Another issue with the use of transport maps of this ilk, is that in
-% order to construct such a map, we require a representative sample from
-% the posterior distribution. Since these methods are designed to sample
-% from distributions which are inherently difficult to sample from, be
-% it due to multimodality, as looked at in \cite{cotter2015parallel}, or
-% as in this paper, due to the density being highly concentrated close
-% to a curved manifold in parameter space, it could be argued that accessing this initial set of
-% samples remains a hard problem. This remains an open problem, which we
-% are aiming to address in future work.
-
-\bibliographystyle{siam}
-\bibliography{bibliography}
-
-\end{document}

### 2019-12-04 Resubmission/TETAIS.tex
Started resubmission

--- 2019-01-17 TETAIS.tex
@@
 \newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 \renewcommand{\b}[1]{{\bf #1}}
+\newcommand{\edit}[1]{{\color{red} #1}} 
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 \end{abstract}
+\begin{keywords}Importance sampling, ensemble, transport map,
+  multiscale, stochastic reaction networks, Bayesian inverse problems
+\end{keywords}
 
@@
 
-In this paper, we particularly concern ourselves with the subject of
-inverse problems in stochastic reaction networks. This type of model
-is used in situations where ODE or diffusion\cite{gillespie2000chemical}
-approximations of chemical reaction kinetics are not viable, usually
-because of the low concentration of one or more chemical
-species\cite{gillespie1977exact}. Biochemical networks typically fall into
-this category, because of the small volume of the reactors involved (a
-single cell).
-
-One challenge with networks arising from these applications is that
-they often have some reactions which are occurring many times on a
-timescale in which other reactions in the system are unlikely to
-occur. This multiscale behaviour throws up two main challenges, both
-in observability in an experimental setting, but also in the context
-of modelling, where multiscale systems can make the simulation of the
-continuous time Markov chain computationally intractable.
-
-A range of numerical methods have been suggested to tackle this second
-problem, including tau-leaping\cite{rathinam2003stiffness,cao2006efficient}, diffusion approximations\cite{erban2006gene,cotter2011constrained,cotter2016error}, quasi-steady state
-approximations\cite{weinan2007nested,cao2005slow}, and other averaging
-arguments\cite{cotter2016constrained}. Some of these approaches are more
-computationally intensive than others, but this can be mitigated with the
-use of analytical results about the invariant densities of certain
-types of fast subsystems\cite{jahnke2007solving,anderson2010product,anderson2016product}.
-
-We would not expect the fast processes in such a system to be
-experimentally observable. In such a situation where only the slow
-variables in a system are identifiable, in an ideal world, we would be
-be able to integrate over the space of all possible trajectories of
-the fast processes. Since this is an insurmountable computational
-task, we might instead wish to approximate the likelihoods arising
-from such a set of observations by the use of an appropriate
-multiscale methodology.
-
-The partial observability of these systems lead to only
-partial-observability of the unknown parameters in the system, for
-example the reaction rates of the fast reactions. The resulting
-multiscale approximations, and the resulting approximations of the
-likelihoods arising from the inverse problem for such a network, 
-often exhibit 
-insensitivity to changes in certain directions of
-parameter space. This causes problems when aiming to sample from such
-a target, since the density is concentrated on a manifold whose
-tangential vectors point in directions of insensitivity. This
-\emph{sloppiness}\cite{gutenkunst2007universally,apgar2010sloppy} of the likelihood is a challenging problem
-that appears in many applications in science and engineering\cite{constantine2014active}.
+\edit{
+There are many applications in which complex posterior structures
+often arise. One family of such problems is those for which the
+forward model is insensitive to changes in certain directions of
+parameter space, sometimes called
+\emph{sloppiness}\cite{gutenkunst2007universally,apgar2010sloppy}. This
+sloppiness in the forward model appears in many applications in
+science and engineering\cite{constantine2014active}. Where one is
+aiming to solve an inverse problem based on such a model, this can
+lead to the posterior distribution being highly concentrated on lower
+dimensional manifolds, which represent the directions of
+sloppiness. This is particularly prevalent in partially observable
+systems, for example multiscale systems such
+as stochastic reaction networks, where only slowly changing quantities
+can be reliably observed.}
+
+% In this paper, we particularly concern ourselves with the subject of
+% inverse problems in stochastic reaction networks. This type of model
+% is used in situations where ODE or diffusion\cite{gillespie2000chemical}
+% approximations of chemical reaction kinetics are not viable, usually
+% because of the low concentration of one or more chemical
+% species\cite{gillespie1977exact}. Biochemical networks typically fall into
+% this category, because of the small volume of the reactors involved (a
+% single cell).
+
+% One challenge with networks arising from these applications is that
+% they often have some reactions which are occurring many times on a
+% timescale in which other reactions in the system are unlikely to
+% occur. This multiscale behaviour throws up two main challenges, both
+% in observability in an experimental setting, but also in the context
+% of modelling, where multiscale systems can make the simulation of the
+% continuous time Markov chain computationally intractable.
+
+% A range of numerical methods have been suggested to tackle this second
+% problem, including tau-leaping\cite{rathinam2003stiffness,cao2006efficient}, diffusion approximations\cite{erban2006gene,cotter2011constrained,cotter2016error}, quasi-steady state
+% approximations\cite{weinan2007nested,cao2005slow}, and other averaging
+% arguments\cite{cotter2016constrained}. Some of these approaches are more
+% computationally intensive than others, but this can be mitigated with the
+% use of analytical results about the invariant densities of certain
+% types of fast subsystems\cite{jahnke2007solving,anderson2010product,anderson2016product}.
+
+% We would not expect the fast processes in such a system to be
+% experimentally observable. In such a situation where only the slow
+% variables in a system are identifiable, in an ideal world, we would be
+% be able to integrate over the space of all possible trajectories of
+% the fast processes. Since this is an insurmountable computational
+% task, we might instead wish to approximate the likelihoods arising
+% from such a set of observations by the use of an appropriate
+% multiscale methodology.
+
+****Could move this material to stochastic network section****
+
+% The partial observability of these systems lead to only
+% partial-observability of the unknown parameters in the system, for
+% example the reaction rates of the fast reactions. The resulting
+% multiscale approximations, and the resulting approximations of the
+% likelihoods arising from the inverse problem for such a network, 
+% often exhibit 
+% insensitivity to changes in certain directions of
+% parameter space. This causes problems when aiming to sample from such
+% a target, since the density is concentrated on a manifold whose
+% tangential vectors point in directions of insensitivity. This
+% \emph{sloppiness}\cite{gutenkunst2007universally,apgar2010sloppy} of the likelihood is a challenging problem
+% that appears in many applications in science and engineering\cite{constantine2014active}.
+
+********Did I mean to get rid of this?**********
 
@@
 In Section \ref{sec:ETAIS}, we will briefly reintroduce the ensemble
-transport adaptive importance sampling (ETAIS) method. In Section \ref{sec:map} we show how an appropriate transport map can
+transform adaptive importance sampling (ETAIS) method. In Section \ref{sec:map} we show how an appropriate transport map can
 be constructed from importance samples which maps the posterior close
@@
 
-\section{Ensemble Transport Adaptive Importance Sampling}\label{sec:ETAIS}
-Ensemble transport adaptive importance sampling
+\section{Ensemble Transform Adaptive Importance Sampling}\label{sec:ETAIS}
+Ensemble transform adaptive importance sampling
 (ETAIS)\cite{cotter2015parallel} is an adaptive importance sampling framework
@@
 formula
-\[\mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
-  w_kf(\theta^{(k)}),\]
+\begin{equation} \mathbb{E}_\pi(f) \approx \frac{1}{\bar{w}} \sum_{k=1}^N
+  w_kf(\theta^{(k)}),\end{equation}
 where $\bar{w} = \sum_{k=1}^N w_k$ is the total weight.
@@
 \cdot, \beta)$ then 
-\[\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)},
-\beta).\]
+\begin{equation}\chi^{(k)} = \frac{1}{M} \sum_{i=1}^M q(\cdot ; \theta_i^{(k)},
+\beta).\end{equation}
 Usually the kernel $q(\cdot ;
@@
 target density, i.e.
-\[ \theta_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\]
+\begin{equation} \hat{\theta}_i^{(k)} \sim q(\cdot ; \theta_i^{(k)},\beta)\end{equation}
 for each $i = 1,2,\ldots,M$.  We now have a weighted sample, and it
@@
 	\For{$k=1,\ldots,N$}{
-		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_j^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:ETAIS_propose}\;
+		Sample $\hat{\theta}_i^{(k)} \sim q(\cdot; \theta_i^{(k)}, \beta)$, for $i = 1,\dots, M$.\label{algline:ETAIS_propose}\;
 		Calculate $w^{(k)} = (w_1^{(k)}, \ldots, w_M^{(k)})^\top$, where
-			\[
+			\begin{equation}
 				w^{(k)}_i = \frac{\pi(\hat{\theta}^{(k)}_i)}{\chi^{(k)}(\hat{\theta}^{(k)}_i; \theta^{(k)}, \beta)}.
-			\]
+			\end{equation}
 		\label{algline:ETAIS_weights}
@@
 	is equal to the target density $\pi(\theta)$ for all $\theta \in \mathcal{X}$. The pullback is defined in terms of the determinant of the Jacobian of $T$,
-	\[
+	\begin{equation}
 		|J_T(\theta)| = \text{det}\begin{bmatrix} \partial_{\theta_1} T_1(\theta) & \dots & \partial_{\theta_d} T_1(\theta) \\ \vdots & \ddots & \vdots \\ \partial_{\theta_1} T_d(\theta) & \dots & \partial_{\theta_d} T_d(\theta) \end{bmatrix}.
-	\]
+	\end{equation}
 \end{dfn}
@@
 % divergence is given by
-% \[D_\text{KL}(\pi_1\|\pi_2) = 
-% 		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\]
+% \begin{equation}D_\text{KL}(\pi_1\|\pi_2) = 
+% 		\mathbb{E}_{\pi_1}\left[\log\left(\frac{\pi_1(\theta)}{\pi_2(\theta)}\right)\right].\end{equation}
 % The KL divergence is not itself a norm, since
@@
 invertibility if we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
-\[
+\begin{equation}
 	\tilde{T}(\theta_1, \dots, \theta_n) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \\ \vdots \\
 		T_n(\theta_1, \dots, \theta_n) \end{bmatrix},
-\]
+\end{equation}
 where $T_i\colon \mathbb{R}^i \to \mathbb{R}$. % We assume that the target and reference probability densities are absolutely continuous on
@@
 % regularisation (which will be detailed later), we are guaranteed a unique invertible map $\tilde{T}$ with the property that
-% \[
+% \begin{equation}
 % 	\tilde{T}\mu_{\theta} \approx \mu_r.
-% \]
+% \end{equation}
 % % Relaxing the equality constraint to finding the approximate map
@@
 to be the Kullback-Leibler (KL) divergence between the posterior density and the pullback density,
-\[
+\begin{equation}
 	D_\text{KL}(\pi\|\tilde{\pi}) =
 		\mathbb{E}_\pi\left[\log\left(\frac{\pi(\theta)}{\tilde{\pi}(\theta)}\right)\right].
-\]
+\end{equation}
 This divergence results in some nice properties which we will explore in the following derivation.
 The KL divergence is not a true metric since it is not symmetric, however it is commonly used to measure the distance between probability distributions due to its relatively simple form, and because it provides a bound for the square of the Hellinger distance by Pinsker's inequality~\cite{pinsker1960information},
-\[
+\begin{equation}
 	D_{KL}(p\|q) \geq D_H^2(p,q),
-\]
+\end{equation}
 which is a true distance metric on probability distributions.
 Given the form of the pullback in Equation~\eqref{eq:pullback}, now taken through an approximate map $\tilde{T}$, the divergence becomes
-\[
+\begin{equation}
 	D_\text{KL}(\pi\|\tilde{\pi}) = \mathbb{E}_\pi\left[\log\pi(\theta) - \log\pi_r(\tilde{T}(\theta)) -
 		\log\left|J_{\tilde{T}}(\theta)\right|\right].
-\]
+\end{equation}
 We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
@@
 To find the best coupling, $\tilde{T} \in \mathcal{T}^{\ltri}$, we solve the optimisation problem,
-\[
+\begin{equation}
 	\tilde{T} = \arg\min_{T \in \mathcal{T}^{\smallltri}} \mathbb{E}_\pi\left[-\log\pi_r(T(\theta)) -
 		\log\left|J_T(\theta)\right|\right]
-\]
+\end{equation}
 which has a unique solution since the cost function is convex.
@@
 \end{equation}
-The parameter $\beta>0$ does not need to be tuned, as experimentation has shown that the choice
-$\beta=1$ is sufficient for most problems. This expectation can be
-approximated by using an MCMC approximation. The form of the penalisation term promotes maps which are
+\edit{The parameter $\beta>0$ does not need to be tuned, as experimentation has shown that the choice
+$\beta=1$ is sufficient for most problems. Note that this
+regularisation term is necessary to avoid overfitting of the transport
+map to what is often an incomplete and poorly converged sample from
+the posterior, and smaller values of $\beta$ may lead to maps which
+hinder further exploration of the state space beyond that already
+represented in the sample used to construct the map.
+
+The expectation in \eqref{eq:gen_map_optim}} can be
+approximated by using a Monte Carlo approximation. The form of the penalisation term promotes maps which are
 closer to the identity, and so prevents overfitting when the quality
@@
 
-\subsection{The structure of the map}
+\subsection{The structure of the map}*****CHECK HOW MUCH OF THIS IS
+IDENTICAL TO MARZOUK PAPER******
 
@@
 The Jacobian of $\tilde{T}$ is a lower triangular matrix,
-\[
+\begin{equation}
 	J_T(\theta) = \begin{bmatrix}
@@
 	\end{bmatrix}
-\]
+\end{equation}
 since $\partial_{\theta_n} \tilde{T}_k(\theta) = 0$ for all $n > k$.
@@
 \mathcal{J}_i \subset \mathbb{N}_0^i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
-\[
+\begin{equation}
 	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
-\]
+\end{equation}
 and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$.
@@
 to use polynomials of total order $p$,
-\[
+\begin{equation}
 	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
-\]
+\end{equation}
 which is optimal in terms of the amount of information captured by the map about the target.
@@
 particular $\theta^{(k)}$. If we organise our basis evaluations into two matrices,
-\[
+\begin{equation}
 	(F_i)_{k,\mathbf{j}} = \psi_\mathbf{j}(\theta^{(k)}), \quad \text{and} \quad (G_i)_{k,\mathbf{j}} =
 \frac{\partial\psi_\mathbf{j}}{\partial\theta_i}(\theta^{(k)}),
-\]
+\end{equation}
 for all $\mathbf{j}
 \in \mathcal{J}_i^\text{TO}$, and $k = 1,\dots,K$, then we have that
-\[
+\begin{equation}
 	T_i(\theta^{(k)}) = (F_i)_{k\cdot}\gamma_i \quad \text{and} \quad \frac{\partial T_i}{\partial \theta_i}(\theta^{(k)}; \gamma_i) = (G_i)_{k\cdot}\gamma_i,
-\]
+\end{equation}
 so \eqref{eq:gamma_map_optim} becomes
@@
 The regularisation term in Equation~\eqref{eq:blas_map_optim} can be approximated using Parseval's identity,
-\[
+\begin{equation}
 	\frac{1}{\bar{w}}\sum\limits_{k=1}^K \! w_k
@@
 		\sum\limits_{\mathbf{j}\in\mathcal{J}_i^\text{TO}} (\gamma_{i,\mathbf{j}}-\iota_\mathbf{j})^2,
-\]
+\end{equation}
 where $\iota$ is the vector of coefficients for the identity map. This is of course only true when
@@
 identity matrix. The first derivative of $C_i(T)$ is
-\[
+\begin{equation}
 	\nabla C_i(\gamma_i) = \frac{1}{\bar{w}}\left[F_i^\top WF_i\gamma_i - G_i^\top
 		W[G_i\gamma_i]^{-1}\right] + \beta(\gamma_i - \iota),
-\]
+\end{equation}
 again $[G_i\gamma_i]^{-1}$ is taken element-wise.
@@
 
-\begin{table}
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\BlankLine
-Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
-Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
-\For{$k \leftarrow 1, \dots, L-1$}{
-	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
-	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
-	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
-	Calculate:
-	\[
-		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
-	\]
-
-	Resample $\theta^{(k+1)} \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{\hat{\theta}^{(k)}_j}(\cdot)$.\;
-
-	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
-		\For{$i \leftarrow 1, \dots, n$}{
-			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
-				and update $\gamma_i^{(k+1)}$.\;
-		}
-	}{
-		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
-	}
-}
-\caption{ETAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportETAIS1}}
-\end{algorithm}
-\end{table}
-
-The first option is given in Algorithm~\ref{alg:TransportETAIS1}. We
-denote the ensembles of states in target space $\theta^{(k)} =
-\{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the
-reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble
-size. Similarly, the proposal states are denoted $r' =
-\{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) =
-\{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M,
-\hat{\theta}^{(k)}_M)\}$, where these pairs are the states which
-together form our sample from the target distribution. As in the standard version of the ETAIS algorithm we use the deterministic mixture weights.
+% \begin{table}
+% \begin{algorithm}[H]
+% \DontPrintSemicolon
+% \BlankLine
+% Initialise state $\theta^{(1)}_i = \theta_0$, \quad $i = 1,\dots,M$.\;
+% Initialise map $\bar{\gamma}^{(1)} = \iota$.\;
+% \For{$k \leftarrow 1, \dots, L-1$}{
+% 	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
+% 	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
+% 	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
+% 	Calculate:
+% 	\begin{equation}
+% 		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
+% 	\end{equation}
+
+% 	Resample $\theta^{(k+1)} \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{\hat{\theta}^{(k)}_j}(\cdot)$.\;
+
+% 	\eIf{$k\ \text{mod}\ K_U = 0$ and $k < K_\text{stop}$}{
+% 		\For{$i \leftarrow 1, \dots, n$}{
+% 			Solve \eqref{eq:weighted_map_optim} with $\{(w^{(1)},\hat{\theta}^{(1)}), \dots, (w^{(k+1)},\hat{\theta}^{(k+1)})\}$
+% 				and update $\gamma_i^{(k+1)}$.\;
+% 		}
+% 	}{
+% 		$\bar{\gamma}^{(k+1)} = \bar{\gamma}^{(k)}$.\;
+% 	}
+% }
+% \caption{ETAIS algorithm with adaptive transport map. Option 1.\label{alg:TransportETAIS1}}
+% \end{algorithm}
+% \end{table}
 
@@
 	Calculate:
-	\[
+	\begin{equation}
 		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
-	\]
+	\end{equation}
 
@@
 
-The second option, Algorithm~\ref{alg:TransportETAIS2}, is similar to
+The second option is given in Algorithm~\ref{alg:TransportETAIS2}. We
+denote the ensembles of states in target space $\theta^{(k)} =
+\{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the
+reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble
+size. Similarly, the proposal states are denoted $r' =
+\{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) =
+\{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M,
+\hat{\theta}^{(k)}_M)\}$, where these pairs are the states which
+together form our sample from the target distribution. As in the standard version of the ETAIS algorithm we use the deterministic mixture weights.
+
+The second option, is similar to
 the first except on Line~\ref{algline:TETAIS_resample} where rather
@@
 members. This means that the resampling step will be more efficient in
-moderately higher dimensions, which we discuss in Section~\ref{sec:TETAIS_higher_dim}.
+moderately higher dimensions, which we discuss in
+Section~\ref{sec:TETAIS_higher_dim}.
+
+\edit{The choice of ensemble size can be problem dependent, but the
+  behaviour of the algorithm is quite clear when the ensemble size is
+  too small, because the proposal distribution poorly represents the
+  targets, and the variance of the weights is high. The ensemble size
+  can also be made to be adaptive, so that when the variances grow too
+high the ensemble size can increase. The ensemble size needs to be
+controlled since the cost of the resampling step is often polynomial
+in the ensemble size.}
 
@@
 since otherwise the map will not be surjective. This results in a map of the form
-\[
+\begin{equation}
 	T(\theta_1, \theta_2) = \begin{bmatrix} T_1(\theta_1) \\ T_2(\theta_1, \theta_2) \end{bmatrix},
-\]
+\end{equation}
 where
-\begin{align*}
+\begin{align}
 		T_1(\theta_1) &= \gamma_{1,1} + \gamma_{1,2}\theta_1 + \gamma_{1,3}\theta_1^2 + \gamma_{1,4}\theta_1^3, \\
@@
 					 \gamma_{2,10}\theta_2^3.
-\end{align*}
+\end{align}
 Clearly even with only basis functions of total order 3, we have a large number of unknowns in our optimisation problem, $\bar{\gamma} \in \mathbb{R}^{14}$.
@@
 	Statistic \quad / \quad Algorithm & Transport M-H &
-                                                            Alg. \ref{alg:TransportETAIS1} & Alg. \ref{alg:TransportETAIS2}  \\ \cmidrule(lr){1-4}
+                                                           Transport
+                                                            ETAIS
+                                                            option 1 & Alg. \ref{alg:TransportETAIS2}  \\ \cmidrule(lr){1-4}
 	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
@@
 options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
-\subfigure[Comparison of Transport ETAIS option (3) with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
+\subfigure[Comparison of Transport ETAIS option 2 with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
 \caption{Convergence of algorithms; Transport M-H,
-  algorithm \ref{alg:TransportETAIS1}, algorithm \ref{alg:TransportETAIS2} for
+  Transport ETAIS option 1, algorithm \ref{alg:TransportETAIS2} for
   density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the MT algorithm.}
@@
 
-\section{Multiscale Methods for Stochastic Chemical Reaction
+\section{Inverse Problems for Multiscale Stochastic Chemical Reaction
   Networks}\label{sec:multi}
-In this Section, we discuss some recent advances in multiscale methods
-for stochastic reaction networks. Inverse problems arising in this
-area often lead to complex posterior
-distributions, which traditional MCMC methods can struggle to sample
-from. We will then go on to solve some inverse problems related to
-this in Section \ref{sec:num}, using the transport map versions of the
-ETAIS algorithm, as described in Section \ref{sec:TETAIS}.
+\edit{In this Section, we briefly introduce the idea of a continuous
+  time Markov chain model for stochastic reaction networks. We consider the formulation of the Bayesian inference of
+  the reaction parameters given exact data. In particular we consider
+  the case where only slow variables in the system are observable, and
+  hence the true likelihood function is intractable. In this case, the
+  likelihood can be approximated using multiscale methods. The data in
+  these scenarios is often insufficient to identify all of the
+  parameters, and this can lead to posterior distributions which are
+  heavily concentrated close to lower dimensional manifolds, which are
+  challenging to sample from, and which motivated the development of
+  the methodology outlined in this paper.}
 
@@
 model the dynamics through a discrete stochastic model, such as a
-continuous time Markov chain.
+continuous time Markov chain. \edit{Further details about continuous time
+  Markov chains are given in the supplementary material.} ****ARE THEY?****
 
@@
 likely that reaction is to fire, defined by
-\[\alpha_j(X(t)) = \lim_{dt \to 0} \mathbb{P}(\text{Reaction $R_j$ in
-    the time interval  } \quad s \in [t, t+ dt] ).\] If a system
-satisfies what is called \emph{mass action kinetics}, then the form of
-the function $\alpha_j$ is determined, up to a rate constant, by the
-reactants involved in that reaction:
-\begin{equation}\label{eq:MAK}
-\alpha_j(X) = k_j \prod_{m=1}^{N_s} \prod_{n=0}^{\nu_{j,m} -1} (X_m - n),
-\end{equation}
-where $\nu_{j,m}$ is the $m$th component of the stoichiometric vector
-$\nu_j$, $X_m$ is the $m$th component of the state vector $X$, and the
-$k_j$ are rate constants.
+\begin{equation}\alpha_j(X(t)) = \lim_{dt \to 0} \mathbb{P}(\text{Reaction $R_j$ in
+    the time interval  } \quad s \in [t, t+ dt] ).\end{equation} % If a system
+% satisfies what is called \emph{mass action kinetics}, then the form of
+% the function $\alpha_j$ is determined, up to a rate constant, by the
+% reactants involved in that reaction:
+% \begin{equation}\label{eq:MAK}
+% \alpha_j(X) = k_j \prod_{m=1}^{N_s} \prod_{n=0}^{\nu_{j,m} -1} (X_m - n),
+% \end{equation}
+% where $\nu_{j,m}$ is the $m$th component of the stoichiometric vector
+% $\nu_j$, $X_m$ is the $m$th component of the state vector $X$, and the
+% $k_j$ are rate constants.
 
@@
 \begin{equation}\label{eq:RTC}
-X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j \int_0^t \alpha_j(X(s)) ds.
+X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j Y_j \left (\int_0^t
+  \alpha_j(X(s)) \right ) ds.
 \end{equation}
@@
 sampled exactly, for instance using the Gillespie SSA\cite{gillespie1977exact}, or
-its variants\cite{gillespie2007stochastic,cao2004efficient,anderson2007modified}. However, if the system is stiff, i.e. there
-are some reactions which are firing many times on a timescale for
-which others are unlikely to fire at all, then trajectories can become
-prohibitively expensive to simulate, since these methods simulate
-every single reaction event with the same cost. In such a system, one
-might employ multiscale methods in order to approximate trajectories
-at a lower cost than the exact algorithms.
-
-One common approach is to induce the quasi-equilibrium approximation
-QEA. This approximation makes the assumption that fast reactions
+its
+variants\cite{gillespie2007stochastic,cao2004efficient,anderson2007modified}.
+
+% However, if the system is stiff, i.e. there
+% are some reactions which are firing many times on a timescale for
+% which others are unlikely to fire at all, then trajectories can become
+% prohibitively expensive to simulate, since these methods simulate
+% every single reaction event with the same cost. In such a system, one
+% might employ multiscale methods in order to approximate trajectories
+% at a lower cost than the exact algorithms.
+
+% One common approach is to induce the quasi-equilibrium approximation
+% QEA. This approximation makes the assumption that fast reactions
+% enter quasi-equilibrium on a timescale which is negligible with
+% respect to the timescale on which the slow dynamics in the system are
+% evolving. This amounts to taking the asymptotic limit that the
+% propensities of the ``slow'' reaction channels are equal to zero. This
+% allows us to sample approximate trajectories of the slow reactions
+% without the need to compute many fast reaction events.
+
+% Multiscale methods are not only of use when forward evaluations of the
+% dynamics are costly, but can also be of use where we attempt to solve
+% an inverse problem where the fast variables are unobservable. One
+% approach in this situation would be to integrate over all possible
+% trajectories of the fast variables, but this would almost always be
+% prohibitively expensive. Another approach would be to use an
+% appropriate multiscale approximation, so that the effective dynamics
+% of the slow variables can be approximate without the need to consider
+% the rapid fluctuations of the fast variables in the system.
+
+\subsection{Likelihoods arising from multiscale stochastic reaction networks}
+Suppose that we are able to exactly observe the number of molecules of
+each chemical species in a system that can be well approximated by
+\eqref{eq:RTC}. \edit{Each reaction in the system involves the firing
+  of a Poisson process with rate which is a function of the number of
+  reactant molecules present, and the reaction rates. The likelihood
+  of each reaction is simply the product of the probability density of
+  an exponential random
+  variable which gives the waiting time since the last reaction, with
+  the mass function of a
+  multinomial distribution describing the likelihood of each type of
+  reaction from that state. Since the
+  reactions are assumed to be independent, the likelihood of the whole
+  path is a product of all the individual reaction likelihoods. This
+  can be computed in a more efficient way by considering all reactions
+  in the data from the same point in the state space at the same
+  time. A detailed derivation of this is given in the
+  ******supplementary material******.
+
+If we are only able to observe the slow species, then we cannot write
+down an explicit likelihood for the path. One approach is to try to
+simultaneously infer the
+missing data of the fast variables, but this is computationally
+intractable in all but the simplest of cases. The other option is to
+invoke a multiscale approximation to the dynamics. This allows us to
+approximate effective propensities for the slow reactions, and hence
+to approximate the likelihood of paths of the slow variables. A
+detailed derivation of this likelihood is also given in the
+******supplementary material******.
+
+In this work, we will consider two different multiscale
+approximations. The first is the quasi-equillibrium approximation
+which makes the assumption that fast reactions
 enter quasi-equilibrium on a timescale which is negligible with
 respect to the timescale on which the slow dynamics in the system are
-evolving. This amounts to taking the asymptotic limit that the
-propensities of the ``slow'' reaction channels are equal to zero. This
-allows us to sample approximate trajectories of the slow reactions
-without the need to compute many fast reaction events.
-
-This approach can work well when there is a large timescale gap
+evolving. This approximation is the basis of several
+methods\cite{weinan2007nested,cao2005slow}. This approach can work well when there is a large timescale gap
 between the fast and slow reactions in a system, but where the
@@
 
-Multiscale methods are not only of use when forward evaluations of the
-dynamics are costly, but can also be of use where we attempt to solve
-an inverse problem where the fast variables are unobservable. One
-approach in this situation would be to integrate over all possible
-trajectories of the fast variables, but this would almost always be
-prohibitively expensive. Another approach would be to use an
-appropriate multiscale approximation, so that the effective dynamics
-of the slow variables can be approximate without the need to consider
-the rapid fluctuations of the fast variables in the system.
-
-\subsection{Likelihoods arising from stochastic reaction networks}
-Suppose that we are able to exactly observe the number of molecules of
-each chemical species in a system which satisfies mass action
-kinetics, and which can be well approximated by
-\eqref{eq:RTC}. Suppose that we wish to be able to infer the value of
-the rate constants of each reaction from these observations. Even with no observational noise, since the dynamics
-of the system are stochastic, this still leads to a Bayesian inverse
-problem where we can only infer a joint probability distributions on the
-reaction parameters.
-
-Suppose that we are in state $X(t) = X_0$. There are two independent
-univariate random variables which decide when and what the next
-reaction in the system is. There is the $j$th waiting time $\tau_j$ to the
-next reaction, which is given by an exponential random variable
-\[\tau_j \sim \exp \left ( \frac{1}{\alpha_0(X(t))} \right ),\]
-where $\alpha_0(X(t)) =\sum_{i}^{N_r} \alpha_i(X(t))$ is the total
-propensity in the system. The second is a multinomial random variable $r_j$
-which dictates which reaction has occurred during the $j$th reaction,
-which takes the value
-$i \in \{1,2,\ldots,N_r\}$ with probability
-\[\mathbb{P}(r_j = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\]
-As such, in order to compute the likelihood of a particular trajectory
-given a possible realisation of the reaction parameters, it is
-sufficient to have access to the total time spent in each state, and
-the frequency of each reaction which led to leaving each state.
-
-From this formulation, we see that the random variables $(\tau_j, r_j)$ only depend on the states $\mathbf{X}(t_{j-1})$ and so are Markovian.
-This conditional independence means that we can group events together by what state the system was in when the event happened.
-We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
-\[
-	T(\mathbf{Y}) = \sum\limits_{j=1}^M \tau_j\; \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y}).
-\]
-This random variable, $T(\mathbf{Y})$, is a sum of exponentially
-distributed random variable, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
-\begin{equation}\label{eqn:chem_time_dist}
-	T(\mathbf{Y}) \sim \text{Gamma}\left(\alpha=K(\mathbf{Y}),~\beta = \alpha_0(\mathbf{Y}; \mathbf{k})\right),
-\end{equation}
-where $K(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y})$.
-
-Similarly, we can define the reactions which occurred when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_{N_r}(\mathbf{Y})]^\top$ where
-\[
-	r_i(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(r_j = i \textbf{ and }\mathbf{X}(t_{j-1}) = \mathbf{Y}).
-\]
-Here all the random variables $r_j$ follow the same multinomial distribution, and so
-\begin{equation}\label{eqn:chem_react_dist}
-	\mathbf{r}(\mathbf{Y}) \sim \text{Multinomial}(K(\mathbf{Y}),~\mathbf{p}(\mathbf{Y})). 
-\end{equation}
-The random variables defined in Equations~\eqref{eqn:chem_time_dist} and \eqref{eqn:chem_react_dist} are sufficient statistics for the posterior distribution $\pi(\mathbf{k}|D)$.
-With these definitions we define two new structures
-\[
-	\mathbf{T} = [T(\mathbf{Y}_1), \dots, T(\mathbf{Y}_K)]^\top, \quad \text{and} \quad \mathbf{R} = [\mathbf{r}(\mathbf{Y}_1), \dots, \mathbf{r}(\mathbf{Y}_K)]^\top,
-\]
-where $K = |\mathcal{S}|$, the number of states in $\mathcal{S}$, and each state $\mathbf{Y}_i \in \mathcal{S}$ has been enumerated. We use these structures to define shorter notation,
-\[
-	\mathbf{T}_i = T(\mathbf{Y}_i), \quad \mathbf{R}_{ij} = r_j(\mathbf{Y}_i), \quad \text{and} \quad \mathbf{K}_i = K(\mathbf{Y}_i).
-\]
-
-To construct the posterior distribution for the reaction rates, $\mathbf{k}$, in the chemical system, we formulate the likelihood using the sufficient statistics derived in the previous section.
-Due to the non-negativity of these reaction rates, we assign a Gamma prior distribution to each rate.
-Given the distributions in Equations~\eqref{eqn:chem_time_dist} and~\eqref{eqn:chem_react_dist} for our data, the likelihood of observing the data $\mathbf{R}$ and $\mathbf{T}$ is
-\[
-	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{R}_{i\cdot}; \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
-\]
-where again $\mathbf{Y}_i \in \mathcal{S}$ and the propensities
-$\alpha_i$ and probabilities $p_i$ depend on the reaction rates
-$\mathbf{k} = [k_1, k_2, \ldots k_{N_r}]^\top$ through the concept of
-mass action kinetics given in \eqref{eq:MAK}.
-
-Our choice of Gamma prior distributions with hyper-parameters $(a_i, b_i)$ results in the posterior distribution of the form
-\begin{align}
-	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
-	\prod_{i=1}^{N_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
-		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
-				\mathbf{K}_i\log \alpha_0(\mathbf{Y}_i; \mathbf{k}) - \mathbf{T}_i\alpha_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
-%		& \qquad\qquad\qquad
-				+\sum_{j=1}^{N_r} \mathbf{R}_{ij}\log\mathbf{p}_j(\mathbf{Y}_i; \mathbf{k})
-			\Bigg]  \notag\\
-		&	\qquad\qquad\qquad{} + \sum_{i=1}^{N_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}. \label{eqn:chem_posterior}
-\end{align}
-
-\subsection{Approximation of likelihoods in multiscale chemical
-  networks}
-Suppose now that we only observe the slower variables in a
-multiscale chemical system of this type. Suppose that $n_r < N_r$ is the number of slow
-reactions in the system, and the slow reactions are given by $\{R_{s_1},
-R_{s_2}, \ldots R_{s_{n_r}} \} \subset \{R_1, R_2, \ldots , R_{N_r}
-\}$. The propensities of the slow
-reactions $\alpha_{s_i}$ may depend on the value of the fast variables which is
-unknown. However, the invariant distribution of the fast variables
-conditioned on the slow variables in the system can be approximated
-using a multiscale method, such as the QEA or the CMA, as described
-briefly in Section \ref{sec:multi}. Once the approximations have been
-made, we arrive at approximate \emph{effective} propensities $\bar{\alpha}_{s_i}$, which
-have been averaged over the computed invariant distributions of the
-fast variables. Then the approximate effective dynamics on the slow variable $S(t)$ are given by 
-\begin{equation}\label{eq:RTCS}
-S(t) = S(0) + \sum_{j=1}^{n_r} \nu_{s_j} \int_0^t \bar{\alpha}_{s_j}(S(s)) ds.
-\end{equation}
-
-In turn, the approximate posterior distribution on the reaction
-parameters $\mathbf{k}$ is then given by
-\begin{align}
-	\pi(\mathbf{k}|\mathbf{R}, \mathbf{T}) &\propto \ell(\mathbf{R}, \mathbf{T}|\mathbf{k})
-	\prod_{i=1}^{n_r} \! \text{Gamma}(k_i; a_i, b_i) \notag \\
-		&\propto \exp\Bigg\{\sum\limits_{i=1}^K \Bigg[
-				\mathbf{K}_i\log
-                  \bar{\alpha}_0(\mathbf{Y}_i; \mathbf{k}) -
-                  \mathbf{T}_i \bar{\alpha}_0(\mathbf{Y}_i; \mathbf{k}) %\notag \\
-%		& \qquad\qquad\qquad
-				+\sum_{j=1}^{n_r} \mathbf{R}_{ij}\log \bar{\mathbf{p}}_j(\mathbf{Y}_i; \mathbf{k})
-			\Bigg]  \notag\\
-		&	\qquad\qquad\qquad{} + \sum_{i=1}^{n_r} ((a_i-1)\log k_i - b_ik_i)\Bigg\}, \label{eq:Apos}
-\end{align}
-where $\bar{\alpha}_0(\mathbf{Y}_i; \mathbf{k}) = \sum_{j=1}^{n_r}
-\bar{\alpha}_j(\mathbf{Y}_i; \mathbf{k})$ is the total of the
-effective propensities, $n_r$ is the number of slow reactions,
-$\bar{\mathbf{p}}(\mathbf{Y}_i; \mathbf{k})$ is the multiscale
-approximation of the expected proportion of each slow reaction at
-state $\mathbf{Y}_i$ and with reaction rates $\mathbf{k}$ and the
-data is limited only to changes in the slow variable due to the
-occurrence of slow reactions.
+}
 \section{Numerical Examples}\label{sec:num}
@@
 The effective propensity $\bar{\alpha}_4(S)$ can be  approximated through
-application of a multiscale approximation, as detailed in Section
-\ref{sec:multi}, and in more detail in \cite{cotter2016constrained}. These effective
+application of a multiscale approximation, as detailed in \edit{the
+  supplementary material}, and in more detail in \cite{cotter2016constrained}. These effective
 propensities can often be computed without the need for expensive
@@
 the effective rate of degradation of the slow variable $S$ is given by
-\[
+\begin{equation}
 	\bar{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3}.
-\]
+\end{equation}
 Similarly, the analysis of the constrained system as discussed in~\cite{cotter2016constrained} yields the effective propensity
@@
 We aim to recover three different posterior distributions. In the
-first, which is of the form \eqref{eqn:chem_posterior}, we assume that we can fully and exactly observe the whole state
+first, we assume that we can fully and exactly observe the whole state
 vector for the whole of the observation time window $t \in [0,T]$. In
@@
 variable $S = X_1 + X_2$. In these last two examples, we approximate the
-posterior using \eqref{eq:Apos} with QEA and CMA approximations of the
+posterior using QEA and CMA approximations of the
 effective dynamics respectively.
@@
 \in \mathcal{X} = \mathbb{R}_+^4$. These four parameters are assigned Gamma prior distributions,
-\[
+\begin{equation}
 	k_i \sim \text{Gamma}(\cdot; \alpha_i, \beta_i), \quad \text{for} \quad i = 1, \dots, 4.
-\]
+\end{equation}
 The hyper-parameters corresponding to each of these prior distributions are given in Table~\ref{tab:multiscale_priors}. These priors are the same for each of the three posterior distributions.
@@
 \centering
-\begin{tabular}{|l|R{2em}R{2em}R{2em}R{2em}|}
-	\hline
-	Parameter$\backslash$Dimension $i$ & 1 & 2 & 3 & 4 \\ \hline
+% \begin{tabular}{|l|R{2em}R{2em}R{2em}R{2em}|}
+
+\begin{tabular}{lrrrr} 
+	\toprule
+	Parameter$\backslash$Dimension $i$ & 1 & 2 & 3 & 4  \\ \cmidrule(lr){1-5} 
 	$\alpha_i$ & 150 & 5 & 5 & 3 \\
-	$\beta_i$ & $\frac{15}{9}$ & $\frac{5}{12}$ & $\frac{5}{12}$ & $1$ \\ \hline 
+	$\beta_i$ & $\frac{15}{9}$ & $\frac{5}{12}$ & $\frac{5}{12}$ & $1$ \\ \bottomrule 
 \end{tabular}
@@
 \includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
-\caption{CMA approximation \eqref{eq:Apos} of the posterior arising from observations of
+\caption{CMA approximation of the posterior arising from observations of
   the slow variable $S = X_1 + X_2$ for system
@@
 
-\begin{figure}
-	\centering
-	\begin{tikzpicture}
-		\pgfmathsetseed{1}
-
-		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
-		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$R$};
-
-		\draw (A) \irregularcircle{1cm}{2.5mm};
-		\draw (B) \irregularcircle{1cm}{0.5mm};
-
-		\draw [->] (A) -- (B);
-
-		\node at (2.5,1) {$T\colon\mathcal{X}\rightarrow R$};
-
-		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
-		\node at (5,-1.5) {$r\in R\subset\mathbb{R}^d$};
-	\end{tikzpicture}
-
-	\begin{tikzpicture}
-		\pgfmathsetseed{1}
-
-		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
-		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
-		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};
-
-		\draw (A) \irregularcircle{1cm}{2.5mm};
-		\draw (B) \irregularcircle{1cm}{1.1mm};
-		\draw (C) \irregularcircle{1cm}{0.5mm};
-
-		\draw [->] (A) -- (B);
-		\draw [->] (B) -- (C);
-
-		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
-		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};
-
-		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
-		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
-		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
-	\end{tikzpicture}
-	\caption{Couplings between parameter space and reference
-          when using the transport map with (bottom) and without (top)
-        a log preconditioner.}
-	\label{fig:chem_coupling}
-\end{figure}
-
-Figure~\ref{fig:chem_coupling} (top) shows how we define a bijective map,
-$\hat{T}$, between parameter space $\mathcal{X}$ and a reference space
-$R$.
-When using the transport map proposal distribution, we prefix the proposal method with a T, e.g.
-MH-RW  and MH-TRW, as well as ETAIS-RW and ETAIS-TRW.
+% \begin{figure}
+% 	\centering
+% 	\begin{tikzpicture}
+% 		\pgfmathsetseed{1}
+
+% 		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
+% 		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$R$};
+
+% 		\draw (A) \irregularcircle{1cm}{2.5mm};
+% 		\draw (B) \irregularcircle{1cm}{0.5mm};
+
+% 		\draw [->] (A) -- (B);
+
+% 		\node at (2.5,1) {$T\colon\mathcal{X}\rightarrow R$};
+
+% 		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
+% 		\node at (5,-1.5) {$r\in R\subset\mathbb{R}^d$};
+% 	\end{tikzpicture}
+
+% 	\begin{tikzpicture}
+% 		\pgfmathsetseed{1}
+
+% 		\node[inner sep=1.5cm,outer sep=0] (A) at (0,0) {$\mathcal{X}$};
+% 		\node[inner sep=1.5cm,outer sep=0] (B) at (5,0) {$\mathcal{Y}$};
+% 		\node[inner sep=1.5cm,outer sep=0] (C) at (10,0) {$R$};
+
+% 		\draw (A) \irregularcircle{1cm}{2.5mm};
+% 		\draw (B) \irregularcircle{1cm}{1.1mm};
+% 		\draw (C) \irregularcircle{1cm}{0.5mm};
+
+% 		\draw [->] (A) -- (B);
+% 		\draw [->] (B) -- (C);
+
+% 		\node at (2.5,1) {$\log\colon\mathcal{X}\rightarrow\mathcal{Y}$};
+% 		\node at (7.5,1) {$T\colon\mathcal{Y}\rightarrow R$};
+
+% 		\node at (0,-1.5) {$\mathbf{k}\in\mathcal{X}\subset\mathbb{R}_+^d$};
+% 		\node at (5,-1.5) {$X\in\mathcal{Y}\subset\mathbb{R}^d$};
+% 		\node at (10,-1.5) {$r\in R \subset\mathbb{R}^d$};
+% 	\end{tikzpicture}
+% 	\caption{Couplings between parameter space and reference
+%           when using the transport map with (bottom) and without (top)
+%         a log preconditioner.}
+% 	\label{fig:chem_coupling}
+% \end{figure}
+
+% Figure~\ref{fig:chem_coupling} (top) shows how we define a bijective map,
+% $\hat{T}$, between parameter space $\mathcal{X}$ and a reference space
+% $R$.
 In practice we cannot ensure that the approximate map $\tilde{T}$
@@
 state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in
-$R$ do not necessarily map back onto $\mathcal{X}$.
-
-Therefore we also consider how these algorithms perform when the transport map
-is preconditioned by the logarithmic function applied to the parameters. We choose to use this log
-preconditioner since it converts our strictly positive parameter
-space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset
-\mathbb{R}^d$. This allows us to define $T$ between two subsets of
-$\mathbb{R}^d$, which means that all proposals made on the reference
-space are mapped to valid
-proposals on the true parameter space. It also reduces the work required of the map, which since
-we use only a finite dimensional approximation, can improve
-performance and stability.
-As before, the proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
-MH-logRW and MH-logTRW, with, ETAIS-logRW and ETAIS-logTRW.
-
-Figure~\ref{fig:chem_coupling} (bottom) displays the composition of maps
-when we include this $\log$ preconditioner leading to an intermediate space $\mathcal{Y}$.
-The inclusion of this additional map means that we must again alter our importance weight definition to reflect the pullback from $R$ through $\mathcal{Y}$.
-The weight is now
-\[
+$R$ do not necessarily map back onto $\mathcal{X}$. \edit{This can be
+  avoided, however, by employing a logarithmic preconditioner for the
+  transport map.
+The proposal distributions are labelled with a T for transport map and when using the intermediate space we prepend `log' to the proposal method, e.g.
+MH-logRW and MH-logTRW, with, ETAIS-logRW and ETAIS-logTRW.}
+
+% Therefore we also consider how these algorithms perform when the transport map
+% is preconditioned by the logarithmic function applied to the parameters. We choose to use this log
+% preconditioner since it converts our strictly positive parameter
+% space, $\mathcal{X}$, into an intermediate space $\mathcal{Y} \subset
+% \mathbb{R}^d$. This allows us to define $T$ between two subsets of
+% $\mathbb{R}^d$, which means that all proposals made on the reference
+% space are mapped to valid
+% proposals on the true parameter space. It also reduces the work required of the map, which since
+% we use only a finite dimensional approximation, can improve
+% performance and stability.
+
+% Figure~\ref{fig:chem_coupling} (bottom) displays the composition of maps
+% when we include this $\log$ preconditioner leading to an intermediate space $\mathcal{Y}$.
+
+\edit{ The inclusion of the composition with the logarithmic map} means that we must alter our importance weight definition to reflect the pullback from $R$ through $\mathcal{Y}$. The weight is now
+\begin{equation}
 	w_i(\theta) = \frac{\pi(\theta|\mathbf{R},\mathbf{T})}{\chi(\tilde{T}\circ\log(\theta)|\tilde{T}\circ\log(\theta^{(i-1)}))|J_{\tilde{T}\circ\log}(\theta)|},
-\]
+\end{equation}
 where $\theta$ is a proposed new state, $\theta^{(i-1)}$ is the ensemble of states from the previous iteration, and $J_{\tilde{T}\circ\log}(\theta)$ is the Jacobian of the composition of the two maps.
 This Jacobian is straightforward to calculate,
-\[
+\begin{equation}
 	|J_{\tilde{T}\circ\log}(\theta')| = |J_{\tilde{T}}(\log(\theta'))||J_{\log}(\theta')|,
-\]
+\end{equation}
 where the first determinant is computed as in
 \eqref{eqn:separable_jacobian}, and the second is given by
-\[
+\begin{equation}
 	|J_{\log}(\theta')| = \prod\limits_{i=1}^d \frac{1}{\theta'_i}.
-\]
+\end{equation}
 
@@
 For this problem, we continue to use monomials in each dimension in our transport map construction. We use polynomials of total order $p=3$ as the basis functions, i.e.
-\[
+\begin{equation}
 	T_i(\theta) = \sum_{\mathbf{j}\in\mathcal{J}^{\text{TO}}_i(p)} \gamma_{i,\mathbf{j}}\psi_{\mathbf{j}}(\theta) \quad \text{where} \quad \psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \theta_k^{j_k},
-\]
+\end{equation}
 and
-\[
+\begin{equation}
 	\mathcal{J}^{\text{TO}}_i(p) = \{\mathbf{j} \in \mathbb{N}^d_0\ |\ \|\mathbf{j}\|_1 \leq p, \ \text{and}\ j_k = 0\ \forall k > i\}.
-\]
+\end{equation}
 
@@
 have so far discussed. Convergence is shown only for the CMA
-approximation of the posterior \eqref{eq:Apos} with effective rate for
+approximation of the posterior with effective rate for
 the degradation of $S$ given by \eqref{eqn:chem_CMA_rate}, but we
@@
 	& & $\delta$ & & $\delta$ & ESS \\ \midrule
-	RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
-	TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
+	% RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
+	% TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
 	logRW & & 1.7e-1 & & 1.2e-0 & 6.0e-2 \\
@@
 \centering
-\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
-\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}}
-\caption{Convergence of eight different MCMC algorithms for the CMA
+% \subfigure[Sampling algorithms with no preconditioner for
+    %     $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
+\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}
+\caption{Convergence of four different MCMC algorithms for the CMA
   approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}.}
@@
 
-Convergence of the eight algorithms for this example are shown in
+\edit{
+Convergence of the four algorithms for this example are shown in
 Figure~\ref{fig:chem_multiscale_L2}. We first note the poor
@@
 
-The ETAIS variants with RW and TRW proposals perform similarly on both
-sample spaces. When sampling without a preconditioner, the transport map is not
-quite as efficient, largely due to the difficulties discussed in the
-previous section i.e. many proposals are made which do result in
-negative reaction rates. Sampling with transport maps with a $\log$
-preconditioner leads to more comparable Monte Carlo errors, with the
-logTRW being apparently slightly less stable. This proposal method
+% The ETAIS variants with RW and TRW proposals perform similarly on both
+% sample spaces. When sampling without a preconditioner, the transport map is not
+% quite as efficient, largely due to the difficulties discussed in the
+% previous section i.e. many proposals are made which do result in
+% negative reaction rates. Sampling with transport maps with a $\log$
+% preconditioner leads to more comparable Monte Carlo errors, with the
+% logTRW being apparently slightly less stable. 
+This proposal method
 becomes more stable as we increase either the ensemble size, or the
@@
 of likelihood evaluations coming from the ETAIS-logTRW algorithm.
+}
 
@@
 \centering
-\subfigure[Push forward of the posterior sample through $\hat{T}$ with
-no log
-preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
-\\
+% \subfigure[Push forward of the posterior sample through $\hat{T}$ with
+% no log
+% preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
+% \\
 \subfigure[Push forward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
-\caption{Push forward of the target measure on reference space for the the CMA
-  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}, found using the TRW and logTRW proposal distributions.
-Components are linked by the relation $r_i = T_i(k_i)$ in (a) and $r_i = T_i\circ\log(k_i)$ in (b).}
+\caption{\edit{Push forward of the target measure on reference space for the the CMA
+  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}, found using the logTRW proposal distributions.
+Components are linked by the relation $r_i = T_i\circ\log(k_i)$.}}
 \label{fig:chem_reference_spaces}
@@
 
-We now look at the proposal distributions of the transport map accelerated algorithms.
-In Figure~\ref{fig:chem_reference_spaces}, we see the reference spaces found by; (a) mapping the posterior through the map $\hat{T}$, (b) by mapping the posterior through $T\circ\log$.
-For the most part, each of these marginal distributions can be recognised as a Gaussian.
+\edit{
+We now look at the proposal distributions of the transport map accelerated algorithm.
+In Figure~\ref{fig:chem_reference_spaces}, we see the reference space found by mapping the posterior through $T\circ\log$.} For the most part, each of these marginal distributions can be recognised as a Gaussian.
 However, with the exception of $\mathbb{P}(r_2,r_3)$, we would not consider them to be close to a standardised $\mathcal{N}(0, \text{I})$ distribution.
@@
 
-In Figure~\ref{fig:chem_reference_spaces} (b) we see that
+\edit{We see that
 $\text{var}(r_1)$ and $\text{var}(r_4)$ are much smaller than
 $\text{var}(r_2)$ and $\text{var}(r_3)$. To combat this we have a
-number of choices. We might wish to learn the covariance structure of
+number of choices.} We might wish to learn the covariance structure of
 the push forward of the target onto the reference space, and
 incorporate this information into the covariances of the mixture
-components in an adaptive scheme.
-Another option is to increase the total order of our index set.
-For these numerics we have chosen $p=3$, but we can obtain reference spaces which are closer to $\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
+components in an adaptive scheme. Another option is to increase the
+total order of our index set. For these numerics we have chosen $p=3$,
+but we can obtain reference spaces which are closer to
+$\mathcal{N}_d(0, \text{I})$ by choosing a larger $p$.
+
+\edit{At this point, we note that if the transport map found is equal
+  or close to the Knothe-Rosenblatt rearrangement, then the most
+  efficient sampling strategy is clearly to independently sample from the
+  reference measure and pull these samples back through the map onto
+  (or close to) the target. However we see from Figure
+  \ref{fig:chem_reference_spaces} that in practice for very complex
+  targets, the approximation of the transport map is not as good, and
+  as such the push forward of the target onto the reference
+  space differs greatly from the the intended reference measure. In
+  this case, the pull back of the reference measure will be far from
+  the target distribution, and this independent sampling strategy
+  will be inefficient. This motivates the use of the ETAIS algorithm
+  to efficiently sample from the more complex distributions arising
+  from the push forward of the target distribution. ****HIGHLIGHT THIS
+  POINT ELSEWHERE -DEFINITELY IN THE CONCLUSIONS AND MAYBE THE INTRO/ABSTRACT?*****
+  }
 
 \subsubsection{Comparison of the Constrained and QEA approaches}
+****GREATLY REDUCE THIS SECTION****
+****Jump straight to Bayes' factors and Figure 7.5****
 
@@
 arising in this problem. Recall that the approaches differed only in the form of the effective degradation propensity $\hat{\alpha}_4$,
-\[
+\begin{equation}
 	\hat{\alpha}_4^{\text{QEA}}(s) = \frac{k_2k_4s}{k_2+k_3} \quad \text{and} \quad \hat{\alpha}_4^{\text{CMA}}(s) = \frac{k_2k_4s}{k_2+k_3+k_4}.
-\]
+\end{equation}
 This difference in the denominator causes a shift in the manifold on
-which the posterior is concentrated, as can be seen in
-Figure~\ref{fig:chem_diff}, in which we plot the difference in the 
-posteriors densities
-\begin{equation}\label{eqn:chem_diff}\pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
-\end{equation}
-
-\begin{figure}[htb]
-\centering
-\includegraphics[width=0.9\textwidth]{"images/Applications/Diff_CMA_QEA"}
-\caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
-\label{fig:chem_diff}
-\end{figure}
-
-Since the two posteriors have been approximated using an MCMC sample,
-there is a visible amount of Monte Carlo static, particularly in the tails of
-the distributions. We can see that the differences in the marginals
-for $k_1, k_2,$ and $k_3$ are relatively small, but the marginal for
-$k_4$ varies by a significant amount.
+which the posterior is concentrated.
+  % , as can be seen in
+% Figure~\ref{fig:chem_diff}, in which we plot the difference in the 
+% posteriors densities
+% \begin{equation}\label{eqn:chem_diff}\pi^{\text{CMA}}(\mathbf{k}|\mathbf{R},\mathbf{T}) - \pi^{\text{QEA}}(\mathbf{k}|\mathbf{R},\mathbf{T}).
+% \end{equation}
+
+% \begin{figure}[htb]
+% \centering
+% \includegraphics[width=0.9\textwidth]{"images/Applications/Diff_CMA_QEA"}
+% \caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
+% \label{fig:chem_diff}
+% \end{figure}
+
+% Since the two posteriors have been approximated using an MCMC sample,
+% there is a visible amount of Monte Carlo static, particularly in the tails of
+% the distributions. We can see that the differences in the marginals
+% for $k_1, k_2,$ and $k_3$ are relatively small, but the marginal for
+% $k_4$ varies by a significant amount.
 
@@
 
-A conventional way to compare two models under a Bayesian framework is
-to calculate the Bayes factors~\cite{chen2012monte}. The Bayes factor,
-$B_{1,2}$, between two models, $\mathcal{M}_1$ and $\mathcal{M}_2$,
-given data $D$, can be interpreted as a ratio of the normalisation constants of the posterior distributions given each model,
-\[
-	B_{1,2} = \frac{\mathbb{P}(D|\mathcal{M}_1)}{\mathbb{P}(D|\mathcal{M}_2)}, \quad \text{where} \quad \mathbb{P}(D|\mathcal{M}_k) = \int_\mathcal{X} \mathbb{P}(D|\theta_k,\mathcal{M}_k)\mathbb{P}(\theta_k|\mathcal{M}_k) \, \text{d}\theta_k.
-\]
-Under the ETAIS framework, it is straightforward to calculate these factors using the Monte Carlo estimator for the normalisation constants.
+    %     The Bayes factor,
+% $B_{1,2}$, between two models, $\mathcal{M}_1$ and $\mathcal{M}_2$,
+% given data $D$, can be interpreted as a ratio of the normalisation constants of the posterior distributions given each model,
+% \begin{equation}
+% 	B_{1,2} = \frac{\mathbb{P}(D|\mathcal{M}_1)}{\mathbb{P}(D|\mathcal{M}_2)}, \quad \text{where} \quad \mathbb{P}(D|\mathcal{M}_k) = \int_\mathcal{X} \mathbb{P}(D|\theta_k,\mathcal{M}_k)\mathbb{P}(\theta_k|\mathcal{M}_k) \, \text{d}\theta_k.
+% \end{equation}
+% Under the ETAIS framework, it is straightforward to calculate these factors using the Monte Carlo estimator for the normalisation constants.
 From~\cite{robert2013monte} these normalisation constants take the form
-\[
-	Z_k \approx \hat{Z}_k\frac{1}{NM}\sum_{i=1}^N\sum_{j=1}^M w_j^{(i)}(\mathcal{M}_k),
-\]
-where $w_j^{(i)}(\mathcal{M}_k)$ is the weight under model $\mathcal{M}_k$ corresponding to the $j$-th ensemble member on the $i$-th iteration.
+% \begin{equation}
+% 	Z_k \approx \hat{Z}_k\frac{1}{NM}\sum_{i=1}^N\sum_{j=1}^M w_j^{(i)}(\mathcal{M}_k),
+% \end{equation}
+% where $w_j^{(i)}(\mathcal{M}_k)$ is the weight under model $\mathcal{M}_k$ corresponding to the $j$-th ensemble member on the $i$-th iteration.
 Hence, $B_{1,2} \approx \hat{Z}_1/\hat{Z}_2$.
@@
 
-We now label the constrained model as $\mathcal{M}_1$, the model
-arising from the QEA as $\mathcal{M}_2$, and the full data model as
-$\mathcal{M}_0$. Model $\mathcal{M}_1$ is dependent on the parameter
-$\theta_1 = (k_1, \hat{k}_4^{\text{CMA}})^\top$, and model
-$\mathcal{M}_2$ is dependent on the parameters $\theta_2 = (k_1,
-\hat{k}_4^{\text{QEA}})^\top$. The full model $\mathcal{M}_0$ is
-dependent on all four parameters $\theta_0 = (k_1, k_2, k_3, k_4)^\top$.
-The marginal densities for the data, evaluated at the observed data, given the models are displayed in Table~\ref{tab:chem_Bayes_marginals}.
+% We now label the constrained model as $\mathcal{M}_1$, the model
+% arising from the QEA as $\mathcal{M}_2$, and the full data model as
+% $\mathcal{M}_0$. Model $\mathcal{M}_1$ is dependent on the parameter
+% $\theta_1 = (k_1, \hat{k}_4^{\text{CMA}})^\top$, and model
+% $\mathcal{M}_2$ is dependent on the parameters $\theta_2 = (k_1,
+% \hat{k}_4^{\text{QEA}})^\top$. The full model $\mathcal{M}_0$ is
+% dependent on all four parameters $\theta_0 = (k_1, k_2, k_3, k_4)^\top$.
+
+A conventional way to compare two models under a Bayesian framework is
+to calculate the Bayes factors~\cite{chen2012monte}. The marginal densities for the data, evaluated at the observed data, given the models are displayed in Table~\ref{tab:chem_Bayes_marginals}.
 
@@
                                   (\mathbf{R},\mathbf{T})^\top|\mathcal{M}_k)$
-  & $B_{0,k}$ \\ \cmidrule(lr){1-4}
+  & Bayes' factor  \\ \cmidrule(lr){1-4}
 	Full & 0 & 6.8e-3 & 1 \\
@@
 \end{tabular}
-\caption{Marginal distributions and Bayes factors for the data $(\mathbf{R},\mathbf{T})^\top$ for each model considered in Section~\ref{sec:chem_multiscale}.}
+\caption{Marginal distributions and Bayes factors for the data
+  $(\mathbf{R},\mathbf{T})^\top$ for each model considered in
+  Section~\ref{sec:chem_multiscale}. The Bayes' factors are computed
+  with respect to the ``full'' model.}
 \label{tab:chem_Bayes_marginals}
@@
 Computing the Bayes factors from Table~\ref{tab:chem_Bayes_marginals}
-we see that we should of course prefer the model $\mathcal{M}_0$ in
-which we observe all reactions perfectly. However this model is not
+we see that we should of course prefer the model \edit{ in
+which we observe all reactions perfectly and are able to explicitly
+compute the likelihood}. However this model is not
 significantly better than the CMA model ($B_{0,1} = 2.09 < 3.2$,
@@
 We create trajectories of this system, using the Gillespie SSA, with the following parameters:
-\begin{eqnarray*}
+\begin{eqnarray}
 k_1 = 0.04, \qquad
@@
 k_8 = 0.05,
-\end{eqnarray*}
+\end{eqnarray}
 over the time period $t \in [0,500]$.
@@
 As alluded to in the previous section, we aim to sample from the CMA
-approximation of the posterior distribution as defined in
-Equation~\eqref{eqn:chem_posterior}. This example is more involved
+approximation of the posterior distribution. This example is more involved
 than the previous one, since we have an eight dimensional parameter
@@
 \centering
-\begin{tabular}{|l|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|}
+%\begin{tabular}{|l|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|R{1.8em}|}
+\begin{tabular}{lrrrrrrrr}
 	\hline
@@
 
-We apply the eight RW and logRW proposal distributions, with and
+We apply the \edit{four logRW} proposal distributions, with and
 without transport map acceleration, which were discussed in
@@
 The transport map setup has not changed from the previous
-example. However, due to the scaling of the weight we have had to be
+example. However, due to the scaling of the weights we have had to be
 more careful about which samples we use to build our map. We require
@@
 
-In all eight algorithms we adapt the proposal variances during the
+In all \edit{four} algorithms we adapt the proposal variances during the
 burn-in phase of the algorithms to avoid the need for a huge ensemble size to approximate the posterior with a mixture of isotropic Gaussian kernels.
@@
 	\centering
-	\begin{tabular}{lrrrrrrrr}
+	\begin{tabular}{lrrrr}
 	\toprule
-		 & \multicolumn{4}{c}{MH} & \multicolumn{4}{c}{ETAIS} \\ \cmidrule(lr){2-5} \cmidrule(lr){6-9}
-		& RW & logRW & TRW & log-TRW & RW & logRW & TRW & log-TRW \\ \midrule
-		$\beta_{\%}^*$	 	& 1.0e-1 & 3.9e-1 & 1.0e-0 &6.0e-1 & - & - & - & - \\
-		$\beta_{\text{ESS}}^*$	& - 	        & -           & -            & -            & 1.0e-0 & 1.3e-0 & 8.0e-1 & 5.0e-1 \\
-		ESS/$M$		 		& - 	        & -           & -            & -            & 3.5e-4 & 4.7e-2 & 4.8e-2 & 1.6e-1  \\
+		 & \multicolumn{2}{c}{MH} & \multicolumn{2}{c}{ETAIS} \\ \cmidrule(lr){2-3} \cmidrule(lr){4-5}
+		& logRW & log-TRW & logRW & log-TRW \\ \midrule
+		$\beta_{\%}^*$	 	& 3.9e-1 &6.0e-1 & - & - \\
+		$\beta_{\text{ESS}}^*$	       & -            & -          & 1.3e-0 & 5.0e-1 \\
+		ESS/$M$		 	 & -            & -            & 4.7e-2 & 1.6e-1  \\
 	\bottomrule
@@
 the efficiency of the algorithm both by transforming the parameter
-space into something closer to Gaussian, and additionally by utilising
-a log preconditioner.
+space into something closer to Gaussian.
 
@@
 \centering
-\subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
-\subfigure[Sampling algorithms with a $\log$ preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}}
-\caption{Convergence of eight different MCMC algorithms for the CMA
+% \subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
+\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}
+\caption{Convergence of four different MCMC algorithms for the CMA
   approximation of the  posterior arising from the example in Section \ref{sec:grn}.}
@@
 
-We note that in this 
-example the ETAIS-TRW algorithm does not perform as well as the ETAIS-RW
-algorithm. This is due to the lack of log preconditioner. The
-posterior has support only on the positive quadrant, and without the
-preconditioner, the transport map has to work hard to enforce
-this. With a relatively low-dimensional approximation being used for
-the transport map, this makes it more likely that the map produced
-may hinder sampling rather than improving it. This highlights the
-importance of using preconditioners with transport maps in order to
-ensure absolute continuity of the posterior distribution with respect
-to the pullback of the proposal distribution on reference space.
-
-In contrast, when we look at the convergence of these two algorithms
-with the log preconditioner, the transport map version exhibits more
+% We note that in this 
+% example the ETAIS-TRW algorithm does not perform as well as the ETAIS-RW
+% algorithm. This is due to the lack of log preconditioner. The
+% posterior has support only on the positive quadrant, and without the
+% preconditioner, the transport map has to work hard to enforce
+% this. With a relatively low-dimensional approximation being used for
+% the transport map, this makes it more likely that the map produced
+% may hinder sampling rather than improving it. This highlights the
+% importance of using preconditioners with transport maps in order to
+% ensure absolute continuity of the posterior distribution with respect
+% to the pullback of the proposal distribution on reference space.
+
+\edit{The transport map version exhibits more
 consistent and stable convergence, with some of the 32 repeats for the
-ETAIS-logRW algorithm producing rare large weights which hinder convergence.
+ETAIS-logRW algorithm producing rare large weights which hinder convergence.}
 The large increase in the effective sample size
@@
 
-A similar pattern is seen in the MH algorithms, where the MH-TRW
-performs the worst, and the MH-logTRW algorithm performs the best,
-although only marginally.
+\edit{A similar pattern is seen in the MH algorithms, where the MH-logTRW algorithm performs the best,
+although only marginally. }
 

### 2020-01-14 Resubmission/response.tex
First draft of resubmission complete

@@
 We have also rebalanced the introduction to make the main focus the MCMC methodology.
-Any required references for the material regarding multiscale methods for stochastic reaction networks are given Section 6.
+Any required references for the material regarding multiscale methods for stochastic reaction networks are given in Section 6.
 
@@
 
-****DO SUPPLEMENTARY MATERIAL*****
+% I have reduced the introduction and it also flows better now.
+%supplementary material done
 
-%I have reduced the introduction and it also flows better now.
-
-\comment{Regularization.
-In regards to the regularization terms in equation (3.2), the authors claim that "experimentation has shown that the choice $\beta=1$ is sufficient for most problems." I would like to see more discussion on this choice.
+\comment{Regularisation.
+In regards to the regularisation terms in equation (3.2), the authors claim that "experimentation has shown that the choice $\beta=1$ is sufficient for most problems." I would like to see more discussion on this choice.
 With a sufficiently large number of samples, decreasing $\beta$ should result in a more accurate transport map, so why does a smaller value of $\beta$ not yield a more efficient algorithm?
@@
 However, as demonstrated in the examples in the paper, for example Figure 7.4, the pushforward distribution for complex examples is often quite far from the Gaussian reference distribution, which would lead to a poorly performing algorithm.
-The whole basis of combining the ETAIS approaach with the transport maps is that for complex examples, the push forward density itself has complex structure which benefits from adaptive importance sampling.
-We have added a substantial discussion on this topic to just before Section 7.1.1, and also to the conclusions and abstract, since this is a key justification for the approach *****DO THIS*******.
+The whole basis of combining the ETAIS approach with the transport maps is that for complex examples, the push forward density itself has complex structure which benefits from adaptive importance sampling.
+We have added a discussions on this topic to just before Section 5.2, just before 7.1.1, and also to the introduction, since this is a key justification for the approach.
 
-  The referee suggests using the Rosenbrock example to demonstrate this, but as can be seen in Figure ****, the transport map does a very good job for this example, and so experiments of the type suggested would not be sufficient to demonstrate this point.
-Conducting these experiments on the stochastic reaction networks example would be a substantial body of work, for which unfortunately we no longer have the resources to undertake.
-*****check and rewrite this******}
+The referee suggests a comparison of proposal types.
+This was done in detail for our first paper ``Ensemble Transport Adaptive Importance Sampling'', and was published in detail in Paul Russell's PhD thesis.
+We compared RW proposals with MALA and even HMC proposals.
+The performance of all of the proposals were similar, but the MALA and HMC proposals come at additional costs.
+We have added a brief discussion about proposal types at the end of Section 2.}
 
-%First part done, need to think more carefully about the second part. We REALLY don't want to do more numerics...
+%DONE
 
@@
 When reading the manuscript, I felt like I walked off the end of the manuscript.
-I think a sentence or two that ties things together would provide a better closing statement.}{We have rewritten this Section and ******.}
+I think a sentence or two that ties things together would provide a better closing statement.}{We have added a paragraph to further explore our argument for combining AIS schemes and transport maps.}
 
-%REDO THE CONCLUSIONS AFTER EVERYTHING ELSE HAS BEEN REWRITTEN
+%DONE
 
@@
 
-\comment{When discussing monotonicity of the map, it might be worth pointing out the monotone parameterizations introduced in "An Introduction to Sampling via Measure Transport" by Marzouk et al.}{Thanks for this suggestion, we have added a couple of sentences on this to ******.}
-
-%STILL TO DO
-
-\comment{In Section 4, the notation for proposed points on the reference space (marked with an apostrophe) and proposed points on the target space (marked with a $\hat{ }$) are different.
-It might be easier to follow if these were consistent, but that is up to the authors.}{We have made a thorough review of the notation in this Section *******}
-%STILL TO DO
-
-\comment{Equation (6.2) $Y_j$ is used in the text but $X$ is used in the equation. Is that a typo?}{Correct, thanks.}
+\comment{When discussing monotonicity of the map, it might be worth pointing out the monotone parameterisations introduced in "An Introduction to Sampling via Measure Transport" by Marzouk et al.}{Thanks for this suggestion, we have added a comment and a reference.}
 
@@
 
-\comment{Sentence before equation (6.3). "variable" should be plural "variables"}{Corrected, thanks.}
+\comment{In Section 4, the notation for proposed points on the reference space (marked with an apostrophe) and proposed points on the target space (marked with a $\hat{ }$) are different.
+It might be easier to follow if these were consistent, but that is up to the authors.}{We have done as the referee suggests, both are now marked with a $\hat{ }$.}
+%DONE
+
+\comment{Equation (6.2) $Y_j$ is used in the text but $X$ is used in the equation. Is that a typo?}{Corrected, thank you.}
+
+%DONE
+
+\comment{Sentence before equation (6.3). "variable" should be plural "variables"}{Corrected, thank you.}
 
@@
 \comment{Also, it should be made more clear what exactly is new within the stochastic reaction networks portion of the paper}{Thanks to the referee for this comment.
-The main novel result in this portion of the paper is that summarised by the plots in Fig *****.
+The main novel result in this portion of the paper is that summarised by the plots in Fig 7.4.
 The multiscale methods themselves have been presented in other work which was already heavily cited in this paper.
@@
 But the ability to sample from the complex posteriors arising from these problems, and therefore the ability to compare the posteriors arising from different multiscale approximations is new.
-Following this comment and comments from referee 1, we have heavily reduced the volume of material discussing the multiscale approximations in Section 6, which now appears in supplementary material.}
+Following this comment and comments from referee 1, we have heavily reduced the volume of material discussing the multiscale approximations in Section 6, which now appears in supplementary material.
+We have also heavily reduced the material in 7.1.1, with a simple presentation of Figure 7.4, the Bayes factors, and a short discussion.}
 
-%DONE?
+%DONE
 
@@
 And relative to what?
-Similarly, can you define $\delta_{ESS}$ as a function of the ESS?}{Thanks for this comment, we have added more thorough descriptions of these quantities to *******.}
+Similarly, can you define $\delta_{ESS}$ as a function of the ESS?}{The relative $L^2$ error in this context refers to the error made in approximating the PDF of the target density with a histogram of the samples.
+We have added in definitions for this.
+The two deltas, as described in the caption in the table, are the optimal scaling parameters, one optimised for ESS and the other for our relative $L^2$ measure.
+We have expanded on this in the caption, and hope that this is clearer now.}
+
+%DONE
 
@@
 which case I think plotting all the runs in the same figure would be more illustrative).
-These are Monte Carlo schemes, so I expect the error in any given running estimate to be noisy.}{These plots are produced by averaging over ***** repeats.
+These are Monte Carlo schemes, so I expect the error in any given running estimate to be noisy.}{These plots are produced by averaging over 32 repeats.
 The plots show comparison of average convergence for several different algorithms.
 We tried creating the plots that the referee has suggested but the sheer number of lines led to plots which were in our opinion far less informative.
-We have given more thorough details of how the plots were created in ******.}
+We have added details of the averaging to the captions of all relevant plots.}
+
+%DONE
 
@@
 Similarly, what is meant by \# of samples in the ETIAS error plots?
-Using “samples” without context is confusing, as there are many intermediate populations of samples at hand.}{For the MCMC methods we mean the number of iterations, and for ETAIS, we mean the number of iterations multiplied by the ensemble size.
-We have added a description along these lines to ******.}
+Using “samples” without context is confusing, as there are many intermediate populations of samples at hand.}{We literally meant the number of samples produced.
+For the MCMC methods we mean the number of iterations, since there is one sample per iteration.
+For ETAIS, we mean the number of iterations multiplied by the ensemble size.
+We have added a description along these lines below the first such example, Figure 5.3.}
+
+%DONE
 
@@
 We understand the reasons that the referee has asked for this metric to be considered, not least because it's use is ubiquitous but we believe that the one we have used is actually a more rigorous measure.}
+
+%DONE
 
@@
 If by the true map one means the KR rearrangement, then it is invertible.
-Maybe I am confused about what is meant by “true map?”}{Thanks to the referee for highlighting our confusion about this, we have ammended the discussions in ****** and ****** to reflect this.}
+Maybe I am confused about what is meant by “true map?”}{Thanks to the referee for their comments here.
+We have amended the discussion following Definition 3.2, and we have removed Section 8.2.}
+
+%DONE}
 
@@
 More broadly, this reflects the close link between independence Metropolis algorithms and importance sampling.
-Some comments (please agree or disagree) on this front would be illuminating.}{The first referee made a similar comment on this on Page ***** of this document, which we hope addresses this comment also.
-We also refere the referee to the discussion we have added on this in the manuscript at ********.
+Some comments (please agree or disagree) on this front would be illuminating.}{The first referee made a similar comment on this, which we hope addresses this comment also.
+We also refer the referee to the discussion we have added on this in the manuscript just before Section 5.2, just before 7.1.1, and also to the introduction.
 The problems we are attempting to sample from in this work are highly challenging, with the density highly concentrated on a low dimensional curved manifold, and in the majority of cases the low dimensional approximations of the transport map are not sufficient to make independent sampling efficient.}
+
+%DONE
 
@@
 
+%DONE
+
 \comment{Here are several suggestions for content that could be removed entirely or shortened, along with a few areas that could be expanded:}{Thanks for these suggestions, we will address each in turn below.}
+
+%DONE
 
@@
 Log- preconditioning is well recognized as a good thing to do.}{Thanks to the referee for these very helpful suggestions.
-We agree that the log transformation is pretty standard, and so we have vastly reduced the discussion and numerical experiments presented on this topic.}
+We agree that the log transformation is pretty standard, and so we have vastly reduced the discussion and numerical experiments presented on this topic to just a single comparison.}
 
-%Have I?
+%DONE
 
 \comment{Suggestion for removal: A lot of Section 3 could be abridged by appropriate referencing of [46].}{
-We have removed as much as possible, but we wanted the paper to remain as standalone as possible to make it more accessible and easier to read.
-We also need to highlight the differences in which we construct the map, the material in section 6.****.
-However we have removed some of the details from sections *** and ****.}
+  We have removed as much as possible, but we wanted the paper to remain as standalone as possible to make it more accessible and easier to read, and also to highlight the differences in finding in the map with importance samples.
+We have removed quite a bit, in particular Section 3.2.}
+
+%DONE
 
@@
 
-\comment{Suggestion for removal: The description of the ETIAS algorithm in Section 2 is very nice, but here an illustrative figure might help.}{****Illustrative figure? Like what?*****}
+%DONE
 
-\comment{End of Section 2, “learn local covariances across the whole of the domain”: I believe I under- stand what is meant, but this awkwardly phrased}{Thanks, we have rerwritten this *****.}
+\comment{Suggestion for removal: The description of the ETIAS algorithm in Section 2 is very nice, but here an illustrative figure might help.}{There is a nice illustration in a paper by Bugallo et al which demonstrates how AIS works, so rather than making a poor reproduction, we have cited this paper.}
+
+%DONE
+
+\comment{End of Section 2, “learn local covariances across the whole of the domain”: I believe I under- stand what is meant, but this awkwardly phrased}{Thanks, we have rewritten this.}
+
+%DONE
 

### 2020-01-14 Resubmission/Supplementary.tex
First draft of resubmission complete

@@
+\documentclass[final]{siamltex}
+%test change
+\usepackage{cite}
+\usepackage{graphicx,bbm,pstricks,soul}
+\usepackage{pifont}
+\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
+\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
+\usepackage{tikz,hhline}
+\usepackage{tabularx}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\arraybackslash\hspace{0pt}}m{#1}}
+
+\setlength{\parindent}{0in}
+\usepackage{amsmath,amsfonts,amsbsy,amssymb}
+\newcommand{\RARR}[3]{#1
+  \;\displaystyle\mathop{\displaystyle\longrightarrow}^{#3}\; #2}
+\newcommand{\RARRlong}[3]{#1
+  \;\displaystyle\mathop{-\!\!\!-\!\!\!-\!\!\!-\!\!\!-\!\!\!\!\displaystyle
+  \longrightarrow}^{#3}\; #2}
+\newcommand{\LARR}[3]{#1
+  \;\displaystyle\mathop{\displaystyle\longleftarrow}^{#3}\; #2}
+\newcommand{\LRARR}[4]{{\mbox{ \raise 0.4 mm \hbox{$#1$}}} \;
+  \mathop{\stackrel{\displaystyle\longrightarrow}\longleftarrow}^{#3}_{#4}
+  \; {\mbox{\raise 0.4 mm\hbox{$#2$}}}}
+\newcommand{\bX}{{\bf X}}
+\newcommand{\vecx}{{\mathbf x}}
+\newcommand{\vecy}{{\mathbf y}}
+\newcommand{\vecz}{{\mathbf z}}
+\newcommand{\vecq}{{\mathbf q}}
+\newcommand{\bs}{{\mathbf s}}
+\newcommand{\vecr}{{\mathbf r}}
+\newcommand{\vecX}{{\mathbf X}}
+\newcommand{\vecv}{{\mathbf v}}
+\newcommand{\tick}{\ding{52}}
+\newcommand{\cross}{\ding{54}}
+\newcommand{\vecn}{{\mathbf n}}
+\newcommand{\vecp}{{\mathbf p}}
+\newcommand{\cT}{{\mathcal T}}
+\newcommand{\dt}{{\mbox{d}t}}
+\newcommand{\dx}{{\mbox{d} \vecx}}
+\newcommand{\boldnu}{{\boldsymbol \nu}}
+\newcommand{\er}{{\mathbb R}}
+\renewcommand{\div}{{\rm div}}
+\newcommand{\bnu}{{\bf \nu}}
+\newcommand{\divergence}{\mathop{\mbox{div}}}
+\renewcommand{\P}{\mathbb{P}}
+\newcommand{\FP}{P_{\rm{FP}}}
+\newcommand{\ME}{P_{\rm{ME}}}
+\newcommand{\MEs}{P_{\rm{ME}_{S}}}
+\newcommand{\D}{\mathcal{D}}
+\newcommand{\G}{\mathcal{G}}
+\newcommand{\N}{\mathcal{N}}
+\newcommand{\X}{{\mathbf X}}
+\newcommand{\Y}{{\mathbf Y}}
+\newcommand{\W}{{\mathbf W}}
+\newcommand{\data}{D}
+\newcommand{\neff}{n_{\text{eff}}}
+\newcommand{\E}{{\mathbb E}}
+\renewcommand{\b}[1]{{\bf #1}}
+\newcommand{\edit}[1]{{\color{red} #1}} 
+\DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator*{\argmax}{arg\,max}
+
+\newcommand{\picturesAB}[6]{
+\centerline{
+\hskip #4
+\raise #3 \hbox{\raise 0.9mm \hbox{(a)}}
+\hskip #5
+\epsfig{file=#1,height=#3}
+\hskip #6
+\raise #3 \hbox{\raise 0.9mm \hbox{(b)}}
+\hskip #5
+\epsfig{file=#2,height=#3}
+}}
+\newcommand{\picturesCD}[6]{
+\centerline{
+\hskip #4
+\raise #3 \hbox{\raise 0.9mm \hbox{(c)}}
+\hskip #5
+\epsfig{file=#1,height=#3}
+\hskip #6
+\raise #3 \hbox{\raise 0.9mm \hbox{(d)}}
+\hskip #5
+\epsfig{file=#2,height=#3}
+}}
+
+\makeatletter  
+\newcommand{\xleftrightarrows}[2][]{\mathrel{%  
+ \raise.40ex\hbox{$  
+       \ext@arrow 3095\leftarrowfill@{\phantom{#1}}{#2}$}%  
+ \setbox0=\hbox{$\ext@arrow 0359\rightarrowfill@{#1}{\phantom{#2}}$}%  
+ \kern-\wd0 \lower.4ex\box0}}
+
+\newcommand{\xrightleftarrows}[2][]{\mathrel{%  
+ \raise.40ex\hbox{$\ext@arrow 3095\rightarrowfill@{\phantom{#1}}{#2}$}%  
+ \setbox0=\hbox{$\ext@arrow 0359\leftarrowfill@{#1}{\phantom{#2}}$}%  
+ \kern-\wd0 \lower.4ex\box0}}
+
+\def\leftrightarrowfill@{%
+ \arrowfill@\leftarrow\relbar\rightarrow%
+ }
+\newcommand*{\centerfloat}{%
+  \parindent \z@
+  \leftskip \z@ \@plus 1fil \@minus \textwidth
+  \rightskip\leftskip
+  \parfillskip \z@skip}
+\makeatother
+\makeatother
+
+\newcommand\irregularcircle[2]{% radius, irregularity
+  \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
+  +(0:\len pt)
+  \foreach \a in {10,20,...,350}{
+    \pgfextra {\pgfmathsetmacro\len{(#1)+rand*(#2)}}
+    -- +(\a:\len pt)
+  } -- cycle
+}
+
+\newcommand{\into}{\operatornamewithlimits{\longrightarrow}}
+\newtheorem{dfn}{Definition}[section]
+
+\newcommand{\ltri}{%
+\,\resizebox{!}{0.25\baselineskip}{%
+\begin{tikzpicture}%
+\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
+\end{tikzpicture}%  
+}\xspace%
+}%
+
+\newcommand{\smallltri}{%
+\,\resizebox{!}{0.15\baselineskip}{%
+\begin{tikzpicture}%
+\draw[line width=1mm](0,0) -- (0,2) -- (2,0)  -- (0,0);
+\end{tikzpicture}%  
+}\xspace%
+}%
+
+\author{Simon L. Cotter\thanks{School of
+    Mathematics, University of Manchester, Manchester, UK. e:
+    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
+    grant award EP/L023393/1. SLC would like to thank the Isaac Newton
+    Institute for Mathematical Sciences for support and hospitality
+    during the programme ``Uncertainty quantification for complex systems: theory and methodologies'' when work on this paper was undertaken. This work was supported by:
+EPSRC grant number EP/K032208/1.} \and Ioannis
+G. Kevrekidis\thanks{Chemical and Biomolecular Engineering \& Applied
+  Mathematics and Statistics, Johns Hopkins University, Baltimore, MD
+  21218. The work of IGK was partially supported by the US National
+  Science Foundation, by the US AFOSR, and by DARPA.} \and Paul
+  Russell\thanks{School of
+    Mathematics, University of Manchester, Manchester, UK.}}
+
+\title{Multiscale Methods for Stochastic Chemical Reaction
+  Networks\\Supplementary material}
+\begin{document}
+\maketitle
+In this supplementary material, we discuss some recent advances in multiscale methods
+for stochastic reaction networks. Inverse problems arising in this
+area often lead to complex posterior
+distributions, which traditional MCMC methods can struggle to sample
+from.
+
+We consider chemical reaction networks of $N_s$ chemical species $\{S_j\}_{j=1}^{N_s}$,
+with population numbers given by $X(t) = [X_1(t), X_2(t), \ldots, X_{N_s}(t)]^\top \in
+\mathbb{N}_0^{N_s}$ reacting in a small reactor, through $N_r$ different reaction
+channels. When population numbers of one or more of the chemical
+species in the reactor is small, as is the case with chemical
+reactions occurring within living cells, the sporadic and discrete way
+in which reactions occur can not be well modelled by deterministic
+continuous models, such as ODEs. In such a situation, we may wish to
+model the dynamics through a discrete stochastic model, such as a
+continuous time Markov chain.
+
+For each reaction $R_j$, for $j = 1,2,\ldots N_r$, there is a
+propensity or hazard function $\alpha_j(X(t))$ which indicates how
+likely that reaction is to fire, defined by
+\[\alpha_j(X(t)) = \lim_{dt \to 0} \mathbb{P}(\text{Reaction $R_j$ in
+    the time interval  } \quad s \in [t, t+ dt] ).\] If a system
+satisfies what is called \emph{mass action kinetics}, then the form of
+the function $\alpha_j$ is determined, up to a rate constant, by the
+reactants involved in that reaction:
+\begin{equation}\label{eq:MAK}
+\alpha_j(X) = k_j \prod_{m=1}^{N_s} \prod_{n=0}^{\nu_{j,m} -1} (X_m - n),
+\end{equation}
+where $\nu_{j,m}$ is the $m$th component of the stoichiometric vector
+$\nu_j$, $X_m$ is the $m$th component of the state vector $X$, and the
+$k_j$ are rate constants.
+
+Following each reaction $R_j$ there is an instantaneous change in the
+current state, as the reactants of the reaction are consumed, and the
+products produced. This is modelled by a change in the state vector
+$X(t) = X(t) + \nu_j$ where $\nu_j$ is that stoichiometric vector for
+reaction $R_j$.
+
+The model can be represented as the following expression involving
+$N_r$ different unit rate Poisson
+processes\cite{anderson2011continuous} $Y_j$, given by:
+\begin{equation}\label{eq:RTC}
+X(t) = X(0) + \sum_{j=1}^{N_r} \nu_j \int_0^t \alpha_j(X(s)) ds.
+\end{equation}
+
+The master equation for this model is only solvable in certain
+situations, for example monomolecular networks\cite{jahnke2007solving}, or for
+steady-state solutions for weakly reversible deficiency
+zero networks\cite{anderson2010product,anderson2016product}. Trajectories for this system can be
+sampled exactly, for instance using the Gillespie SSA\cite{gillespie1977exact}, or
+its variants\cite{gillespie2007stochastic,cao2004efficient,anderson2007modified}. However, if the system is stiff, i.e. there
+are some reactions which are firing many times on a timescale for
+which others are unlikely to fire at all, then trajectories can become
+prohibitively expensive to simulate, since these methods simulate
+every single reaction event with the same cost. In such a system, one
+might employ multiscale methods in order to approximate trajectories
+at a lower cost than the exact algorithms.
+
+One common approach is to induce the quasi-equilibrium approximation
+QEA. This approximation makes the assumption that fast reactions
+enter quasi-equilibrium on a timescale which is negligible with
+respect to the timescale on which the slow dynamics in the system are
+evolving. This amounts to taking the asymptotic limit that the
+propensities of the ``slow'' reaction channels are equal to zero. This
+allows us to sample approximate trajectories of the slow reactions
+without the need to compute many fast reaction events.
+
+This approach can work well when there is a large timescale gap
+between the fast and slow reactions in a system, but where the
+timescale gap is less pronounced, it can lead to significant
+errors\cite{cotter2016error}. Another approach is the
+constrained multiscale algorithm (CMA)\cite{cotter2011constrained,cotter2016constrained}, based in part on the
+equation-free approach to multiscale computations\cite{kevrekidis2003equation,erban2006gene}. This approach also assumes quasi-equilibrium in the
+system, but takes into account the effect that the slow reactions can
+have on the invariant distribution of the fast variables in the
+system. For a more detailed description of this method, please refer
+to the literature\cite{cotter2016constrained}.
+
+Multiscale methods are not only of use when forward evaluations of the
+dynamics are costly, but can also be of use where we attempt to solve
+an inverse problem where the fast variables are unobservable. One
+approach in this situation would be to integrate over all possible
+trajectories of the fast variables, but this would almost always be
+prohibitively expensive. Another approach would be to use an
+appropriate multiscale approximation, so that the effective dynamics
+of the slow variables can be approximate without the need to consider
+the rapid fluctuations of the fast variables in the system.
+
 \subsection{Likelihoods arising from stochastic reaction networks}
@@
 next reaction, which is given by an exponential random variable
-\begin{equation}\tau_j \sim \exp \left ( \frac{1}{\alpha_0(X(t))} \right ),\end{equation}
+\[\tau_j \sim \exp \left ( \frac{1}{\alpha_0(X(t))} \right ),\]
 where $\alpha_0(X(t)) =\sum_{i}^{N_r} \alpha_i(X(t))$ is the total
@@
 $i \in \{1,2,\ldots,N_r\}$ with probability
-\begin{equation}\mathbb{P}(r_j = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\end{equation}
+\[\mathbb{P}(r_j = i) = \frac{\alpha_i(X(t))}{\alpha_0(X(t))}.\]
 As such, in order to compute the likelihood of a particular trajectory
@@
 We define two new random variables which depend on a state $\mathbf{Y} \in \mathcal{S}$, first the total time spent in state $\mathbf{Y}$,
-\begin{equation}
+\[
 	T(\mathbf{Y}) = \sum\limits_{j=1}^M \tau_j\; \text{I}(\mathbf{X}(t_{j-1}) = \mathbf{Y}).
-\end{equation}
+\]
 This random variable, $T(\mathbf{Y})$, is a sum of exponentially
-distributed random variables, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
+distributed random variable, each with the rate $\alpha_0(\mathbf{Y};\mathbf{k})$, and hence follows the Gamma distribution,
 \begin{equation}\label{eqn:chem_time_dist}
@@
 Similarly, we can define the reactions which occurred when the system was in state $\mathbf{Y}$ as $\mathbf{r}(\mathbf{Y}) = [r_1(\mathbf{Y}), \ldots, r_{N_r}(\mathbf{Y})]^\top$ where
-\begin{equation}
+\[
 	r_i(\mathbf{Y}) = \sum\limits_{j=1}^M \text{I}(r_j = i \textbf{ and }\mathbf{X}(t_{j-1}) = \mathbf{Y}).
-\end{equation}
+\]
 Here all the random variables $r_j$ follow the same multinomial distribution, and so
@@
 With these definitions we define two new structures
-\begin{equation}
+\[
 	\mathbf{T} = [T(\mathbf{Y}_1), \dots, T(\mathbf{Y}_K)]^\top, \quad \text{and} \quad \mathbf{R} = [\mathbf{r}(\mathbf{Y}_1), \dots, \mathbf{r}(\mathbf{Y}_K)]^\top,
-\end{equation}
+\]
 where $K = |\mathcal{S}|$, the number of states in $\mathcal{S}$, and each state $\mathbf{Y}_i \in \mathcal{S}$ has been enumerated. We use these structures to define shorter notation,
-\begin{equation}
+\[
 	\mathbf{T}_i = T(\mathbf{Y}_i), \quad \mathbf{R}_{ij} = r_j(\mathbf{Y}_i), \quad \text{and} \quad \mathbf{K}_i = K(\mathbf{Y}_i).
-\end{equation}
+\]
 
@@
 Given the distributions in Equations~\eqref{eqn:chem_time_dist} and~\eqref{eqn:chem_react_dist} for our data, the likelihood of observing the data $\mathbf{R}$ and $\mathbf{T}$ is
-\begin{equation}
+\[
 	\ell(\mathbf{R}, \mathbf{T}|\mathbf{k}) \propto \prod\limits_{i=1}^K \text{Multi}(\mathbf{R}_{i\cdot}; \mathbf{K}_i, \mathbf{p}(\mathbf{Y}_i))\text{Gamma}(\mathbf{T}_i; \mathbf{K}_i, \alpha_0(\mathbf{Y}_i)),
-\end{equation}
+\]
 where again $\mathbf{Y}_i \in \mathcal{S}$ and the propensities
@@
 conditioned on the slow variables in the system can be approximated
-using a multiscale method, such as the QEA or the CMA, as described
-briefly in Section \ref{sec:multi}. Once the approximations have been
+using a multiscale method, such as the QEA or the CMA. Once the approximations have been
 made, we arrive at approximate \emph{effective} propensities $\bar{\alpha}_{s_i}$, which
@@
 occurrence of slow reactions.
+
+\bibliographystyle{siam}
+\bibliography{bibliography}
+
+\end{document}

### 2020-01-14 Resubmission/TETAIS.tex
First draft of resubmission complete

@@
 Other approaches to challenging posterior structure have been
-proposed, including the work of Marzouk and Parno\cite{parno2014transport}, who
+proposed, including the work of Marzouk and Parno\cite{parno2018transport}, who
 demonstrated how invertible transport maps can be constructed which
@@
 
-****Could move this material to stochastic network section****
+% ****Could move this material to stochastic network section****
 
@@
 
-********Did I mean to get rid of this?**********
+% ********Did I mean to reintroduce this last paragraph?********** NOOOOO!!!
 
@@
 arising from Bayesian inverse problems on multiscale stochastic
-reaction networks.
+reaction networks. \edit{In a perfect scenario, the numerical
+  approximation of the optimal transport map would capture all of the
+  posterior structure, and independent samples could be drawn by
+  sampling from the reference measure. However in many cases the push
+  forward of the posterior measure can still exhibit complex structure
+  which means that efficient adaptive importance sampling can be
+  preferential to this independent sampling approach.}
 
@@
 history of the algorithm up to the current point, or to use the
-current state of a whole ensemble of $M$ particles in the system.
+current state of a whole ensemble of $M$ particles in the system.  \edit{A nice illustrative figure of how adaptive
+  importance sampling algorithms work is given in \cite{bugallo2017adaptive}.}
 
@@
 
+\edit{The choice of proposal kernel $q$ can affect the convergence of
+  the algorithm. In particular, it is important that the tails of the
+  proposal distribution are as fat as the target distribution. In this
+  work we will primarily use Gaussian kernels, but we have previously
+  made comparisons with more complex kernels, such as MALA or
+  HMC\cite{cotter2015parallel,russ2017parallel}. There is minimal
+  benefit from using these more complex kernels, which come at
+  additional cost.}
+
 One problem with the ETAIS and other similar methods can become apparent if
@@
 
-In this paper, we investigate the use of transport maps to learn local
-covariances across the whole of the domain, in order to stabilise
-adaptive importance sampling methods, and make these methods more applicable to a wider
-range of more challenging inference problems.
+\edit{In this paper, we investigate the use of transport maps which allows
+us to learn the local correlation structures of the posterior in
+currently explored regions. Using these transport maps, we can
+stabilise adaptive importance sampling methods, and make these methods
+more applicable to a wider range of more challenging inference problems.}
 
@@
 investigate the posterior distribution. This
-methodology was adapted in~\cite{parno2014transport} to form a new proposal method for MH
+methodology was adapted in~\cite{parno2018transport} to form a new proposal method for MH
 algorithms. In this case, rather than transforming a sample from the prior into a sample from the target
@@
 In this Section we outline the methodology in
-\cite{parno2014transport} for approximately coupling the target
+\cite{parno2018transport} for approximately coupling the target
 measure with density
@@
 Not all maps satisfy these conditions, so we define a smaller space of
-maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$ which contains all
-invertible maps. This space does not necessarily contain an exact
-coupling between target and reference space, and so we are motivated to formulate an optimisation problem to find the map $\tilde{T}
+maps, $\mathcal{T}^\uparrow \subset \mathcal{T}$, \edit{ where $\mathcal{T}$ which contains all
+invertible maps. It is an intractable problem to find the
+  Knothe–Rosenblatt rearrangement, and so we numerically approximate
+  it by formulating an optimisation problem to find the map $\tilde{T}
 \in \mathcal{T}^\uparrow$ which most closely maps the target density
-to the reference density.
+to the reference density.}
 
@@
 % measure) is minimised, for $T \in \mathcal{T}^\uparrow$. As in
-% \cite{parno2014transport} we aim to minimise the Kullback-Liebler (KL)
+% \cite{parno2018transport} we aim to minimise the Kullback-Liebler (KL)
 % divergence between the density of $\mu_{\theta}$ and the pullback of
@@
 
-As in previous work in \cite{parno2014transport}, we can ensure
-invertibility if we restrict the map to be lower triangular, i.e. $\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$. This lower triangular map has the form,
+As in previous work in \cite{parno2018transport}, we can ensure
+invertibility if we restrict the map to be \edit{lower triangular\cite{bogachev2005triangular}}, i.e.
+$\tilde{T} \in \mathcal{T}^{\ltri}\subset\mathcal{T}^\uparrow$.
+\edit{There are a number of different
+  approaches to guaranteeing monotonicity\cite{y.17:_handb_uncer_quant}, but here we follow
+  the approach taken in \cite{parno2018transport}.} This lower triangular map has the form,
 \begin{equation}
@@
 density and the pullback of the reference density is minimised. As in
-\cite{parno2014transport}, we choose the cost function
+\cite{parno2018transport}, we choose the cost function
 to be the Kullback-Leibler (KL) divergence between the posterior density and the pullback density,
@@
 closer to the identity, and so prevents overfitting when the quality
-or size of the current sample from the posterior is not sufficient.
-
-\subsection{The structure of the map}*****CHECK HOW MUCH OF THIS IS
-IDENTICAL TO MARZOUK PAPER******
-
-Before we continue with the derivation of the optimisation problem, we consider the structure
-of the map in more detail.
-The lower triangular structure of the map not only guarantees monotonicity, it also allows for efficient calculation of the pullback density, as well as the inverse of the map, $\tilde{T}^{-1}$.
-The Jacobian of $\tilde{T}$ is a lower triangular matrix,
-\begin{equation}
-	J_T(\theta) = \begin{bmatrix}
-		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & \partial_{\theta_d} \tilde{T}_1(\theta)\\
-		\vdots & \ddots & \vdots \\
-		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
-	\end{bmatrix} = \begin{bmatrix}
-		\partial_{\theta_1} \tilde{T}_1(\theta) & \dots & 0\\
-		\vdots & \ddots & \vdots \\
-		\partial_{\theta_1} \tilde{T}_d(\theta) & \dots & \partial_{\theta_d} \tilde{T}_d(\theta)
-	\end{bmatrix}
-\end{equation}
-since $\partial_{\theta_n} \tilde{T}_k(\theta) = 0$ for all $n > k$.
-This lower triangular structure means that the determinant of the Jacobian is a product of the diagonal elements which, when we take logs, becomes
+or size of the current sample from the posterior is not
+sufficient. \edit{Moreover, we can exploit the lower triangular
+  structure, for example in the evaluation of 
 \begin{equation}\label{eqn:separable_jacobian}
@@
 \end{equation}
-where we note that this term is separable in terms of the dimension $i$.
-
-Inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map.
-The map component $\tilde{T}_1(\theta)$ is a univariate polynomial in $\theta_1$, so we can find the inverse of this function by solving the equation $T_1(\theta_1) = r_1$.
-This inversion tells us the value of $\theta_1$, which means the next component is again a univariate polynomial, $T_2(\theta_2; \theta_1)=r_2$.
-We can then perform $d$ root finding problems instead of a full $d$ dimensional non-linear solve.
-
-We require that the first derivatives of the map are continuous, which is easy to enforce by the choice of basis functions.
-Here we assume that the map will be built from a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not necessarily orthogonal with respect to the target distribution.
-Each component of the map is defined as a multivariate polynomial expansion,
+where we note that this term is separable in terms of the dimension
+$i$. Similarly, inverting $\tilde{T}$ at a point $r$ is simplified by the lower triangular structure of the map.
+
+As in \cite{parno2018transport}, we parameterise each component of the
+map $\tilde{T}_i$ with a multivariate polynomial expansion,
 \begin{equation}\label{eq:map_defn}
@@
 \end{equation}
-The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
+built from
+a family of orthogonal polynomials, $\mathcal{P}(\theta)$, not
+necessarily orthogonal with respect to the target distribution. The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
 $\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in
-\mathcal{J}_i \subset \mathbb{N}_0^i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
-\begin{equation}
-	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
-\end{equation}
-and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$.
+\mathcal{J}_i \subset \mathbb{N}_0^i$. 
+Here we stick with polynomials of total order $p$ since we work with
+low dimensional problems with the ETAIS algorithm. More details about
+the structure of the maps can be found in  \cite{parno2018transport}.}
+
+% Each component of the map is defined as a multivariate polynomial expansion,
+% \begin{equation}\label{eq:map_defn}
+% 	\tilde{T}_i(\theta; \gamma_i) = \sum\limits_{\mathbf{j}\in\mathcal{J}_i} \!
+% \gamma_{i,\mathbf{j}}\psi_\mathbf{j}(\theta).
+% \end{equation}
+% The parameter $\gamma_i \in \mathbb{R}^{M_i}$ is a vector of coefficients. Each component of $\gamma_i$ corresponds to a basis function
+% $\psi_\mathbf{j}$, indexed by the multi-index $\mathbf{j} \in
+% \mathcal{J}_i \subset \mathbb{N}_0^i$. A multi-index defines a product of univariate polynomials in $\theta_k$,
+% \begin{equation}
+% 	\psi_\mathbf{j}(\theta) = \prod\limits_{k=1}^i \! \varphi_{j_k}(\theta_k), \quad \text{for} \quad \mathbf{j} \in \mathcal{J}_i,
+% \end{equation}
+% and where $\varphi_{j_k}(\theta_k) \in \mathcal{P}(\theta_k)$.
 Since $\tilde{T}$ is lower triangular, a multi-index $\mathbf{j}\in\mathcal{J}_i$ only contains entries for univariate polynomials in $\theta_k$ for $k\leq i$.
 
-The cardinalities of the multi-index sets, $M_i = \text{card}(\mathcal{J}_i)$, give the number of unknowns in our
-optimisation problem, and so we would like to keep this number as small as possible. One option is
-to use polynomials of total order $p$,
-\begin{equation}
-	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
-\end{equation}
-which is optimal in terms of the amount of information captured by the map about the target.
+% The cardinalities of the multi-index sets, $M_i = \text{card}(\mathcal{J}_i)$, give the number of unknowns in our
+% optimisation problem, and so we would like to keep this number as small as possible. One option is
+% to use polynomials of total order $p$,
+% \begin{equation}
+% 	\mathcal{J}_i^\text{TO} = \left\{\mathbf{j}:\|\mathbf{j}\|_1 \leq p, j_k = 0\ \forall k > i\right\},
+% \end{equation}
+% which is optimal in terms of the amount of information captured by the map about the target.
 The cardinality of $\mathcal{J}_i^\text{TO}$ is $M_i = \binom{i+p}{p}$ which increases rapidly in $d$ and $p$, where $i = 1, \dots, d$.
@@
 These index sets are discussed
-in~\cite{parno2014transport}.
-Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}.
-Here we stick with polynomials of total order $p$ since we work with low dimensional problems with the ETAIS algorithm.
+% in~\cite{parno2018transport}. Increased information with a slower increase in the number of map parameters can be achieved with the composition of maps discussed in~\cite{parno2015transport}.
 
@@
 We now discuss how we can evaluate the cost function in
-Equation~\eqref{eq:gen_map_optim}. In \cite{parno2014transport}, this
+Equation~\eqref{eq:gen_map_optim}. In \cite{parno2018transport}, this
 expectation is approximated using an MCMC estimator, such that
@@
 
-We now return to the structure of the map components given in Equation~\eqref{eq:map_defn}. Since the basis functions are
+We now return to the structure of the map components given in equation
+\ref{eq:map_defn}. Since the basis functions are
 fixed, the optimisation problem in \eqref{eq:gen_map_optim} is really over the map components $\bar{\gamma} = (\gamma_1, \dots,
@@
 This optimisation problem can be efficiently solved using Newton iterations. It is suggested
-in~\cite{parno2014transport} that this method usually converges in around 10-15 iterations, and we
+in~\cite{parno2018transport} that this method usually converges in around 10-15 iterations, and we
 have seen no evidence that this is not a reasonable estimate. When calculating the map several times
@@
 implement an ensemble importance sampling algorithm which uses
-these maps to propose new states. In \cite{parno2014transport} it was
+these maps to propose new states. In \cite{parno2018transport} it was
 shown how approximate transport maps can be used to accelerate
@@
 with the samples produced from the target measure. Convergence of this
-adaptation is shown in~\cite{parno2014transport}. In this Section, we
+adaptation is shown in~\cite{parno2018transport}. In this Section, we
 will show how similarly, these maps can be used to construct highly
@@
 	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
-	Sample $r'_i \sim q_r(\cdot; r_i)$.\;
-	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(r'_i; \bar{\gamma}^{(k)})$.\;
+	Sample \edit{$\hat{r}_i \sim q_r(\cdot; r_i)$.\;
+	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(\hat{r}_i; \bar{\gamma}^{(k)})$.\;
 	Calculate:
 	\begin{equation}
-		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(r_i'; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
-	\end{equation}
-
-	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{r'_j}(\cdot)$.\label{algline:TETAIS_resample}\;
+		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(\hat{r}_i; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
+	\end{equation}
+
+	Resample $r^* \leftarrow \|w^{(k)}\|^{-1}\sum\limits_{j=1}^M \! w_j^{(k)}\delta_{\hat{r}_j}(\cdot)$.\label{algline:TETAIS_resample}\;}
 	Invert $\theta^{(k+1)}_i = \tilde{T}^{-1}(r^*_i)$.\;
@@
 \{\theta^{(k)}_1,\dots,\theta^{(k)}_M\}$, and the states in the
-reference space, $r = \{r_1,\dots,r_M\}$, where $M$ is the ensemble
-size. Similarly, the proposal states are denoted $r' =
-\{r'_1,\dots,r'_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) =
+reference space, \edit{$r = \{r_1,\dots,r_M\}$}, where $M$ is the ensemble
+size. Similarly, the proposal states are denoted $\hat{r} =
+\{\hat{r}_1,\dots,\hat{r}_M\}$ and $(w^{(k)}, \hat{\theta}^{(k)}) =
 \{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M,
@@
 reference space. This inversion is well defined in the sampling
-region, although not necessarily outside\cite{parno2014transport}.
+region, although not necessarily outside\cite{parno2018transport}.
+
+\edit{For more challenging posterior distributions, the push forward
+  through the numerical
+approximation of the optimal map often leads to a distribution on
+reference space which still has complex structure. It has previously
+been proposed that in more favourable scenarios, the optimal approach
+is to use independent Gaussian samples on reference space, pulled back
+through the inverse map. However, when the push forward of the
+posterior differs greatly from this distribution, as we will see in
+later examples, this is not an
+efficient approach. Using ETAIS in conjunction with the transport map
+allows us to sample efficiently from the push forward measure on
+reference space.}
 
@@
 the ETAIS algorithm, and by tuning the relative $L^2$ error in the MH
-algorithm. There is currently no
+algorithm. 
+\edit{The relative $L^2$ error of the PDF of the measure can be
+  approximated by $E$ where
+\begin{equation}\label{eqn:L2_error}
+	E^2 = \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(s|D) \, \mbox{d}s - vB_i\right]^2 \Big/ \sum\limits_{i=1}^{n_b}\left[\displaystyle\int_{R_i} \! \pi(s|D) \, \mbox{d}s\right]^2,
+\end{equation}
+where the regions $\{R_i\}_{i=1}^{n_b}$ are the $d$-dimensional
+histogram bins, so that $\bigcup_i R_i \subseteq X$ and
+$R_i\cap R_j=\emptyset$, $n_b$ is the number of bins, $v$ is the
+volume of each bin, and $B_i$ is the value of the $i$th bin. This
+metric converges to the standard definition of the relative $L^2$
+error as $v\rightarrow 0$.}
+
+There is currently no
 guidance on the best way of tuning the MH algorithm with transport map
@@
 \end{tabular}
-\caption{Optimal scaling parameters for the transport map based
+\caption{Optimal scaling parameters \edit{$\delta$ for the proposal
+    kernels with the} transport map based
   algorithms applied to $R_1$, optimising for the $L^2$ error and
@@
   Transport ETAIS option 1, algorithm \ref{alg:TransportETAIS2} for
-  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling performed using the MT algorithm.}
+  density \eqref{eqn:R2_repeat}. Ensemble size $M=150$, resampling
+  performed using the MT algorithm. \edit{Plots show average
+    convergence over 32 repeats.}}
 \label{fig:R2_l2_convergence}
@@
 The convergence of the three algorithms is displayed in
-Figure~\ref{fig:R2_l2_convergence}. Figure (a) shows that the two
+Figure~\ref{fig:R2_l2_convergence}. \edit{These convergence plots show
+  the error as a function of the number of samples output by each
+  algorithm. In the case of standard Metropolis-Hastings methods, this
+  corresponds to the number of iterations. In the case of ETAIS, this
+  corresponds to the number of iterations multiplied by the ensemble
+  size.} Figure (a) shows that the two
 variations of the transport based ETAIS algorithms converge with
@@
 model the dynamics through a discrete stochastic model, such as a
-continuous time Markov chain. \edit{Further details about continuous time
-  Markov chains are given in the supplementary material.} ****ARE THEY?****
+continuous time Markov chain.
 
@@
   time. A detailed derivation of this is given in the
-  ******supplementary material******.
+  supplementary material.
 
@@
 detailed derivation of this likelihood is also given in the
-******supplementary material******.
+supplementary material.
 
 In this work, we will consider two different multiscale
-approximations. The first is the quasi-equillibrium approximation
+approximations. The first is the quasi-equilibrium approximation
 which makes the assumption that fast reactions
@@
 \caption{Convergence of four different MCMC algorithms for the CMA
-  approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}.}
+  approximation of the  posterior arising from the example in Section
+  \ref{sec:chem_multiscale}. \edit{Plots show average
+    convergence over 32 repeats.}}
 \label{fig:chem_multiscale_L2}
@@
   to efficiently sample from the more complex distributions arising
-  from the push forward of the target distribution. ****HIGHLIGHT THIS
-  POINT ELSEWHERE -DEFINITELY IN THE CONCLUSIONS AND MAYBE THE INTRO/ABSTRACT?*****
+  from the push forward of the target distribution. % ****HIGHLIGHT THIS
+  % POINT ELSEWHERE -DEFINITELY IN THE CONCLUSIONS AND MAYBE THE INTRO/ABSTRACT?*****
   }
@@
 \subsubsection{Comparison of the Constrained and QEA approaches}
-****GREATLY REDUCE THIS SECTION****
-****Jump straight to Bayes' factors and Figure 7.5****
+% ****GREATLY REDUCE THIS SECTION****
+% ****Jump straight to Bayes' factors and Figure 7.5****
 
@@
 \caption{Convergence of four different MCMC algorithms for the CMA
-  approximation of the  posterior arising from the example in Section \ref{sec:grn}.}
+  approximation of the  posterior arising from the example in Section
+  \ref{sec:grn}. \edit{Plots show average
+    convergence over 32 repeats.}}
 \label{fig:grn_L2}
@@
 
-In
-particular we looked at the application of these methods to problems
-arising in inverse problems for multiscale stochastic chemical
-networks, where the posterior distributions which arise can be highly
-concentrated along curved manifolds in parameter space. Through these
-results we also demonstrated the improvement in accuracy of inference
-for multiscale problems when more sophisticated multiscale
-approximations are employed.
-Below we highlight two areas for future work.
+\edit{If good transport maps can be found, such that the push forward
+  of the posterior is close to a reference Gaussian, then the most
+  efficient approach is to draw independent samples pulled back
+  through the map onto the target space. However, for complex
+  posteriors, in particular the type that we have observed here where
+  due to sloppiness in the model, the posterior is concentrated on a
+  low-dimensional manifold, a good map is hard to find. The numerical
+  approximation of the optimal map leads to a measure on the reference
+  space which is far removed from the reference Gaussian, and requires
+  more considered exploration. Adaptive importance sampling schemes
+  such as ETAIS are able to efficiently sample from such a measure
+  more efficiently than Metropolis-Hastings equivalents.}
+
+In particular we looked at the application of these methods to
+problems arising in inverse problems for multiscale stochastic
+chemical networks, where the posterior distributions which arise can
+be highly concentrated along curved manifolds in parameter
+space. Through these results we also demonstrated the improvement in
+accuracy of inference for multiscale problems when more sophisticated
+multiscale approximations are employed. Below we highlight two areas for future work.
 
@@
 
-\subsection{Invertibility of the transport map}
-In this paper, we have considered an approach similar to that
-presented in \cite{parno2014transport}, where we aim to find an
-invertible map which approximates the true transport map between the
-reference measure and the target measure. However, this true map is
-invariably itself not invertible. However, by considering embedding
-the map in higher dimensions\cite{whitney_self-intersections_1994,nash1956imbedding,takens_detecting_1981}, these issues of non-invertibility may be
-avoided, and higher quality maps may be found which better aid
-exploration and sampling of the whole state space.
+% \subsection{Invertibility of the transport map}
+% In this paper, we have considered an approach similar to that
+% presented in \cite{parno2018transport}, where we aim to find an
+% invertible map which approximates the true transport map between the
+% reference measure and the target measure. However, this true map is
+% invariably itself not invertible. However, by considering embedding
+% the map in higher dimensions\cite{whitney_self-intersections_1994,nash1956imbedding,takens_detecting_1981}, these issues of non-invertibility may be
+% avoided, and higher quality maps may be found which better aid
+% exploration and sampling of the whole state space.
 

### 2020-02-06 Resubmission/response.tex
Resubmitted version

@@
 We also attach a version of the paper with the changes in text highlighted in red.
-Since a lot of the larger changes involved removal of sections of text and material, we have also attached a copy of the originally submitted paper with the removed sections highlighted in blue.
+Note that a lot of the changes involved removal of material which is not visible in this document but is outlined in the responses below.
 
@@
 With the length of this paper in mind, I would suggest shortening Sections 6 and 7 in order to focus more on numerical comparisons of the new algorithm and less on stochastic reaction networks.
-Details of the multiscale material could be moved into supplementary material or even into another manuscript.}{We agree that the paper is lengthy, and welcome suggestions as to how to shorten it to make it more accessible.
+Details of the multiscale material could be moved into supplementary material or even into another manuscript.}{We agree that the paper is lengthy, we shortened it considerably, and we welcome the suggestions as to how to make it more accessible.
 The method was designed with the application of inference of multiscale stochastic networks in mind, and the main examples in the manuscript concern this topic.
@@
 The value used has been shown to work well, as demonstrated in the numerical experiments presented.
-A discussion along these lines has been added to after equation (3.2).}
+A discussion along these lines has been added after equation (3.2).}
 
@@
 This was done in detail for our first paper ``Ensemble Transport Adaptive Importance Sampling'', and was published in detail in Paul Russell's PhD thesis.
-We compared RW proposals with MALA and even HMC proposals.
-The performance of all of the proposals were similar, but the MALA and HMC proposals come at additional costs.
+We compared RW proposals with MALA and HMC.
+The performance of all of the proposals were similar, but the RW proposal is the cheapest.
 We have added a brief discussion about proposal types at the end of Section 2.}
@@
 The authors mention that a smaller ensemble can be used when the transport map is accurate, but are there any guidelines on how to choose the ensemble size given the map?}{It is relatively easy to assess when the ensemble size is too small, since it leads to poor importance samples and larger variances in the weights.
-If the transport map is doing a good, then the push forward distribution will be close to Gaussian, and then rules of thumb regarding the number of samples required to characterise an $n$-dimensional Gaussian can be used to suggest ensemble sizes.
+If the transport map is doing a good job, then the push forward distribution will be close to Gaussian, and then rules of thumb regarding the number of samples required to characterise an $n$-dimensional Gaussian can be used to suggest ensemble sizes.
 If variances of weights are very high, then it is likely that ensemble sizes will need to be increased.
@@
 The two deltas, as described in the caption in the table, are the optimal scaling parameters, one optimised for ESS and the other for our relative $L^2$ measure.
-We have expanded on this in the caption, and hope that this is clearer now.}
+We have expanded on this in the caption, and hope that this is clearer now.
+Moreover, our notation wasn't consistent throughout the paper, as $\delta$ is actually the $\beta$ scaling parameter in the proposal distributions $q(\cdot;\cdot,\beta)$.
+This has now been corrected throughout.}
 
@@
 This is challenging both in terms of computing it, but it is also a challenging metric for the Monte Carlo methods, since it requires global convergence of the chains.
-We understand the reasons that the referee has asked for this metric to be considered, not least because it's use is ubiquitous but we believe that the one we have used is actually a more rigorous measure.}
+We understand the reasons that the referee has asked for this metric to be considered, not least because it's use is ubiquitous, but we believe that the one we have used is actually a more rigorous measure.}
 
@@
 This is more or less the Knothe–Rosenblatt rearrangement; see one of Villani’s books, or [Bogachev, Kolesnikov, Medvedev 2005] for a more formal treatment.
-Similarly, the statement in Section 8.2 about the true map being “invariably itself not in- vertible” seemed odd to me.
+Similarly, the statement in Section 8.2 about the true map being “invariably itself not invertible” seemed odd to me.
 If by the true map one means the KR rearrangement, then it is invertible.

### 2020-02-06 Resubmission/TETAIS.tex
Resubmitted version

@@
 sampling (ETAIS) method, and allows for efficient sampling with
-smaller ensemble sizes.
+smaller ensemble sizes. \edit{This approach allows us to take
+  advantage of the large increases of efficiency when using adaptive
+  importance sampling methods for previously intractable Bayesian
+  inverse problems with complex posterior structure.}
 \end{abstract}
@@
 which a map which describes the manifold is constructed using
-principle component analysis. These ideas have also been applied to
+Principle Component Analysis. These ideas have also been applied to
 stochastic reaction
@@
 convergence. Better informed proposal densities
-leads to more stable and faster converging importance sampling
+lead to more stable and faster converging importance sampling
 schemes, and we will demonstrate this through examples
 arising from Bayesian inverse problems on multiscale stochastic
-reaction networks. \edit{In a perfect scenario, the numerical
+reaction networks.
+
+\edit{In a perfect scenario, the numerical
   approximation of the optimal transport map would capture all of the
   posterior structure, and independent samples could be drawn by
-  sampling from the reference measure. However in many cases the push
-  forward of the posterior measure can still exhibit complex structure
+  sampling from the reference measure. However in many cases the pushforward of the posterior measure can still exhibit complex structure
   which means that efficient adaptive importance sampling can be
@@
 In ETAIS, the proposal distribution $\chi$ is chosen to be the equally
-weighted mixture of set of proposal kernels. If $\theta^{(k)} = [\theta_1^{(k)},
+weighted mixture of a set of proposal kernels. If $\theta^{(k)} = [\theta_1^{(k)},
 \theta_2^{(k)}, \ldots, \theta_M^{(k)}]^\top$ is the current state of the
@@
   work we will primarily use Gaussian kernels, but we have previously
-  made comparisons with more complex kernels, such as MALA or
-  HMC\cite{cotter2015parallel,russ2017parallel}. There is minimal
-  benefit from using these more complex kernels, which come at
+  made comparisons with more complex kernels, such as MALA and HMC\cite{cotter2015parallel,russ2017parallel}. There are minimal
+  benefits from using these more complex kernels, which also come at
   additional cost.}
@@
 
-Proposing new states in this way allows us to make large, informed
+Proposing new states in this way allows us to make large, well-informed
 moves, even when sampling from probability distributions with complex structure.
 If a very good such map can be found, it is also possible that the
-push forward of the posterior density through the map is close enough
+pushforward of the posterior density through the map is close enough
 to the reference Gaussian that we can efficiently
@@
 Our aim is now to find the lower triangular map $\tilde{T} \in
-\mathcal{T}^{\ltri}$ such that the difference between target
+\mathcal{T}^{\ltri}$ such that the difference between the target
 density and the pullback of the reference density is minimised. As in
@@
 \end{equation}
-We note the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
+We note that the posterior density is independent of $\tilde{T}$, and so it is not necessary for us to compute it when optimising this cost function.
 This expression is a complicated integral with respect to the target distribution, for which the normalisation constant is unknown.
@@
 \end{equation}
-\edit{The parameter $\beta>0$ does not need to be tuned, as experimentation has shown that the choice
+\edit{The expectation in \eqref{eq:gen_map_optim} can be
+approximated by using a Monte Carlo approximation. The form of the regularisation term promotes maps which are
+closer to the identity. The parameter $\beta>0$ does not need to be tuned, as experimentation has shown that the choice
 $\beta=1$ is sufficient for most problems. Note that this
@@
 map to what is often an incomplete and poorly converged sample from
-the posterior, and smaller values of $\beta$ may lead to maps which
+the posterior in the initial phases. Smaller values of $\beta$ may lead to maps which
 hinder further exploration of the state space beyond that already
-represented in the sample used to construct the map.
-
-The expectation in \eqref{eq:gen_map_optim}} can be
-approximated by using a Monte Carlo approximation. The form of the penalisation term promotes maps which are
-closer to the identity, and so prevents overfitting when the quality
-or size of the current sample from the posterior is not
-sufficient. \edit{Moreover, we can exploit the lower triangular
+represented in the sample used to construct the map.}
+
+\edit{Moreover, we can exploit the lower triangular
   structure, for example in the evaluation of 
@@
 
-We now discuss how we can evaluate the cost function in
+\edit{We now discuss how we can evaluate the cost function in
 Equation~\eqref{eq:gen_map_optim}. In \cite{parno2018transport}, this
-expectation is approximated using an MCMC estimator, such that
-\begin{align}
-	C(T) &= \mathbb{E}_\pi\left[ -\log\pi_r(T(\theta)) - \log|J_T(\theta)|\right] +
-			\beta\mathbb{E}(T(\theta)-\theta)^2 \notag \\
-		&\approx \frac{1}{K}\sum\limits_{i=1}^d \! \sum\limits_{k=1}^K \left[-\log\pi_r(T_i(\theta^{(k)})) -
-			\log\left|\frac{\partial T_i}{\partial \theta_i}(\theta^{(k)})\right| + \beta(T_i(\theta^{(k)})-\theta^{(k)})^2\right]. 
-\end{align}
-Here we diverge from previous work, as we aim to build a map from
-samples from an importance sampling scheme. Such samples no longer
+expectation is approximated using an Monte Carlo estimator with samples from
+a Metropolis-Hastings method, which have equal weighting.
+Here we aim to build a map from
+samples from an importance sampling scheme.} Such samples no longer
 carry equal weight, and as such the Monte Carlo estimator becomes
@@
 positive and so this functional is always finite.
-In practice it is infeasible to enforce this condition across the whole parameter space.
+In practice it is infeasible to enforce this condition across the entire parameter space.
 We instead enforce this condition by ensuring that the derivatives are positive at each sample point.
@@
 have seen no evidence that this is not a reasonable estimate. When calculating the map several times
-during a Monte Carlo run, using previous guesses of the optimal map to seed the Newton algorithm
+during a Monte Carlo run, using previous guesses of the optimal map to
+seed the Newton algorithm 
 results in much faster convergence, usually taking only a couple of iterations to satisfy the stopping
@@
 Metropolis-Hastings methods, with the map being periodically updated
-with the samples produced from the target measure. Convergence of this
+through the samples produced from the target measure. Convergence of this
 adaptation is shown in~\cite{parno2018transport}. In this Section, we
@@
 	Compute $r_i = \tilde{T}(\theta^{(k)}_i; \bar{\gamma}^{(k)})$, \quad $i = 1,\dots,M$.\;
-	Sample \edit{$\hat{r}_i \sim q_r(\cdot; r_i)$.\;
+	Sample \edit{$\hat{r}_i \sim q_r(\cdot; r_i, \beta)$.\;
 	Invert $\hat{\theta}_i^{(k)} = \tilde{T}^{-1}(\hat{r}_i; \bar{\gamma}^{(k)})$.\;
@@
 	\begin{equation}
-		w_i^{(k)} = \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \! q_r(\hat{r}_i; r_j)\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
+		w_i^{(k)} =
+                \frac{\pi(\hat{\theta}_i^{(k)})}{\left(\sum_{j=1}^M \!
+                    \edit{q_r(\hat{r}_i; r_j, \beta)}\right)|J_{\tilde{T}}(\hat{\theta}_i^{(k)};\bar{\gamma}^{(k)})|}.
 	\end{equation}
@@
 \{(w^{(k)}_1, \hat{\theta}^{(k)}_1),\dots,(w^{(k)}_M,
-\hat{\theta}^{(k)}_M)\}$, where these pairs are the states which
-together form our sample from the target distribution. As in the standard version of the ETAIS algorithm we use the deterministic mixture weights.
+\hat{\theta}^{(k)}_M)\}$, where these \edit{pairs constitute the
+  weighted samples} from the target distribution. As in the standard version of the ETAIS algorithm, we use the deterministic mixture weights.
 
 The second option, is similar to
-the first except on Line~\ref{algline:TETAIS_resample} where rather
+the first except on Line~\ref{algline:TETAIS_resample}, where rather
 than resampling in target space we resample in reference
-space. Assuming a good transport map, the push forward of the target
+space. Assuming a good transport map, the pushforward of the target
 measure onto 
@@
 
-\edit{The choice of ensemble size can be problem dependent, but the
-  behaviour of the algorithm is quite clear when the ensemble size is
+\edit{The choice of ensemble size can be problem dependent, but it is quite clear when the ensemble size is
   too small, because the proposal distribution poorly represents the
@@
 We draw 1 million samples from the density \eqref{eqn:R2_repeat}, and use this sample in the framework of Section~\ref{sec:map} to build a transport map.
-We use this map to push forward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian.
+We use this map to pushforward the original sample onto the reference space, where we will be able to see how well the map has performed at converting the original sample to a standard Gaussian.
 We then pull the sample back on to target space using the inverse map to check that our map is invertible and well behaved.
@@
 Clearly even with only basis functions of total order 3, we have a large number of unknowns in our optimisation problem, $\bar{\gamma} \in \mathbb{R}^{14}$.
-If we were to increase the dimension of $\theta$ further we would need to reduce the number of terms we include in the expansion by, for example, removing all the ``cross'' terms.
-This reduces the quality of our map but since we only require an approximate map we can afford to reduce the accuracy.
+If we were to increase the dimension of $\theta$ further, we would need to reduce the number of terms we include in the expansion by, for example, removing all the ``cross'' terms.
+This reduces the quality of our map, but since we only require an approximate map we can afford to reduce the accuracy.
 
@@
 \subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
-\subfigure[Push forward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
-\subfigure[Pull back of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
+\subfigure[Pushforward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
+\subfigure[Pullback of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
 \caption{The effect of the approximate transport map $\tilde{T}$ on a
@@
 the monomial basis down to 4 and 10 terms in respective dimensions,
-the push forward of the sample is still a unimodal distribution
+the pushforward of the sample is still a unimodal distribution
 centred at the origin with standard deviation 1. As one moves out into
 the tails of the reference density more non-Gaussian features are
-clearly visible. However, overall, the push forward of the target
+clearly visible. However, overall, the pushforward of the target
 density does not look a challenging one to sample from, with even
@@
 reference space. This inversion is well defined in the sampling
-region, although not necessarily outside\cite{parno2018transport}.
-
-\edit{For more challenging posterior distributions, the push forward
+region, although not necessarily outside it\cite{parno2018transport}.
+
+\edit{For more challenging posterior distributions, the pushforward
   through the numerical
@@
 is to use independent Gaussian samples on reference space, pulled back
-through the inverse map. However, when the push forward of the
+through the inverse map. However, when the pushforward of the
 posterior differs greatly from this distribution, as we will see in
@@
 efficient approach. Using ETAIS in conjunction with the transport map
-allows us to sample efficiently from the push forward measure on
+allows us to sample efficiently from the pushforward measure on
 reference space.}
@@
   transport map based algorithms on the Rosenbrock density}
-
+\edit{The performance of Metropolise-Hastings methods and adaptive
+importance sampling methods depends on the scaling parameter
+$\beta^2$ for the proposal distributions $q(\cdot ;
+\cdot, \beta)$. For the RW method, for example, the variance of the
+proposal is equal to $\beta^2$.}
 We first find the optimal scaling parameters for the individual
@@
 guidance on the best way of tuning the MH algorithm with transport map
-proposals although one might expect results similar to the standard MH
+proposals, although one might expect results similar to the standard MH
 results, especially if adaptation of the map is stopped after a given
@@
                                                             option 1 & Alg. \ref{alg:TransportETAIS2}  \\ \cmidrule(lr){1-4}
-	$\delta_{L^2}$				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
-	$\delta_{\text{ESS}}$				 & - & 1.0e-1 & 5.2e-1 \\ \cmidrule(lr){1-4}
+	\edit{$\beta_{L^2}$}				 & 1.0e-0 & 1.1e-1 & 3.5e-1 \\
+	\edit{$\beta_{\text{ESS}}$}				 & - & 1.0e-1 & 5.2e-1 \\ \cmidrule(lr){1-4}
 	Acc. rate							 & 0.23 & - & - \\
@@
 \end{tabular}
-\caption{Optimal scaling parameters \edit{$\delta$ for the proposal
-    kernels with the} transport map based
-  algorithms applied to $R_1$, optimising for the $L^2$ error and
+\caption{\edit{Optimal scaling parameters $\beta$ for the proposal
+    kernels with the transport map based
+  algorithms applied to $R_1$, optimising for the approximate
+    $L^2$ error~\eqref{eqn:L2_error} and
   the effective sample size (ESS) for ETAIS algorithms, and for average
-  acceptance rate for MH algorithms.}
+  acceptance rate for MH algorithms.}}
 \label{tab:R2_opt_scaling}
@@
 and its accelerated version is relatively negligible, in part because
-this version of the Rosenbrock is not overtly pathological. The story
+this version of the Rosenbrock density is not overtly tortuous. The story
 is different for the ETAIS algorithms. The slowly decaying tails of
@@
 channels. When population numbers of one or more of the chemical
-species in the reactor is small, as is the case with chemical
+species in the reactor are small, as is the case with chemical
 reactions occurring within living cells, the sporadic and discrete way
@@
 \begin{equation}\alpha_j(X(t)) = \lim_{dt \to 0} \mathbb{P}(\text{Reaction $R_j$ in
-    the time interval  } \quad s \in [t, t+ dt] ).\end{equation} % If a system
+    the time interval  } \, s \in [t, t+ dt] ).\end{equation} % If a system
 % satisfies what is called \emph{mass action kinetics}, then the form of
@@
 products produced. This is modelled by a change in the state vector
-$X(t) = X(t) + \nu_j$ where $\nu_j$ is that stoichiometric vector for
+$X(t) = X(t) + \nu_j$ where $\nu_j$ is the stoichiometric vector for
 reaction $R_j$.
@@
   of a Poisson process with rate which is a function of the number of
-  reactant molecules present, and the reaction rates. The likelihood
+  reactant molecules present, and the reaction rate constants. The likelihood
   of each reaction is simply the product of the probability density of
@@
 \section{Numerical Examples}\label{sec:num}
-In this section we look at two examples of chemical systems to
+In this section we consider two examples of chemical systems to
 demonstrate the effectiveness of transport ETAIS algorithms for
@@
 which is often the case\cite{anderson2010product,anderson2016product}. For this very
-simple example, the dependance of these effective
+simple example, the dependence of these effective
 propensities on the original rate parameters can be explicitly
@@
 $\bar{\alpha}_4$. Although this posterior does not appear to be
-pathologically difficult to sample using MCMC from these plots, it most certainly is. The posterior
+impenetrably difficult to sample using MCMC from these plots, it most certainly is. The posterior
 density is concentrated close to a curved manifold. Since the manifold
-is curved, the joint marginals do not look unreasonable, where in fact
-the density has very differing derivatives in different directions
-when evaluated close to this manifold.
+is curved, the joint marginals do not look unreasonable. However,
+the distribution
+is highly peaked around this curved manifold, making it very difficult
+to sample from efficiently.
 
@@
 this is a Gaussian mixture proposal distribution, and in the MH
-algorithm this is equivalent to a vanilla random walk. The proposal
+algorithm this is equivalent to a ``vanilla'' random walk. The proposal
 distribution uses a covariance matrix which has been constructed using
@@
 truly bijective. This leads to problems for our strictly positive
-state space, $\mathcal{X} \subset \mathbb{R}_+^d$, since proposals in
+state space, $\mathcal{X} \subset \mathbb{R}_+^d$, and so proposals in
 $R$ do not necessarily map back onto $\mathcal{X}$. \edit{This can be
@@
 
+\edit{
 \begin{table}[!h]
 	\centering
-\begin{tabular}{rrrrrr}
+	\begin{tabular}{lrrrr}
 	\toprule
-	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{ETAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
-	& & $\delta$ & & $\delta$ & ESS \\ \midrule
-	% RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
-	% TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
-	logRW & & 1.7e-1 & & 1.2e-0 & 6.0e-2 \\
-	logTRW & & 2.7e-2 & & 1.5e-1 & 3.5e-1 \\
+		 & \multicolumn{2}{c}{MH} & \multicolumn{2}{c}{ETAIS} \\ \cmidrule(lr){2-3} \cmidrule(lr){4-5}
+		& logRW & log-TRW & logRW & log-TRW \\ \midrule
+		$\beta_{\%}^*$	 	& 1.7e-1 &2.7e-2 & - & - \\
+		$\beta_{\text{ESS}}^*$	       & -            & -          & 1.2e-0 & 1.5e-1 \\
+		ESS/$M$		 	 & -            & -            & 6.0e-2 & 3.5e-1  \\
 	\bottomrule
 	\end{tabular}
-\caption{Optimal scaling parameters for the MH and ETAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}.
+	\caption{Optimal scaling parameters for the MH and ETAIS
+          algorithms applied to the constrained multiscale problem in
+          Section~\ref{sec:chem_multiscale}. MH parameters optimised
+          by acceptance rate, and ETAIS parameters optimised using
+          effective sample size. ESS per sample also shown for the
+          ETAIS algorithms.}
+\label{tab:chem_multiscale_scaling}
+\end{table}}
+
+% \begin{table}[!h]
+% \centering
+% \begin{tabular}{rrrrrr}
+% \toprule
+% 	\multicolumn{1}{l}{Algorithm} & \multicolumn{3}{c}{MH} & \multicolumn{2}{c}{ETAIS} \\ \cmidrule(lr){2-4} \cmidrule(lr){5-6}
+% 	& & $\delta$ & & $\delta$ & ESS \\ \midrule
+% 	% RW & & 5.5e-3 & & 1.0e-0 & 9.0e-3 \\
+% 	% TRW & & 1.2e-0 & & 4.0e-1 & 1.2e-3 \\
+% 	logRW & & 1.7e-1 & & 1.2e-0 & 6.0e-2 \\
+% 	logTRW & & 2.7e-2 & & 1.5e-1 & 3.5e-1 \\
+% \bottomrule
+% \end{tabular}
+% \caption{Optimal scaling parameters for the MH and ETAIS algorithms applied to the constrained multiscale problem in Section~\ref{sec:chem_multiscale}.
 MH parameters optimised by acceptance rate, and ETAIS parameters optimised using effective sample size.}
-\label{tab:chem_multiscale_scaling}
-\end{table}
+% %\label{tab:chem_multiscale_scaling}
+% \end{table}
 
@@
 since they relate to algorithms operating on different
-transformations. The ESS is highest for the logTRW variant
-of the ETAIS algorithm, as we would hope. We note here that as
+transformations. \edit{The ESS per sample is also displayed in
+Table~\ref{tab:chem_multiscale_scaling} for the ETAIS implementations,
+and is highest for the logTRW variant
+of the ETAIS algorithm, as we would hope.} We note here that, as
 outlined in \cite{cotter2015parallel}, it is possible to construct
@@
 $T$. Overall we see the smallest Monte Carlo errors for a given number
-of likelihood evaluations coming from the ETAIS-logTRW algorithm.
+of likelihood evaluations coming from the ETAIS-logTRW algorithm. This
+is due to weight collapse in some of the 32 repeats of the ETAIS-logRW
+algorithm, demonstrating that the transport map stabilises the ETAIS methodology.
 }
@@
 % \\
-\subfigure[Push forward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
-\caption{\edit{Push forward of the target measure on reference space for the the CMA
+\subfigure[Pushforward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
+\caption{\edit{Pushforward of the target measure on reference space for the CMA
   approximation of the  posterior arising from the example in Section \ref{sec:chem_multiscale}, found using the logTRW proposal distributions.
@@
 number of choices.} We might wish to learn the covariance structure of
-the push forward of the target onto the reference space, and
+the pushforward of the target onto the reference space, and
 incorporate this information into the covariances of the mixture
@@
   targets, the approximation of the transport map is not as good, and
-  as such the push forward of the target onto the reference
-  space differs greatly from the the intended reference measure. In
-  this case, the pull back of the reference measure will be far from
+  as such the pushforward of the target onto the reference
+  space differs greatly from the intended reference measure. In
+  this case, the pullback of the reference measure will be far from
   the target distribution, and this independent sampling strategy
@@
   to efficiently sample from the more complex distributions arising
-  from the push forward of the target distribution. % ****HIGHLIGHT THIS
+  from the pushforward of the target distribution. % ****HIGHLIGHT THIS
   % POINT ELSEWHERE -DEFINITELY IN THE CONCLUSIONS AND MAYBE THE INTRO/ABSTRACT?*****
@@
 interesting correlation structures, although several dimensions have
-leptokurtic tails which are difficult for the standard ETAIS algorithm
+leptokurtic tails, which are difficult for the standard ETAIS algorithm
 to sample from. However as with the previous example, the density is
-concentrated close to a lower-dimensional manifold. The marginal densities also vary over very different
+concentrated close to a lower-dimensional curved manifold. The marginal densities also vary over very different
 scales, which could lead to problems with the algorithms which do not
@@
 converted into an estimate which is twice as accurate after 1 million
-samples. This demonstrates our claim that the transport map can
+samples. This corroborates with our claim that the transport map can
 produce stable ETAIS implementations with smaller ensemble sizes than without.
@@
 
-\edit{If good transport maps can be found, such that the push forward
+\edit{If good transport maps can be found, such that the pushforward
   of the posterior is close to a reference Gaussian, then the most
@@
   more considered exploration. Adaptive importance sampling schemes
-  such as ETAIS are able to efficiently sample from such a measure
+  such as ETAIS are able to sample from such a measure
   more efficiently than Metropolis-Hastings equivalents.}
@@
 distribution close to a Gaussian reference measure, which is
-uncorrelated in all dimensions. This lack of correlations allows us to
+approximately uncorrelated in all dimensions. This lack of correlations allows us to
 resample in each dimension separately. Resampling in a

### 2020-07-27 TETAIS/TETAIS.tex
Final preparation of files

--- 2020-02-06 Resubmission/TETAIS.tex
@@
 \renewcommand{\b}[1]{{\bf #1}}
-\newcommand{\edit}[1]{{\color{red} #1}} 
+\newcommand{\edit}[1]{#1} 
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 \centering
-\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_1"}}
-\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_marginal_2"}}
-\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_posterior}}
+\subfigure[Marginal density function for $\theta_1$.]{\includegraphics[width=0.31\textwidth]{"images/R2_marginal_1"}}
+\subfigure[Marginal density function for $\theta_2$.]{\includegraphics[width=0.31\textwidth]{"images/R2_marginal_2"}}
+\subfigure[Contour plot for Rosenbrock density.]{\includegraphics[width=0.31\textwidth]{images/R2_posterior}}
 \caption{Visualisation of the Rosenbrock density as given in Equation~\eqref{eqn:R2_repeat}.}
@@
 \centering
-\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_orig"}}\quad
-\subfigure[Pushforward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/TPAIS/R2_transport_ref"}}\quad
-\subfigure[Pullback of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/TPAIS/R2_transport_pullback}}
+\subfigure[Original sample $\theta$ from MH-RW algorithm.]{\includegraphics[width=0.31\textwidth]{"images/R2_transport_orig"}}\quad
+\subfigure[Pushforward of $\theta$ onto reference space.]{\includegraphics[width=0.31\textwidth]{"images/R2_transport_ref"}}\quad
+\subfigure[Pullback of reference sample onto target space.]{\includegraphics[width=0.31\textwidth]{images/R2_transport_pullback}}
 \caption{The effect of the approximate transport map $\tilde{T}$ on a
@@
 \subfigure[Comparison of the two Transport ETAIS
-options.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2}}
-\subfigure[Comparison of Transport ETAIS option 2 with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/TPAIS/R1_L2_all}}
+options.]{\includegraphics[width=0.45\textwidth]{images/R1_L2}}
+\subfigure[Comparison of Transport ETAIS option 2 with the standard algorithms.]{\includegraphics[width=0.45\textwidth]{images/R1_L2_all}}
 \caption{Convergence of algorithms; Transport M-H,
@@
 \centering
-\includegraphics[width=0.9\textwidth]{"images/Applications/CMA_posterior"}
+\includegraphics[width=0.9\textwidth]{"images/CMA_posterior"}
 \caption{CMA approximation of the posterior arising from observations of
@@
 % \subfigure[Sampling algorithms with no preconditioner for
-    %     $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_param_space"}}
-\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_L2_log_space"}
+    %     $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/CMA_L2_param_space"}}
+\includegraphics[width=0.495\textwidth]{"images/CMA_L2_log_space"}
 \caption{Convergence of four different MCMC algorithms for the CMA
@@
 % no log
-% preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHTRW"}}
+% preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/CMA_ref_space_MHTRW"}}
 % \\
-\subfigure[Pushforward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/Applications/CMA_ref_space_MHlogTRW"}}
+\subfigure[Pushforward of the posterior sample through $\hat{T}$ with log preconditioner.]{\includegraphics[width=0.8\textwidth]{"images/CMA_ref_space_MHlogTRW"}}
 \caption{\edit{Pushforward of the target measure on reference space for the CMA
@@
 % \centering
-% \includegraphics[width=0.9\textwidth]{"images/Applications/Diff_CMA_QEA"}
+% \includegraphics[width=0.9\textwidth]{"images/Diff_CMA_QEA"}
 % \caption{Difference between the CMA and QEA posteriors as defined in Equation~\eqref{eqn:chem_diff}.}
@@
 \centering
-\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/QEA_approx_marginal"}}
-\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/CMA_approx_marginal"}}
+\subfigure[Marginal for $\hat{k}_4^{\text{QEA}}$.]{\includegraphics[width=0.495\textwidth]{"images/QEA_approx_marginal"}}
+\subfigure[Marginal for $\hat{k}_4^{\text{CMA}}$.]{\includegraphics[width=0.495\textwidth]{"images/CMA_approx_marginal"}}
 \caption{Comparison of the approximate marginal densities for the
@@
 \centering
-\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/Applications/GRN_posterior"}}%
+\makebox[\textwidth][c]{\includegraphics[width=1.2\textwidth]{"images/GRN_posterior"}}%
 \caption{Marginal densities of the CMA approximation of the posterior
@@
 \centering
-% \subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence"}}
-\includegraphics[width=0.495\textwidth]{"images/Applications/GRN_convergence_log"}
+% \subfigure[Sampling algorithms with no preconditioner for $\tilde{T}$.]{\includegraphics[width=0.495\textwidth]{"images/GRN_convergence"}}
+\includegraphics[width=0.495\textwidth]{"images/GRN_convergence_log"}
 \caption{Convergence of four different MCMC algorithms for the CMA
@@
   more efficiently than Metropolis-Hastings equivalents.}
+
+One issue that we have not looked at in depth, is that of the choice
+of basis functions for the transport map, which is discussed in more
+depth in \cite{parno2018transport}. It seems reasonable to
+expect that this choice could have an impact on the quality of the
+transport map, which in turn has an impact on the number of ensemble
+members which are required to stably sample from the target
+distribution. We explored some different combinations, including higher
+order but sparse bases, but we did not observe a big difference in
+performance. It would be an interesting avenue to explore how other
+approaches, for example tailored, anisotropic, or even adaptively
+chosen parameterisations, could enable the map to more accurately
+capture the structure of the target density, reducing the required
+ensemble size.
 

### 2020-07-27 TETAIS/TETAIS_supplementary.tex
Final preparation of files

--- 2020-01-14 Resubmission/Supplementary.tex
@@
 \title{Multiscale Methods for Stochastic Chemical Reaction
-  Networks\\Supplementary material}
+  Networks:\\Supplementary material}
 \begin{document}
 \maketitle
+\stepcounter{section}
+\def\thesection{\alph{section}}
 In this supplementary material, we discuss some recent advances in multiscale methods
@@
 
-One common approach is to induce the quasi-equilibrium approximation
+One common approach is to invoke the quasi-equilibrium approximation
 QEA. This approximation makes the assumption that fast reactions
